---
title: "Maxfield Meadow Ipomopsis Volatiles"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>


```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)
library(vegan)
library(broom)
library(emmeans)
library(viridis)
library(ggvegan)
library(gridExtra)
library(dendsort)
library(pheatmap)
library(RColorBrewer)
library(DT)
library(lme4)
library(lmerTest)
library(bouquet)
library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, 
                      fig.path = "images_volatiles2020/", dev="svglite", dev.args=list(fix_text_size=FALSE), fig.height=8, fig.width=8)
options(digits=4) # for kables
```

```{r rename_figs, eval=FALSE}
figs <- read_tsv("data/volatiles/volatiles_fig_paths.tsv")
walk2(figs$path, figs$fig, ~system(paste0("cp ", .x, " ./figures_volatiles2020/Figure", .y, ".",tools::file_ext(.x))))
walk2(figs$path, figs$fig, ~system(paste0("inkscape -o ./figures_volatiles2020/Figure", .y, ".eps ", .x)))
```

```{r loadscents}
load("data/maxfield_data.rda") #load trait measurements, etc
load("data/volatiles/maxfield_volatiles_lessdupes.rda") # from maxfield_gc.R, loads maxfmeta and maxf (wide GC data)

fitnesstraits <- c("seeds_est", "seeds_per_flower", "prop_uninfested", "eggs_per_flower")
fitnessnames <- c(traitnames, prop_uninfested = "Prop. uneaten fruits",
                  eggs_per_flower="Fly eggs per flower")[fitnesstraits]

#check for duplicates in vt and maxfmeta
#vt %>% filter(!exclude %in% c("flowers", "incomplete", "resample")) %>% count(sampledate.file, plot.file, subplot.file, plant.file) %>% filter(n>1)  #exclude="resample" #no duplicates in vt
#I think these plantid duplicates are all just two samples taken on the same day, except maybe 3D858
# maxfmeta %>% count(sampledate,plantid) %>% filter(n>1)
# maxfmeta %>% count(sampledate,plantid, other) %>% filter(n>1)# other column has info on vial, run position, etc

volsmeta <- maxfmeta %>% mutate(sampledate = as.Date(sampledate), year = factor(year(sampledate))) %>% 
  drop_na(plotid) %>% #cut out ambients, leaving 459 chromatograms
  filter(year %in% c("2018","2019","2020")) %>% droplevels() %>% 
  mutate(plot=str_sub(plotid, 1,1), subplot=str_sub(plotid, 2,2),
         sequence.date = date(sequence.start),
         sequence.year = year(sequence.date)) %>% 
  rename(sampledate.file=sampledate, plot.file = plot, subplot.file=subplot, plant.file=plant) %>% 
  left_join(vt %>% select(-c(colnames(treatments),"type","plantid")) %>% 
              filter(!exclude %in% c("flowers", "incomplete", "resample"))) %>% 
  left_join(treatments) %>% #assume that the written metadata is correct not the info in the  .file name
  mutate(total_time = end - bag,
         total_time = replace_na(total_time, median(total_time, na.rm=T)),# fill in 45 minutes when missing metadata
         pump_id=as.character(pump_id))
volsmeta <- left_join(volsmeta, volsmeta %>% count(sequence.date, sequence.start) %>% group_by(sequence.date) %>% mutate(sequence.12 = paste(sequence.date, row_number()))) #add the batch number for that day (1 or 2)
rownames(volsmeta) <- volsmeta$index 
```


# Sample size

Calculated from two datasets that will eventually match: datasheets from sampling and GC-MS filenames.

```{r samplesizes, fig.width=10, fig.height=5}
bind_rows(sampling=vt %>% group_by(year, type) %>% tally(),
  filenames=maxfmeta %>% mutate(year=factor(year(sampledate))) %>% group_by(year, type) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year,type), values_from = n) %>% kable()

bind_rows(sampling=vt %>% filter(type=="floral") %>% group_by(year, water, snow) %>% tally(),
  filenames=volsmeta %>% mutate(year=factor(year(sampledate.file))) %>% group_by(year, water, snow) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year), values_from = n) %>% kable()

bind_rows(sampling = vt %>% filter(type=="floral") %>% group_by(sampledate.file) %>% tally(),
         filenames = volsmeta %>% drop_na(sampledate.file) %>% group_by(sampledate.file) %>% tally(), .id="source") %>% 
  ggplot(aes(x=yday(sampledate.file), y=n, fill=source)) + geom_col(position=position_dodge(preserve="single")) +
  facet_wrap(vars(year(sampledate.file)), ncol=1) +
  labs(x="VOC sampling date (day of year)", y="Floral samples")
```

## Matching datasheets and filenames

```{r matching}
# diagnostics for matching up vt (datasheets) and volsmeta (filenames)
volsmeta.samp <- volsmeta %>% mutate(samp = paste(sampledate.file, plot.file, subplot.file, plant.file)) %>% arrange(samp) %>% pull(samp)
vt.samp <- vt %>% mutate(samp = paste(sampledate.file, plot.file, subplot.file, plant.file)) %>% arrange(samp) %>% pull(samp)
#intersect(volsmeta.samp, vt.samp) #in both volsmeta and vt
cat("Have files but not datasheet entries:")
setdiff(volsmeta.samp, vt.samp)
cat("Have datasheet entries but no files:")
setdiff(vt.samp, volsmeta.samp)

vt.volsmeta <- full_join(vt %>% select(vt.n, sampledate.file, plot.file, subplot.file, plant.file) %>% mutate(in.vt=T),
                         volsmeta %>% select(FileName, sampledate.file, plot.file, subplot.file, plant.file) %>% mutate(in.volsmeta=T)) %>% 
  mutate(across(c(in.vt, in.volsmeta), ~replace_na(.,FALSE)))
vt.volsmeta.count <- vt.volsmeta  %>% count(sampledate.file, in.vt, in.volsmeta) %>% mutate(status=paste("datasheet",ifelse(in.vt,"yes","no"),"+","GCfile", ifelse(in.volsmeta,"yes","no")))
#View(vt.volsmeta.count %>% filter(!in.vt | !in.volsmeta) %>% arrange(sampledate.file))

ggplot(vt.volsmeta.count, aes(x=as.Date(sampledate.file),y=n, fill=status))+ geom_col() +
  facet_wrap(vars(year(sampledate.file)), scales="free_x", ncol=1) +
  scale_x_date(breaks=scales::date_breaks(width = "7 days"), minor_breaks = scales::date_breaks(width = "1 days")) +
  theme_minimal() + theme(legend.position="top") + scale_fill_manual(values=c("red","orange","darkgreen")) +
  labs(x="Sampling date (according to GCfile)", y="Samples", fill="Data status")

ggplot(volsmeta %>% filter(year %in% c(2018, 2020) | sequence.year==year), aes(x=sequence.date - sampledate)) + 
  geom_histogram() + facet_wrap(vars(year), ncol=1, scales="free_x") + theme_minimal()
```


# Soil moisture

```{r sm, fig.height=5,fig.width=5}
ggplot(vt %>% drop_na(water) %>% mutate(date=sampledate), aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol = 1) +
  scale_x_continuous(limits=c(160, 230)) +
  geom_col(data=daily_precip_est, aes(y=EPA_NOAA_filled), fill=brewer.pal(7, "Set3")[7]) +
  geom_point(aes(y=VWC, color=water), size=0.5) + 
  geom_line(aes(y=VWC, color=water),
            data=vt %>% drop_na(water, VWC) %>% group_by(year, water, date=sampledate) %>% summarize(VWC=mean(VWC)), size=1) + 
  geom_line(aes(y=VWC, color=water),
            data=sm %>% group_by(year, water, date) %>% summarize(VWC=mean(VWC)), size=0.5, alpha=0.5) + 
  scale_color_manual(values=water_pal) + 
  labs(x="Day of year", y="Soil moisture (%VWC) | Daily precipitation (mm)", color="Precipitation") + theme_bw()
```

# Ambient controls
```{r ambients}
vt.ambi %>% filter(type=="ambient") %>% mutate(year=year(sampledate)) %>% 
ggplot(aes(x=yday(sampledate))) + geom_bar() + facet_wrap(vars(year), ncol=1) + 
  labs(x="Day of year", y="Ambient samples") + scale_fill_manual(values=water_snow_pal)+  theme_minimal() 

print("dates missing ambient controls")
setdiff(vt.ambi %>% filter(type=="floral") %>% pull(sampledate) %>% unique, 
        vt.ambi %>% filter(type=="ambient") %>% pull(sampledate) %>% unique) %>% as.Date(origin=ymd("1970-01-01"))

vt.ambi %>% count(type) #TODO check if all of these were run and used for filtering
volsmeta %>% count(year, type)
volsmeta %>% count(type)

ggplot(vt.ambi %>% filter(type=="ambient"), aes(x=hour(bag), fill=year)) + 
  geom_histogram(show.legend = F, binwidth=1) + facet_wrap(vars(year), ncol=1) + 
  scale_fill_manual(values=year_pal) + theme_minimal() + 
  labs(x="Ambient control bagged (hr)", caption= paste(format(c(min(volsmeta$bag, na.rm=T), max(volsmeta$bag, na.rm=T)), "%H:%M"), collapse=" - "))
```


# GCMS data checks

## Bouquet

```{r bouquetprep}
# load short names and standard regessions
ipochems <- read_csv("data/volatiles/Ipo volatile compounds - chemsf_ipo.csv") %>% 
  select(name, shortname, standard, verdict) %>%  filter(verdict != "") %>% 
  mutate(standard = fct_recode(standard, "Methyl_salicylate"="Benzaldehyde"), # benzaldehyde regressions not reliable
         class = fct_recode(standard, Aliphatic="Hexenol", Benzenoid="Methyl_salicylate", Benzenoid="Indole", 
                            Sesquiterpene="Caryophyllene", Monoterpene="alpha_Pinene", Monoterpene="Linalool")) %>%  
  left_join(read_csv("data/volatiles/regressions_181921_filtered_slopes.csv") %>% 
              pivot_wider(id_cols=standard, names_from="year", names_prefix="area_per_ng", values_from=area_per_ng)) 
class_pal <- set_names(c("#BC0060","#027B8C","#E56E00","#86D400"), levels(ipochems$class))

#shorten chemical names and merge compounds with multiple names
shortnames <- ipochems %>% select(name, shortname) %>% filter(shortname!="") %>% deframe()
#shortnames[shortnames %in% shortnames[duplicated(shortnames)]]

greekify <- function(names) {
  names %>% 
    str_replace("^a-","\U03B1-") %>% str_replace("-a-","-\U03B1-") %>% 
    str_replace("^b-","\U03B2-") %>% str_replace("-b-","-\U03B2-") %>% 
    str_replace("^g-","\U03B3-") %>% str_replace("-g-","-\U03B3-")
}

metadata <- maxfmeta %>% 
  mutate(sampledate = as.Date(sampledate), year = year(sampledate)) %>% 
  filter(year != 2021) %>% mutate(year=factor(year)) %>% #drop samples collected in 2021
  as.data.frame() %>% #tibbles break bouquet's add_count_freqs
  select(-sample) %>% #name conflict with bouquet
  load_metadata(date="Date", sample="FileName", type="type")#group="year"

load("data/volatiles/maxfield210827.Rdata") #load long maxf.data
longdata <- maxf.data %>% 
  filter(batch != "heather_maxfield190813_new.txt", #duplicates - subset of heather_maxfield190813_new001.txt
         Filename %in% metadata$sample, #match list in metadata
         Name != "") %>% #drop unidentified peaks
  mutate(Name = droplevels(recode(Name, !!!shortnames))) %>% 
  load_longdata(sample="Filename", RT="Ret.Time", name="Name", area="Area", match = "SI", maxmatch=100)
remove(maxf.data)

vol.all <- make_sampletable(longdata, metadata)
chems <- make_chemtable(longdata, metadata)
```

```{r bouquet}
chemsf <- chems %>%
  filter_RT(2, 17) %>% 
  filter_match(0.8) %>% 
  filter_freq(0.1) %>% 
  filter_contaminant(cont.list = c("Butanenitrile, 2-methyl-","2-Oxepanone, 7-methyl-", #RT order
                                   "Ethylbenzene", "Hexyl chloroformate", "L-Lactic acid",
                                   "1,5-Dimethyl-6-oxa-bicyclo[3.1.0]hexane","Cyclopentanone, 2-ethyl-",
                                   "trisiloxane, 1,1,1,5,5,5-hexamethyl-3-[(trimethylsilyl)oxy]-",
                                   "Cyclobutane, 1,2-bis(1-methylethenyl)-, trans-",
                                   "N-.alpha.,N-.omega.-Di-cbz-L-arginine", "3-Ethyl-3-methylheptane",
                                   "Trichloroacetic acid, 6-ethyl-3-octyl ester", "11-Methyldodecanol",
                                   "8-Hexadecenal, 14-methyl-, (Z)-",
                                   "Tritetracontane","Carbonic acid, decyl undecyl ester",
                                   "Eicosane, 2,4-dimethyl-", "1-Decanol, 2-octyl-",
                                   "linalyl acetate", #kovats doesn't match
                                   "[(Z)-hex-3-enyl] 2-methylpropanoate",  #merge into [(E/Z)-hex-3-enyl] butanoate
                                   "[(E)-hex-4-enyl] butanoate", #merge into [(E/Z)-hex-3-enyl] butanoate
                                   "[(Z)-hex-3-enyl] 2-methylbutanoate", #passed filters, only in 50 samples after quant
                                   "2-butyloctan-1-ol",#absent in 2019
                                   "(2E)-2,7-dimethylocta-2,6-dien-1-ol" #a quant integration that coelutes with b-myrcene
                                   )) %>%
  filter_area(min_maximum = 1e5) %>% #not used for final filtering
  filter_ambient_ratio(vol.all, metadata, ratio = 4) %>%
  combine_filters() 

#see below for add_quant1 and add_quant2 to generate this list
add_quant2 <- c("(4E,6Z)-alloocimene", "methyl benzoate", "2-methylbenzonitrile", "(Z)-b-ocimene", "(2E)-2,7-dimethylocta-2,6-dien-1-ol", "cis-a-bergamotene", "g-terpinene", "trans-b-bergamotene", "(1Z,4Z,7Z)-1,5,9,9-tetramethylcycloundeca-1,4,7-triene", "pseudoionone")

chemsf$filter_final <- with(chemsf, filter_RT=="OK" & filter_match=="OK" & 
                              filter_freq.floral=="OK" & filter_contaminant=="OK" |
                             name %in% c("[(E)-hex-3-enyl] butanoate", add_quant2)) #include as counterpart to Z isomer

#now run t-tests after rarity filtering
#don't filter by ambient ratio first, see paper on problems: https://doi.org/10.1186/1471-2105-11-450
chemsf <- bind_rows(chemsf %>% filter(filter_final) %>% 
  filter_ambient_ttest(prune_sampletable(vol.all, chemsf, metadata), 
                       metadata, alpha = 0.05, adjust = "fdr"), #not used for final filtering
  chemsf %>% filter(!filter_final)) %>% 
  mutate(filter_ambient_ttest=fct_na_value_to_level(filter_ambient_ttest, "notchecked"),
         filter_final = filter_final & filter_ambient_ratio =="OK") %>% 
  arrange(match(name, colnames(vol.all)))

bouquet::plot_filters(chemsf, option="rarity", yrange=3.5)
bouquet::plot_filters(chemsf, option="ambient", yrange=3.5)
bouquet::plot_filters(chemsf, option="volcano", yrange=2.5, alpha_excluded = 0.5)
bouquet::plot_filters(chemsf, option="prop")
```

```{r quant}
#vols <- maxf[volsmeta$index,]
vols <- vol.all[volsmeta$FileName,]
vols.cut <- vols[,colSums(vols)>3e8]
rownames(vols.cut) <- volsmeta$index
vol <- vols[volsmeta$FileName, chemsf$name[chemsf$filter_final]]#don't cut ambients yet
rownames(vol) <- volsmeta$index

# Shimadzu Quantitative Table ---------------------------------------------
# source("read_shimadzu.R")
# quantpath <- "~/MyDocs/MEGA/UCI/Schiedea/Analysis/scent/rmbl/RMBL Batches/quant_round4/"
# bfiles <-  list.files(quantpath)
# quant.full <- map_dfr(set_names(paste0(quantpath, bfiles), bfiles), read.shimadzu.quant, .id="batch") %>% 
#   mutate(Name = trimws(Name), Area=replace_na(Area, 0)) %>% 
#   distinct(Filename, Name, .keep_all=T)

# quant <- quant.full %>% select(batch, Dirname, Filename, Name, Area) %>% 
#   mutate(Dirname = fct_relabel(Dirname, str_remove, "C:/GCMSsolution/Data/Project1_190815/")) %>% 
#   filter(!Name %in% c("1,6,10-Dodecatrien-3-ol, 3,7,11-trimethyl-", 
#          "4,7-Methano-1H-indene, 2,4,5,6,7,7a-hexahydro-",
#          "Tricyclo[3.1.0.0(2,4)]hexane, 3,6-diethyl-3,6-dimethyl-, trans-")) %>% 
#   mutate(Name = recode(Name, !!!shortnames)) %>% 
#   pivot_wider(names_from = "Name", values_from="Area") %>% as.data.frame()
# 
# rownames(quant) <- quant$Filename
# quant[,c("batch","Dirname","Filename")] <- NULL
# quant.all <- quant[rownames(vols),]
# write_tsv(quant.all %>% rownames_to_column("Filename"), "data/volatiles/quant_all.tsv")
quant.all <- read_tsv("data/volatiles/quant_all.tsv") %>% column_to_rownames("Filename")

#add_quant1 <- chemsf %>% filter(name %in% setdiff(colnames(quant.all), colnames(vol)), 
#                  ambient_ratio > 4) %>% pull(name) 
# setdiff(colnames(quant.all), colnames(vol))
# setdiff(colnames(vol), colnames(quant.all)) 

# pheatmap(as.matrix(t(quant.all))^(1/4), 
#          show_colnames=F,
#          clustering_method="mcquitty", clustering_distance_rows="correlation",
#          clustering_callback = function(hc, ...){dendsort(hc, type="average")},
#          scale="none", color=mako(512),main="Quant volatiles")

quant.slopes <- set_names(rep(0,ncol(quant.all)), colnames(quant.all))
for(compvol in colnames(quant.all)) {
  nozeros <- bind_cols(select(vols,qual=compvol), select(quant.all,quant=compvol)) %>% 
    filter(!(qual==0 | quant == 0))
  quant.slopes[compvol] = coef(lm(quant ~ 0 + qual, data=nozeros))["qual"]
  # quant_wins <- sum(vols[,compvol] == 0 & quant.all[,compvol] > 0, na.rm=T)
  # qual_wins <-  sum(quant.all[,compvol] == 0 & vols[,compvol] > 0, na.rm=T)
  # plot(x=sqrt(vols[,compvol]), y=sqrt(quant.all[,compvol]),
  #      main=compvol, sub = paste0("quant:qual ",quant_wins,":",qual_wins,
  #                                 " slope ",round(quant.slopes[compvol],3)),
  #      xlab="qual",ylab="quant")
  # abline(a=0, b=sqrt(quant.slopes[compvol]))
}
write_tsv(enframe(quant.slopes) %>% left_join(enframe(shortnames, name="Name", value="name")),
          "data/volatiles/quant_slopes.tsv")

quant.scaled <- sweep(quant.all, 2, quant.slopes, FUN = '/')

quant.scaled.vol <- quant.scaled[volsmeta$FileName, 
                                 colnames(quant.scaled) %in% 
                                   intersect(colnames(quant.all), colnames(vol))] 

#swap in scaled quant integrations
#plot(unlist(vol[,colnames(quant.scaled.vol)]), unlist(quant.scaled.vol))
vol[,colnames(quant.scaled.vol)] <- quant.scaled.vol

#do this after swapping in add_quant1 integrations, exclude those <100
# add_quant2 <- left_join(chemsf, sort(colSums(decostand(vol, "pa"))) %>% enframe()) %>% drop_na(value) %>%
#   filter(value>100, !filter_final) %>% pull(name)
```

```{r retention, fig.width=12}
longdata.RT <- longdata %>% filter(name %in% chemsf$name[chemsf$filter_final]) %>% 
  left_join(select(metadata, sample, year)) %>%
  group_by(name) %>%  mutate(n=n(), RT.sd=sd(RT)) %>% ungroup() %>% 
  mutate(name=fct_reorder(name, RT))

RT.year <- longdata.RT %>% group_by(name, year) %>% summarize(RT=median(RT)) %>% pivot_wider(names_from=year, values_from=RT) 
RT.year %>% filter(name %in% setdiff(colnames(vol), colnames(quant.all))) %>% arrange(`2018`)


RT.tol <- 0.17  # 10 seconds before or after in quantitative params
ggplot(longdata.RT, aes(x=RT, y=name,color=RT.sd)) + geom_point(position=position_jitter(width=0, height=0.5)) + 
  scale_color_viridis_c(option="magma") + theme_dark()+
  geom_text(aes(x=4.5, label=n), color="grey80") #+
  #stat_summary(fun="median", fun.max=~median(.x)+RT.tol, fun.min=~median(.x)-RT.tol, geom="pointrange", color="white", size=0.5, shape=15)

ggplot(longdata.RT,  aes(y=RT, x=factor(year), color=RT.sd)) + 
  facet_wrap(vars(name), scales="free_y") + geom_boxplot() + scale_color_viridis_c(option="magma") + theme_dark() #+
#  geom_pointrange(data=RT.year, aes(y=mean1821, ymin=mean1821-0.1, ymax=mean1821+0.1, x="2019"), color="blue")

#RT fixes
longdata.butanoate <- longdata %>% filter(str_detect(name , "\\)-hex-3-enyl\\] (2-methylpropa|buta)noate")) %>% 
    left_join(select(metadata, sample, year)) 
ggplot(longdata.butanoate, aes(x=RT, fill=factor(year))) + geom_histogram(binwidth=0.02) + 
  facet_wrap(vars(name), ncol=1) + scale_x_continuous(limits=c(NA,NA))
longdata.butanoate %>% mutate(EZ=ifelse(RT>11, "Z","E")) %>% group_by(EZ, year) %>% 
  summarize(RT=median(RT)) %>% kable(digits=2)

longdata.pentanol <- longdata %>% filter(str_detect(name , "^(3-methylbutan-1-ol|pentan-1-ol)")) %>% 
    left_join(select(metadata, sample, year)) 
ggplot(longdata.pentanol, aes(x=RT, fill=factor(year))) + geom_histogram(binwidth=0.02) + 
  facet_wrap(vars(name), ncol=1) + geom_vline(xintercept=3.75)
```

```{r diverse, eval=FALSE}
tointegrate <- c("hexan-1-ol","sabinene","3-carene","verbenone","[(Z)-hex-3-enyl] 2-methylbutanoate","2-butyloctan-1-ol","b-caryophyllene oxide")
best <- list()
i <- 1
vol.left <- vol[,tointegrate]^(1/4) # halfway between areas and presence-absence
while(ncol(vol.left)>0) {
  print(dim(vol.left))
  best[[i]] <- which.max(rowSums(vol.left))
  vol.left <- vol.left[ , !decostand(vol.left[best[[i]],], method="pa"), drop=FALSE]
  i <- i+1
}
best.vol <- as.data.frame(t(vol[unlist(best),tointegrate]))
colnames(best.vol) <- metadata[colnames(best.vol),"sample"]
best.vol
```

## Filtering

This section is outdated, see bouquet filters above.

OLD Filtering criteria (verdict="fvoc" in the table below):

* Present in 10% of samples
* Maximum peak area is at least 4000000
* Passes t-test against ambient controls with FDR correction OR mean exceeds 4x mean of controls
* Not a known contaminant (Oxybenzone, Ethanol, 2,2'-oxybis-, Butylated Hydroxytoluene, Caprolactam, [1,1'-Bicyclopentyl]-2-one, L-Lactic acid, Pentadecane, 8-hexyl-, Decane, 2,8,8-trimethyl-, 2-Methyltetradecan-2-ol)

Note that these potential floral compounds did not meet the tests against ambient controls: 2-phenylacetaldehyde, 3-carene, D-limonene, camphene, benzaldehyde, a-thujene, sabinene, abietadiene, 2-ethylhexan-1-ol 

Plus (verdict other than "fvoc"): 

* bis14: compounds included in Bischoff et al. (2014)
* keep: kept since they are known floral volatiles
* keep_leaf: kept since they are known leaf volatiles
* merge: a compound that is a synonym of one of the floral volatiles (b-caryophyllene, b-ocimene, verbenone)

```{r filtering}
# ipochems <- read_csv("data/volatiles/Ipo volatile compounds - chemsf_ipo.csv") %>% 
#   filter(verdict %in% c("bis14", "fvoc", "keep","keep_leaf","merge"),
#          name %in% colnames(vols)[colSums(vols)>0], name != "Anisole") %>% #filter empty compounds
#   mutate(standard = fct_recode(standard, "Methyl_salicylate"="Benzaldehyde"), # benzaldehyde regressions not reliable
#          class = fct_recode(standard, Aliphatic="Hexenol", Benzenoid="Methyl_salicylate", Benzenoid="Indole", 
#                             Sesquiterpene="Caryophyllene", Monoterpene="alpha_Pinene", Monoterpene="Linalool")) %>%  
#   left_join(read_csv("data/volatiles/regressions_181921_filtered_slopes.csv") %>% 
#               pivot_wider(id_cols=standard, names_from="year", names_prefix="area_per_ng", values_from=area_per_ng)) 
# class_pal <- set_names(c("#BC0060","#027B8C","#E56E00","#86D400"), levels(ipochems$class))
# 
# ipochems %>% select(verdict, name, Ret.Time, mean, max, match, freq.floral, freq.ambient, ambient_ratio, filter_ambient_ttest, filters_passed, standard) %>%
#   arrange(desc(freq.floral)) %>% mutate(across(starts_with("freq."), round, 3)) %>% 
#   DT::datatable(caption="Volatile compounds to use for analysis")

#ipogood <- ipochems$name
ipogood <- colnames(vol)
#vols.good <- vols[,ipogood] %>% 
vols.good <- vol %>% 
  #sweep(2, ipochems$area_per_ng, FUN = '/') %>%  # convert areas to nanograms 
  sweep(1, as.numeric(volsmeta$total_time, units="hours"),  FUN = '/') #divide by equilibration + pumping time
ipochemsf <- ipochems[match(colnames(vols.good), ipochems$shortname),]
for(yr in unique(volsmeta$sequence.year)) {
  thisyr <- volsmeta$sequence.year==yr
  #TODO these slopes are leading to most of the year effects. try using one year for now. 
  vols.good[thisyr,] <- sweep(vols.good[thisyr,], 2, pull(ipochemsf, paste0("area_per_ng","2019")), FUN = '/')# convert areas to nanograms 
}

# merge peaks identified as multiple compounds
# give volatiles shorter names too
# vols.good <- left_join(ipochems %>% select(name, shortname), 
#          vols.good %>% t %>% as.data.frame %>% rownames_to_column("name")) %>% 
#   select(-name) %>% group_by(shortname) %>% 
#   summarize(across(everything(), sum)) %>% 
#   column_to_rownames("shortname") %>% t %>% as.data.frame

ipochemsf <- ipochemsf %>% #group_by(name=shortname, verdict, standard, class) %>%
  #summarize(.groups = "drop") %>%
  left_join(tibble(shortname = colnames(vols.good), # the following values are new, computed on the filtered dataset
                            freq.floral = map_dbl(as.data.frame(vols.good>0), mean),
                            mean.floral = map_dbl(vols.good, mean),
                            max.floral =  map_dbl(vols.good, max))) %>% 
  select(-name) %>% rename(name=shortname) %>%
  mutate(class=droplevels(class))#no benzenoids
rownames(ipochemsf) <- ipochemsf$name
#ipogood == ipochemsf$name

k <- 2 #split into this many k-means cluster per year
volsmeta$cluster <- 0
for(yr in unique(year(volsmeta$sequence.start))) {
  thisyr <- volsmeta$sequence.year==yr
  chems.include <- setdiff(colnames(vols.cut), ipogood) #drop alpha-pinene so this is all nonfloral
  set.seed(1)
  volsmeta$cluster[thisyr] <- kmeans(decostand(vols.cut[thisyr,chems.include], method="log"), k, nstart=3)$cluster
}

volsmeta <- volsmeta %>% 
  mutate(year.cluster = paste0(year, if_else(sequence.year == 2018, "", paste0(".",cluster))),
         total = rowSums(vols.good),
         sampledate = coalesce(sampledate, sampledate.file))#get sampling date from filename when no metadata

# Cut contaminated samples
#nonzero <-rowSums(vols.good) > 0 #one sample ("55" on row 44) has no volatiles
# set.seed(1) # ensures the contaminated samples are in cluster.good = 1
# volsmeta$cluster.good <-factor(kmeans((vols.good)^(1/3), centers=3, nstart=3)$cluster)
# set.seed(1) # ensures the contaminated samples are in cluster.cut = 1
# volsmeta$cluster.cut <-factor(kmeans((vols.cut)^(1/3), centers=3, nstart=3)$cluster)

#Cuts 2 samples with extemely low emissions that fly off of nmds blob
keep.samples <- rowSums(vols.good)>3#nonzero#unname(nonzero & !(volsmeta$year.cluster %in% c("2019.1", "2020.1")))
volsmeta <- volsmeta[keep.samples,]
vols <- vols[keep.samples,]
vols.cut <- vols.cut[keep.samples,]
vols.good <- vols.good[keep.samples,]
```

## Averaging 

```{r averaging}
volsmeta %>% count(year, plantid, name="n_samples") %>% count(n_samples) %>% kable(caption="number of samples per plant")

volsmeta.avg <- volsmeta %>% 
  group_by(across(all_of(c("type", "plantid", colnames(treatments))))) %>% 
  summarize(across(all_of(c("VWC", "total")), ~mean(.x, na.rm=T)), 
            index=index[1], .groups="drop")

vols.good.avg <- vols.good %>% bind_cols(volsmeta) %>% 
  group_by(across(all_of(c("plantid", colnames(treatments))))) %>% 
  summarize(across(all_of(colnames(vols.good)), ~mean(.x, na.rm=T)), .groups="drop") %>% 
  select(-any_of(colnames(volsmeta)))
rownames(volsmeta.avg) <- rownames(vols.good.avg) <- volsmeta.avg$index
```


```{r topvols}
minsamples <- nrow(vols.good) * 0.75 #top 12 most common volatiles, in 75% of samples
ipogood.top <- vols.good[colSums(decostand(vols.good,"pa")) > minsamples] %>% 
  colSums() %>% sort(decreasing=T) %>% names()
vols.good.top <- vols.good.avg[,ipogood.top]

volsmeta.good.top <- bind_cols(volsmeta.avg, vols.good.top) %>%  
  pivot_longer(-all_of(colnames(volsmeta.avg))) %>% 
  mutate(name = fct_relevel(name, ipogood.top))

mnps.plantyr.traits <- mnps.plantyr %>% 
  select(year, plotid, plantid, all_of(traits), eggs_per_flower) %>%
  mutate(prop_uninfested = 1 - prop_infested) %>% 
  mutate(plantid = str_remove(plantid, "_[12][890]")) %>% #disambiguations on the plantid already handled by year column, except...
  group_by(year, plotid, plantid) %>% #...two plants need to be combined in 2019
  summarize(across(everything(), ~.x[!is.na(.x)][1]), .groups = "drop")

traits.good.top <- volsmeta.good.top %>% 
  left_join(mnps.plantyr.traits) %>% 
  mutate(name = fct_relevel(name, ipogood.top))

vols.class <- vols.good.avg %>% t() %>% as.data.frame() %>% rownames_to_column("name") %>% 
  left_join(select(ipochemsf, name=name, class)) %>% 
  group_by(class) %>% summarize(across(-name, sum)) %>% column_to_rownames("class") %>%  t() %>% as.data.frame()
volsmeta.class <- bind_cols(volsmeta.avg, vols.class) %>%  
  pivot_longer(-all_of(colnames(volsmeta.avg)))
traits.class    <- volsmeta.class %>% left_join(mnps.plantyr.traits)
volsmeta.traits <- volsmeta.avg %>%       left_join(mnps.plantyr.traits) #wide format
vols.traits <- bind_cols(volsmeta.avg, vols.good.avg) %>% left_join(mnps.plantyr.traits)
```


## Compound amounts

```{r relamounts}
vols.good %>% decostand(method = "tot") %>% #relative amount in each sample
  pivot_longer(everything()) %>% group_by(name) %>% summarize(mean_rel = mean(value)) %>% 
  arrange(desc(mean_rel)) %>% kable(caption = "Mean relative amounts of each volatile")

vols.good.avg %>% #average by plant first
  pivot_longer(everything()) %>% group_by(name) %>% summarize(mean_emissions = mean(value), sd_emissions=sd(value)) %>% 
  left_join(ipochemsf) %>%  arrange(class, desc(mean_emissions)) %>% 
  mutate(meansd = paste(sprintf(mean_emissions, fmt = '%#.1f'), "\U00B1", sprintf(sd_emissions, fmt = '%#.1f')),
         name=greekify(name)) %>% 
  select(class, name, meansd) %>%  kable(caption = "Mean emission rates of each volatile")
```


## Sample size after filtering

```{r samplesizes_filtered, fig.width=10, fig.height=5}
volsmeta %>% count(year, name="samples") %>% 
  left_join(volsmeta %>% count(year, plantid) %>% count(year, name="plants")) %>% 
  kable(caption = "Samples and plants by year")
volsmeta %>% count(year, water, snow) %>% 
  pivot_wider(names_from=c(year), values_from = n) %>% 
  kable(caption = "Samples by year and treatment")
volsmeta %>% count(year, sampledate) %>% 
  kable(caption = "Samples by date")
volsmeta %>% count(year, sampledate) %>% count(year) %>% 
  kable(caption="Number of sampling dates")

volsmeta %>% count(year, plantid) %>% ggplot(aes(x=n, fill=year)) + facet_wrap(vars(year), ncol=1) +geom_histogram(binwidth=1) +
  xlab("samples per plant") + scale_fill_manual(values=year_pal, guide="none") + theme_minimal()
```

## Time of day

```{r sample_times}
ggplot(volsmeta, aes(x=bag, fill=year)) + geom_histogram(show.legend = F) + facet_wrap(vars(year), ncol=1) + 
  scale_fill_manual(values=year_pal) + theme_minimal() + 
  labs(x="Flower bagged", caption= paste(format(c(min(volsmeta$bag, na.rm=T), max(volsmeta$bag, na.rm=T)), "%H:%M"), collapse=" - "))
ggplot(volsmeta, aes(x=bag+total_time, fill=year)) + geom_histogram(show.legend = F) + facet_wrap(vars(year), ncol=1) + 
  scale_fill_manual(values=year_pal) + theme_minimal() + 
  labs(x="Pump stopped", caption= paste(format(c(min(volsmeta$bag+volsmeta$total_time, na.rm=T), max(volsmeta$bag+volsmeta$total_time, na.rm=T)), "%H:%M"), collapse=" - "))

volsmeta %>% count(beforenoon=hour(bag+total_time) <13)

ggplot(volsmeta, aes(x=end-bag)) + geom_histogram() + labs(x="Equilibration and pumping time (min)")
```

## Distributed among treatments

```{r random_sampling, fig.width=5, fig.height=4}
ggplot(volsmeta, aes(x=yday(sampledate), fill=paste(water, snow))) + geom_bar() + facet_wrap(vars(year), ncol=1) + 
  labs(fill="Precipitation, snowmelt", x="Day of year", y="Samples") + scale_fill_manual(values=water_snow_pal)+ theme_minimal() 

list(water=chisq.test(with(vt, table(sampledate, water))),
     snow=chisq.test(with(vt, table(sampledate, snow))),
     water_snow=chisq.test(with(vt, table(sampledate, paste(water, snow))))) %>% 
  map_dfr(tidy, .id="treatment") %>% kable(caption="Test of randomness of treatments on each sampling date")
```

## Flowers and buds

```{r buds}
volsmeta %>% count(flowers)
volsmeta %>% count(buds>0)
cat(paste("Buds included in", round(sum(volsmeta$buds>0, na.rm=T)/sum(volsmeta$buds>=0, na.rm=T)*100), "% of samples"))
```

## Heatmap

```{r heatmap, dev="png", dev.args=list(),fig.width=16, fig.height=12}
annot_pal <- list(
  sequence.date = volsmeta %>% mutate(yr = sequence.year) %>% group_by(yr) %>% summarize(n = length(unique(sequence.date))) %>% 
    pull(n) %>% map(scales::hue_pal()) %>% unlist() %>% set_names(unique(volsmeta$sequence.date)),
  cluster = set_names(brewer.pal(8, "Set1")[1:length(unique(volsmeta$cluster))], unique(volsmeta$cluster)))

pheatmap(as.matrix(t(vols.cut))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),main="Unfiltered top volatiles",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors =  annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))

pheatmap(as.matrix(t(vols.good))^(1/4), 
         cluster_cols=T, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),main="Filtered volatiles",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors = annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))

pinene_correlates <- cor(sqrt(vols[,colSums(vols)>1e7])) %>% as.data.frame() %>%
  filter(`a-pinene`>0.2) %>% rownames()
pheatmap(as.matrix(t(vols[,pinene_correlates]))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),
                  annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors = annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0), main="Volatiles that correlate with alpha-pinene")

pheatmap(as.matrix(t(vols.class))^(1/4), 
         cluster_cols=F, cluster_rows=F, show_colnames=F,
         scale="none", color=mako(512),main="Filtered volatiles classes",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors =  annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))
```

## Chromatograms by GCMS run date

### Univariate

```{r gcms_date, fig.height=8, fig.width=7, eval=FALSE}
plot_vols_rundate <- function(type, checkvols) {
volsmeta %>% mutate(yr=year(Desorb.Start.Time)) %>% 
  bind_cols(vols %>% select(checkvols)) %>% pivot_longer(cols=checkvols) %>% 
ggplot(aes(x=Desorb.Start.Time)) + facet_grid(name ~ yr, scales="free") + 
  geom_point(aes(y=value, color=name), alpha=0.3)  + 
  guides(color="none") + theme_bw() + labs(x="GCMS run time", y="Area", title=type)
}

plot_vols_rundate("Terpenes", c("a-pinene", "b-myrcene","(E)-b-ocimene","b-caryophyllene","linalool","longifolene"))

plot_vols_rundate("Benzenoids", c("benzophenone","methyl salicylate","anisole"))

plot_vols_rundate("Green leaf volatiles", c("(E)-hex-3-en-1-ol"))
```

### Multivariate

```{r gcms_sample_date, fig.height=6, fig.width=10, eval=FALSE}
for(categories in list(c("water","snow"),"plot","observer","pump_id","sequence.12", "sampledate", c("sampledate", "sequence.12"), c("sampledate.observer", "sequence.12"))) {
  cluster_counts <- volsmeta %>% 
    mutate(sampledate = coalesce(sampledate, sampledate.file),
           sampledate.observer = paste(sampledate, observer)) %>% 
    count(!!!syms(c("sequence.year", categories, "cluster")))
  p <- ggplot(cluster_counts, aes(x=factor(paste(!!!syms(categories), sep=" x ")), 
                                  y=n, fill=factor(paste0(sequence.year," C",cluster)))) + 
    geom_col() + facet_wrap(vars(sequence.year), scales="free_x") +
    theme(axis.text.x = element_text(angle=90)) + labs(fill="Cluster", x=paste(categories, collapse=" x ")) +
    scale_fill_brewer(palette="Paired")
  print(p)
  
  if(length(categories)>1) {
    (p2 <- cluster_counts %>% filter(sequence.year !="2018") %>% mutate(cluster=factor(cluster), across(where(~ !is.numeric(.)),factor)) %>% 
     ggplot(aes_string(x=categories[[1]], y = categories[[2]], size="n", color="cluster")) + geom_point(shape=1, stroke=1) +
      facet_wrap(vars(sequence.year), scales="free") + theme_minimal()+ scale_size_area(max_size=15)+
      theme(axis.text.x = element_text(angle=90)) + scale_color_manual(values=c("darkgreen","magenta")))
    print(p2)
  }
}

ggplot(volsmeta, aes(x=Desorb.Start.Time, y=cluster))  + geom_line() + geom_point(aes(color=factor(mday(sampledate))))+
  facet_wrap(vars(sequence.year), scales="free_x", ncol=1) + scale_color_manual(values=sample(rainbow(30)))+
  scale_x_datetime(date_breaks="1 day", guide=guide_axis(angle=90)) + scale_y_discrete()
```


```{r gcms_date_ord, eval=FALSE}
for(yr in unique(year(volsmeta$sequence.start))) {
  print(yr)
  yrfilter <- year(volsmeta$sequence.start) == yr
  vols.cap <- capscale(sqrt(vols.cut[yrfilter,]) ~ factor(date(sequence.start)), data=volsmeta[yrfilter,],
                       distance="bray")
  print(vols.cap)
  print(anova(vols.cap, by="term"))
  # vols.cap.points <-  fortify(vols.cap) %>% as_tibble()
  # vols.cap.samples <- vols.cap.points %>% filter(Score=="sites") %>% 
  #   left_join(volsmeta, by=c("Label"="index"))
  # vols.cap.chems <- vols.cap.points %>% filter(Score=="species")
  # p <- vols.cap.samples %>% 
  #   mutate(sequence.date = date(sequence.start)) %>% 
  #   filter(sequence.year == yr) %>% 
  #   ggplot(aes(x=CAP1, y=CAP2)) + theme_classic()+
  #   geom_point(aes(color=factor(sequence.date)), size=3) + scale_color_hue("Run date")
  # print(p)
}
```

## Ordinations

### NMDS

```{r nmds, fig.height=9, fig.width=6}
(vols.nmds <- metaMDS(sqrt(vols.cut), autotransform=FALSE, trace=F))
vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
  left_join(volsmeta, by=c("Label"="index"))
vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")

(scents.nmds <- metaMDS(sqrt(vols.good), autotransform=FALSE, trace=F)) #had noshare = FALSE
nmds.points <-  fortify(scents.nmds) %>% as_tibble()
nmds.samples <- nmds.points %>% filter(Score=="sites") %>% left_join(volsmeta, by=c("Label"="index"))
nmds.chems <- nmds.points %>% filter(Score=="species") %>% left_join(ipochemsf, by=c("Label"="name"))

# By year
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() +
  geom_point(aes(color=year)) + scale_color_manual(values=year_pal)+
  labs(title="Unfiltered NMDS", color="Year")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  #geom_text(data=nmds.chems, aes(label=Label)) +
  geom_point(aes(color=year))+ scale_color_manual(values=year_pal)+
  #geom_point(aes(color=water, shape=year)) + scale_color_manual("Precipitation", values=water_pal) +
  labs(title="Filtered NMDS", color="Year")
grid.arrange(vols.plot, vols.good.plot)

# By a-Pinene
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2, shape=year)) + theme_classic() +
  geom_point(aes(color=vols.good$`a-pinene`^(1/6))) + 
  labs(title="Unfiltered NMDS", color="a-Pinene", shape="Year") + scale_color_viridis_c()
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2, shape=year)) + theme_classic() + 
  geom_point(aes(color=vols.good$`a-pinene`^(1/6)))+ 
  labs(title="Filtered NMDS", color="a-Pinene", shape="Year")+ scale_color_viridis_c()
grid.arrange(vols.plot, vols.good.plot)
```

```{r nmds_chems, fig.height=12, fig.width=12}
ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=year))+ scale_color_manual(values=year_pal)+
    geom_text(data=nmds.chems, aes(label=Label, alpha=freq.floral^(1/3))) +
  labs(title="Filtered NMDS", color="Year")
```


```{r nmds_clusters, eval=FALSE}
#Cluster based on filtered data
volsmeta %>% count(cluster.good, year) %>% arrange(year) %>% 
  pivot_wider(names_from="year", values_from="n") %>% 
  kable(caption = "Contaminated samples cluster = 1 based on filtered data")
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.good, year)))+ 
  labs(title="Unfiltered NMDS by year", color="Filtered cluster")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.good, year)))+ 
  labs(title="Filtered NMDS by year", color="Filtered cluster")
grid.arrange(vols.plot, vols.good.plot)

#cluster based on unfiltered data
set.seed(1)
volsmeta$cluster.cut <-factor(kmeans((vols.cut)^(1/3), centers=3, nstart=3)$cluster)
volsmeta %>% count(cluster.cut, year) %>% arrange(year) %>% 
  pivot_wider(names_from="year", values_from="n") %>% 
  kable(caption = "Contaminated samples cluster = 1 based on unfiltered data")
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.cut, year)))+ 
  labs(title="Unfiltered NMDS by year", color="Unfiltered cluster")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.cut, year)))+ 
  labs(title="Filtered NMDS by year", color="Unfiltered cluster")
grid.arrange(vols.plot, vols.good.plot)

# Ordinations for each year
for(yr in unique(year(volsmeta$sequence.start))) {
  thisyr <- year(volsmeta$sequence.start)==yr
  (vols.nmds <- metaMDS(sqrt(vols.cut[thisyr,]), autotransform=FALSE, trace=F))
  vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
  vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
    left_join(volsmeta, by=c("Label"="index")) %>% filter(sequence.year == yr)
  vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")
  
  (scents.nmds <- metaMDS(sqrt(vols.good[thisyr,]), autotransform=FALSE, trace=F))
  nmds.points <-  fortify(scents.nmds) %>% as_tibble()
  nmds.samples <- nmds.points %>% filter(Score=="sites") %>% 
    left_join(volsmeta, by=c("Label"="index")) %>% filter(sequence.year == yr)
  nmds.chems <- nmds.points %>% filter(Score=="species")
  
  vols.plot <- vols.nmds.samples %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Unfiltered NMDS run",yr))+
    geom_text(aes(color=factor(sequence.date), label=cluster), size=4) + scale_color_hue("Run date")
  vols.good.plot <- nmds.samples %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Filtered NMDS run",yr))+
        geom_text(aes(color=factor(sequence.date), label=cluster), size=4)  + scale_color_hue("Run date")
  grid.arrange(vols.plot, vols.good.plot)
}
```

## Correlations

Heatmap of Pearson correlations between compound emission rates, for compounds present in > 10% of samples. Positive correlations are shown in red and negative correlations in blue. For each compound, the frequency in floral samples, retention time in minutes, and compound class are shown in the sidebars. The dendrogram is a hierarchical clustering of the compound correlations.

```{r corr, fig.height=8, fig.width=10}
corr.annot <- ipochemsf %>% select(name, Class=class, Frequency=freq.floral) %>% #`Retention time`=RT,
  column_to_rownames("name") %>%  
  filter(Frequency>0.1)
corr.vols <- cor(vols.good[, rownames(corr.annot)], method="pearson")
pheatmap(corr.vols, 
         scale="none", clustering_method = "mcquitty",
         color=colorRampPalette(c("blue","white","red"))(200), breaks=seq(-1,1,by=0.01),
         annotation_col=corr.annot, annotation_row=corr.annot, 
         annotation_colors = list(Class=class_pal),
         labels_row=greekify(rownames(corr.vols)), labels_col=greekify(colnames(corr.vols)))

(corr.vols*lower.tri(corr.vols)) %>% as.data.frame() %>% rownames_to_column("name1") %>% 
  pivot_longer(cols=-name1, values_to="cor") %>% arrange(desc(cor)) %>% filter(cor!=1, cor!=0) %>% 
  mutate(cor.cut = cut_width(cor, width=0.25, boundary=0)) %>% count(cor.cut)

```


# Plasticity analyses

## Plasticity to treatments

### CAP

```{r cap}
(scents.cap <- capscale(sqrt(vols.good.avg) ~ year * water * snow, data=volsmeta.avg, distance="bray"))
anova(scents.cap, by="term") %>% kable()

scents.cap <- capscale(sqrt(vols.good.avg) ~ year + water, data=volsmeta.avg, distance="bray")
plot(scents.cap, type="n")
points(scents.cap, display = "sites", col=water_pal[volsmeta.avg$water], pch=as.integer(volsmeta.avg$year)+14)
text(scents.cap, display="cn", labels=c(levels(volsmeta.avg$year), names(water_pal)), col=c(rep("black",3),water_pal), font=2, cex=1.5)

plot(scents.cap, type="n")
text(scents.cap, display="species", col="black", cex=0.5)

scents.cap <- capscale(sqrt(vols.good.avg) ~ year + snow, data=volsmeta.avg, distance="bray")
plot(scents.cap, type="n")
points(scents.cap, display = "sites", col=snow_pal[volsmeta.avg$snow], pch=as.integer(volsmeta.avg$year)+14)
text(scents.cap, display="cn", labels=c(levels(volsmeta.avg$year), names(snow_pal)), col=c(rep("black",3),snow_pal), font=2, cex=1.5)

plot(scents.cap, type="n")
text(scents.cap, display="species", col="black", cex=0.5)
```

### Joint species distribution model

```{r jsdm_splitplot, eval=F}
library(Hmsc)

# model volatile abundance (emission rates) plasticity to year, water, and snow
# include random effect of plot
model.abu <- Hmsc(Y=vols.good,
                XData = volsmeta %>% select(year, water, snow) %>% mutate_all(as.factor), 
                XFormula= ~ year + water + snow, 
                studyDesign = volsmeta %>% select(plot) %>% mutate_all(as.factor),
                ranLevels = list(plot = HmscRandomLevel(units = levels(treatments$plot))),
                distr="lognormal poisson")

# Fit model with MCMC
thin <- 1
(model.abu <- sampleMcmc(model.abu, thin = thin, samples = 100, transient = thin*50, nChains = 3, nParallel = 3,
                        initPar="fixed effects")) #chains start from ML solution to fixed-effects model
```

```{r jsdm_splitplot_results, eval=F}
# Model convergence
mpost <- convertToCodaObject(model.abu)[c("Beta","Gamma","V","Sigma")]
ess <- map(mpost, effectiveSize)
par(mfrow=c(2,2))
walk(names(ess), ~hist(ess[[.x]], xlab = expression("Effective sample size" ~ beta ~ ""), main=.x))
psrf <- map(mpost, gelman.diag, multivariate=FALSE)
walk(names(psrf), ~hist(psrf[[.x]]$psrf, xlab = expression("Potential scale reduction factor"), main=.x))

# Parameters
postBeta = getPostEstimate(model.abu, parName="Beta")
plotBeta(model.abu, post=postBeta, param="Support", supportLevel = 0.95, 
         spNamesNumbers = c(T,F), covNamesNumbers = c(T,F))

Gradient = constructGradient(model.abu,focalVariable = "water")
predY = predict(model.abu,Gradient = Gradient, expected = TRUE)
prob = c(0.25,0.5,0.75)
plotGradient(model.abu, Gradient, pred=predY, measure="S", showData = TRUE, q = prob, ylabel="Total emissions") 

Gradient = constructGradient(model.abu,focalVariable = "snow")
predY = predict(model.abu,Gradient = Gradient, expected = TRUE)
plotGradient(model.abu, Gradient, pred=predY, measure="S", showData = TRUE, q = prob, ylabel="Total emissions")

# Model fit (explanatory power)
predY = computePredictedValues(model.abu)
MF = evaluateModelFit(hM=model.abu, predY=predY)
tibble(name = ipogood, SR2 = MF$SR2) %>% mutate(name=fct_reorder(name, SR2)) %>% 
  filter(SR2>0.05) %>% 
  ggplot(aes(y=name, x=SR2))+ geom_point() + xlim(c(0,NA))+
  labs(x="squared Spearman correlation between obs and pred values (SR2)", 
       title="Volatiles with highest R2 in plasticity model") + theme_minimal()

# cross validation (predictive power)
partition = createPartition(model.abu, nfolds = 2)
#predY = computePredictedValues(model.abu, partition = partition, nParallel = 3)
#MFCV = evaluateModelFit(hM=model.abu, predY=predY)
#plot(MF$SR2, MFCV$SR2)
#abline(0,1)
```

### Boxplots

```{r boxplots, fig.width=16, fig.height=10}
treatment_boxplots <- function(volsmeta.vol, treatment) {
  ggplot(volsmeta.vol %>% 
           mutate(water_snow = factor(paste(water, snow), levels=names(water_snow_pal))),
         aes_string(x = "year", fill = treatment, y = "value")) + 
    facet_wrap(vars(name), scales="free_y", labeller = as_labeller(greekify)) + geom_boxplot(outlier.size=0.2) + scale_y_sqrt() +
    scale_fill_manual(values=list(water=water_pal, snow=snow_pal,
                                  water_snow=water_snow_pal)[[treatment]]) + theme_bw() + 
    labs(x="Year", y ="Emissions (ng/hr/flower)", 
         fill=c(water="Precipitation",snow="Snowmelt", 
                water_snow="Precipitation, snowmelt")[[treatment]]) 
}
treatment_boxplots(volsmeta.good.top, "water_snow")
```

### Emissions by class

```{r boxplots_class, fig.width=7, fig.height=5}
treatment_boxplots(volsmeta.class, "water_snow")

options(contrasts = c("contr.sum", "contr.poly"))
treatment_effects <- function(class) {
  lmer(sqrt(value) ~ year * water * snow + (1|plot) + (1|plotid), 
       data = filter(volsmeta.class, name==class))
}

mod.split <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, treatment_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.split %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to treatments")
```

## Plasticity to snowmelt date and precipitation

### CAP

```{r cap_abs, fig.height=6.5, fig.width=6.5}
(scents.cap <- capscale(sqrt(vols.good.avg) ~ sun_date * precip_est_mm, data=volsmeta.avg, distance="bray"))
anova(scents.cap, by="term") %>% kable()

library(ggvegan)
scents.cap <- update(scents.cap, ~ sun_date + precip_est_mm)
cap.df <- fortify(scents.cap) #drop non-signif interaction for plotting
prop.expl <- summary(scents.cap)$cont$importance
ax.labs <- paste0(colnames(prop.expl)[1:2], " (",round(100*prop.expl[2,1:2]),"% explained)")
ggplot(mapping=aes(x=CAP1, y=CAP2, label=greekify(Label))) + 
  geom_point(data=cap.df %>% filter(Score=="sites") %>% bind_cols(volsmeta.avg), aes(color=year))+
  geom_segment(data=cap.df %>% filter(Score=="biplot"), aes(x=0,y=0,xend=CAP1*1.8, yend=CAP2*1.8), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(data=cap.df %>% filter(Score=="biplot") %>% 
              mutate(Label=c("Snowmelt", "Precipitation")),
            aes(x=CAP1*2, y=CAP2*2, angle=atan(CAP2/CAP1)*180/pi), hjust=c(0,1), fontface="bold") +
  geom_text(data=cap.df %>% left_join(ipochemsf, by=c("Label"="name")) %>% filter(Score=="species", mean.floral>1.5), aes(x=CAP1*3, y=CAP2*3), size=3)+ labs(x=ax.labs[1], y=ax.labs[2])+
  scale_color_manual("Year", values=year_pal) + coord_fixed() + 
  theme_bw() + theme(panel.grid = element_blank())
```

### Emissions by class

```{r mod_abs}
options(contrasts = c("contr.sum", "contr.poly"))
abs_effects <- function(class) {
  lmer(sqrt(value) ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), 
       data = filter(volsmeta.class, name==class))
}

mod.abs <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, abs_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.abs %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to snowmelt date and estimated precipitation")
```

### Scatterplots

#### Snowmelt date

```{r vols_abs, fig.height=5, fig.width=7}
precip.breaks <- seq(25,175, by=25)
grid.points <- list(sun_date=range(treatments$sun_date), precip_est_mm=precip.breaks)
options(contrasts = c("contr.sum", "contr.poly"))
#TODO this code partially repeats chunk above
mod.class.abs <- volsmeta.class %>% group_by(name) %>% nest() %>% 
  mutate(model = map(data, ~lmer(value ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), data=.)),
         test = map(model, ~tidy(anova(.x))),
         emm.grid = map(model, ~ summary(emmeans(.x, ~ precip_est_mm * sun_date, at=grid.points))))

ggplot(volsmeta.class, aes(x = sun_date, y = value, color=precip_est_mm)) + facet_wrap(vars(name), scales="free_y") + 
  geom_point(aes(shape=year), size=2) + 
    geom_line(data = select(mod.class.abs, name, emm.grid) %>% unnest(emm.grid),
            aes(y=emmean, group=precip_est_mm))+
  theme_minimal()+ scale_y_sqrt()  + 
  labs(x="Snowmelt date", y ="Emissions (ng/hr/flower)", shape="Year", color="Precipitation") +
  scale_color_gradientn("Summer precipitation (mm)", colors=rev(water_pal), 
                        values=rev(c(1, (treatments %>% group_by(water) %>% summarize(across(precip_est_mm, mean)) %>%
                                           pull(precip_est_mm) %>% scales::rescale(from=range(precip.breaks)))[2],0)),
                        breaks=precip.breaks)
```

## Plasticity to plant-level soil moisture

### CAP

```{r cap_sm}
hasVWC <- !is.na(volsmeta.avg$VWC)
(scents.cap <- capscale(sqrt(vols.good.avg[hasVWC,]) ~ year * VWC, data=volsmeta.avg[hasVWC,], distance="bray"))
anova(scents.cap, by="term") %>% kable()

cap.df <- fortify(scents.cap)
ggplot(mapping=aes(x=CAP1, y=CAP2, label=Label)) + 
  geom_point(data=cap.df %>% filter(Score=="sites") %>% bind_cols(volsmeta.avg[hasVWC,]), aes(color=year))+
  geom_segment(data=cap.df %>% filter(Score=="biplot"), aes(x=0,y=0,xend=CAP1*3.7, yend=CAP2*3.7), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(data=cap.df %>% filter(Score=="biplot") ,
            aes(x=CAP1*4, y=CAP2*4, angle=atan(CAP2/CAP1)*180/pi), hjust=0, fontface="bold") +
  geom_text(data=cap.df %>% left_join(ipochemsf, by=c("Label"="name")) %>% filter(Score=="species", mean.floral>1.5), aes(x=CAP1*3, y=CAP2*3), size=3)+
  scale_color_manual("Year", values=year_pal) + coord_fixed(xlim=c(-3,NA)) + 
  theme_bw() + theme(panel.grid = element_blank())
```

### Emissions by class

```{r mod_sm}
options(contrasts = c("contr.sum", "contr.poly"))
sm_effects <- function(class) {
  lmer(sqrt(value) ~ year * VWC + (1|plotid), 
       data = filter(volsmeta.class, name==class))
}

mod.sm <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, sm_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.sm %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to soil moisture")
```

### Scatterplots

```{r sm_vols_top, fig.width=14, fig.height=10}
ggplot(volsmeta.good.top, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") +
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt()  + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year") +
  scale_color_manual(values=year_pal)
```

```{r sm_vols, fig.width=5, fig.height=5}
ggplot(volsmeta.class, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") + 
  geom_point(size=0.5) + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year")+
  scale_color_manual(values=year_pal)

ggplot(volsmeta.avg, aes(x = VWC, color = year, y = total)) + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Total filtered emissions (ng/hr/flower)", color="Year") +
  scale_color_manual(values=year_pal)
```

## Plasticity to subplot-level soil moisture

### CAP

```{r cap_sm_subplot}
(scents.cap <- capscale(sqrt(vols.good.avg) ~ year * VWC, data=left_join(select(volsmeta.avg,-VWC), sm.subplotyear),
                        distance="bray"))
anova(scents.cap, by="term") %>% kable()
cap.df <- fortify(scents.cap)
ggplot(mapping=aes(x=CAP1, y=CAP2, label=Label)) + 
  geom_point(data=cap.df %>% filter(Score=="sites") %>% bind_cols(volsmeta.avg), aes(color=year))+
  geom_segment(data=cap.df %>% filter(Score=="biplot"), aes(x=0,y=0,xend=CAP1*3.7, yend=CAP2*3.7), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(data=cap.df %>% filter(Score=="biplot") ,
            aes(x=CAP1*4, y=CAP2*4, angle=atan(CAP2/CAP1)*180/pi), hjust=0, fontface="bold") +
  geom_text(data=cap.df %>% left_join(ipochemsf, by=c("Label"="name")) %>% filter(Score=="species", mean.floral>1.5), aes(x=CAP1*3, y=CAP2*3), size=3)+
  scale_color_manual("Year", values=year_pal) + coord_fixed(xlim=c(-3,NA)) + 
  theme_bw() + theme(panel.grid = element_blank())
```

### Emissions by class

```{r mod_sm_subplot}
options(contrasts = c("contr.sum", "contr.poly"))
sm_effects <- function(class) {
  lmer(sqrt(value) ~ year * VWC + (1|plotid), 
       data = filter(left_join(select(volsmeta.class,-VWC), sm.subplotyear), name==class))
}

mod.sm <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, sm_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.sm %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to subplot-level soil moisture")
```

### Scatterplots

```{r sm_subplot_vols_top, fig.width=14, fig.height=10}
ggplot(left_join(select(volsmeta.good.top,-VWC), sm.subplotyear), aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") +
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ 
  scale_y_sqrt()  + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year") +
  scale_color_manual(values=year_pal)
```

```{r sm_subplot_vols, fig.width=5, fig.height=5}
ggplot(left_join(select(volsmeta.class,-VWC), sm.subplotyear), aes(x = VWC, color = year, y = value)) + 
  facet_wrap(vars(name), scales="free_y") + 
  geom_point(size=0.5) + geom_smooth(se=F, span=1)+ theme_minimal()+ 
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year")+
  scale_color_manual(values=year_pal)

ggplot(left_join(select(volsmeta.avg,-VWC), sm.subplotyear), aes(x = VWC, color = year, y = total)) + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ 
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Total filtered emissions (ng/hr/flower)", color="Year") +
  scale_color_manual(values=year_pal)
```


# Selection analyses

## Fitness measures

## Plots of selection in treatments and years

```{r selection, fig.width=16, fig.height=10}
walk(fitnesstraits, ~ 
       print(ggplot(traits.good.top %>% mutate(prop_uninfested = if_else(fruits_nonaborted > 4, prop_uninfested, NA)), 
                    aes_string(y=.x, x="value", color="year")) + 
               geom_point() + geom_smooth(method="lm", se=F) + 
               facet_wrap(vars(name), scales="free_x") + 
               labs(x="Emissions (ng/hr/flower)", y=fitnessnames[.x], color="Year")+
               theme_bw() + scale_x_sqrt() + scale_color_manual(values=year_pal)))


walk(fitnesstraits, ~ 
       print(ggplot(traits.good.top %>% mutate(prop_uninfested = if_else(fruits_nonaborted > 4, prop_uninfested, NA)), 
                    aes(x=value, color=paste(water,snow), linetype=year, shape=year)) + 
               geom_point(aes_string(y=.x)) + geom_smooth(aes_string(y=.x), method="lm",se=F) + 
               facet_wrap(vars(name), scales="free_x") + 
               labs(x="Emissions (ng/hr/flower)", y=fitnessnames[.x], 
                    color="Precipitation and snowmelt", linetype="Year", shape="Year")+
               theme_bw() + scale_x_sqrt() + scale_color_manual(values=water_snow_pal)))
```

### Emissions by class

```{r selection_class}
walk(fitnesstraits, ~ 
       print(ggplot(traits.class %>% mutate(prop_uninfested = if_else(fruits_nonaborted > 4, prop_uninfested, NA)), 
                    aes(x=value, color=paste(water,snow), linetype=year, shape=year)) + 
               geom_point(aes_string(y=.x)) + geom_smooth(aes_string(y=.x), method="lm",se=F) + 
               facet_wrap(vars(name), scales="free_x") + 
               labs(x="Emissions (ng/hr/flower)", y=fitnessnames[.x], 
                    color="Precipitation and snowmelt", linetype="Year", shape="Year")+
               theme_bw() + scale_x_sqrt() + scale_color_manual(values=water_snow_pal)))
```

```{r selection_class_gradient, fig.width=9.5, fig.height=6}
trait.class.wide <- bind_cols(volsmeta.avg, vols.class) %>% left_join(mnps.plantyr.traits) %>% 
  select(plantid, year, water, snow, all_of(fitnesstraits), all_of(names(class_pal))) 
 #mutate(across(all_of(names(class_pal)), sqrt))

selection_effects <- function(fitnesstrait) {
  options(contrasts = c("contr.sum", "contr.poly"))
  lm(relfitness ~ year * water * snow + 
     Monoterpene * snow + Monoterpene * water + Monoterpene * year +
     Sesquiterpene * snow + Sesquiterpene * water + Sesquiterpene * year + 
     Benzenoid * snow + Benzenoid * water + Benzenoid * year +
     Aliphatic * snow + Aliphatic * water + Aliphatic * year,
       data = trait.class.wide %>% rename(fitness=fitnesstrait) %>% 
       mutate(relfitness = fitness/mean(fitness, na.rm=T)))
}

fitnesstraits.sd <- map_dbl(fitnesstraits, ~sd(mnps.plantyr.traits[[.x]], na.rm=T)) %>% set_names(fitnesstraits)
class.sd <- map_dbl(names(class_pal), ~sd(trait.class.wide[[.x]], na.rm=T)) %>% set_names(names(class_pal)) 

get_emt_class <- function(mod) {
  map_dfr(set_names(names(class_pal)), ~emtrends(mod, c("year","water","snow"), var=.x) %>% 
            summary %>% as_tibble %>% rename("trend" = ends_with("trend")), .id="class")
}

mod.selection <- tibble(fitnesstrait=fitnesstraits) %>% 
  mutate(model = map(fitnesstrait, selection_effects),
         emt = map(model, get_emt_class),
         test = map(model, ~ tidy(car::Anova(.x, type=3))))

mod.selection %>% unnest(test) %>% select(fitnesstrait, term, p.value) %>% 
  pivot_wider(names_from=fitnesstrait, values_from=p.value) %>%
  filter(term!="Residuals") %>% 
  kable(caption="Selection of emissions in each treatment", digits=4)

mod.selection %>% select(fitnesstrait, emt) %>% unnest(emt) %>% 
  mutate(classsd = class.sd[class],
         trend.std = trend * classsd,
         uSE = (trend + SE) * classsd,
         lSE = (trend - SE) * classsd,
         water_snow = factor(paste(water, snow), levels=names(water_snow_pal))) %>% 
  ggplot(aes(x=class, color=water_snow, y=trend.std, ymin=lSE, ymax=uSE)) +
  facet_grid(cols=vars(fitnesstrait), rows=vars(year), labeller=as_labeller(c(set_names(2018:2020), fitnessnames))) +
  geom_hline(yintercept=0)+
  geom_pointrange(position=position_dodge(width=0.5), size=0.2) +
  scale_color_manual(values=water_snow_pal) +
  labs(x="Compound class", y = "Standardized selection gradient", color="Precipitation, snowmelt") +
  theme_bw() + theme(axis.text.x = element_text(angle=90, hjust=1))
```


## Joint species distribution model

```{r jsdm_selection, eval=F}
library(Hmsc)

# model Y "species" (seeds, seeds per flower, prop fly fruits, fly eggs) 
# vs. XRRRData "environment" (volatiles) with reduced-rank regression

fitness.matrix <- volsmeta.traits %>% select(all_of(fitnesstraits))
fitness.complete <- !is.na(rowSums(fitness.matrix)) & !is.infinite(rowSums(fitness.matrix))
  
(model.sel <- Hmsc(Y = fitness.matrix %>% filter(fitness.complete) %>% as.matrix(),
                   XData = volsmeta %>% filter(fitness.complete) %>% select(year),
                   XFormula = ~ year,
                   XRRRData = vols.good %>% filter(fitness.complete), 
                   XRRRScale = TRUE, ncRRR = 4, 
                   studyDesign = volsmeta %>% filter(fitness.complete) %>% select(plot) %>%
                     mutate_all(as.factor),
                   ranLevels = list(plot = HmscRandomLevel(units = levels(treatments$plot))),
                   distr=c("poisson", "poisson", "probit","poisson")))

# Fit model with MCMC
thin <- 1
(model.sel <- sampleMcmc(model.sel, thin = thin, samples = 100, transient = thin*50, 
                         nChains = 3, nParallel = 3)) 
```

```{r jsdm_selection_results, eval=F}
# Model convergence
mpost <- convertToCodaObject(model.sel)[c("Beta","Gamma","V","Sigma")]
ess <- map(mpost, effectiveSize)
par(mfrow=c(2,2))
walk(names(ess), ~hist(ess[[.x]], xlab = expression("Effective sample size" ~ beta ~ ""), main=.x))

# Parameters
postBeta = getPostEstimate(model.sel, parName="Beta")
plotBeta(model.sel, post=postBeta, param="Support", supportLevel = 0.95, 
         spNamesNumbers = c(T,F), covNamesNumbers = c(T,F))

# Model fit (explanatory power)
predY = computePredictedValues(model.sel)
MF = evaluateModelFit(hM=model.sel, predY=predY)
tibble(name = fitnesstraits, SR2 = MF$SR2) %>% 
  ggplot(aes(y=name, x=SR2))+ geom_point() +
  labs(x="squared Spearman correlation between obs and pred values (SR2)", 
       title="Model fit (pseudo R2)") + theme_minimal()
```

## Selection gradients

Following method of [Chong et al. 2018](https://doi.org/10.1002/evl3.63) for reconstructing selection estimates from estimates of selection on PCs: Beta = E * A.

```{r calc_gradients}
prcomp_std <- function(data) {
  data %>% mutate(across(everything(), ~ (.x-mean(.x, na.rm=T))/sd(.x, na.rm=T))) %>% #center and scale
  prcomp(center = FALSE, scale. = FALSE)
}

calculate_beta <- function(E, slopes, resid.df) {
  data.frame(B = E %*% slopes$A,
             SE = sqrt(E ^ 2 %*% slopes$SE ^ 2)) %>% 
    mutate(p.value = 2*pt(abs(B)/SE, df=resid.df, lower.tail=F)) %>% #calculate t statistic from difference in means and SE, then look up one-tailed p in the upper tail, double to get a 2-tailed test
    rownames_to_column("trait")
}

selection_pca <- function(year=names(year_pal), water=names(water_pal), snow=names(snow_pal), 
                          fitnesstrait, n.PCs, PCA, mdata) {
  if(ncol(PCA$x) <= n.PCs) {
    warning(paste("Only ", ncol(PCA$x), "PCs"))
    n.PCs <- ncol(PCA$x)
  }
  std.traits <- mdata %>% bind_cols(as.data.frame(PCA$x)) %>% 
    filter(year == .env$year, water %in% .env$water, snow %in% .env$snow) %>% 
    select(-any_of(fitnesstrait)) %>% 
    left_join(mnps.plantyr.traits %>% select(year, plotid, plantid, all_of(fitnesstrait)), by=c("year", "plotid", "plantid")) %>% 
    select(all_of(fitnesstrait), any_of(names(treatments)), starts_with("PC")) %>% 
    rename("fitness" = all_of(fitnesstrait)) %>% 
    group_by(year) %>% mutate(relfitness = fitness/mean(fitness, na.rm=T), .keep="unused") %>% ungroup %>% #relative fitness within year
    drop_na(relfitness) %>% 
    select(where(~sum(is.na(.))==0))
  
  # Regress relative fitness on PCs
  selection.on.PCs <- lm(formula(paste("relfitness ~", paste0("PC",1:n.PCs, collapse=" + "))), data = std.traits)

  # Convert selection gradients on PCs to gradients on traits
  calculate_beta(E = PCA$rotation[,1:n.PCs],
                 slopes = tidy(selection.on.PCs) %>% filter(term != "(Intercept)") %>%
                   rename(A=estimate, SE=std.error), 
                 resid.df = aov(selection.on.PCs)$df.residual)
}
calculate_all_gradients <- function(trts) {
  mutate(trts, PCA=list(vols.PCA), mdata = list(volsmeta.avg), n.PCs = 5) %>% 
  bind_rows(mutate(., 
                   n.PCs = if_else(year=="2018", 4, 5),
                   PCA =   if_else(year=="2018", list(mnps.PCA.nosepal), list(mnps.PCA.withsepal)),
                   mdata = if_else(year=="2018", list(meta.pca.nosepal), list(meta.pca.withsepal)))) %>% 
  mutate(betas = pmap(., selection_pca)) %>% select(-PCA, -mdata) %>% unnest(betas)
}

# Run PCAs
vols.PCA <- prcomp_std(vols.good.top)
summary(vols.PCA) # PCs 1 - 5 explain 82% of the variation, 12 PCs total (12 top volatiles)

(floraltraits.pca <- floraltraits[-8]) #drop total flowers
mnps.PCA.withsepal <- mnps.plantyr %>% select(all_of(floraltraits.pca)) %>% drop_na() %>% prcomp_std()
mnps.PCA.nosepal <- mnps.plantyr %>% select(all_of(floraltraits.pca[-4])) %>% drop_na() %>% prcomp_std()

summary(mnps.PCA.withsepal) # PCs 1 - 5 explain 83% of the variation, 7 PCs total
summary(mnps.PCA.nosepal)   # PCs 1 - 4 explain 83% of the variation, 6 PCs total

meta.pca.withsepal <- mnps.plantyr %>% filter(mnps.plantyr %>% select(all_of(floraltraits.pca)) %>% vctrs::vec_detect_complete())
meta.pca.nosepal <- mnps.plantyr %>% filter(mnps.plantyr %>% select(all_of(floraltraits.pca[-4])) %>% vctrs::vec_detect_complete())

#Calculate selection gradients in each treatment category
all.beta.yr <- expand_grid(year=names(year_pal), fitnesstrait=fitnesstraits) %>% 
  calculate_all_gradients()
all.beta.water <- expand_grid(year=names(year_pal), fitnesstrait=fitnesstraits, water=names(water_pal)) %>% 
  calculate_all_gradients()
all.beta.snow <- expand_grid(year=names(year_pal), fitnesstrait=fitnesstraits, snow=names(snow_pal)) %>% 
  calculate_all_gradients()
all.beta.water.snow <- expand_grid(year=names(year_pal), fitnesstrait=fitnesstraits, 
                                   water=names(water_pal), snow=names(snow_pal)) %>% 
  calculate_all_gradients()
```

```{r gradients, fig.width=11, fig.height=9}
plot_beta <- function(betas, colorby) {
  bs <- betas %>% left_join(ipochemsf, by=c("trait"="name")) %>% 
        mutate(trait = recode(trait, !!!traitnames) %>%  
            fct_reorder(replace_na(mean.floral,1000))) #order graph by mean emissions
  if(colorby=="water_snow")  bs <- bs %>%  mutate(water_snow = paste(water, snow))
  ggplot(bs, aes(y=trait, x = B, xmin=B-SE, xmax=B+SE, alpha=p.value<0.05, shape=year)) + 
    facet_wrap(vars(fitnesstrait), nrow=1, labeller=as_labeller(fitnessnames))+
    geom_vline(xintercept=0)+ 
    geom_pointrange(aes_string(color=colorby), position=position_dodge(width=0.5)) + 
    labs(x = "Selection gradient (beta) estimated from gradients on PCs", y = "Trait", shape="Year", alpha="P < 0.05",
         color=list(year="Year", water="Precipitation", snow="Snowmelt", 
                    water_snow = "Precipitation and snowmelt")[[colorby]]) +
    scale_color_manual(values=list(year=year_pal, water=water_pal, snow=snow_pal,
                                   water_snow=water_snow_pal)[[colorby]]) +
    theme_bw() 
}

plot_beta(all.beta.yr, colorby="year")
plot_beta(all.beta.water, colorby="water")
plot_beta(all.beta.snow, colorby="snow")
plot_beta(all.beta.water.snow, colorby="water_snow") + coord_cartesian(xlim=c(-2,3.5))
```

```{r gradient_heatmap, fig.width=11}
all.beta.yr %>% left_join(ipochemsf, by=c("trait"="name")) %>% 
        mutate(trait = recode(trait, !!!traitnames) %>% fct_reorder(replace_na(mean.floral,1000))) %>% 
  ggplot(aes(x=year, y=trait, fill=B)) + geom_tile() + 
  scale_fill_gradient2()+ geom_point(aes(alpha=p.value < 0.05))+ scale_alpha_discrete(range=c(0,1))+
  facet_wrap(vars(fitnesstrait), nrow=1, labeller=as_labeller(fitnessnames)) + theme_minimal()
```

## Adaptive plasticity

### Selection gradient vs. plasticity

```{r adaptive, fig.height=5, fig.width=7}
adaptive.plasticity.data <- 
  bind_rows(volsmeta.good.top %>% mutate(index=as.integer(index)), 
            mnps.plantyr %>% pivot_longer(all_of(floraltraits.pca)) %>% drop_na(value)) %>% 
  group_by(name, year) %>% nest() %>% 
  mutate(model = map(data, ~lm(value ~ snow + water, data=.)), # + (1|plotid) + (1|snow:plot)
         emm = map(model, ~ summary(emmeans(.x, ~ snow + water))%>% filter(water=="Control", snow=="Normal")), 
         test =  map(model, ~tidy(multcomp:::summary.glht(multcomp::glht(.x, 
                           linfct = multcomp::mcp(water = c("Addition - Control = 0",
                                                            "Reduction - Control = 0",
                                                            "Reduction - Addition = 0"),
                                                  snow = c("Early - Normal = 0"))), multcomp::adjusted(type="none"))))) %>%#adjust later 
  select(year, name, emm, test) %>% unnest(c(emm, test)) %>% 
  mutate(estimate=estimate/emmean, std.error=std.error/emmean, #divide estimated difference in means by emm in Normal Control subplots
         contrast = str_replace(str_remove_all(contrast, "[a-z\\s]"),"-","_"),
         newenv = recode(contrast, "A_C"="Addition", "R_C"="Reduction","E_N"="Early"),
         multiplecomps = contrast %in% c("A_C","R_C")) %>% #these two reuse the same control data
  group_by(multiplecomps, name, year) %>% #two comparisons to Control made within each volatile and year 
  mutate(plasticity.p.value = ifelse(multiplecomps, p.adjust(p.value, method="fdr"), p.value)) %>% 
  ungroup %>% select(year, name, newenv, plasticity.p.value, estimate, std.error) %>% 
  left_join(bind_rows(rename(all.beta.water, newenv=water),
                    rename(all.beta.snow, newenv=snow)) %>% 
            filter(fitnesstrait!="eggs_per_flower") %>% 
  rename(name=trait)) %>% left_join(ipochemsf %>% select(-starts_with("area"))) %>% 
  mutate(class = fct_na_value_to_level(class, "Other trait"))

envnames <- c(Addition="additional precipitation",
             Reduction="reduced precipitation",
             Early="early snowmelt")

library(ggpubr)
adaptive.plasticity.data %>% filter(newenv!="R_A") %>% 
  mutate(fitnesstrait=factor(fitnesstrait, levels = sort(fitnesstraits)),
         plasticity.signif = plasticity.p.value < 0.05, selection.signif= p.value < 0.05,
         plasticity.selection.signif = interaction(factor(plasticity.signif), factor(selection.signif))) %>% 
  group_by(newenv) %>% nest() %>% 
  pwalk(function(newenv, data) { 
    print(ggplot(data, aes(x=estimate, y=B, color=class)) + 
            labs(x=paste("Emissions change with",envnames[newenv]), 
                 y=paste("Selection gradient \U03B2 under",envnames[newenv])) + 
            facet_grid(rows=vars(year), cols=vars(fitnesstrait), scales="fixed", 
                       labeller=as_labeller(c(fitnessnames, set_names(levels(treatments$year)))))+
            geom_vline(xintercept=0) + geom_hline(yintercept=0) +
            annotate("rect", xmin = 0, xmax = Inf, ymin = 0, ymax = -Inf, fill= "black", alpha=0.1) + 
            annotate("rect", xmin = 0, xmax = -Inf, ymin = Inf, ymax = 0, fill= "black", alpha=0.1) +
            geom_errorbar(aes(ymin=B-SE, ymax=B+SE)) + 
            geom_errorbarh(aes(xmin=estimate-std.error, xmax=estimate+std.error)) +
            geom_point(aes(shape = plasticity.selection.signif), size=2.5) + 
            scale_shape_manual("adjusted P<0.05", values=c(1,17,19,15), guide=guide_legend(override.aes = list(size=5)),
                               labels=c("neither", "plasticity", "selection","both"), drop=FALSE)+
            scale_x_continuous(labels=scales::percent, breaks=c(-4:4)/ifelse(max(data$estimate)<1,2,1)) + 
            coord_cartesian(ylim = c(min(data$B)*1.05, max(data$B)*1.05), xlim = c(min(data$estimate)*1.05, max(data$estimate)*1.05))+
            scale_color_manual("Compound class", values=c(class_pal,"Other trait"="black"), 
                               guide=guide_legend(override.aes = list(size=5))) + 
            theme_bw() + theme(axis.text = element_text(size=7)))})

adaptive.plasticity.data %>% filter(newenv != "R_A") %>% count(p.value < 0.05, plasticity.p.value < 0.05)

ggplot(adaptive.plasticity.data %>% filter(newenv!="R_A"), aes(x=estimate, xmin=estimate-std.error, xmax=estimate+std.error,
                          y=name, color=year, alpha=plasticity.p.value<0.05)) +  facet_wrap(vars(newenv)) +
  geom_vline(xintercept = 0) + geom_pointrange() + scale_color_manual(values=year_pal) + 
  theme_bw() + theme(panel.grid.major.y=element_blank(), axis.title.y=element_blank()) +
  labs(x="Plastic change in new environment compared to control")
```

```{r adaptive_quadrants}
adaptive.plasticity.data %>% 
  mutate(adaptive = sign(B) == sign(estimate)) %>% filter(newenv !="R_A") %>% 
  group_by(newenv, fitnesstrait) %>% summarize(adaptive=sum(adaptive), n=n(), .groups="drop") %>% 
  mutate(prop = adaptive/n,
         p.value=map2_dbl(adaptive, n, ~ prop.test(.x, .y)$p.value),
         fitnesstrait = fitnessnames[fitnesstrait]) %>% 
  kable(caption=paste("Proportion of", ncol(vols.good.top), "volatiles and", length(floraltraits.pca), 
                      "other traits whose plasticity is adaptive across 3 yr"), digits=2)
```


## Plasticity to snowmelt vs. precipitation

```{r snow_v_water, fig.height=6, fig.width=4.5}
compare.plasticity <- full_join(
  adaptive.plasticity.data %>% filter(newenv=="R_A") %>% 
    select(year, name, class, mean.floral, estimate, std.error) %>%
    rename(R_A=estimate, R_A.SE=std.error),
  adaptive.plasticity.data %>% filter(newenv=="Early") %>% 
    select(year, name, class, mean.floral, estimate, std.error) %>%
    rename(E_N=estimate, E_N.SE=std.error) %>% distinct()) #3 rows each, one for each fitness trait

ggplot(compare.plasticity, aes(x=R_A, xmin=R_A-R_A.SE, xmax=R_A+R_A.SE, 
                               y=E_N, ymin=E_N-E_N.SE, ymax=E_N+E_N.SE, color=class)) + 
  facet_wrap(vars(year), ncol=1)+
  geom_vline(xintercept=0) + geom_hline(yintercept=0) +
  annotate("rect", xmin = 0, xmax = Inf, ymin = 0, ymax = -Inf, fill= "black", alpha=0.1) + 
  annotate("rect", xmin = 0, xmax = -Inf, ymin = Inf, ymax = 0, fill= "black", alpha=0.1) +
  scale_alpha_manual(values=c(0,1), guide="none")+ 
  geom_errorbar() + geom_errorbarh() + geom_point(aes(size=replace_na(mean.floral, 10))) + 
  scale_size_area("Mean emissions\n(ng/hr/flower)", max_size=3) +
  scale_x_continuous(labels=scales::percent) + 
  scale_y_continuous(labels=scales::percent) + 
  coord_cartesian(ylim = c(min(compare.plasticity$E_N)*1.05, max(compare.plasticity$E_N)*1.05), 
                  xlim = c(min(compare.plasticity$R_A)*1.05, max(compare.plasticity$R_A)*1.05))+
  scale_color_manual("Compound class", values=c(class_pal, "Other trait"="black"), 
                     guide=guide_legend(override.aes = list(size=5))) + theme_bw() +
  labs(x="Plasticity to reduced precipitation (vs. additional)",
       y="Plasticity to early snowmelt (vs. normal)")

compare.plasticity %>% group_by(year) %>%  
  summarize(snow_water_cor = cor(R_A, E_N, use="complete"),
            snow_water_p = cor.test(R_A, E_N, use="complete")$p.value, n=n()) %>% 
  kable(caption="correlation between plasticity to water and snowmelt in each year")

```


```{r adaptive_diff, fig.height=14, fig.width=10, eval=F}
### Change in selection gradient vs. plasticity
diffbeta <- vols.beta.water %>% rename(name=trait) %>% 
  pivot_wider(names_from=water, values_from=c(B,SE)) %>% 
  mutate(B_A_R = B_Addition - B_Reduction)

#TODO get this working again with compare.plasticity
plasticity.water %>% left_join(diffbeta) %>% 
  ggplot(aes(x=A_R, y=B_A_R, color=class)) + facet_wrap(vars(year), ncol=1)+
  geom_vline(xintercept=0)+geom_hline(yintercept=0) + geom_text(aes(label=name)) + 
  scale_x_continuous(labels=scales::percent, limits=c(NA,NA)) + theme_bw()  + theme(legend.position = "top") +
  labs(x="Emissions change in precipitation addition vs. reduction", y="Change in selection gradient (beta) in precipitation addition vs. reduction")
```

## Selection vs. absolute snowmelt, precip

```{r selection_abs_model}
selection_mod_abs <- function(trait, response, usetrait=TRUE) {
  if(trait %in% ipogood) data <- vols.traits 
  else data <- mnps.plantyr %>% mutate(prop_uninfested = 1 - prop_infested)
  vm.abs <- data %>% 
    select(all_of(c(response, trait)), sun_date, precip_est_mm) %>% 
    rename("trait" = all_of(trait), "response" = all_of(response)) %>% 
    filter(!is.infinite(response), !is.nan(response), !is.na(response)) %>% 
    mutate(response = response/mean(response, na.rm=T), #relative fitness
           trait = (trait-mean(trait, na.rm=T))/sd(trait, na.rm=T)) #center and scale
  options(contrasts = c("contr.sum", "contr.poly"))
  if(usetrait) lm(response ~ trait * sun_date + trait * precip_est_mm, data = vm.abs)
  else         lm(response ~ sun_date + precip_est_mm, data = vm.abs)
}

mod.selection.abs <- expand.grid(list(response = fitnesstraits, trait = c(ipogood, floraltraits))) %>%
  mutate(model = pmap(., selection_mod_abs),
         coefs = map(model, tidy))

mod.selection.tests <- mod.selection.abs %>% 
  left_join(ipochemsf, by=c("trait"="name")) %>% 
  mutate(freq.floral = replace_na(freq.floral, 1)) %>% #morph/nectar traits
  filter(freq.floral>0.1) %>% 
    mutate(tests = map(model, ~ tidy(car::Anova(.x, type=3)))) %>% 
  select(-model) %>% unnest(coefs) %>% 
  filter(str_detect(term, "trait"))
```

```{r abs_model}
#ignore trait and get the direct effects on fitness
tibble(response = fitnesstraits, trait = floraltraits[1], usetrait=FALSE) %>% #dummy trait not used
  mutate(model = pmap(., selection_mod_abs),
         coefs = map(model, tidy)) %>% select(response, coefs) %>% unnest(coefs) %>% 
  filter(term !="(Intercept)") %>% kable()
```


```{r selection_abs, fig.width=9, fig.height=7}
mod.selection.abs %>% select(-model) %>% unnest(coefs) %>% 
  ggplot(aes(x=response, y=estimate)) + facet_wrap(vars(term), scales="free_y") +
  geom_hline(yintercept=0) + geom_boxplot()

library(ggrepel)
term_labels <- c(trait="Selection on trait (\U03B2)",
                 `trait:precip_est_mm`="Selection x summer precipitation (\U03B2 / mm)",
                 `trait:sun_date`="Selection x snowmelt date (\U03B2 / day)")
ggplot(mod.selection.tests, aes(x=factor(response, levels=sort(fitnesstraits)), y=estimate, label=greekify(trait))) + 
  facet_wrap(vars(term), labeller = as_labeller(term_labels), scales="free_y", ncol=1) + theme_bw() +
  geom_point(aes(size=replace_na(mean.floral, 1),  #default size for non-vol traits
                 color=fct_na_value_to_level(class, "Other trait"))) + geom_hline(yintercept=0) + 
  geom_text_repel(data = filter(mod.selection.tests, p.value < 0.05) %>% 
                    mutate(trait=recode(trait, !!!c("(1Z,4Z,7Z)-1,5,9,9-tetramethylcycloundeca-1,4,7-triene"=
                                          "(1Z,4Z,7Z)-1,5,9,9-\ntetramethylcycloundeca-1,4,7-triene", traitnames))), 
                  show.legend = FALSE, size=3.5, box.padding=0, point.padding=0, direction="y", lineheight=0.7)+
  scale_size_area(max_size = 10) + scale_x_discrete(labels=fitnessnames) +
  scale_color_manual(values=c(class_pal,"Other trait"="grey50"), guide=guide_legend(override.aes = list(size=5))) + 
  labs(x="Fitness measure", size="Mean emissions\n(ng/hr/flower)", color="Class") + 
  theme(axis.title.y=element_blank())

mod.selection.tests %>% filter(trait %in% c(ipogood.top, floraltraits)) %>% 
  mutate(trait = trait %>% greekify() %>% recode(!!!traitnames) %>% 
           factor(levels = rev(c(traitnames, greekify(ipogood.top))))) %>% 
  ggplot(aes(x=estimate,y=trait, color=response, alpha=p.value < 0.05, xmin=estimate-std.error, xmax=estimate+std.error)) + 
  facet_wrap(vars(term), labeller = as_labeller(term_labels), scales="free_x", nrow=1) + 
  geom_vline(xintercept=0)+ 
  geom_pointrange(position = position_dodge2(width=0.6, reverse=T))+
  scale_color_brewer("Fitness measure", palette = "Dark2", labels = fitnessnames)+
  scale_alpha_discrete("P < 0.05", range=c(0.3, 1), guide="none")+ 
  scale_x_continuous(expand = expansion(c(0,0)))+
  theme_bw() + theme(axis.title=element_blank(), legend.position = "top",
                     panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())

mod.selection.abs %>% filter(trait=="a-pinene") %>% 
  mutate(tests = map(model, ~ tidy(car::Anova(.x, type=3)))) %>% 
  select(-c(model, coefs)) %>% unnest(tests) %>% filter(str_detect(term, "trait")) %>% kable()

library(skimr)
skim(mod.selection.tests %>% group_by(term), estimate)
```

# Volatiles vs. floral traits

## Overlap of datasets

For the plants that have seeds counted

```{r overlap}
samples_per_plant <- volsmeta %>% count(year, plantid, name="scent") %>% as_tibble() %>% 
  full_join(mnps.plantyr.traits %>% drop_na(corolla_length) %>% count(year,plantid, name="morph")) %>% 
  full_join(mnps.plantyr.traits %>% drop_na(nectar_24_h_ul) %>% count(year,plantid, name="nectar")) %>% 
  full_join(mnps.plantyr.traits %>% drop_na(height_cm) %>% count(year,plantid, name="height")) %>% 
  full_join(mnps.plantyr.traits %>% drop_na(seeds_est) %>% count(year,plantid, name="seeds")) %>% 
  drop_na(seeds)

library(ggvenn) # numbers outside all circles have seed estimates
samples_per_plant %>% mutate(across(c(scent, morph, nectar), ~!is.na(.x))) %>% ggvenn()
samples_per_plant %>% mutate(across(c(scent, morph, nectar, height), ~!is.na(.x))) %>% ggvenn()

traits.m <- c("corolla_length", "corolla_width", "style_length") #exclude sepal_width (no 2018), anther min/max
traits.n <- c("nectar_24_h_ul", "nectar_conc")
# drop plants with missing data for combinations of measurements
combos <- list(v=vols.good.top, 
               vm=vols.traits %>% select(all_of(c(ipogood.top, traits.m, "height_cm"))) %>% drop_na(), 
               vn=vols.traits %>% select(all_of(c(ipogood.top, traits.n, "height_cm"))) %>% drop_na(), 
               vmn=vols.traits %>% select(all_of(c(ipogood.top, traits.m, traits.n, "height_cm"))) %>% drop_na(),
               mn=mnps.plantyr.traits %>% select(all_of(c(traits.m, traits.n, "height_cm"))) %>% drop_na(),
               n=mnps.plantyr.traits %>% select(all_of(c(traits.n, "height_cm"))) %>% drop_na(),
               m=mnps.plantyr.traits %>% select(all_of(c(traits.m, "height_cm"))) %>% drop_na())

all.traits.combos <- full_join(bind_cols(volsmeta.avg, vols.good.avg), 
                               left_join(mnps.plantyr.traits, treatments)) %>% 
  mutate(v =   vctrs::vec_detect_complete(.[,colnames(combos$v)]),
         vm =  vctrs::vec_detect_complete(.[,colnames(combos$vm)]),
         vn =  vctrs::vec_detect_complete(.[,colnames(combos$vn)]),
         vmn = vctrs::vec_detect_complete(.[,colnames(combos$vmn)]),
         mn =  vctrs::vec_detect_complete(.[,colnames(combos$mn)]),
         n =  vctrs::vec_detect_complete(.[,colnames(combos$n)]),
         m =  vctrs::vec_detect_complete(.[,colnames(combos$m)]))

all.traits.combos %>% group_by(year, water, snow) %>% 
  summarize(across(all_of(names(combos)), sum)) %>% kable(caption="plants with each set of measurements")
```

## Correlations

```{r vmnt_corr, fig.width=7, fig.height=5}
corr.annot <- tibble(name=colnames(combos$vmn)) %>% left_join(ipochemsf) %>% column_to_rownames("name") %>% 
  mutate(class = fct_na_value_to_level(class, "Other trait")) %>% select(Class=class)
corr.labs <- rownames(corr.annot) %>% greekify() %>% recode(!!!traitnames)
pheatmap(cor(combos$vmn, method="pearson"), 
         scale="none", clustering_method = "mcquitty",
         color=colorRampPalette(c("blue","white","red"))(200), breaks=seq(-1,1,by=0.01),
         annotation_col=corr.annot, annotation_row=corr.annot, 
         annotation_colors = list(Class=c(class_pal, "Other trait"="black")),
         labels_row=corr.labs, labels_col=corr.labs)
```

## PCA

```{r vmnt_pca}
#TODO consider sqrt transformation: https://bcrc.bio.umass.edu/projects/selection_analysis/Main.htm
vmn.pca <- combos$vmn %>% mutate(across(all_of(ipogood.top), sqrt)) %>% prcomp_std()

vmn.pca.vars <- as.data.frame(vmn.pca$rotation) %>% rownames_to_column("variable") %>% 
  mutate(class=fct_collapse(variable, !!!deframe(select(filter(ipochemsf, name %in% ipogood.top), class, name)), 
                            other_level="Other trait")) 
vmn.pca.labs <- paste0("PC", 1:2, " (", round(100*vmn.pca$sdev[1:2]^2/sum(vmn.pca$sdev^2), 0), "% explained)") 
zoom<-15
ggplot(bind_cols(combos$vmn, as_tibble(vmn.pca$x)), aes(x=PC1, y=PC2)) +
  geom_point(shape=21, color="grey80", size=2) + #, fill=`a-pinene` + scale_fill_viridis_c(option = "plasma")
  geom_segment(data=vmn.pca.vars, aes(x=0, y=0, xend=PC1*zoom, yend=PC2*zoom), 
               arrow = arrow(length = unit(1/2, "picas")), color="grey80") +
  geom_text(data=vmn.pca.vars, show.legend = F,
            aes(label = greekify(recode(variable, !!!traitnames)), 
                x = PC1*zoom*1.1, y = PC2*zoom*1.1, color = class,
                angle =  (180/pi) * atan(PC2/PC1), hjust=if_else(PC1>0,0,1)), size=3)+
  coord_fixed()+ scale_color_manual(values=c(class_pal, "Other trait"="black")) + 
  theme_minimal()+ theme(panel.grid=element_blank(), legend.position = "top") + 
  geom_point(data=vmn.pca.vars, aes(color=class), alpha=0)+ 
  guides(color=guide_legend(override.aes = list(alpha=1, size=3)))+
  labs(color="", x=vmn.pca.labs[1], y=vmn.pca.labs[2]) 
```

## Selection gradients

Following method of [Chong et al. 2018](https://doi.org/10.1002/evl3.63) for reconstructing selection estimates from estimates of selection on PCs: Beta = E * A.

```{r vmnt_gradients, fig.width=9, fig.height=9}
meta.pca <- #treatments used in selection_pca function
  volsmeta.avg[!is.na(rowSums(vols.traits[,colnames(combos$vmn)])),] #match rows in vmn

#TODO consider sqrt transformation: https://bcrc.bio.umass.edu/projects/selection_analysis/Main.htm
vols.PCA <- combos$vmn %>% prcomp_std() #mutate(across(all_of(ipogood.top), sqrt))
summary(vols.PCA) # PCs 1 - 5 explain 66% of the variation, 18 PCs total (12 volatiles + 6 traits)
heatmap(vols.PCA$rotation[,1:5], Colv = NA, col=viridis(256, option="magma"), scale="none")
pca <- vols.PCA

vols.beta.yr <- expand_grid(year=unique(volsmeta$year), fitnesstrait=fitnesstraits)  %>% 
  mutate(PCA=list(vols.PCA), mdata=list(meta.pca), n.PCs = 5) %>% 
  mutate(betas = pmap(., selection_pca)) %>% select(-PCA, -mdata) %>% unnest(betas)
vols.beta.water <- expand_grid(year=unique(volsmeta$year), water=names(water_pal), fitnesstrait=fitnesstraits)  %>% 
  mutate(PCA=list(vols.PCA), mdata=list(meta.pca), n.PCs = 5) %>% 
  mutate(betas = pmap(., selection_pca)) %>% select(-PCA, -mdata) %>% unnest(betas)
vols.beta.snow <- expand_grid(year=unique(volsmeta$year), snow=names(snow_pal), fitnesstrait=fitnesstraits)  %>% 
  mutate(PCA=list(vols.PCA), mdata=list(meta.pca), n.PCs = 5) %>% 
  mutate(betas = pmap(., selection_pca)) %>% select(-PCA, -mdata) %>% unnest(betas)

plot_beta(vols.beta.yr, colorby = "year")
plot_beta(vols.beta.water, colorby = "water")
plot_beta(vols.beta.snow, colorby = "snow")
```

### Sensitivity to number of PCs


```{r numberPCs, fig.width=10}
vols.beta.yr.PCs <- expand_grid(year=unique(volsmeta$year), fitnesstrait=fitnesstraits, 
                                n.PCs=2:17) %>% 
  mutate(PCA=list(vols.PCA), mdata=list(meta.pca)) %>% 
  mutate(betas = pmap(., selection_pca)) %>% select(-PCA, -mdata) %>% unnest(betas)
ggplot(vols.beta.yr.PCs, aes(x=n.PCs)) + facet_wrap(vars(fitnesstrait), scales="free_y") + 
  geom_line(aes(y=B, color=ifelse(trait %in% traits, "Other trait", "Volatile"), group=paste(trait, year))) +
  labs(color="")
```

### Sensitivity to more observations

```{r moreobs, fig.width=10}
map(combos, dim)

combos.PCA <- map(combos,  ~ .x %>% prcomp_std()) #%>% mutate(across(any_of(ipogood.top), sqrt))

#combos.meta <- map(names(combos), ~ volsmeta.avg[pull(vols.traits.combos, .x),])
combos.meta <- map(names(combos), ~ all.traits.combos[pull(all.traits.combos, .x),])

vols.beta.yr.combos <- expand_grid(year=unique(volsmeta$year), fitnesstrait=fitnesstraits, 
                                n.PCs=5)  %>% 
  cross_join(tibble(PCA = combos.PCA, mdata=combos.meta, 
                    combo = factor(names(combos), levels=names(combos)))) %>% 
  mutate(betas = pmap(select(., -combo), selection_pca)) %>% select(-PCA, -mdata) %>% unnest(betas)

ggplot(vols.beta.yr.combos, aes(x=combo)) + facet_wrap(vars(fitnesstrait), scales="free_y") + 
  geom_line(aes(y=B, color=ifelse(trait %in% traits, "Other trait", "Volatile"), group=paste(trait, year))) +
  labs(color="")
```

