--- 
title: "Maxfield Meadow Ipomopsis Volatiles"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>


```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)
library(vegan)
library(broom)
library(viridis)
library(ggvegan)
library(gridExtra)
library(dendsort)
library(pheatmap)
library(RColorBrewer)
library(DT)

library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, 
                      fig.path = "images_volatiles2020/", dev="svglite", dev.args=list(fix_text_size=FALSE), fig.height=8, fig.width=8)
options(digits=4) # for kables

load("data/volatiles/maxfield_volatiles.rda")
load("data/maxfield_data.rda")

#check for duplicates
#View(vt %>% filter(!exclude %in% c("flowers", "incomplete", "resample")) %>% count(sampledate.file, plot.file, subplot.file, plant.file) %>% filter(n>1) %>% left_join(vt)) #exclude="resample"
#View(maxfmeta %>% count(sampledate,plantid) %>% filter(n>1) %>% left_join(maxfmeta))
#TODO fix these GCMS filename duplicates

volsmeta <- maxfmeta %>% mutate(sampledate = as.Date(sampledate), year = factor(year(sampledate))) %>% 
  drop_na(plotid) %>% #cut out ambients, leaving 459 chromatograms
  filter(year %in% c("2018","2019","2020")) %>% droplevels() %>% 
  mutate(plot=str_sub(plotid, 1,1), subplot=str_sub(plotid, 2,2),
         sequence.date = date(sequence.start),
         sequence.year = year(sequence.date)) %>% 
  rename(sampledate.file=sampledate, plot.file = plot, subplot.file=subplot, plant.file=plant) %>% 
  left_join(vt %>% select(-c(colnames(treatments),"type","plantid")) %>% 
              filter(!exclude %in% c("flowers", "incomplete", "resample"))) %>% 
  left_join(treatments) %>% #assume that the written metadata is correct not the info in the  .file name
  mutate(total_time = end - bag,
         total_time = replace_na(total_time, median(total_time, na.rm=T)),# fill in 45 minutes when missing metadata
         pump_id=as.character(pump_id))
volsmeta <- left_join(volsmeta, volsmeta %>% count(sequence.date, sequence.start) %>% group_by(sequence.date) %>% mutate(sequence.12 = paste(sequence.date, row_number()))) #add the batch number for that day (1 or 2)
rownames(volsmeta) <- volsmeta$index 
```

# Sample size

Calculated from two datasets that will eventually match: datasheets from sampling and GC-MS filenames.

```{r samplesizes, fig.width=10, fig.height=5}
bind_rows(sampling=vt %>% group_by(year, type) %>% tally(),
  filenames=maxfmeta %>% mutate(year=factor(year(sampledate))) %>% group_by(year, type) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year,type), values_from = n) %>% kable()

bind_rows(sampling=vt %>% filter(type=="floral") %>% group_by(year, water, snow) %>% tally(),
  filenames=volsmeta %>% mutate(year=factor(year(sampledate.file))) %>% group_by(year, water, snow) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year), values_from = n) %>% kable()

bind_rows(sampling = vt %>% filter(type=="floral") %>% group_by(sampledate.file) %>% tally(),
         filenames = volsmeta %>% drop_na(sampledate.file) %>% group_by(sampledate.file) %>% tally(), .id="source") %>% 
  ggplot(aes(x=yday(sampledate.file), y=n, fill=source)) + geom_col(position=position_dodge(preserve="single")) +
  facet_wrap(vars(year(sampledate.file)), ncol=1) +
  labs(x="VOC sampling date (day of year)", y="Floral samples")
```

## Matching datasheets and filenames

```{r matching}
# diagnostics for matching up vt (datasheets) and volsmeta (filenames)
volsmeta.samp <- volsmeta %>% mutate(samp = paste(sampledate.file, plot.file, subplot.file, plant.file)) %>% arrange(samp) %>% pull(samp)
vt.samp <- vt %>% mutate(samp = paste(sampledate.file, plot.file, subplot.file, plant.file)) %>% arrange(samp) %>% pull(samp)
#intersect(volsmeta.samp, vt.samp) #in both volsmeta and vt
cat("Have files but not datasheet entries:")
setdiff(volsmeta.samp, vt.samp)
cat("Have datasheet entries but no files:")
setdiff(vt.samp, volsmeta.samp)

vt.volsmeta <- full_join(vt %>% select(vt.n, sampledate.file, plot.file, subplot.file, plant.file) %>% mutate(in.vt=T),
                         volsmeta %>% select(FileName, sampledate.file, plot.file, subplot.file, plant.file) %>% mutate(in.volsmeta=T)) %>% 
  mutate(across(c(in.vt, in.volsmeta), ~replace_na(.,FALSE)))
vt.volsmeta.count <- vt.volsmeta  %>% count(sampledate.file, in.vt, in.volsmeta) %>% mutate(status=paste("datasheet",ifelse(in.vt,"yes","no"),"+","GCfile", ifelse(in.volsmeta,"yes","no")))
#View(vt.volsmeta.count %>% filter(!in.vt | !in.volsmeta) %>% arrange(sampledate.file))

ggplot(vt.volsmeta.count, aes(x=as.Date(sampledate.file),y=n, fill=status))+ geom_col() +
  facet_wrap(vars(year(sampledate.file)), scales="free_x", ncol=1) +
  scale_x_date(breaks=scales::date_breaks(width = "7 days"), minor_breaks = scales::date_breaks(width = "1 days")) +
  theme_minimal() + theme(legend.position="top") + scale_fill_manual(values=c("red","orange","darkgreen")) +
  labs(x="Sampling date (according to GCfile)", y="Samples", fill="Data status")

ggplot(volsmeta %>% filter(year %in% c(2018, 2020) | sequence.year==year), aes(x=sequence.date - sampledate)) + 
  geom_histogram() + facet_wrap(vars(year), ncol=1, scales="free_x") + theme_minimal()
```


# Soil moisture

```{r sm}
ggplot(vt %>% drop_na(water), aes(x=factor(sampledate), y=VWC, color=water)) + facet_wrap(vars(year), ncol = 1, scales="free_x") +
  geom_boxplot() + scale_color_manual(values=water_pal) + scale_x_discrete(guide = guide_axis(angle=90)) + 
  labs(x="Date", y="Soil moisture (%VWC)", color="Precipitation")
```

# Ambient controls
```{r ambients}
vt.ambi %>% filter(type=="ambient") %>% mutate(year=year(sampledate)) %>% 
ggplot(aes(x=yday(sampledate))) + geom_bar() + facet_wrap(vars(year), ncol=1) + 
  labs(x="Day of year", y="Ambient samples") + scale_fill_manual(values=water_snow_pal)+  theme_minimal() 

setdiff(vt.ambi %>% filter(type=="floral") %>% pull(sampledate) %>% unique, 
        vt.ambi %>% filter(type=="ambient") %>% pull(sampledate) %>% unique) %>% as.Date(origin=ymd("1970-01-01"))

vt.ambi %>% count(type) #TODO check if all of these were run and used for filtering
volsmeta %>% count(year, type)
volsmeta %>% count(type)

ggplot(vt.ambi %>% filter(type=="ambient"), aes(x=hour(bag), fill=year)) + 
  geom_histogram(show.legend = F, binwidth=1) + facet_wrap(vars(year), ncol=1) + 
  scale_fill_manual(values=year_pal) + theme_minimal() + 
  labs(x="Ambient control bagged (hr)", caption= paste(format(c(min(volsmeta$bag, na.rm=T), max(volsmeta$bag, na.rm=T)), "%H:%M"), collapse=" - "))
```


# GCMS data checks

## Filtering

Filtering criteria (verdict="fvoc" in the table below):

* Present in 10% of samples
* Maximum peak area is at least 4000000
* Passes t-test against ambient controls with FDR correction OR mean exceeds 4x mean of controls
* Not a known contaminant (Oxybenzone, Ethanol, 2,2'-oxybis-, Butylated Hydroxytoluene, Caprolactam, [1,1'-Bicyclopentyl]-2-one, L-Lactic acid, Pentadecane, 8-hexyl-, Decane, 2,8,8-trimethyl-, 2-Methyltetradecan-2-ol)

Note that these potential floral compounds did not meet the tests against ambient controls: 2-phenylacetaldehyde, 3-carene, D-limonene, camphene, benzaldehyde, a-thujene, sabinene, abietadiene, 2-ethylhexan-1-ol 

Plus (verdict other than "fvoc"): 

* bis14: compounds included in Bischoff et al. (2014)
* keep: kept since they are known floral volatiles
* keep_leaf: kept since they are known leaf volatiles
* merge: a compound that is a synonym of one of the floral volatiles (b-caryophyllene, b-ocimene, verbenone)

```{r filtering}
vols <- maxf[volsmeta$index,]
vols.cut <- vols[,colSums(vols)>3e8]

ipochems <- read_csv("data/volatiles/chemsf_ipo.csv") %>% 
  filter(verdict %in% c("bis14", "fvoc", "keep","keep_leaf","merge"),
         name %in% colnames(vols)[colSums(vols)>0], name != "Anisole") %>% #filter empty compounds
  mutate(standard = fct_recode(standard, "Methyl_salicylate"="Benzaldehyde"), # benzaldehyde regressions not reliable
         class = fct_recode(standard, Aliphatic="Hexenol", Benzenoid="Methyl_salicylate", Benzenoid="Indole", 
                            Sequiterpene="Caryophyllene", Monoterpene="alpha_Pinene", Monoterpene="Linalool")) %>%  
  left_join(read_csv("data/volatiles/regressions_181921_filtered_slopes.csv") %>% 
              pivot_wider(id_cols=standard, names_from="year", names_prefix="area_per_ng", values_from=area_per_ng)) 
ipogood <- ipochems$name
vols.good <- vols[,ipogood] %>% 
  #sweep(2, ipochems$area_per_ng, FUN = '/') %>%  # convert areas to nanograms 
  sweep(1, as.numeric(volsmeta$total_time, units="hours"),  FUN = '/') #divide by equilibration + pumping time
for(yr in unique(volsmeta$sequence.year)) {
  thisyr <- volsmeta$sequence.year==yr
  vols.good[thisyr,] <- sweep(vols.good[thisyr,], 2, pull(ipochems, paste0("area_per_ng",yr)), FUN = '/')# convert areas to nanograms 
}

k <- 2 #split into this many k-means cluster per year
volsmeta$cluster <- 0
for(yr in unique(year(volsmeta$sequence.start))) {
  thisyr <- volsmeta$sequence.year==yr
  chems.include <- setdiff(colnames(vols.cut), ipogood) #drop alpha-pinene so this is all nonfloral
  set.seed(1)
  volsmeta$cluster[thisyr] <- kmeans(decostand(vols.cut[thisyr,chems.include], method="log"), k, nstart=3)$cluster
}
ipochems %>% select(verdict, name, RT, mean, max, match, freq.floral, freq.ambient, ambient_ratio, filter_ambient_ttest, filters_passed, standard) %>%
  arrange(desc(freq.floral)) %>% mutate(across(starts_with("freq."), round, 3)) %>% 
  DT::datatable(caption="Volatile compounds to use for analysis")

volsmeta <- volsmeta %>% 
  mutate(year.cluster = paste0(year, if_else(sequence.year == 2018, "", paste0(".",cluster))),
         total = rowSums(vols.good),
         sampledate = coalesce(sampledate, sampledate.file))#get sampling date from filename when no metadata

# Cut contaminated samples
nonzero <-rowSums(vols.good) > 0 #one sample ("55" on row 44) has no volatiles
set.seed(1) # ensures the contaminated samples are in cluster.good = 1
volsmeta$cluster.good <-factor(kmeans((vols.good)^(1/3), centers=3, nstart=3)$cluster)
set.seed(1) # ensures the contaminated samples are in cluster.cut = 1
volsmeta$cluster.cut <-factor(kmeans((vols.cut)^(1/3), centers=3, nstart=3)$cluster)

keep.samples <- unname(nonzero & !(volsmeta$year.cluster %in% c("2019.1", "2020.1")))
volsmeta <- volsmeta[keep.samples,]
vols <- vols[keep.samples,]
vols.cut <- vols.cut[keep.samples,]
vols.good <- vols.good[keep.samples,]

minsamples <- 70
vols.good.top <- vols.good[,colSums(decostand(vols.good,"pa"))> minsamples]#occur in more than minsamples
ipogood.top <- names(sort(colSums(vols.good.top), decreasing=T))
volsmeta.good.top <- bind_cols(volsmeta, vols.good.top) %>%  
  pivot_longer(-all_of(colnames(volsmeta))) %>% 
  mutate(name = fct_relevel(name, ipogood.top))

traits.good.top <- volsmeta.good.top %>% 
  left_join(mnps.plantyr %>% select(year, plotid, plant, all_of(traits))) %>% 
  mutate(name = fct_relevel(name, ipogood.top)) %>% 
  left_join(ph.plantyr %>% select(year, plotid, plant, open, eggs) %>% mutate(eggs_per_open = eggs/open))

vols.drydown <- vols[,c("1,3-Octadiene","Benzyl alcohol", "(1R)-2,6,6-Trimethylbicyclo[3.1.1]hept-2-ene",".alpha.-Farnesene", "Bicyclo[3.1.1]heptane, 6,6-dimethyl-2-methylene-, (1S)-", ".beta.-Ocimene")]
volsmeta.drydown <- bind_cols(volsmeta, vols.drydown) %>%  pivot_longer(-all_of(colnames(volsmeta)))

vols.class <- vols.good %>% t() %>% as.data.frame() %>% rownames_to_column("name") %>% 
  left_join(select(ipochems, name, class)) %>% 
  group_by(class) %>% summarize(across(-name, sum)) %>% column_to_rownames("class") %>%  t() %>% as.data.frame()
volsmeta.class <- bind_cols(volsmeta, vols.class) %>%  
  pivot_longer(-all_of(colnames(volsmeta)))
traits.class <- volsmeta.class %>% 
  left_join(mnps.plantyr %>% select(year, plotid, plant, all_of(traits), eggs)) %>% 
  mutate(eggs_per_open = eggs/open)
volsmeta.traits <- volsmeta %>% #wide format
  left_join(mnps.plantyr %>% select(year, plotid, plant, all_of(traits), eggs)) %>% 
  mutate(eggs_per_open = eggs/open)
```

## Sample size after filtering

```{r samplesizes_filtered, fig.width=10, fig.height=5}
volsmeta %>% count(year) %>% 
  kable(caption = "Samples sizes by year")
volsmeta %>% count(year, water, snow) %>% 
  pivot_wider(names_from=c(year), values_from = n) %>% 
  kable(caption = "Samples sizes by year and treatment")
volsmeta %>% count(year, sampledate) %>% 
  kable(caption = "Samples sizes by date")
#TODO investigate dates with only one sample - likely errors
volsmeta %>% count(year, sampledate) %>% filter(n>1) %>% count(year) %>% 
  kable(caption="More than one sample taken on these dates")
```

## Time of day

```{r sample_times}
ggplot(volsmeta, aes(x=bag, fill=year)) + geom_histogram(show.legend = F) + facet_wrap(vars(year), ncol=1) + 
  scale_fill_manual(values=year_pal) + theme_minimal() + 
  labs(x="Flower bagged", caption= paste(format(c(min(volsmeta$bag, na.rm=T), max(volsmeta$bag, na.rm=T)), "%H:%M"), collapse=" - "))
ggplot(volsmeta, aes(x=bag+total_time, fill=year)) + geom_histogram(show.legend = F) + facet_wrap(vars(year), ncol=1) + 
  scale_fill_manual(values=year_pal) + theme_minimal() + 
  labs(x="Pump stopped", caption= paste(format(c(min(volsmeta$bag+volsmeta$total_time, na.rm=T), max(volsmeta$bag+volsmeta$total_time, na.rm=T)), "%H:%M"), collapse=" - "))

volsmeta %>% count(beforenoon=hour(bag+total_time) <13)

ggplot(volsmeta, aes(x=end-bag)) + geom_histogram() + labs(x="Equilibration and pumping time (min)")
```

## Distributed among treatments

```{r random_sampling}
ggplot(volsmeta, aes(x=yday(sampledate), fill=paste(water, snow))) + geom_bar() + facet_wrap(vars(year), ncol=1) + 
  labs(fill="Precipitation, snowmelt", x="Day of year", y="Samples") + scale_fill_manual(values=water_snow_pal)+ theme_minimal() 

list(water=chisq.test(with(vt, table(sampledate, water))),
     snow=chisq.test(with(vt, table(sampledate, snow))),
     water_snow=chisq.test(with(vt, table(sampledate, paste(water, snow))))) %>% 
  map_dfr(tidy, .id="treatment") %>% kable(caption="Test of randomness of treatments on each sampling date")
```

## Flowers and buds

```{r buds}
volsmeta %>% count(flowers)
volsmeta %>% count(buds>0)
cat(paste("Buds included in", round(sum(volsmeta$buds>0, na.rm=T)/sum(volsmeta$buds>=0, na.rm=T)*100), "% of samples"))
```

## Heatmap

```{r heatmap, fig.width=16, fig.height=12}
annot_pal <- list(
  sequence.date = volsmeta %>% mutate(yr = sequence.year) %>% group_by(yr) %>% summarize(n = length(unique(sequence.date))) %>% 
    pull(n) %>% map(scales::hue_pal()) %>% unlist() %>% set_names(unique(volsmeta$sequence.date)),
  cluster = set_names(brewer.pal(8, "Set1")[1:length(unique(volsmeta$cluster))], unique(volsmeta$cluster)))

pheatmap(as.matrix(t(vols.cut))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),main="Unfiltered top volatiles",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors =  annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))

pheatmap(as.matrix(t(vols.good))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),main="Filtered volatiles",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors = annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))

pinene_correlates <- cor(sqrt(vols[,colSums(vols)>1e7])) %>% as.data.frame() %>% filter(`(1R)-2,6,6-Trimethylbicyclo[3.1.1]hept-2-ene`>0.15) %>% rownames()
pheatmap(as.matrix(t(vols[,pinene_correlates]))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),
                  annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors = annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0), main="Volatiles that correlate with alpha-pinene")

pheatmap(as.matrix(t(vols.class))^(1/4), 
         cluster_cols=F, cluster_rows=F, show_colnames=F,
         scale="none", color=mako(512),main="Filtered volatiles classes",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors =  annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))
```

## Chromatograms by GCMS run date

### Univariate

```{r gcms_date, fig.height=8, fig.width=7, eval=FALSE}
plot_vols_rundate <- function(type, checkvols) {
volsmeta %>% mutate(yr=year(Desorb.Start.Time)) %>% 
  bind_cols(vols %>% select(checkvols)) %>% pivot_longer(cols=checkvols) %>% 
ggplot(aes(x=Desorb.Start.Time)) + facet_grid(name ~ yr, scales="free") + 
  geom_point(aes(y=value, color=name), alpha=0.3)  + 
  guides(color="none") + theme_bw() + labs(x="GCMS run time", y="Area", title=type)
}

plot_vols_rundate("Terpenes", c("(1R)-2,6,6-Trimethylbicyclo[3.1.1]hept-2-ene", ".beta.-Myrcene",".beta.-Ocimene","Linalool","Caryophyllene", "Longifolene"))

plot_vols_rundate("Benzenoids", c("Benzophenone","Methyl salicylate","Anisole"))

plot_vols_rundate("Green leaf volatiles", c("4-Hexen-1-ol, (Z)-", "3-Hexen-1-ol, (E)-"))
```

### Multivariate

```{r gcms_sample_date, fig.height=6, fig.width=10, eval=FALSE}
for(categories in list(c("water","snow"),"plot","observer","pump_id","sequence.12", "sampledate", c("sampledate", "sequence.12"), c("sampledate.observer", "sequence.12"))) {
  cluster_counts <- volsmeta %>% 
    mutate(sampledate = coalesce(sampledate, sampledate.file),
           sampledate.observer = paste(sampledate, observer)) %>% 
    count(!!!syms(c("sequence.year", categories, "cluster")))
  p <- ggplot(cluster_counts, aes(x=factor(paste(!!!syms(categories), sep=" x ")), 
                                  y=n, fill=factor(paste0(sequence.year," C",cluster)))) + 
    geom_col() + facet_wrap(vars(sequence.year), scales="free_x") +
    theme(axis.text.x = element_text(angle=90)) + labs(fill="Cluster", x=paste(categories, collapse=" x ")) +
    scale_fill_brewer(palette="Paired")
  print(p)
  
  if(length(categories)>1) {
    (p2 <- cluster_counts %>% filter(sequence.year !="2018") %>% mutate(cluster=factor(cluster), across(where(~ !is.numeric(.)),factor)) %>% 
     ggplot(aes_string(x=categories[[1]], y = categories[[2]], size="n", color="cluster")) + geom_point(shape=1, stroke=1) +
      facet_wrap(vars(sequence.year), scales="free") + theme_minimal()+ scale_size_area(max_size=15)+
      theme(axis.text.x = element_text(angle=90)) + scale_color_manual(values=c("darkgreen","magenta")))
    print(p2)
  }
}

ggplot(volsmeta, aes(x=Desorb.Start.Time, y=cluster))  + geom_line() + geom_point(aes(color=factor(mday(sampledate))))+
  facet_wrap(vars(sequence.year), scales="free_x", ncol=1) + scale_color_manual(values=sample(rainbow(30)))+
  scale_x_datetime(date_breaks="1 day", guide=guide_axis(angle=90)) + scale_y_discrete()
```


```{r gcms_date_ord, eval=FALSE}
for(yr in unique(year(volsmeta$sequence.start))) {
  print(yr)
  yrfilter <- year(volsmeta$sequence.start) == yr
  vols.cap <- capscale(sqrt(vols.cut[yrfilter,]) ~ factor(date(sequence.start)), data=volsmeta[yrfilter,])
  print(vols.cap)
  print(anova(vols.cap, by="term"))
  # vols.cap.points <-  fortify(vols.cap) %>% as_tibble()
  # vols.cap.samples <- vols.cap.points %>% filter(Score=="sites") %>% 
  #   left_join(volsmeta, by=c("Label"="index"))
  # vols.cap.chems <- vols.cap.points %>% filter(Score=="species")
  # p <- vols.cap.samples %>% 
  #   mutate(sequence.date = date(sequence.start)) %>% 
  #   filter(sequence.year == yr) %>% 
  #   ggplot(aes(x=CAP1, y=CAP2)) + theme_classic()+
  #   geom_point(aes(color=factor(sequence.date)), size=3) + scale_color_hue("Run date")
  # print(p)
}
```

## Ordinations

### NMDS

```{r nmds, fig.height=9, fig.width=6}
(vols.nmds <- metaMDS(sqrt(vols.cut), autotransform=FALSE, trace=F))
vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
  left_join(volsmeta, by=c("Label"="index"))
vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")

(scents.nmds <- metaMDS(sqrt(vols.good), autotransform=FALSE, trace=F)) #had noshare = FALSE
nmds.points <-  fortify(scents.nmds) %>% as_tibble()
nmds.samples <- nmds.points %>% filter(Score=="sites") %>% left_join(volsmeta, by=c("Label"="index"))
nmds.chems <- nmds.points %>% filter(Score=="species") %>% left_join(ipochems, by=c("Label"="name"))

# By year
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() +
  geom_point(aes(color=year)) + scale_color_manual(values=year_pal)+
  labs(title="Unfiltered NMDS", color="Year")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  #geom_text(data=nmds.chems, aes(label=Label)) +
  geom_point(aes(color=year))+ scale_color_manual(values=year_pal)+
  #geom_point(aes(color=water, shape=year)) + scale_color_manual("Precipitation", values=water_pal) +
  labs(title="Filtered NMDS", color="Year")
grid.arrange(vols.plot, vols.good.plot)

# By a-Pinene
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2, shape=year)) + theme_classic() +
  geom_point(aes(color=vols.good$`(1R)-2,6,6-Trimethylbicyclo[3.1.1]hept-2-ene`^(1/6))) + 
  labs(title="Unfiltered NMDS", color="a-Pinene", shape="Year") + scale_color_viridis_c()
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2, shape=year)) + theme_classic() + 
  geom_point(aes(color=vols.good$`(1R)-2,6,6-Trimethylbicyclo[3.1.1]hept-2-ene`^(1/6)))+ 
  labs(title="Filtered NMDS", color="a-Pinene", shape="Year")+ scale_color_viridis_c()
grid.arrange(vols.plot, vols.good.plot)
```

```{r nmds_chems, fig.height=12, fig.width=12}
ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=year))+ scale_color_manual(values=year_pal)+
    geom_text(data=nmds.chems, aes(label=Label, alpha=freq.floral^(1/3))) +
  labs(title="Filtered NMDS", color="Year")
```


```{r nmds_clusters, eval=FALSE}
#Cluster based on filtered data
volsmeta %>% count(cluster.good, year) %>% arrange(year) %>% 
  pivot_wider(names_from="year", values_from="n") %>% 
  kable(caption = "Contaminated samples cluster = 1 based on filtered data")
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.good, year)))+ 
  labs(title="Unfiltered NMDS by year", color="Filtered cluster")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.good, year)))+ 
  labs(title="Filtered NMDS by year", color="Filtered cluster")
grid.arrange(vols.plot, vols.good.plot)

#cluster based on unfiltered data
set.seed(1)
volsmeta$cluster.cut <-factor(kmeans((vols.cut)^(1/3), centers=3, nstart=3)$cluster)
volsmeta %>% count(cluster.cut, year) %>% arrange(year) %>% 
  pivot_wider(names_from="year", values_from="n") %>% 
  kable(caption = "Contaminated samples cluster = 1 based on unfiltered data")
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.cut, year)))+ 
  labs(title="Unfiltered NMDS by year", color="Unfiltered cluster")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.cut, year)))+ 
  labs(title="Filtered NMDS by year", color="Unfiltered cluster")
grid.arrange(vols.plot, vols.good.plot)

# Ordinations for each year
for(yr in unique(year(volsmeta$sequence.start))) {
  thisyr <- year(volsmeta$sequence.start)==yr
  (vols.nmds <- metaMDS(sqrt(vols.cut[thisyr,]), autotransform=FALSE, trace=F))
  vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
  vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
    left_join(volsmeta, by=c("Label"="index")) %>% filter(sequence.year == yr)
  vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")
  
  (scents.nmds <- metaMDS(sqrt(vols.good[thisyr,]), autotransform=FALSE, trace=F))
  nmds.points <-  fortify(scents.nmds) %>% as_tibble()
  nmds.samples <- nmds.points %>% filter(Score=="sites") %>% 
    left_join(volsmeta, by=c("Label"="index")) %>% filter(sequence.year == yr)
  nmds.chems <- nmds.points %>% filter(Score=="species")
  
  vols.plot <- vols.nmds.samples %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Unfiltered NMDS run",yr))+
    geom_text(aes(color=factor(sequence.date), label=cluster), size=4) + scale_color_hue("Run date")
  vols.good.plot <- nmds.samples %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Filtered NMDS run",yr))+
        geom_text(aes(color=factor(sequence.date), label=cluster), size=4)  + scale_color_hue("Run date")
  grid.arrange(vols.plot, vols.good.plot)
}
```


# Plasticity analyses

## Plasticity to treatments

### CAP

```{r cap}
(scents.cap <- capscale(sqrt(vols.good) ~ year * water * snow, data=volsmeta))
anova(scents.cap, by="term")

scents.cap <- capscale(sqrt(vols.good) ~ year + water, data=volsmeta)
plot(scents.cap, type="n")
points(scents.cap, display = "sites", col=water_pal[volsmeta$water], pch=as.integer(volsmeta$year)+14)
text(scents.cap, display="cn", labels=c(levels(volsmeta$year), names(water_pal)), col=c(rep("black",5),water_pal), font=2, cex=1.5)

plot(scents.cap, type="n")
text(scents.cap, display="species", col="black", cex=0.5)

scents.cap <- capscale(sqrt(vols.good) ~ year + snow, data=volsmeta)
plot(scents.cap, type="n")
points(scents.cap, display = "sites", col=snow_pal[volsmeta$snow], pch=as.integer(volsmeta$year)+14)
text(scents.cap, display="cn", labels=c(levels(volsmeta$year), names(snow_pal)), col=c(rep("black",5),water_pal), font=2, cex=1.5)

plot(scents.cap, type="n")
text(scents.cap, display="species", col="black", cex=0.5)
```

### Joint species distribution model

```{r jsdm_splitplot}
library(Hmsc)

# model volatile abundance (emission rates) plasticity to year, water, and snow
# include random effect of plot
model.abu <- Hmsc(Y=vols.good,
                XData = volsmeta %>% select(year, water, snow) %>% mutate_all(as.factor), 
                XFormula= ~ year + water + snow, 
                studyDesign = volsmeta %>% select(plot) %>% mutate_all(as.factor),
                ranLevels = list(plot = HmscRandomLevel(units = levels(treatments$plot))),
                distr="lognormal poisson")

# Fit model with MCMC
thin <- 1
(model.abu <- sampleMcmc(model.abu, thin = thin, samples = 100, transient = thin*50, nChains = 3, nParallel = 3,
                        initPar="fixed effects")) #chains start from ML solution to fixed-effects model
```

```{r jsdm_splitplot_results}
# Model convergence
mpost <- convertToCodaObject(model.abu)[c("Beta","Gamma","V","Sigma")]
ess <- map(mpost, effectiveSize)
par(mfrow=c(2,2))
walk(names(ess), ~hist(ess[[.x]], xlab = expression("Effective sample size" ~ beta ~ ""), main=.x))
psrf <- map(mpost, gelman.diag, multivariate=FALSE)
walk(names(psrf), ~hist(psrf[[.x]]$psrf, xlab = expression("Potential scale reduction factor"), main=.x))

# Parameters
postBeta = getPostEstimate(model.abu, parName="Beta")
plotBeta(model.abu, post=postBeta, param="Support", supportLevel = 0.95, 
         spNamesNumbers = c(T,F), covNamesNumbers = c(T,F))

Gradient = constructGradient(model.abu,focalVariable = "water")
predY = predict(model.abu,Gradient = Gradient, expected = TRUE)
prob = c(0.25,0.5,0.75)
plotGradient(model.abu, Gradient, pred=predY, measure="S", showData = TRUE, q = prob, ylabel="Total emissions") 

Gradient = constructGradient(model.abu,focalVariable = "snow")
predY = predict(model.abu,Gradient = Gradient, expected = TRUE)
plotGradient(model.abu, Gradient, pred=predY, measure="S", showData = TRUE, q = prob, ylabel="Total emissions")

# Model fit (explanatory power)
predY = computePredictedValues(model.abu)
MF = evaluateModelFit(hM=model.abu, predY=predY)
tibble(name = ipogood, SR2 = MF$SR2) %>% mutate(name=fct_reorder(name, SR2)) %>% 
  filter(SR2>0.05) %>% 
  ggplot(aes(y=name, x=SR2))+ geom_point() + xlim(c(0,NA))+
  labs(x="squared Spearman correlation between obs and pred values (SR2)", 
       title="Volatiles with highest R2 in plasticity model") + theme_minimal()

# cross validation (predictive power)
partition = createPartition(model.abu, nfolds = 2)
#predY = computePredictedValues(model.abu, partition = partition, nParallel = 3)
#MFCV = evaluateModelFit(hM=model.abu, predY=predY)
#plot(MF$SR2, MFCV$SR2)
#abline(0,1)
```


### Boxplots

```{r boxplots, fig.width=16, fig.height=10}
treatment_boxplots <- function(volsmeta.vol, treatment) {
  ggplot(volsmeta.vol, aes_string(x = "year", fill = treatment, y = "value")) + 
    facet_wrap(vars(name), scales="free_y") + geom_boxplot(outlier.size=1) + scale_y_sqrt() +
    scale_fill_manual(values=list(water=water_pal, snow=snow_pal)[[treatment]]) + theme_bw() + 
    labs(x="Year", y ="Emissions (ng/hr)", fill=c(water="Precipitation",snow="Snowmelt")[[treatment]]) 
}
treatment_boxplots(volsmeta.good.top, "water")
#treatment_boxplots(volsmeta.drydown, "water")
treatment_boxplots(volsmeta.class, "water")
treatment_boxplots(volsmeta.good.top, "snow")
#treatment_boxplots(volsmeta.drydown, "snow")
treatment_boxplots(volsmeta.class, "snow")

ggplot(volsmeta, aes(x=year, fill=factor(paste(water, snow),levels=names(water_snow_pal)), y=total))  + 
  geom_boxplot(outlier.size=1) + 
  scale_fill_manual(values=water_snow_pal) + theme_minimal() +
  labs(x="Year", y ="Total filtered emissions (ng/hr)", fill="Precipiation, snowmelt")
```

## Plasticity to snowmelt date and precipitation

### CAP

```{r cap_abs}
(scents.cap <- capscale(sqrt(vols.good) ~ sun_date * precip_est_mm, data=volsmeta))
anova(scents.cap, by="term")

scents.cap <- capscale(sqrt(vols.good) ~ sun_date + precip_est_mm, data=volsmeta)
plot(scents.cap, type="n")
points(scents.cap, display = "sites", col=year_pal[volsmeta$year], pch=19)
text(scents.cap, display="bp", labels=c("Snowmelt date", "Precipitation"),  font=2)
legend("topleft", names(year_pal), fill=year_pal, title="Year")

plot(scents.cap, type="n", xlim=c(-1,2), ylim=c(-0.5,0.5))
text(scents.cap, display="species", col="black", cex=0.5, xpd=T)
```

### Scatterplots

```{r vols_abs, fig.width=16, fig.height=10}
precip.breaks <- seq(25,175, by=25)
water_scale <- scale_fill_gradientn("Summer precipitation (mm)", colors=rev(water_pal), 
                        values=rev(c(1, (treatments %>% group_by(water) %>% summarize(across(precip_est_mm, mean)) %>%
                                           pull(precip_est_mm) %>% scales::rescale(from=range(precip.breaks)))[2],0)),
                        breaks=precip.breaks)
year_scale <- scale_shape_manual(values=21:23, guide=guide_legend(override.aes=list(fill="black"))) 

ggplot(volsmeta.good.top, aes(x = sun_date, y = value)) + facet_wrap(vars(name), scales="free_y") + 
  geom_point(aes(fill=precip_est_mm, shape=year), color="white", size=2) + 
  geom_smooth(aes(color=water), se=F)+ theme_minimal()+ scale_y_sqrt()  + 
  labs(x="Snowmelt date", y ="Total filtered emissions (ng/hr)", shape="Year", color="Precipitation") +
  water_scale + year_scale + scale_color_manual(values=water_pal)

ggplot(volsmeta.class, aes(x = sun_date, y = value)) + facet_wrap(vars(name), scales="free_y") + 
  geom_point(aes(fill=precip_est_mm, shape=year), color="white", size=2) + 
  geom_smooth(aes(color=water), se=F)+ theme_minimal()+ scale_y_sqrt()  + 
  labs(x="Snowmelt date", y ="Total filtered emissions (ng/hr)", shape="Year", color="Precipitation") +
  water_scale + year_scale + scale_color_manual(values=water_pal)

ggplot(volsmeta, aes(x = sun_date, y = total)) + 
  geom_point(aes(fill=precip_est_mm, shape=year), color="white", size=2) + 
  geom_smooth(aes(color=water), se=F)+ theme_minimal()+ scale_y_sqrt()  + 
  labs(x="Snowmelt date", y ="Total filtered emissions (ng/hr)", shape="Year", color="Precipitation") +
  water_scale + year_scale + scale_color_manual(values=water_pal)
```


## Plasticity to soil moisture

### CAP

```{r cap_sm}
hasVWC <- !is.na(volsmeta$VWC)
(scents.cap <- capscale(sqrt(vols.good[hasVWC,]) ~ year.cluster * VWC, data=volsmeta[hasVWC,]))
anova(scents.cap, by="term")
```

### Scatterplots

```{r sm_vols, fig.width=16, fig.height=10}
ggplot(volsmeta.good.top, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt()  + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr)", color="Year") +
  scale_color_manual(values=year_pal)

ggplot(volsmeta.drydown, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr)", color="Year")+
  scale_color_manual(values=year_pal)

ggplot(volsmeta.class, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr)", color="Year")+
  scale_color_manual(values=year_pal)

ggplot(volsmeta, aes(x = VWC, color = year, y = total)) + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Total filtered emissions (ng/hr)", color="Year") +
  scale_color_manual(values=year_pal)
```

# Selection analyses

## Plots of selection in treatments and years
```{r selection, fig.width=16, fig.height=10}
ggplot(traits.good.top, aes(x=value, y=seeds_est, color=year)) + 
  geom_point() + geom_smooth(method="lm", se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Seed production", color="Year")+
  theme_bw()+ scale_x_sqrt()+
  scale_color_manual(values=year_pal)

ggplot(traits.good.top, aes(x=value, y=seeds_est, color=water, linetype=year, shape=year)) + 
  geom_point() + geom_smooth(method="lm",se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Seed production", color="Precipitation", linetype="Year", shape="Year")+
  theme_bw() + scale_color_manual(values=water_pal)+ scale_x_sqrt()

ggplot(traits.good.top, aes(x=value, y=seeds_est, color=snow, linetype=year, shape=year)) + 
  geom_point() + geom_smooth(method="lm",se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Seed production", color="Snowmelt", linetype="Year", shape="Year")+
  theme_bw() + scale_color_manual(values=snow_pal)+ scale_x_sqrt()

ggplot(traits.class, aes(x=value, y=seeds_est, color=water, linetype=year, shape=year)) + 
  geom_point() + geom_smooth(method="lm",se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Seed production", color="Precipitation", linetype="Year", shape="Year")+
  theme_bw() + scale_color_manual(values=water_pal)+ scale_x_sqrt()

ggplot(traits.class, aes(x=value, y=seeds_est, color=snow, linetype=year, shape=year)) + 
  geom_point() + geom_smooth(method="lm",se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Seed production", color="Snowmelt", linetype="Year", shape="Year")+
  theme_bw() + scale_color_manual(values=snow_pal)+ scale_x_sqrt()

ggplot(traits.class, aes(x=total, y=seeds_est, color=water, linetype=year, shape=year)) + 
  geom_point() + geom_smooth(method="lm",se=F) + 
  labs(x="Total emissions", y="Seed production", color="Precipitation", linetype="Year", shape="Year")+
  theme_bw() + scale_color_manual(values=water_pal)+ scale_x_sqrt()

ggplot(traits.class, aes(x=total, y=seeds_est, color=snow, linetype=year, shape=year)) + 
  geom_point() + geom_smooth(method="lm",se=F) + 
  labs(x="Total emissions", y="Seed production", color="Snowmelt", linetype="Year", shape="Year")+
  theme_bw() + scale_color_manual(values=snow_pal)+ scale_x_sqrt()
```

## Fly infestation vs. volatiles

```{r flies, fig.width=16, fig.height=10}
ggplot(traits.good.top %>% filter(fruits_nonaborted>4), aes(x=value, y=prop_infested, color=year)) + 
  geom_point() + geom_smooth(method="lm", se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y=paste(traitnames.units["prop_infested"], "(>4 fruits)"), color="Year")+
  theme_bw()+ scale_x_sqrt() + scale_color_manual(values=year_pal)

ggplot(traits.good.top, aes(x=value, y=eggs_per_open, color=year)) + 
  geom_point() + geom_smooth(method="lm",se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Mean fly eggs / mean open flowers", color="Year")+
  theme_bw()+ scale_x_sqrt() + ylim(0,1.2) +  scale_color_manual(values=year_pal)

ggplot(traits.class %>% filter(fruits_nonaborted>4), aes(x=value, y=prop_infested, color=year)) + 
  geom_point() + geom_smooth(method="loess",se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y=paste(traitnames.units["prop_infested"], "(>4 fruits)"), color="Year")+
  theme_bw()+  scale_x_sqrt()+  scale_color_manual(values=year_pal)

ggplot(traits.class, aes(x=value, y=eggs_per_open, color=year)) + 
  geom_point() + geom_smooth(method="loess",span=0.8,se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Mean fly eggs / mean open flowers")+
  theme_bw()+ scale_x_sqrt() + ylim(0,1.2)+  scale_color_manual(values=year_pal)

ggplot(traits.class %>% filter(fruits_nonaborted>4), aes(x=total, y=prop_infested, color=year)) + 
  geom_point() + geom_smooth(method="loess",span=0.8,se=F) + 
  labs(x="Total emissions", y=paste(traitnames.units["prop_infested"], "(>4 fruits)"), color="Year")+
  theme_bw()+  scale_x_sqrt()+  scale_color_manual(values=year_pal)

ggplot(traits.class, aes(x=total, y=eggs_per_open, color=year)) + 
  geom_point() + geom_smooth(method="loess",span=0.8,se=F) + 
  labs(x="Emissions", y="Mean fly eggs / mean open flowers", color="Year")+
  theme_bw()+  scale_x_sqrt() + ylim(0,1.2)+  scale_color_manual(values=year_pal)
```


## Joint species distribution model

```{r jsdm_selection}
library(Hmsc)

# model Y "species" (seeds, seeds per flower, prop fly fruits, fly eggs) 
# vs. XRRRData "environment" (volatiles) with reduced-rank regression

fitness.matrix <- volsmeta.traits %>% select(seeds_est, seeds_per_flower, prop_infested, eggs_per_open)
fitness.complete <- !is.na(rowSums(fitness.matrix)) & !is.infinite(rowSums(fitness.matrix))
  
(model.sel <- Hmsc(Y = fitness.matrix %>% filter(fitness.complete) %>% as.matrix(),
                   XData = volsmeta %>% filter(fitness.complete) %>% select(year),
                   XFormula = ~ year,
                   XRRRData = vols.good %>% filter(fitness.complete), 
                   XRRRScale = TRUE, ncRRR = 4, 
                   studyDesign = volsmeta %>% filter(fitness.complete) %>% select(plot) %>%
                     mutate_all(as.factor),
                   ranLevels = list(plot = HmscRandomLevel(units = levels(treatments$plot))),
                   distr=c("poisson", "poisson", "probit","poisson")))

# Fit model with MCMC
thin <- 1
(model.sel <- sampleMcmc(model.sel, thin = thin, samples = 100, transient = thin*50, 
                         nChains = 3, nParallel = 3)) 
```


```{r jsdm_selection_results}
# Model convergence
mpost <- convertToCodaObject(model.sel)[c("Beta","Gamma","V","Sigma")]
ess <- map(mpost, effectiveSize)
par(mfrow=c(2,2))
walk(names(ess), ~hist(ess[[.x]], xlab = expression("Effective sample size" ~ beta ~ ""), main=.x))

# Parameters
postBeta = getPostEstimate(model.sel, parName="Beta")
plotBeta(model.sel, post=postBeta, param="Support", supportLevel = 0.95, 
         spNamesNumbers = c(T,F), covNamesNumbers = c(T,F))

# Model fit (explanatory power)
predY = computePredictedValues(model.sel)
MF = evaluateModelFit(hM=model.sel, predY=predY)
tibble(name = c("seeds_est", "seeds_per_flower", "prop_infested", "eggs_per_open"), SR2 = MF$SR2) %>% 
  ggplot(aes(y=name, x=SR2))+ geom_point() +
  labs(x="squared Spearman correlation between obs and pred values (SR2)", 
       title="Model fit (pseudo R2)") + theme_minimal()
```

## Selection gradients

Following method of [Chong et al. 2018](https://doi.org/10.1002/evl3.63) for reconstructing selection estimates from estimates of selection on PCs: Beta = E * A.

* PCA can't handle missing values so they are removed, but this biases the data and reduces sample size - multiple imputation?

```{r gradients, fig.width=14, fig.height=10}
calculate_beta <- function(E, slopes) {
  data.frame(B = E %*% slopes$A,
             SE = sqrt(E ^ 2 %*% slopes$SE ^ 2)) %>% rownames_to_column("trait")
}

selection_pca <- function(year, water) {
vols.std <- bind_cols(volsmeta, vols.good.top) %>% 
  filter(year == .env$year, water %in% .env$water) %>% 
  left_join(mnps.plantyr %>% select(year, plotid, plant, seeds_est)) %>% 
  select(all_of(ipogood.top),seeds_est, any_of(names(treatments))) %>% 
  mutate(across(all_of(ipogood.top), ~ (.x-mean(.x, na.rm=T))/sd(.x, na.rm=T))) %>% 
  group_by(year) %>% mutate(seeds_rel = seeds_est/mean(seeds_est, na.rm=T), .keep="unused") %>% ungroup %>% 
  drop_na(seeds_rel) %>% 
  select(where(~sum(is.na(.))==0))  #drop_na(all_of(ipogood.top)) 

# Run PCA on traits
vols.PCA <- vols.std %>% select(any_of(ipogood.top)) %>% prcomp() 
summary(vols.PCA) # PCs 1 - 5 explain 51% of the variation
vols.PCA.E <- vols.PCA$rotation[,1:5]
vols.std.PCA <- bind_cols(vols.std, as.data.frame(vols.PCA$x))

# Regress relative fitness on PCs
vols.PCA.regression <- lm(seeds_rel ~ PC1 + PC2 + PC3 + PC4 + PC5, data = vols.std.PCA)
vols.PCA.slopes <- tidy(vols.PCA.regression) %>% filter(term != "(Intercept)") %>% rename(A=estimate, SE=std.error)

# Convert selection gradients on PCs to gradients on traits
calculate_beta(vols.PCA.E, vols.PCA.slopes)
}

plot_beta <- function(betas, colorby="year") {
  betas %>% mutate(trait = fct_reorder(trait, B)) %>% 
    ggplot(aes(y=trait, x = B, xmin=B-SE, xmax=B+SE)) + 
    geom_vline(xintercept=0)+ geom_pointrange(aes_string(color=colorby), position=position_dodge(width=0.5)) + 
    labs(x = "Selection gradient (beta) estimated from gradients on PCs", y = "Trait",
         color=list(year="Year", water="Precipitation", snow="Snowmelt")[[colorby]]) +
    scale_color_manual(values=list(year=year_pal, water=water_pal, snow=snow_pal)[[colorby]]) +
    theme_bw()
}

vols.beta.yr <- map_dfr(set_names(unique(volsmeta$year)), 
                     ~ selection_pca(.x, names(water_pal)), .id="year")
plot_beta(vols.beta.yr)

vols.beta.water <- expand_grid(year=unique(volsmeta$year), water=names(water_pal))  %>% 
  mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

plot_beta(filter(vols.beta.water), colorby="water") + facet_wrap(vars(year), nrow=1)
```

## Adaptive plasticity

### Selection gradient vs. plasticity

```{r adaptive, fig.height=14, fig.width=10}
plasticity <- volsmeta.good.top %>% group_by(year, water, name) %>% summarize(value=mean(value)) %>% 
  pivot_wider(names_from=water, values_from=value) %>% left_join(ipochems) %>% 
  mutate(A_C = (Addition-Control)/Control, 
         R_C = (Reduction-Control)/Control, 
         A_R = (Addition-Reduction)/Reduction, .keep="unused")

plasticity %>% left_join(vols.beta.water %>% rename(name=trait) %>% filter(water=="Addition")) %>% 
  ggplot(aes(x=A_C, y=B, color=class)) + facet_wrap(vars(year), ncol=1)+
  geom_vline(xintercept=0)+geom_hline(yintercept=0) + geom_text(aes(label=name)) + 
  scale_x_continuous(labels=scales::percent, limits=c(NA,NA)) + theme_bw()  + theme(legend.position = "top") +
  labs(x="Emissions change in Addition vs. Control precipitation", y="Selection gradient (beta) under precipitation addition")

plasticity %>% left_join(vols.beta.water %>% rename(name=trait) %>% filter(water=="Reduction")) %>% 
  ggplot(aes(x=R_C, y=B, color=class)) + facet_wrap(vars(year), ncol=1)+
  geom_vline(xintercept=0)+geom_hline(yintercept=0) + geom_text(aes(label=name)) + 
  scale_x_continuous(labels=scales::percent, limits=c(NA,NA)) + theme_bw()  + theme(legend.position = "top") +
  labs(x="Emissions change in Reduction vs. Control precipitation", y="Selection gradient (beta) under precipitation reduction")
```

### Change in selection gradient vs. plasticity

```{r adaptive_diff, fig.height=14, fig.width=10}
diffbeta <- vols.beta.water %>% rename(name=trait) %>% 
  pivot_wider(names_from=water, values_from=c(B,SE)) %>% 
  mutate(B_A_R = B_Addition - B_Reduction)

plasticity %>% left_join(diffbeta) %>% 
  ggplot(aes(x=A_R, y=B_A_R, color=class)) + facet_wrap(vars(year), ncol=1)+
  geom_vline(xintercept=0)+geom_hline(yintercept=0) + geom_text(aes(label=name)) + 
  scale_x_continuous(labels=scales::percent, limits=c(NA,NA)) + theme_bw()  + theme(legend.position = "top") +
  labs(x="Emissions change in precipitation addition vs. reduction", y="Change in selection gradient (beta) in precipitation addition vs. reduction")
```

# Volatiles vs. floral traits

```{r flrsize}
ggplot(traits.class, aes(x=corolla_length, y=value, color=year)) + 
  geom_point() + geom_smooth(se=F, span=1) + 
  facet_wrap(vars(name), scales="free_y") + 
  labs(y="Emissions", x=traitnames.units["corolla_length"], color="Year")+
  theme_bw() + scale_y_sqrt() + scale_color_manual(values=year_pal)

ggplot(traits.class, aes(x=sepal_width, y=value, color=year)) + 
  geom_point() + geom_smooth(se=F, span=1) + 
  facet_wrap(vars(name), scales="free_y") + 
  labs(y="Emissions", x=traitnames.units["sepal_width"], color="Year")+
  theme_bw() + scale_y_sqrt() + scale_color_manual(values=year_pal)

ggplot(traits.class, aes(x=nectar_24_h_ul, y=value, color=year)) + 
  geom_point() + geom_smooth(se=F, span=1) + 
  facet_wrap(vars(name), scales="free_y") + 
  labs(y="Emissions", x=traitnames.units["nectar_24_h_ul"], color="Year")+
  theme_bw() + scale_y_sqrt() + scale_color_manual(values=year_pal)
```

