--- 
title: "Maxfield Meadow Ipomopsis Volatiles"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>


```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)
library(vegan)
library(viridis)
library(ggvegan)
library(pheatmap)
library(DT)

library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images_volatiles2020/", fig.height=8, fig.width=8)
options(digits=4) # for kables

load("data/maxfield_data.rda")
load("data/volatiles/maxfield_volatiles.rda")

vt <- vt %>% rename(sampledate=date) %>% mutate(type=recode(type, air="ambient"))
volsmeta <- maxfmeta %>% mutate(year=factor(year(sampledate))) %>% 
  drop_na(plotid) %>% #cut out ambients
  filter(year %in% c("2018","2019","2020")) %>% droplevels() %>% 
  left_join(treatments) %>% 
  left_join(vt %>% select(observer:notes)) #TODO work on lining these up - 251/459 have no match in vt
rownames(volsmeta) <- volsmeta$index

vols <- maxf[volsmeta$index,]
```

# Sample size

Calculated from two datasets that will eventually match: datasheets from sampling and GC-MS filenames.

```{r samplesizes, fig.width=10, fig.height=5}
bind_rows(sampling=vt %>% group_by(year, type) %>% tally(),
  filenames=maxfmeta %>% mutate(year=factor(year(sampledate))) %>% group_by(year, type) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year,type), values_from = n) %>% kable()

bind_rows(sampling=vt %>% filter(type=="floral") %>% group_by(year, water, snow) %>% tally(),
  filenames=volsmeta %>% mutate(year=factor(year(sampledate))) %>% group_by(year, water, snow) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year), values_from = n) %>% kable()

bind_rows(sampling = vt %>% filter(type=="floral") %>% group_by(sampledate) %>% tally(),
         filenames = volsmeta %>% drop_na(sampledate) %>% group_by(sampledate) %>% tally(), .id="source") %>% 
  ggplot(aes(x=yday(sampledate), y=n, fill=source)) + geom_col(position=position_dodge(preserve="single")) +
  facet_wrap(vars(year(sampledate)), ncol=1) +
  labs(x="VOC sampling date (day of year)", y="Floral samples")
# TODO fix early 2020 date? and check singletons in 2018, 2019
# TODO there are some duplicate file names in maxfmeta, especially sampled 190805
```

# Soil moisture

```{r sm}
ggplot(vt %>% drop_na(water), aes(x=factor(sampledate), y=VWC, color=water)) + facet_wrap(vars(year), ncol = 1, scales="free_x") +
  geom_boxplot() + scale_color_manual(values=water_pal) + scale_x_discrete(guide = guide_axis(angle=90)) + 
  labs(x="Date", y="Soil moisture (%VWC)", color="Precipitation")
```

# Multivariate analyses

## Filtering

Filtering criteria (verdict="fvoc" in the table below):

* Present in 10% of samples
* Maximum peak area is at least 4000000
* Passes t-test against ambient controls with FDR correction OR mean exceeds 4x mean of controls
* Not a known contaminant (Oxybenzone, Ethanol, 2,2'-oxybis-, Butylated Hydroxytoluene, Caprolactam, [1,1'-Bicyclopentyl]-2-one, L-Lactic acid, Pentadecane, 8-hexyl-, Decane, 2,8,8-trimethyl-, 2-Methyltetradecan-2-ol)

Note that these potential floral compounds did not meet the tests against ambient controls: 2-phenylacetaldehyde, 3-carene, D-limonene, camphene, benzaldehyde, a-thujene, sabinene, abietadiene, 2-ethylhexan-1-ol 

Plus (verdict other than "fvoc"): 

* bis14: compounds included in Bischoff et al. (2014)
* keep: kept since they are known floral volatiles
* keep_leaf: kept since they are known leaf volatiles
* merge: a compound that is a synonym of one of the floral volatiles (b-caryophyllene, b-ocimene, verbenone)

```{r filtering}
vols.cut <- vols[,colSums(vols)>1e7]

ipochems <- read_csv("data/volatiles/chemsf_ipo.csv") %>% filter(verdict %in% c("bis14", "fvoc", "keep","keep_leaf","merge")) 
ipogood <- ipochems %>% pull(name)
vols.good <- vols[,intersect(colnames(vols),ipogood)]

ipochems %>% select(verdict, name, RT, mean, max, match, freq.floral, freq.ambient, ambient_ratio, filter_ambient_ttest, filters_passed) %>%
  arrange(desc(freq.floral)) %>% mutate(across(starts_with("freq."), round, 3)) %>% 
  DT::datatable(caption="Volatile compounds to use for analysis")
```

## Heatmap

```{r heatmap, fig.width=16, fig.height=12}
pheatmap(as.matrix(t(vols[,colSums(vols)>3e8]))^(1/4), color=mako(512),
         annotation_col=volsmeta %>% select(year))

pheatmap(as.matrix(t(vols.good))^(1/4), color=mako(512),
         annotation_col=volsmeta %>% select(year))
```

## Boxplots

```{r boxplots, fig.width=16, fig.height=10}
vols.good.top <- vols.good[,colSums(decostand(vols.good,"pa"))>81]#occur in more than N samples
bind_cols(volsmeta, vols.good.top) %>%  pivot_longer(-all_of(colnames(volsmeta))) %>% 
  ggplot(aes(x = year, fill = water, y = value)) + geom_boxplot() + facet_wrap(vars(name), scales="free_y") + 
  scale_y_sqrt() + scale_fill_manual(values=water_pal) + labs(x="Year", y ="Emissions (ng/hr)") + guides(color="none")
```

## Ordinations

### NMDS

```{r nmds}
nonzero <- rowSums(vols.good) > 0 #one sample ("55" on row 44) has no volatiles
(scents.nmds <- metaMDS(sqrt(vols.good[nonzero,]), autotransform=FALSE, trace=F)) #had noshare = FALSE

plot(scents.nmds,type="n")
points(scents.nmds, display = "sites", col=water_pal[volsmeta$water[nonzero]], pch=as.integer(volsmeta$year[nonzero])+14)
legend("topleft", legend=2018:2020, pch=15:17)
legend("topright", legend=names(water_pal), fill=water_pal)

plot(scents.nmds,type="n")
text(scents.nmds, display="species", col="black", cex=0.5)
```

### CAP
```{r cap}
(scents.cap <- capscale(sqrt(vols.good) ~ year * water * snow, data=volsmeta))
anova(scents.cap, by="term")

scents.cap <- capscale(sqrt(vols.good) ~ year + water, data=volsmeta)
plot(scents.cap, type="n")
points(scents.cap, display = "sites", col=water_pal[volsmeta$water], pch=as.integer(volsmeta$year)+14)
text(scents.cap, display="cn", labels=c(2018:2020, names(water_pal)), col=c(rep("black",3),water_pal), font=2, cex=1.5)

plot(scents.cap, type="n")
text(scents.cap, display="species", col="black", cex=0.5)
```

