---
title: "Natural selection on floral volatiles and other traits can change with snowmelt timing and summer precipitation"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>


```{r setup, message=FALSE}
library(tidyverse)
library(vegan)
library(broom)
library(broom.mixed)
library(emmeans)
library(viridis)
library(ggvegan)
library(ggpubr)
library(gridExtra)
library(dendsort)
library(pheatmap)
library(RColorBrewer)
library(lme4)
library(lmerTest)
library(glmnet)
library(bouquet)
library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, 
                      fig.path = "images_volatiles2020_cut/", dev="svglite", dev.args=list(fix_text_size=FALSE), fig.height=8, fig.width=8)
options(digits=4) # for kables
```

```{r rename_figs, eval=FALSE}
figs <- read_tsv("data/volatiles/volatiles_fig_paths.tsv")
walk2(figs$path, figs$fig, ~system(paste0("cp ", .x, " ./figures_volatiles2020_cut/Figure", .y, ".",tools::file_ext(.x))))
walk2(figs$path, figs$fig, ~system(paste0("inkscape -o ./figures_volatiles2020_cut/Figure", .y, ".eps ", .x)))
```

```{r loadscents}
load("data/maxfield_data.rda") #load trait measurements, etc
load("data/volatiles/maxfield_volatiles_lessdupes.rda") # from maxfield_gc.R, loads maxfmeta and maxf (wide GC data)

fitnesstraits <- c("seeds_est", "seeds_initiated_per_flower", "prop_uninfested", "eggs_per_flower")
fitnessnames <- c(traitnames, prop_uninfested = "Prop. uneaten fruits",
                  eggs_per_flower="Fly eggs per flower")[fitnesstraits]

volsmeta <- maxfmeta %>% mutate(sampledate = as.Date(sampledate), year = factor(year(sampledate))) %>% 
  drop_na(plotid) %>% #cut out ambients, leaving 459 chromatograms
  filter(year %in% c("2018","2019","2020")) %>% droplevels() %>% 
  mutate(plot=str_sub(plotid, 1,1), subplot=str_sub(plotid, 2,2),
         sequence.date = date(sequence.start),
         sequence.year = year(sequence.date)) %>% 
  rename(sampledate.file=sampledate, plot.file = plot, subplot.file=subplot, plant.file=plant) %>% 
  left_join(vt %>% select(-c(colnames(treatments),"type","plantid")) %>% 
              filter(!exclude %in% c("flowers", "incomplete", "resample"))) %>% 
  left_join(treatments) %>% #assume that the written metadata is correct not the info in the  .file name
  mutate(total_time = end - bag,
         total_time = replace_na(total_time, median(total_time, na.rm=T)),# fill in 45 minutes when missing metadata
         pump_id=as.character(pump_id))
rownames(volsmeta) <- volsmeta$index 
```

# Soil moisture

```{r sm, fig.height=5,fig.width=5}
ggplot(vt %>% drop_na(water) %>% mutate(date=sampledate), aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol = 1) +
  scale_x_continuous(limits=c(160, 230)) +
  geom_col(data=daily_precip_est, aes(y=EPA_NOAA_filled), fill=brewer.pal(7, "Set3")[7]) +
  geom_point(aes(y=VWC, color=water), size=0.5) + 
  geom_line(aes(y=VWC, color=water),
            data=vt %>% drop_na(water, VWC) %>% group_by(year, water, date=sampledate) %>% summarize(VWC=mean(VWC)), size=1) + 
  geom_line(aes(y=VWC, color=water),
            data=sm %>% group_by(year, water, date) %>% summarize(VWC=mean(VWC)), size=0.5, alpha=0.5) + 
  scale_color_manual(values=water_pal) + 
  labs(x="Day of year", y="Soil moisture (%VWC) | Daily precipitation (mm)", color="Precipitation") + theme_bw()
```

# Process GCMS data

## Filtering with bouquet

```{r bouquetprep}
# load short names and standard regessions
ipochems <- read_csv("data/volatiles/Ipo volatile compounds - chemsf_ipo.csv") %>% 
  select(name, shortname, standard, verdict) %>%  filter(verdict != "") %>% 
  mutate(standard = fct_recode(standard, "Methyl_salicylate"="Benzaldehyde"), # benzaldehyde regressions not reliable
         class = fct_recode(standard, Aliphatics="Hexenol", Benzenoids="Methyl_salicylate", Benzenoids="Indole", 
                            Sesquiterpenes="Caryophyllene", Monoterpenes="alpha_Pinene", Monoterpenes="Linalool")) %>%  
  left_join(read_csv("data/volatiles/regressions_181921_filtered_slopes.csv") %>% 
              pivot_wider(id_cols=standard, names_from="year", names_prefix="area_per_ng", values_from=area_per_ng)) 
class_pal <- set_names(c("#BC0060","#027B8C","#E56E00","#86D400"), levels(ipochems$class))

#shorten chemical names and merge compounds with multiple names
shortnames <- ipochems %>% select(name, shortname) %>% filter(shortname!="") %>% deframe()

greekify <- function(names) {
  names %>% 
    str_replace("^a-","\U03B1-") %>% str_replace("-a-","-\U03B1-") %>% 
    str_replace("^b-","\U03B2-") %>% str_replace("-b-","-\U03B2-") %>% 
    str_replace("^g-","\U03B3-") %>% str_replace("-g-","-\U03B3-")
}

metadata <- maxfmeta %>% 
  mutate(sampledate = as.Date(sampledate), year = year(sampledate)) %>% 
  filter(year != 2021) %>% mutate(year=factor(year)) %>% #drop samples collected in 2021
  as.data.frame() %>% #tibbles break bouquet's add_count_freqs
  select(-sample) %>% #name conflict with bouquet
  load_metadata(date="Date", sample="FileName", type="type")#group="year"

load("data/volatiles/maxfield210827.Rdata") #load long maxf.data
longdata <- maxf.data %>% 
  filter(batch != "heather_maxfield190813_new.txt", #duplicates - subset of heather_maxfield190813_new001.txt
         Filename %in% metadata$sample, #match list in metadata
         Name != "") %>% #drop unidentified peaks
  mutate(Name = droplevels(recode(Name, !!!shortnames))) %>% 
  load_longdata(sample="Filename", RT="Ret.Time", name="Name", area="Area", match = "SI", maxmatch=100)
remove(maxf.data)

vol.all <- make_sampletable(longdata, metadata)
chems <- make_chemtable(longdata, metadata)
```

```{r bouquet}
chemsf <- chems %>%
  filter_RT(2, 17) %>% 
  filter_match(0.8) %>% 
  filter_freq(0.1) %>% 
  filter_contaminant(cont.list = c("Butanenitrile, 2-methyl-","2-Oxepanone, 7-methyl-", #RT order
                                   "Ethylbenzene", "Hexyl chloroformate", "L-Lactic acid",
                                   "1,5-Dimethyl-6-oxa-bicyclo[3.1.0]hexane","Cyclopentanone, 2-ethyl-",
                                   "trisiloxane, 1,1,1,5,5,5-hexamethyl-3-[(trimethylsilyl)oxy]-",
                                   "Cyclobutane, 1,2-bis(1-methylethenyl)-, trans-",
                                   "N-.alpha.,N-.omega.-Di-cbz-L-arginine", "3-Ethyl-3-methylheptane",
                                   "Trichloroacetic acid, 6-ethyl-3-octyl ester", "11-Methyldodecanol",
                                   "8-Hexadecenal, 14-methyl-, (Z)-",
                                   "Tritetracontane","Carbonic acid, decyl undecyl ester",
                                   "Eicosane, 2,4-dimethyl-", "1-Decanol, 2-octyl-",
                                   "linalyl acetate", #kovats doesn't match
                                   "[(Z)-hex-3-enyl] 2-methylpropanoate",  #merge into [(E/Z)-hex-3-enyl] butanoate
                                   "[(E)-hex-4-enyl] butanoate", #merge into [(E/Z)-hex-3-enyl] butanoate
                                   "[(Z)-hex-3-enyl] 2-methylbutanoate", #passed filters, only in 50 samples after quant
                                   "2-butyloctan-1-ol",#absent in 2019
                                   "(2E)-2,7-dimethylocta-2,6-dien-1-ol" #a quant integration that coelutes with b-myrcene
                                   )) %>%
  filter_area(min_maximum = 1e5) %>% #not used for final filtering
  filter_ambient_ratio(vol.all, metadata, ratio = 4) %>%
  combine_filters() 

# removed (2E)-2,7-dimethylocta-2,6-dien-1-ol - quant integration that coelutes with b-myrcene
add_quant2 <- c("(4E,6Z)-alloocimene", "methyl benzoate", "2-methylbenzonitrile", "(Z)-b-ocimene", "(Z)-a-bergamotene", "g-terpinene", "(E)-b-bergamotene", "a-humulene", "pseudoionone")

chemsf$filter_final <- with(chemsf, filter_RT=="OK" & filter_match=="OK" & 
                              filter_freq.floral=="OK" & filter_contaminant=="OK" |
                             name %in% c("[(E)-hex-3-enyl] butanoate", add_quant2)) #include as counterpart to Z isomer

#now run t-tests after rarity filtering
#don't filter by ambient ratio first, see paper on problems: https://doi.org/10.1186/1471-2105-11-450
chemsf <- bind_rows(chemsf %>% filter(filter_final) %>% 
  filter_ambient_ttest(prune_sampletable(vol.all, chemsf, metadata), 
                       metadata, alpha = 0.05, adjust = "fdr"), #not used for final filtering
  chemsf %>% filter(!filter_final)) %>% 
  mutate(filter_ambient_ttest=fct_na_value_to_level(filter_ambient_ttest, "notchecked"),
         filter_final = filter_final & filter_ambient_ratio =="OK") %>% 
  arrange(match(name, colnames(vol.all)))
```

```{r quant}
#vols <- maxf[volsmeta$index,]
vols <- vol.all[volsmeta$FileName,]
vols.cut <- vols[,colSums(vols)>3e8]
rownames(vols.cut) <- volsmeta$index
vol <- vols[volsmeta$FileName, chemsf$name[chemsf$filter_final]]#don't cut ambients yet
rownames(vol) <- volsmeta$index

# Shimadzu Quantitative Table ---------------------------------------------
# source("read_shimadzu.R")
# quantpath <- "~/MyDocs/MEGA/UCI/Schiedea/Analysis/scent/rmbl/RMBL Batches/quant_round4/"
# bfiles <-  list.files(quantpath)
# quant.full <- map_dfr(set_names(paste0(quantpath, bfiles), bfiles), read.shimadzu.quant, .id="batch") %>%
#   mutate(Name = trimws(Name), Area=replace_na(Area, 0)) %>%
#   distinct(Filename, Name, .keep_all=T)
# 
# quant.full %>% drop_na(Ret.Time) %>% mutate(Name = fct_reorder(Name, Ret.Time)) %>%
#   ggplot(aes(x=Ret.Time, y=Name, color=batch)) + geom_jitter(width=0, size=0.2)
# quant.full %>% drop_na(Ret.Time) %>% group_by(Name) %>% mutate(RT.median = median(Ret.Time)) %>%
#   ggplot(aes(x=Ret.Time-RT.median, y=Name, color=batch)) + geom_jitter(width=0, size=0.5)
# 
# quant <- quant.full %>% select(batch, Dirname, Filename, Name, Area) %>%
#   mutate(Dirname = fct_relabel(Dirname, str_remove, "C:/GCMSsolution/Data/Project1_190815/")) %>%
#   filter(!Name %in% c("1,6,10-Dodecatrien-3-ol, 3,7,11-trimethyl-",
#          "4,7-Methano-1H-indene, 2,4,5,6,7,7a-hexahydro-",
#          "Tricyclo[3.1.0.0(2,4)]hexane, 3,6-diethyl-3,6-dimethyl-, trans-")) %>%
#   mutate(Name = recode(Name, !!!shortnames)) %>%
#   pivot_wider(names_from = "Name", values_from="Area") %>% as.data.frame()
# 
# rownames(quant) <- quant$Filename
# quant[,c("batch","Dirname","Filename")] <- NULL
# quant.all <- quant[rownames(vols),]
# write_tsv(quant.all %>% rownames_to_column("Filename"), "data/volatiles/quant_all.tsv")
quant.all <- read_tsv("data/volatiles/quant_all.tsv") %>% column_to_rownames("Filename")

quant.slopes <- set_names(rep(0,ncol(quant.all)), colnames(quant.all))
for(compvol in colnames(quant.all)) {
  nozeros <- bind_cols(select(vols,qual=compvol), select(quant.all,quant=compvol)) %>% 
    filter(!(qual==0 | quant == 0))
  quant.slopes[compvol] = coef(lm(quant ~ 0 + qual, data=nozeros))["qual"]
}
write_tsv(enframe(quant.slopes) %>% left_join(enframe(shortnames, name="Name", value="name")),
          "data/volatiles/quant_slopes.tsv")

quant.scaled <- sweep(quant.all, 2, quant.slopes, FUN = '/')

quant.scaled.vol <- quant.scaled[volsmeta$FileName, 
                                 colnames(quant.scaled) %in% 
                                   intersect(colnames(quant.all), colnames(vol))] 

#swap in scaled quant integrations
vol[,colnames(quant.scaled.vol)] <- quant.scaled.vol
```



```{r filtering}
ipogood <- colnames(vol)
vols.good <- vol %>% 
  sweep(1, as.numeric(volsmeta$total_time, units="hours"),  FUN = '/') #divide by equilibration + pumping time
ipochemsf <- ipochems[match(colnames(vols.good), ipochems$shortname),]
for(yr in unique(volsmeta$sequence.year)) {
  thisyr <- volsmeta$sequence.year==yr
  vols.good[thisyr,] <- sweep(vols.good[thisyr,], 2, pull(ipochemsf, paste0("area_per_ng","2019")), FUN = '/')# convert areas to nanograms 
}

ipochemsf <- ipochemsf %>% 
  left_join(tibble(shortname = colnames(vols.good), # the following values are new, computed on the filtered dataset
                            freq.floral = map_dbl(as.data.frame(vols.good>0), mean),
                            mean.floral = map_dbl(vols.good, mean),
                            max.floral =  map_dbl(vols.good, max))) %>% 
  select(-name) %>% rename(name=shortname) %>%
  mutate(class=droplevels(class))#no Benzenoids
rownames(ipochemsf) <- ipochemsf$name

k <- 2 #split into this many k-means cluster per year
volsmeta$cluster <- 0
for(yr in unique(year(volsmeta$sequence.start))) {
  thisyr <- volsmeta$sequence.year==yr
  chems.include <- setdiff(colnames(vols.cut), ipogood) #drop alpha-pinene so this is all nonfloral
  set.seed(1)
  volsmeta$cluster[thisyr] <- kmeans(decostand(vols.cut[thisyr,chems.include], method="log"), k, nstart=3)$cluster
}

volsmeta <- volsmeta %>% 
  mutate(year.cluster = paste0(year, if_else(sequence.year == 2018, "", paste0(".",cluster))),
         total = rowSums(vols.good),
         sampledate = coalesce(sampledate, sampledate.file))#get sampling date from filename when no metadata

#Cuts 2 samples with extemely low emissions that fly off of nmds blob
keep.samples <- rowSums(vols.good)>3
volsmeta <- volsmeta[keep.samples,]
vols <- vols[keep.samples,]
vols.cut <- vols.cut[keep.samples,]
vols.good <- vols.good[keep.samples,]
```

## Average by plant 

```{r averaging}
volsmeta %>% count(year, plantid, name="n_samples") %>% count(n_samples) %>% kable(caption="number of samples per plant")

volsmeta.avg <- volsmeta %>% 
  group_by(across(all_of(c("type", "plantid", colnames(treatments))))) %>% 
  summarize(across(all_of(c("VWC", "total")), ~mean(.x, na.rm=T)), 
            index=index[1], .groups="drop")

vols.good.avg <- vols.good %>% bind_cols(volsmeta) %>% 
  group_by(across(all_of(c("plantid", colnames(treatments))))) %>% 
  summarize(across(all_of(colnames(vols.good)), ~mean(.x, na.rm=T)), .groups="drop") %>% 
  select(-any_of(colnames(volsmeta)))
rownames(volsmeta.avg) <- rownames(vols.good.avg) <- volsmeta.avg$index
```


```{r topvols}
minsamples <- nrow(vols.good) * 0.75 #top 12 most common volatiles, in 75% of samples
ipogood.top <- vols.good[colSums(decostand(vols.good,"pa")) > minsamples] %>% 
  colSums() %>% sort(decreasing=T) %>% names()
vols.good.top <- vols.good.avg[,ipogood.top]

volsmeta.good.top <- bind_cols(volsmeta.avg, vols.good.top) %>%  
  pivot_longer(-all_of(colnames(volsmeta.avg))) %>% 
  mutate(name = fct_relevel(name, ipogood.top))

mnps.plantyr.traits <- mnps.plantyr %>% 
  select(year, plotid, plantid, all_of(traits), eggs_per_flower) %>%
  mutate(prop_uninfested = 1 - prop_infested) %>% 
  mutate(plantid = str_remove(plantid, "_[12][890]")) %>% #disambiguations on the plantid already handled by year column, except...
  group_by(year, plotid, plantid) %>% #...two plants need to be combined in 2019
  summarize(across(everything(), ~.x[!is.na(.x)][1]), .groups = "drop")

traits.good.top <- volsmeta.good.top %>% 
  left_join(mnps.plantyr.traits) %>% 
  mutate(name = fct_relevel(name, ipogood.top))

vols.class <- vols.good.avg %>% t() %>% as.data.frame() %>% rownames_to_column("name") %>% 
  left_join(select(ipochemsf, name=name, class)) %>% 
  group_by(class) %>% summarize(across(-name, sum)) %>% column_to_rownames("class") %>%  t() %>% as.data.frame()
volsmeta.class <- bind_cols(volsmeta.avg, vols.class) %>%  
  pivot_longer(-all_of(colnames(volsmeta.avg)))
traits.class    <- volsmeta.class %>% left_join(mnps.plantyr.traits)
volsmeta.traits <- volsmeta.avg %>%       left_join(mnps.plantyr.traits) #wide format
vols.traits <- bind_cols(volsmeta.avg, vols.good.avg) %>% left_join(mnps.plantyr.traits)
```


## Compound amounts

```{r relamounts}
quant.RT <- read_tsv("data/volatiles/quant_RT.tsv")
load("data/volatiles/get_kovats.rda") # from RMBL-GCMS
quant.RI.2018 <- quant.RT %>% filter(year==2018) %>% mutate(RI = get_kovats(RT.median, "2018-09-03")) %>% 
  mutate(name = shortnames[Name])

vols.good.avg %>% #average by plant first
  pivot_longer(everything()) %>% group_by(name) %>% summarize(mean_emissions = mean(value), sd_emissions=sd(value)) %>% 
  left_join(ipochemsf) %>%  arrange(class, desc(mean_emissions)) %>% 
  left_join(quant.RI.2018) %>% 
  mutate(meansd = paste(sprintf(mean_emissions, fmt = '%#.1f'), "\U00B1", sprintf(sd_emissions, fmt = '%#.1f')),
         name=greekify(name), RI= round(RI)) %>% 
  select(class, RI, name, meansd) %>%  kable(caption = "Mean emission rates of each volatile")
```


## Sample size

```{r samplesizes_filtered, fig.width=10, fig.height=5}
volsmeta %>% count(year, name="samples") %>% 
  left_join(volsmeta %>% count(year, plantid) %>% count(year, name="plants")) %>% 
  kable(caption = "Samples and plants by year")
volsmeta %>% count(year, sampledate) %>% count(year) %>% 
  kable(caption="Number of sampling dates")
```

## Distribution among treatments

```{r random_sampling, fig.width=5, fig.height=4}
ggplot(volsmeta, aes(x=yday(sampledate), fill=paste(water, snow))) + geom_bar() + facet_wrap(vars(year), ncol=1) + 
  labs(fill="Precipitation, snowmelt", x="Day of year", y="Samples") + scale_fill_manual(values=water_snow_pal)+ theme_minimal() 

vt %>% group_by(water, snow) %>% summarize(bag=mean(bag, na.rm=T)) %>% kable(caption = "Mean bagging time")
```

## Buds

```{r buds}
cat(paste("Buds included in", round(sum(volsmeta$buds>0, na.rm=T)/sum(volsmeta$buds>=0, na.rm=T)*100), "% of samples"))
```

# Plasticity

## Plasticity to treatments

### CAP

```{r cap}
(scents.cap <- capscale(sqrt(vols.good.avg) ~ year * water * snow, data=volsmeta.avg, distance="bray"))
anova(scents.cap, by="term") %>% kable()
```

### Emissions by class

```{r boxplots, fig.width=16, fig.height=10}
treatment_boxplots <- function(volsmeta.vol, treatment) {
  ggplot(volsmeta.vol %>% 
           mutate(water_snow = factor(paste(water, snow), levels=names(water_snow_pal))),
         aes_string(x = "year", fill = treatment, y = "value")) + 
    facet_wrap(vars(name), scales="free_y", labeller = as_labeller(greekify)) + geom_boxplot(outlier.size=0.2) + scale_y_sqrt() +
    scale_fill_manual(values=list(water=water_pal, snow=snow_pal,
                                  water_snow=water_snow_pal)[[treatment]]) + theme_bw() + 
    labs(x="Year", y ="Emissions (ng/hr/flower)", 
         fill=c(water="Precipitation",snow="Snowmelt", 
                water_snow="Precipitation, snowmelt")[[treatment]]) 
}
```

```{r boxplots_class, fig.width=7, fig.height=5}
treatment_boxplots(volsmeta.class, "water_snow")

options(contrasts = c("contr.sum", "contr.poly"))
treatment_effects <- function(class) {
  lmer(sqrt(value) ~ year * water * snow + (1|plot) + (1|plotid), 
       data = filter(volsmeta.class, name==class))
}

mod.split <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, treatment_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.split %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to treatments")
```

## Plasticity to snowmelt date and precipitation

### CAP

```{r cap_abs, fig.height=6.5, fig.width=6.5}
(scents.cap <- capscale(sqrt(vols.good.avg) ~ sun_date * precip_est_mm, data=volsmeta.avg, distance="bray"))
anova(scents.cap, by="term") %>% kable()

scents.cap <- update(scents.cap, ~ sun_date + precip_est_mm)
cap.df <- fortify(scents.cap) #drop non-signif interaction for plotting
prop.expl <- summary(scents.cap)$cont$importance
ax.labs <- paste0(colnames(prop.expl)[1:2], " (",round(100*prop.expl[2,1:2]),"% explained)")
ggplot(mapping=aes(x=CAP1, y=CAP2, label=greekify(label))) + 
  geom_point(data=cap.df %>% filter(score=="sites") %>% bind_cols(volsmeta.avg), aes(color=year))+
  geom_segment(data=cap.df %>% filter(score=="biplot"), aes(x=0,y=0,xend=CAP1*1.8, yend=CAP2*1.8), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(data=cap.df %>% filter(score=="biplot") %>% 
              mutate(label=c("Snowmelt", "Precipitation")),
            aes(x=CAP1*2, y=CAP2*2, angle=atan(CAP2/CAP1)*180/pi), hjust=c(0,1), fontface="bold") +
  geom_text(data=cap.df %>% left_join(ipochemsf, by=c("label"="name")) %>% filter(score=="species", mean.floral>1.5) %>% 
              mutate(nudge = c(0,0,0,0,0,0,1,0,0,0,0,0,1,0,0,0),
                     label = recode(label, "[(E)-hex-3-enyl] butanoate"="unknown ester", "[(E)-hex-3-enyl] butanoate"="unknown ester", "[(E)-hex-3-enyl] acetate"="[(Z)-hex-3-enyl] acetate")),
            aes(x=CAP1*3, y=CAP2*3+nudge*0.1), size=3)+ labs(x=ax.labs[1], y=ax.labs[2])+
  scale_color_manual("Year", values=year_pal) + coord_fixed() + 
  theme_bw() + theme(panel.grid = element_blank())
```

### Emissions by class

```{r mod_abs}
options(contrasts = c("contr.sum", "contr.poly"))
abs_effects <- function(class) {
  lmer(sqrt(value) ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), 
       data = filter(volsmeta.class, name==class))
}

mod.abs <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, abs_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.abs %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to snowmelt date and estimated precipitation")
```

### Snowmelt date

```{r vols_abs, fig.height=5, fig.width=7}
precip.breaks <- seq(25,175, by=25)
grid.points <- list(sun_date=range(treatments$sun_date), precip_est_mm=precip.breaks)
options(contrasts = c("contr.sum", "contr.poly"))
#TODO this code partially repeats chunk above
mod.class.abs <- volsmeta.class %>% group_by(name) %>% nest() %>% 
  mutate(model = map(data, ~lmer(value ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), data=.)),
         test = map(model, ~tidy(anova(.x, ddf="Ken"))),
         emm.grid = map(model, ~ summary(emmeans(.x, ~ precip_est_mm * sun_date, at=grid.points))))

ggplot(volsmeta.class, aes(x = sun_date, y = value, color=precip_est_mm)) + facet_wrap(vars(name), scales="free_y") + 
  geom_point(aes(shape=year), size=2) + 
    geom_line(data = select(mod.class.abs, name, emm.grid) %>% unnest(emm.grid),
            aes(y=emmean, group=precip_est_mm))+
  theme_minimal()+ scale_y_sqrt()  + 
  labs(x="Snowmelt date", y ="Emissions (ng/hr/flower)", shape="Year", color="Precipitation") +
  scale_color_gradientn("Summer precipitation (mm)", colors=rev(water_pal), 
                        values=rev(c(1, (treatments %>% group_by(water) %>% summarize(across(precip_est_mm, mean)) %>%
                                           pull(precip_est_mm) %>% scales::rescale(from=range(precip.breaks)))[2],0)),
                        breaks=precip.breaks)
```

## Plasticity to plant-level soil moisture

### CAP

```{r cap_sm}
hasVWC <- !is.na(volsmeta.avg$VWC)
(scents.cap <- capscale(sqrt(vols.good.avg[hasVWC,]) ~ year * VWC, data=volsmeta.avg[hasVWC,], distance="bray"))
anova(scents.cap, by="term") %>% kable()
```

### Emissions by class

```{r mod_sm}
options(contrasts = c("contr.sum", "contr.poly"))
sm_effects <- function(class) {
  lmer(sqrt(value) ~ year * VWC + (1|plotid), 
       data = filter(volsmeta.class, name==class))
}

mod.sm <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, sm_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.sm %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to soil moisture")
```

```{r sm_vols, fig.width=5, fig.height=5}
ggplot(volsmeta.class, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") + 
  geom_point(size=0.5) + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year")+
  scale_color_manual(values=year_pal)
```

## Plasticity to subplot-level soil moisture

### CAP

```{r cap_sm_subplot}
(scents.cap <- capscale(sqrt(vols.good.avg) ~ year * VWC, data=left_join(select(volsmeta.avg,-VWC), sm.subplotyear),
                        distance="bray"))
anova(scents.cap, by="term") %>% kable()
```

### Emissions by class

```{r mod_sm_subplot}
options(contrasts = c("contr.sum", "contr.poly"))
sm_effects <- function(class) {
  lmer(sqrt(value) ~ year * VWC + (1|plotid), 
       data = filter(left_join(select(volsmeta.class,-VWC), sm.subplotyear), name==class))
}

mod.sm <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, sm_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.sm %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to subplot-level soil moisture")
```

```{r sm_subplot_vols, fig.width=5, fig.height=5}
ggplot(left_join(select(volsmeta.class,-VWC), sm.subplotyear), aes(x = VWC, color = year, y = value)) + 
  facet_wrap(vars(name), scales="free_y") + 
  geom_point(size=0.5) + geom_smooth(se=F, span=1)+ theme_minimal()+ 
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year")+
  scale_color_manual(values=year_pal)
```

# Selection

## Emissions by class

```{r selection_class_gradient, fig.width=5.5, fig.height=6.5}
trait.class.wide <- bind_cols(volsmeta.avg, vols.class) %>% left_join(mnps.plantyr.traits) %>% 
  select(plantid, year, water, snow, all_of(fitnesstraits), all_of(names(class_pal))) 
 #mutate(across(all_of(names(class_pal)), sqrt))

selection_effects <- function(fitnesstrait) {
  options(contrasts = c("contr.sum", "contr.poly"))
  lm(relfitness ~ year * water * snow + 
     Monoterpenes * snow + Monoterpenes * water + Monoterpenes * year +
     Sesquiterpenes * snow + Sesquiterpenes * water + Sesquiterpenes * year + 
     Benzenoids * snow + Benzenoids * water + Benzenoids * year +
     Aliphatics * snow + Aliphatics * water + Aliphatics * year,
       data = trait.class.wide %>% rename(fitness=fitnesstrait) %>% 
       mutate(relfitness = fitness/mean(fitness, na.rm=T)))
}

fitnesstraits.sd <- map_dbl(fitnesstraits, ~sd(mnps.plantyr.traits[[.x]], na.rm=T)) %>% set_names(fitnesstraits)
class.sd <- map_dbl(names(class_pal), ~sd(trait.class.wide[[.x]], na.rm=T)) %>% set_names(names(class_pal)) 
mnps.plantyr.traits %>% pivot_longer(all_of(traits)) %>% group_by(name) %>% 
  summarize(mean=mean(value, na.rm=T), sd=sd(value, na.rm=T),
            n_plants=sum(!is.na(value))) %>% 
    mutate(name = traitnames.units[name]) %>% 
  write_csv("data/trait_means.csv", na="")
bind_rows(volsmeta.avg, vols.good.top) %>% pivot_longer(all_of(ipogood.top)) %>% group_by(name) %>% 
  summarize(mean=mean(value, na.rm=T), sd=sd(value, na.rm=T),
            n_plants=sum(!is.na(value))) %>%  
  write_csv("data/volatiles/volatiles_means.csv", na="")

get_emt_class <- function(mod) {
  map_dfr(set_names(names(class_pal)), ~emtrends(mod, c("year","water","snow"), var=.x) %>% 
            summary %>% as_tibble %>% rename("trend" = ends_with("trend")), .id="class")
}

mod.selection <- tibble(fitnesstrait=fitnesstraits) %>% 
  mutate(model = map(fitnesstrait, selection_effects),
         emt = map(model, get_emt_class),
         test = map(model, ~ tidy(car::Anova(.x, type=3))))

mod.selection %>% unnest(test) %>% select(fitnesstrait, term, p.value) %>% 
  pivot_wider(names_from=fitnesstrait, values_from=p.value) %>%
  filter(term!="Residuals") %>% 
  kable(caption="Selection of emissions in each treatment", digits=4)

mod.selection %>% filter(fitnesstrait != "eggs_per_flower") %>% 
  select(fitnesstrait, emt) %>% unnest(emt) %>% 
  mutate(classsd = class.sd[class],
         trend.std = trend * classsd,
         uSE = (trend + SE) * classsd,
         lSE = (trend - SE) * classsd,
         water_snow = factor(paste(water, snow), levels=names(water_snow_pal))) %>% 
  ggplot(aes(x=class, color=water_snow, y=trend.std, ymin=lSE, ymax=uSE)) +
  facet_grid(cols=vars(fitnesstrait), rows=vars(year), labeller=as_labeller(c(set_names(2018:2020), fitnessnames))) +
  geom_hline(yintercept=0)+
  geom_pointrange(position=position_dodge(width=0.5), size=0.2) +
  scale_color_manual(values=water_snow_pal) +
  labs(x="Compound class", y = "Standardized selection gradient", color="Precipitation,\nsnowmelt") +
  theme_bw() + theme(axis.text.x = element_text(angle=90, hjust=1), legend.position="top")
```

## Selection gradients

```{r calc_gradients}
#Following method of [Chong et al. 2018](https://doi.org/10.1002/evl3.63) for reconstructing selection estimates from estimates of selection on PCs: Beta = E * A.

prcomp_std <- function(data) {
  data %>% mutate(across(everything(), ~ (.x-mean(.x, na.rm=T))/sd(.x, na.rm=T))) %>% #center and scale
  prcomp(center = FALSE, scale. = FALSE)
}

calculate_beta <- function(E, slopes, resid.df) {
  data.frame(B = E %*% slopes$A,
             SE = sqrt(E ^ 2 %*% slopes$SE ^ 2)) %>% 
    mutate(p.value = 2*pt(abs(B)/SE, df=resid.df, lower.tail=F)) %>% #calculate t statistic from difference in means and SE, then look up one-tailed p in the upper tail, double to get a 2-tailed test
    rownames_to_column("trait")
}

selection_pca <- function(year=names(year_pal), water=names(water_pal), snow=names(snow_pal), 
                          fitnesstrait, n.PCs, PCA, mdata) {
  if(ncol(PCA$x) < n.PCs) {
    warning(paste("Only ", ncol(PCA$x), "PCs"))
    n.PCs <- ncol(PCA$x)
  }
  std.traits <- mdata %>% bind_cols(as.data.frame(PCA$x)) %>% 
    filter(year %in% .env$year, water %in% .env$water, snow %in% .env$snow) %>% 
    select(-any_of(fitnesstrait)) %>% 
    left_join(mnps.plantyr.traits %>% select(year, plotid, plantid, all_of(fitnesstrait)), by=c("year", "plotid", "plantid")) %>% 
    select(all_of(fitnesstrait), any_of(names(treatments)), starts_with("PC")) %>% 
    rename("fitness" = all_of(fitnesstrait)) %>% 
    #group_by(year) %>% 
    mutate(relfitness = fitness/mean(fitness, na.rm=T), .keep="unused") %>% #relative fitness across years
    drop_na(relfitness) %>% 
    select(where(~sum(is.na(.))==0))
  
  # Regress relative fitness on PCs
  if(length(year)+length(water)+length(snow)==8) {
      selection.on.PCs <- lm(formula(paste("relfitness ~", paste0("PC",1:n.PCs, collapse=" + "), 
                                           " + precip_est_mm + sun_date")), data = std.traits) 
      #account for direct environmental effects on fitness 
  } else { 
      selection.on.PCs <- lm(formula(paste("relfitness ~", paste0("PC",1:n.PCs, collapse=" + "))), data = std.traits)
  }
  
  # Convert selection gradients on PCs to gradients on traits
  calculate_beta(E = PCA$rotation[,1:n.PCs],
                 slopes = tidy(selection.on.PCs) %>% filter(str_sub(term,1,2) == "PC") %>%
                   rename(A=estimate, SE=std.error), 
                 resid.df = aov(selection.on.PCs)$df.residual)
}

vols.PCA <- prcomp_std(vols.good.top)

floraltraits.pca <- floraltraits[-8] #drop total flowers
mnps.PCA.nosepal <- mnps.plantyr %>% select(all_of(floraltraits.pca[-4])) %>% drop_na() %>% prcomp_std()
meta.pca.nosepal <- mnps.plantyr %>% filter(mnps.plantyr %>% select(all_of(floraltraits.pca[-4])) %>% vctrs::vec_detect_complete())
```

```{r gradients_allyrs, fig.width=9, fig.height=5}
calculate_all_gradients_nosepal <- function(trts) { 
  mutate(trts, PCA=list(vols.PCA), mdata = list(volsmeta.avg), n.PCs = 5) %>% 
  bind_rows(mutate(., 
                   n.PCs = 4,
                   PCA =   list(mnps.PCA.nosepal),
                   mdata = list(meta.pca.nosepal))) %>% 
  mutate(betas = pmap(., selection_pca)) %>% select(-PCA, -mdata) %>% unnest(betas)
}

all.beta.allyrs <- expand_grid(fitnesstrait=fitnesstraits) %>% 
  calculate_all_gradients_nosepal() #run the PC model with continuous environmental variables
```

## Elastic net

```{r elasticnet_morph}
trait_en_reg <- function(fitnesstrait, regtraits, en_alpha, data,
                         year=names(year_pal), water=names(water_pal), snow=names(snow_pal)) {
  std.traits <- data %>% 
    filter(year %in% .env$year, water %in% .env$water, snow %in% .env$snow) %>% 
    mutate(across(all_of(regtraits), ~ (.x-mean(.x))/sd(.x))) %>%  #center and scale traits
    select(-any_of(fitnesstrait)) %>% 
    left_join(mnps.plantyr.traits %>% select(year, plotid, plantid, all_of(fitnesstrait)), by=c("year", "plotid", "plantid")) %>% 
    select(all_of(fitnesstrait), any_of(names(treatments)), all_of(regtraits)) %>% 
    rename("fitness" = all_of(fitnesstrait)) %>% 
    mutate(relfitness = fitness/mean(fitness, na.rm=T), .keep="unused") %>% #relative fitness across years
    drop_na(relfitness) %>% 
    select(where(~sum(is.na(.))==0))
  
  Y = std.traits %>% select(relfitness) %>% as.matrix()
  if(length(year)+length(water)+length(snow)==8) {
    X = std.traits %>% select(all_of(c(regtraits, "precip_est_mm", "sun_date"))) %>% as.matrix()
  } else {
    X = std.traits %>% select(all_of(regtraits)) %>% as.matrix()
  }
  crossvalidate <- cv.glmnet(x=X, y=Y, alpha = en_alpha)
  best_lambda <- crossvalidate$lambda.min
  glmnet(x=X, y=Y, alpha = en_alpha, lambda = best_lambda) %>% tidy() %>% rename(B=estimate, trait=term)
}

enregs <- expand_grid(fitnesstrait=fitnesstraits, en_alpha = seq(0, 1, by=0.25),
                               regtraits = list(c(floraltraits.pca[-4])), data=list(meta.pca.nosepal)) %>% 
  mutate(betas = pmap(., trait_en_reg)) %>% select(-regtraits, -data) %>% 
  unnest(betas) %>% filter(!trait %in% c("(Intercept)", "precip_est_mm", "sun_date")) 
```

```{r elasticnet_vols}
enregs.vol <- expand_grid(fitnesstrait=fitnesstraits, en_alpha = seq(0, 1, by=0.25),
                               regtraits = list(c(ipogood.top)), data = list(bind_cols(volsmeta.avg, vols.good.top))) %>% 
  mutate(betas = pmap(., trait_en_reg)) %>% select(-regtraits, -data) %>% 
  unnest(betas) %>% filter(!trait %in% c("(Intercept)", "precip_est_mm", "sun_date")) %>% 
  mutate(trait = trait %>% greekify() %>% recode(!!!traitnames) %>% 
           factor(levels = rev(c(traitnames, greekify(ipogood.top))))) 
```

```{r elasticnet_treatments, fig.height=5, fig.width=7.5}
calculate_all_gradients_nosepal_en <- function(trts) { 
  mutate(trts,  data = list(bind_cols(volsmeta.avg, vols.good.top)), regtraits = list(ipogood.top)) %>% 
  bind_rows(mutate(trts, data = list(meta.pca.nosepal), regtraits = list(c(floraltraits.pca[-4])))) %>% 
  mutate(en_alpha=0.5) %>% 
  mutate(betas = pmap(., trait_en_reg)) %>% select(-regtraits, -data) %>% unnest(betas)
}

en.beta.water <- expand_grid(fitnesstrait=fitnesstraits, year=names(year_pal), water=names(water_pal)) %>%
  calculate_all_gradients_nosepal_en()
en.beta.snow <- expand_grid(fitnesstrait=fitnesstraits, year=names(year_pal), snow=names(snow_pal)) %>%
  calculate_all_gradients_nosepal_en()

en.beta.water.allyrs <- expand_grid(fitnesstrait=fitnesstraits, water=names(water_pal)) %>% mutate(year=list(names(year_pal))) %>%
  calculate_all_gradients_nosepal_en()
en.beta.snow.allyrs <- expand_grid(fitnesstrait=fitnesstraits, snow=names(snow_pal)) %>% mutate(year=list(names(year_pal))) %>%
  calculate_all_gradients_nosepal_en()
```


## Selection vs. absolute snowmelt, precip

```{r selection_abs_model}
selection_mod_abs <- function(trait, response, use_trait=TRUE, use_env=TRUE) {
  if(trait %in% ipogood) data <- vols.traits 
  else data <- mnps.plantyr %>% mutate(prop_uninfested = 1 - prop_infested)
  vm.abs <- data %>% 
    select(all_of(c(response, trait)), sun_date, precip_est_mm, year) %>% 
    rename("trait" = all_of(trait), "response" = all_of(response)) %>% 
    filter(!is.infinite(response), !is.nan(response), !is.na(response)) %>% 
    mutate(trait = (trait-mean(trait, na.rm=T))/sd(trait, na.rm=T)) %>% #center and scale
    #group_by(year) %>% 
    mutate(response = response/mean(response, na.rm=T)) #%>% ungroup() #relative fitness across years
  options(contrasts = c("contr.sum", "contr.poly"))
  if(use_trait) {
    if(use_env) lm(response ~ trait * sun_date + trait * precip_est_mm, data = vm.abs)
    else        lm(response ~ trait, data = vm.abs)
  } else        lm(response ~ sun_date + precip_est_mm, data = vm.abs)
}

mod.selection.abs <- expand.grid(list(response = fitnesstraits, trait = c(ipogood, floraltraits))) %>%
  mutate(model = pmap(., selection_mod_abs),
         coefs = map(model, tidy))

mod.selection.tests <- mod.selection.abs %>% 
  left_join(ipochemsf, by=c("trait"="name")) %>% 
  mutate(freq.floral = replace_na(freq.floral, 1)) %>% #morph/nectar traits
  filter(freq.floral>0.1) %>% 
    mutate(tests = map(model, ~ tidy(car::Anova(.x, type=3)))) %>% 
  select(-model) %>% unnest(coefs) %>% 
  filter(str_detect(term, "trait"))

#Drop environmental interactions if both not significant
mod.selection.abs.drop <- expand.grid(list(response = fitnesstraits, trait = c(ipogood, floraltraits))) %>%
  left_join(mod.selection.tests %>% select(response, trait, term, p.value) %>% 
              mutate(response=as.character(response)) %>% 
              filter(term != "trait") %>% pivot_wider(names_from = term, values_from = p.value) %>% 
              mutate(use_env = `trait:sun_date` < 0.05 | `trait:precip_est_mm` < 0.05, .keep="unused")) %>% 
  mutate(model = pmap(., selection_mod_abs),
         coefs = map(model, tidy))

mod.selection.tests.drop <- mod.selection.abs.drop %>% 
  left_join(ipochemsf, by=c("trait"="name")) %>% 
  mutate(freq.floral = replace_na(freq.floral, 1)) %>% #morph/nectar traits
  filter(freq.floral>0.1) %>% 
    mutate(tests = map(model, ~ tidy(car::Anova(.x, type=3)))) %>% 
  select(-model) %>% unnest(coefs) %>% 
  filter(str_detect(term, "trait"))
```


```{r selection_abs, fig.width=11, fig.height=7}
term_labels <- c(trait="Selection on trait (\U03B2)",
                 `trait:precip_est_mm`="Selection x summer precipitation (\U03B2 / mm)",
                 `trait:sun_date`="Selection x snowmelt date (\U03B2 / day)",
                 PCA="Direct selection on trait (\U03B2)",
                 en = "Elastic net selection (\U03B2)")

mod.selection.tests.drop %>% 
  complete(response, trait, term) %>% # make dodging consistent 
  #add similar PCA model (no interactions between trait and environment though)
  bind_rows(all.beta.allyrs %>% rename(estimate=B, std.error=SE, response=fitnesstrait) %>% mutate(term = "PCA")) %>% 
  filter(trait %in% c(ipogood.top, floraltraits)) %>% 
  mutate(trait = trait %>% greekify() %>% recode(!!!traitnames) %>% 
           factor(levels = rev(c(traitnames, greekify(ipogood.top))))) %>% 
  ggplot(aes(x=estimate,y=trait, color=response, shape=p.value < 0.05, xmin=estimate-std.error, xmax=estimate+std.error)) + 
  facet_grid(cols=vars(term), labeller = as_labeller(term_labels), scales="free_x") + 
  geom_vline(xintercept=0)+ 
  geom_pointrange(position = position_dodge2(width=0.6, reverse=T, preserve = "single"))+
  scale_color_brewer("Fitness measure", palette = "Dark2", labels = fitnessnames)+
  scale_shape_manual("P < 0.05", values=c(1,19), guide="none")+ 
  scale_x_continuous(expand = expansion(c(0,0)))+
  theme_bw() + theme(axis.title=element_blank(), legend.position = "top",
                     panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())
```

# Adaptive plasticity

```{r indiv_selection_treatments}
selection_mod_treatments <- function(response, trait, 
                                     year=names(year_pal), water=names(water_pal), snow=names(snow_pal)) {
  if(trait %in% ipogood) data <- vols.traits 
  else data <- mnps.plantyr %>% mutate(prop_uninfested = 1 - prop_infested)
  std.traits <- data[, c(response, trait, "year","water","snow")] %>% 
    set_names(c("response","trait","year","water","snow")) %>% 
    filter(year %in% .env$year, water %in% .env$water, snow %in% .env$snow) %>% 
    filter(!is.infinite(response), !is.nan(response), !is.na(response), !is.na(trait)) %>% 
    mutate(trait = (trait-mean(trait))/sd(trait), response = response/mean(response)) 
  options(contrasts = c("contr.sum", "contr.poly"))
  if(nrow(std.traits) == 0) return(NA)
  else return(lm(response ~ trait, data = std.traits))
}
calculate_all_indiv <- function(trts) { 
  trts %>% mutate(model = pmap(., selection_mod_treatments)) %>%
    filter(!is.na(model)) %>% #some combinations with zero rows
    mutate(n = map_int(model, ~nrow(model.frame(.x))),
           coefs = map(model, tidy)) %>% 
    select(-model) %>% unnest(coefs) %>% filter(term=="trait") %>% 
    rename(fitnesstrait=response, B=estimate, SE=std.error)
}

indiv.beta.water <- expand_grid(response=fitnesstraits,  trait = c(ipogood, floraltraits),
                                year=names(year_pal), water=names(water_pal)) %>% calculate_all_indiv()
indiv.beta.snow <- expand_grid(response=fitnesstraits,  trait = c(ipogood, floraltraits),
                                year=names(year_pal), snow=names(snow_pal)) %>% calculate_all_indiv()
```

```{r adaptive, fig.height=5, fig.width=7}
plasticity.data <- 
  bind_rows(volsmeta.good.top %>% mutate(index=as.integer(index)), 
            mnps.plantyr %>% pivot_longer(all_of(floraltraits.pca)) %>% drop_na(value)) %>% 
  group_by(name, year) %>% nest() %>% 
  mutate(model = map(data, ~lm(value ~ snow + water, data=.)), # + (1|plotid) + (1|snow:plot)
         emm = map(model, ~ summary(emmeans(.x, ~ snow + water))%>% filter(water=="Control", snow=="Normal")), 
         test =  map(model, ~tidy(multcomp:::summary.glht(multcomp::glht(.x, 
                                                                         linfct = multcomp::mcp(water = c("Addition - Control = 0",
                                                                                                          "Reduction - Control = 0",
                                                                                                          "Reduction - Addition = 0"),
                                                                                                snow = c("Early - Normal = 0"))),
                                                          multcomp::adjusted(type="none"))))) %>%#adjust later 
  select(year, name, emm, test) %>% unnest(c(emm, test)) %>% 
  mutate(difference=estimate, #difference in glht means
         estimate=estimate/emmean, std.error=std.error/emmean, #divide estimated difference in means by emm in Normal Control subplots
         contrast = str_replace(str_remove_all(contrast, "[a-z\\s]"),"-","_"),
         newenv = recode(contrast, "A_C"="Addition", "R_C"="Reduction","E_N"="Early"),
         multiplecomps = contrast %in% c("A_C","R_C")) %>% #these two reuse the same control data
  group_by(multiplecomps, name, year) %>% #two comparisons to Control made within each volatile and year 
  mutate(plasticity.p.value = ifelse(multiplecomps, p.adjust(p.value, method="fdr"), p.value)) %>% 
  ungroup %>% select(year, name, newenv, plasticity.p.value, difference, estimate, std.error) 

adaptive.plasticity.data <- plasticity.data %>% 
  left_join(bind_rows(rename(indiv.beta.water, newenv=water), #was all.beta.water for the PCA analysis
                      rename(indiv.beta.snow, newenv=snow)) %>% 
              filter(fitnesstrait!="eggs_per_flower") %>% 
              rename(name=trait)) %>% 
  left_join(bind_rows(rename(en.beta.water, newenv=water), 
                      rename(en.beta.snow, newenv=snow)) %>%
              rename(B.en = B, name=trait)) %>% 
  left_join(ipochemsf %>% select(-starts_with("area"))) %>% 
  mutate(class = fct_na_value_to_level(class, "Other trait"))

envnames <- c(Addition="additional precipitation",
             Reduction="reduced precipitation",
             Early="early snowmelt")

adaptive.plasticity.data %>% filter(newenv!="R_A") %>% 
  mutate(fitnesstrait=factor(fitnesstrait, levels = sort(fitnesstraits)),
         plasticity.signif = plasticity.p.value < 0.05, selection.signif= p.value < 0.05,
         en.indiv.same.direction = replace_na(if_else(sign(B) == sign(B.en),"same", "opposite"), "en_zeroed"),
         plasticity.selection.signif = paste0(plasticity.signif,selection.signif, 
                                              if_else(selection.signif, en.indiv.same.direction,"")) %>% 
           fct_relevel("FALSEFALSE","TRUEFALSE"),
         log2FC = log2(estimate+1)) %>%  #log2(fold change) is symmetric for increases and decreases. Add 1 to get ratio.
  group_by(newenv) %>% nest() %>% 
  pwalk(function(newenv, data) { 
    print(ggplot(data, aes(x=log2FC, y=B, color=class)) + 
            labs(x=paste("Plasticity to ",envnames[newenv],"(log2 fold change)"), 
                 y=paste("Selection gradient \U03B2 under",envnames[newenv])) + 
            facet_grid(rows=vars(year), cols=vars(fitnesstrait), scales="fixed", 
                       labeller=as_labeller(c(fitnessnames, set_names(levels(treatments$year)))))+
            annotate("rect", xmin = 0, xmax = Inf, ymin = 0, ymax = -Inf, fill= "black", alpha=0.1) + 
            annotate("rect", xmin = 0, xmax = -Inf, ymin = Inf, ymax = 0, fill= "black", alpha=0.1) +
            geom_errorbar(aes(ymin=B-SE, ymax=B+SE), alpha=0.3) + 
            geom_errorbarh(aes(xmin=log2(estimate+1-std.error), xmax=log2(estimate+1+std.error)), alpha=0.3) +#TODO check math
            geom_point(aes(shape = plasticity.selection.signif), size=2.5) + 
            scale_shape_manual("P < 0.05", values=c(1,17,19,15), guide=guide_legend(override.aes = list(size=5)),
                               labels=c("neither", "plasticity", "selection","selection+EN"), drop=FALSE)+ 
            coord_cartesian(ylim = c(min(data$B)*1.05, max(data$B)*1.05), xlim = c(min(data$log2FC)*1.05, max(data$log2FC)*1.05))+
            scale_color_manual("Compound class", values=c(class_pal,"Other trait"="black"), 
                               guide=guide_legend(override.aes = list(size=5))) + 
            theme_bw() + theme(axis.text = element_text(size=7), panel.grid = element_blank()))})

adaptive.plasticity.data %>% filter(newenv != "R_A") %>% count(p.value < 0.05, plasticity.p.value < 0.05) %>% kable()

#Cohen's d
vols.sd <- map_dbl(ipogood, ~sd(vols.good[[.x]])) %>% set_names(ipogood)
plasticity.cohens.d <- plasticity.data %>% filter(newenv!="R_A") %>% 
  left_join(enframe(c(vols.sd, alltraits.sd), value="trait.sd")) %>% 
  mutate(cohens.d = difference / trait.sd) %>% 
  mutate(traitgroup = if_else(name %in% ipogood.top, "volatile", "other")) 

plasticity.cohens.d.summary <- plasticity.cohens.d %>% 
  group_by(newenv, traitgroup) %>% 
  summarize(absolute_cohen = mean(abs(cohens.d)),
            sd = sd(abs(cohens.d)),
            se = sd/sqrt(n()),
            n = n()) 
kable(plasticity.cohens.d.summary, caption = "Mean absolute Cohen's d for volatiles and other traits", digits=2)
plasticity.cohens.d.summary %>% select(-sd, -n) %>% 
  pivot_wider(names_from=traitgroup, values_from = c(absolute_cohen, se)) %>% 
  mutate(z = (absolute_cohen_other - absolute_cohen_volatile)/sqrt(se_other^2 + se_volatile^2),
         p.value = pnorm(abs(z), mean = 0, sd = 1, lower.tail = FALSE)*2, .keep = "unused") %>% #two-tailed z test 
  kable(caption = "Z test of difference in absolute Cohen's d for volatiles and other traits")
```

```{r adaptive_quadrants}
adaptive.plasticity.data %>%
  mutate(adaptive = sign(B) == sign(estimate)) %>% filter(newenv !="R_A") %>% 
  group_by(newenv, fitnesstrait) %>% summarize(adaptive=sum(adaptive), n=n(), .groups="drop") %>% 
  mutate(prop = adaptive/n,
         p.value=map2_dbl(adaptive, n, ~ prop.test(.x, .y)$p.value),
         fitnesstrait = fitnessnames[fitnesstrait]) %>% 
  kable(caption=paste("Proportion of", ncol(vols.good.top), "volatiles and", length(floraltraits.pca), 
                      "other traits whose plasticity is adaptive across 3 yr"), digits=2)
```


## Plasticity to snowmelt vs. precipitation

```{r snow_v_water, fig.height=6, fig.width=4.5}
compare.plasticity <- full_join(
  adaptive.plasticity.data %>% filter(newenv=="R_A") %>% 
    select(year, name, class, mean.floral, estimate, std.error) %>%
    rename(R_A=estimate, R_A.SE=std.error),
  adaptive.plasticity.data %>% filter(newenv=="Early") %>% 
    select(year, name, class, mean.floral, estimate, std.error) %>%
    rename(E_N=estimate, E_N.SE=std.error) %>% distinct()) #3 rows each, one for each fitness trait

ggplot(compare.plasticity, aes(x=log2(R_A+1), xmin=log2(R_A+1-R_A.SE), xmax=log2(R_A+1+R_A.SE), 
                               y=log2(E_N+1), ymin=log2(E_N+1-E_N.SE), ymax=log2(E_N+1+E_N.SE), color=class)) + 
  facet_wrap(vars(year), ncol=1)+
  geom_vline(xintercept=0) + geom_hline(yintercept=0) +
  annotate("rect", xmin = 0, xmax = Inf, ymin = 0, ymax = -Inf, fill= "black", alpha=0.1) + 
  annotate("rect", xmin = 0, xmax = -Inf, ymin = Inf, ymax = 0, fill= "black", alpha=0.1) +
  geom_errorbar() + geom_errorbarh() + geom_point(aes(size=replace_na(mean.floral, 10))) + 
  scale_size_area("Mean emissions\n(ng/hr/flower)", max_size=3) +
  scale_color_manual("Compound class", values=c(class_pal, "Other trait"="black"), 
                     guide=guide_legend(override.aes = list(size=5))) + theme_bw() +
  labs(x="Plasticity to reduced vs. additional precipitation\n(log2 fold change)", #NaNs from negative xmins
       y="Plasticity to early snowmelt (log2 fold change)")

compare.plasticity %>% group_by(year) %>%  
  summarize(snow_water_cor = cor(log2(R_A+1), log2(E_N+1), use="complete"),
            snow_water_p = cor.test(log2(R_A+1), log2(E_N+1), use="complete")$p.value, n=n()) %>% 
  kable(caption="correlation between plasticity to water and snowmelt in each year")

```


```{r adaptive_diff, fig.height=14, fig.width=10, eval=F}
### Change in selection gradient vs. plasticity
diffbeta <- vols.beta.water %>% rename(name=trait) %>% 
  pivot_wider(names_from=water, values_from=c(B,SE)) %>% 
  mutate(B_A_R = B_Addition - B_Reduction)

#TODO get this working again with compare.plasticity
plasticity.water %>% left_join(diffbeta) %>% 
  ggplot(aes(x=A_R, y=B_A_R, color=class)) + facet_wrap(vars(year), ncol=1)+
  geom_vline(xintercept=0)+geom_hline(yintercept=0) + geom_text(aes(label=name)) + 
  scale_x_continuous(labels=scales::percent, limits=c(NA,NA)) + theme_bw()  + theme(legend.position = "top") +
  labs(x="Emissions change in precipitation addition vs. reduction", y="Change in selection gradient (beta) in precipitation addition vs. reduction")
```

# Combined analysis of volatiles and other floral traits

## Overlap of datasets

```{r overlap}
samples_per_plant <- volsmeta %>% count(year, plantid, name="scent") %>% as_tibble() %>% 
  full_join(mnps.plantyr.traits %>% drop_na(corolla_length) %>% count(year,plantid, name="morph")) %>% 
  full_join(mnps.plantyr.traits %>% drop_na(nectar_24_h_ul) %>% count(year,plantid, name="nectar")) %>% 
  full_join(mnps.plantyr.traits %>% drop_na(height_cm) %>% count(year,plantid, name="height")) %>% 
  full_join(mnps.plantyr.traits %>% drop_na(seeds_est) %>% count(year,plantid, name="seeds")) %>% 
  drop_na(seeds)

traits.m <- c("corolla_length", "corolla_width", "style_length") #exclude sepal_width (no 2018), anther min/max
traits.n <- c("nectar_24_h_ul", "nectar_conc")
overlaps <- c("v","vm","vn","vmn","mn","n","m")
# drop plants with missing data for combinations of measurements
combos <- list(v=vols.good.top, 
               vm=vols.traits %>% select(all_of(c(ipogood.top, traits.m, "height_cm"))) %>% drop_na(), 
               vn=vols.traits %>% select(all_of(c(ipogood.top, traits.n, "height_cm"))) %>% drop_na(), 
               vmn=vols.traits %>% select(all_of(c(ipogood.top, traits.m, traits.n, "height_cm"))) %>% drop_na(),
               mn=mnps.plantyr.traits %>% select(all_of(c(traits.m, traits.n, "height_cm"))) %>% drop_na(),
               n=mnps.plantyr.traits %>% select(all_of(c(traits.n, "height_cm"))) %>% drop_na(),
               m=mnps.plantyr.traits %>% select(all_of(c(traits.m, "height_cm"))) %>% drop_na())

all.traits.combos <- full_join(bind_cols(volsmeta.avg, vols.good.avg), 
                               left_join(mnps.plantyr.traits, treatments)) %>% 
  mutate(v =   vctrs::vec_detect_complete(.[,colnames(combos$v)]),
         vm =  vctrs::vec_detect_complete(.[,colnames(combos$vm)]),
         vn =  vctrs::vec_detect_complete(.[,colnames(combos$vn)]),
         vmn = vctrs::vec_detect_complete(.[,colnames(combos$vmn)]),
         mn =  vctrs::vec_detect_complete(.[,colnames(combos$mn)]),
         n =  vctrs::vec_detect_complete(.[,colnames(combos$n)]),
         m =  vctrs::vec_detect_complete(.[,colnames(combos$m)]))

all.traits.combos %>% group_by(year, water, snow) %>% 
  summarize(across(all_of(names(combos)), sum)) %>% 
  select(Year=year, Precipitation=water, Snowmelt=snow, Volatiles=v, Morphology=m, Nectar=n, `All Traits`=vmn) %>% 
  kable(caption="plants with each set of measurements")

meta.pca <- #treatments used in selection_pca function 
  volsmeta.avg[!is.na(rowSums(vols.traits[,colnames(combos$vmn)])),] #match rows in vmn
```

## Correlations

```{r vmnt_corr, fig.width=7, fig.height=5}
corr.annot <- tibble(name=colnames(combos$vmn)) %>% left_join(ipochemsf) %>% column_to_rownames("name") %>% 
  mutate(class = fct_na_value_to_level(class, "Other trait")) %>% select(Class=class)
corr.labs <- rownames(corr.annot) %>% greekify() %>% recode(!!!traitnames)
pheatmap(cor(combos$vmn, method="pearson"), 
         scale="none", clustering_method = "mcquitty",
         color=colorRampPalette(c("blue","white","red"))(200), breaks=seq(-1,1,by=0.01),
         annotation_col=corr.annot, annotation_row=corr.annot, 
         annotation_colors = list(Class=c(class_pal, "Other trait"="black")),
         labels_row=corr.labs, labels_col=corr.labs)

cor(combos$vmn, method="pearson")[ipogood.top,c(traits.m, traits.n, "height_cm")] %>% abs() %>%  as.vector() %>% mean(na.rm=T) %>%  round(2) %>% paste("Mean correlation among other floral traits: ", .) %>% print()
cor(combos$vmn, method="pearson")[ipogood.top, ipogood.top] %>% as.vector() %>% na_if(1) %>% abs() %>% mean(na.rm=T) %>% round(2) %>% paste("Mean correlation among volatiles: ", .) %>% print()
cor(combos$vmn, method="pearson")[c(traits.m, traits.n, "height_cm"), c(traits.m, traits.n, "height_cm")] %>% as.vector() %>% na_if(1) %>% abs() %>% mean(na.rm=T) %>% round(2) %>% paste("Mean correlation between volatiles and other floral traits: ", .) %>% print()
```

## Elastic net

```{r elasticnet, fig.width=10, fig.height=10}
trait_en_reg_combined <- function(fitnesstrait, regtraits, en_alpha) {
  std.traits <- bind_cols(combos$vmn, meta.pca) %>% 
    mutate(across(all_of(regtraits), ~ (.x-mean(.x))/sd(.x))) %>%  #center and scale traits
    select(-any_of(fitnesstrait)) %>% 
    left_join(mnps.plantyr.traits %>% select(year, plotid, plantid, all_of(fitnesstrait)), by=c("year", "plotid", "plantid")) %>% 
    select(all_of(fitnesstrait), any_of(names(treatments)), all_of(regtraits)) %>% 
    rename("fitness" = all_of(fitnesstrait)) %>% 
    mutate(relfitness = scale(fitness/mean(fitness, na.rm=T), center=T, scale=F), .keep="unused") %>% #relative fitness across years
    drop_na(relfitness) %>% 
    select(where(~sum(is.na(.))==0))
  
  Y = std.traits %>% select(relfitness) %>% as.matrix()
  X = std.traits %>% select(all_of(c(regtraits, "precip_est_mm", "sun_date"))) %>% as.matrix()
  
  crossvalidate <- cv.glmnet(x=X, y=Y, alpha = en_alpha)
  #plot(crossvalidate, main=paste(fitnesstrait, "alpha=", en_alpha))
  best_lambda <- crossvalidate$lambda.min
  
  en_model <- glmnet(x=X, y=Y, alpha = en_alpha, lambda = best_lambda)
  selection.on.traits <- en_model %>% tidy() %>% rename(B=estimate, trait=term)
  
  if(en_alpha == 1) {  #note fixedLassoInf only works for lasso, not ridge or elastic net!
    #the discussion in the selectiveInference help about how to scale lambda is confusing.
    #running with coef(s = lambda/nrow(Y)) and then fixedlassoInf(lambda = best_lamda) works
    #but if example lambda=0.8 is unscaled by n, then it should be
    #coef(s = lambda) and then fixedlassoInf(lambda = best_lamda* nrow(Y)), which gives a polyhedral constraint error
    #See https://stackoverflow.com/questions/68147109/r-how-to-translate-lasso-lambda-values-from-the-cv-glmnet-function-into-the-s

    coefs <- glmnet::coef.glmnet(en_model, x=X, y=Y, s=best_lambda, exact=TRUE, alpha=en_alpha)[-1]

    si <- selectiveInference::fixedLassoInf(x=X, y=Y, beta=coefs, lambda=best_lambda*nrow(Y)) #multiply by sample size to scale
    selection.on.traits <- selection.on.traits %>%
      left_join(data.frame(trait = names(si$vars), p.value = si$pv, 
                           partial = si$coef0, conf_low = si$ci[,1], conf_high = si$ci[,2]), by="trait")
  }
  selection.on.traits
}

enregs.combined <- expand_grid(fitnesstrait=fitnesstraits, en_alpha = seq(0, 1, by=0.25),
                               regtraits = list(c(ipogood.top, floraltraits.pca[-4]))) %>% 
  mutate(betas = pmap(., trait_en_reg_combined)) %>% select(-regtraits) %>% 
  unnest(betas) %>% filter(!trait %in% c("(Intercept)", "precip_est_mm", "sun_date"))
```

```{r elasticnet_abs, fig.width=11, fig.height=7}
mod.selection.tests.drop %>% 
  complete(response, trait, term) %>% # make dodging consistent 
  #add similar PCA model (no interactions between trait and environment though)
  bind_rows(bind_rows(separate = bind_rows(enregs, enregs.vol),
                      combined = enregs.combined, .id="analysis") %>% 
               mutate(term="en") %>% filter(en_alpha==0.5) %>% 
              rename(estimate=B, response=fitnesstrait)) %>% 
  filter(trait %in% c(ipogood.top, floraltraits)) %>% 
  mutate(trait = trait %>% greekify() %>% recode(!!!traitnames) %>% 
           factor(levels = rev(c(traitnames, greekify(ipogood.top))))) %>% 
  ggplot(aes(x=estimate,y=trait, color=response, 
             shape=if_else(!is.na(analysis), analysis, as.character(p.value < 0.05)), 
             xmin=estimate-std.error, xmax=estimate+std.error)) + 
  facet_grid(cols=vars(term), labeller = as_labeller(term_labels), scales="free_x") + 
  geom_vline(xintercept=0)+ 
  geom_path(aes(group=paste(response, trait)), 
            position=position_dodge2(width=0.6, reverse=T, preserve = "single"))+
  geom_pointrange(position = position_dodge2(width=0.6, reverse=T, preserve = "single"))+
  scale_color_brewer("Fitness measure", palette = "Dark2", labels = fitnessnames)+
  scale_shape_manual(values=c(`FALSE`=1,`TRUE`=19, separate=19, combined=18), 
                     na.value=19, guide="none")+ 
  scale_x_continuous(expand = expansion(mult=c(0.02,0.02)))+
  theme_bw() + theme(axis.title=element_blank(), legend.position = "top",
                     panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())
```

### By treatment

```{r elasticnet_separate_combined_treatments, fig.width=7.5, fig.height=6}
calculate_all_gradients_nosepal_en_combined <- function(trts) { 
  mutate(trts, year=list(names(year_pal)),
         data = list(bind_cols(combos$vmn, meta.pca)), 
         regtraits = list(c(ipogood.top, floraltraits.pca[-4])),
         en_alpha=0.5) %>% 
    mutate(betas = pmap(., trait_en_reg)) %>% select(-regtraits, -data) %>% unnest(betas)
}
en.beta.water.allyrs.combined <- expand_grid(fitnesstrait=fitnesstraits, water=names(water_pal)) %>%
  calculate_all_gradients_nosepal_en_combined()
en.beta.snow.allyrs.combined <- expand_grid(fitnesstrait=fitnesstraits, snow=names(snow_pal)) %>%
  calculate_all_gradients_nosepal_en_combined()

plot_beta_en_separate_combined <- function(betas, colorby) {
  betas %>% left_join(ipochemsf, by=c("trait"="name")) %>% 
    mutate(trait = recode(trait, !!!traitnames) %>%  greekify() %>% 
             fct_reorder(replace_na(mean.floral,1000))) %>%  #order graph by mean emissions
    filter(trait != "(Intercept)") %>% 
    rename(treatment=!!colorby) %>% 
    ggplot(aes(y=trait, x = B, shape=analysis, group=paste(fitnesstrait, trait, treatment), color=treatment)) + 
    facet_wrap(vars(fitnesstrait), nrow=1, labeller=as_labeller(fitnessnames))+
    geom_vline(xintercept=0)+ 
    geom_path(position=position_dodge2(width=0.6, reverse=T, preserve = "single"))+
    geom_point(position=position_dodge2(width=0.6, reverse=T, preserve = "single")) + 
    labs(x = "Selection gradient (\U03B2)", y = "Trait",
         color=list(year="Year", water="Precipitation", snow="Snowmelt", 
                    water_snow = "Precipitation and snowmelt")[[colorby]]) +
    scale_shape_manual(values=c(separate=19, combined=18), guide="none")+ 
    scale_color_manual(values=list(year=year_pal, water=water_pal, snow=snow_pal)[[colorby]]) +
    theme_bw() + theme(legend.position = "top")
}
bind_rows(separate=en.beta.water.allyrs, 
          combined=en.beta.water.allyrs.combined, .id="analysis") %>% 
  plot_beta_en_separate_combined(colorby="water")
bind_rows(separate=en.beta.snow.allyrs, 
          combined=en.beta.snow.allyrs.combined, .id="analysis") %>% 
  plot_beta_en_separate_combined(colorby="snow")
```
