--- 
title: "Maxfield Meadow Ipomopsis Experiment"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>

# Purpose
Analyze selection on and plastic responses of floral and vegetative traits of *Ipomopsis aggregata* to manipulations of snowmelt and summer precipitation. Determine whether plasticity is adaptive (improves survival and reproduction). Part of a [larger project](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1654655) to understand the potential for evolutionary rescue in alpine plants under climate change.

# Treatments

```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)
#library(googlesheets4)

library(lme4)
library(lmerTest)
library(MuMIn)
library(glmmTMB)
library(emmeans)
library(broom)
library(broom.mixed)
library(vegan)

library(RColorBrewer)
library(viridis)
library(gridExtra)
library(GGally)
library(ggvegan)
library(ggnewscale)

library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images/")

load("data/maxfield_data.rda")

options(contrasts = c("contr.sum", "contr.poly")) #needed for Type III SS

anova_table <- function(df, mod) { #for lmer models and lmerTest::anova
  df %>% select(c("trait", "term", "p.value")) %>% 
    mutate(p.value=p.value %>% map_chr(~ format(.x,scientific=-2, digits=2 , nsmall=2)) %>% 
             str_replace("e-0(\\d)"," x 10^-\\1^") %>% str_replace("e-(\\d+)"," x 10^-\\1^")) %>%
    pivot_wider(names_from=term, values_from=p.value) %>% 
    mutate(singular = map_lgl(mod, isSingular),
           R2m = map_dbl(mod, ~ MuMIn::r.squaredGLMM(.x)[[1,"R2m"]]))
}
Anova_table <- function(df) { #for glmmTMB models and car::Anova
  df %>% select(c("trait", "term", "p.value")) %>% 
    mutate(p.value=p.value %>% map_chr(~ format(.x,scientific=-2, digits=2 , nsmall=2)) %>% 
             str_replace("e-0(\\d)"," x 10^-\\1^") %>% str_replace("e-(\\d+)"," x 10^-\\1^")) %>%
    pivot_wider(names_from=trait, values_from=p.value)
}
re_table <- function(coefs) {
  coefs %>% filter(effect=="ran_pars") %>% select(trait, group, estimate) %>% pivot_wider(names_from=group, values_from=estimate)
  }
```

```{r map}
subplot_offset <- 1/6
size_meter <- 6
ggplot(treatments_map) + coord_fixed(clip="off")+
  geom_point(aes(x=plot_x, y=plot_y, fill=snow), size=7*size_meter, shape=22, stroke=1)+
  scale_fill_manual("Snowmelt", values=snow_pal, guide=guide_legend(override.aes = list(size = 2*size_meter)))+
  new_scale_fill()+
  geom_point(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, fill=water4), size=2*size_meter, shape=22, stroke=1) + 
    geom_text(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, label=subplot)) + 
  scale_fill_manual("Precipitation", values=water4_pal) +
  geom_text(aes(x=plot_x, y=plot_y, label=plot))+ theme_minimal() + 
  theme(legend.position="right", axis.title=element_blank(), axis.text = element_blank(), 
        axis.ticks = element_blank(), panel.background = element_blank(), panel.grid = element_blank())

ggplot(treatments_map %>% mutate(snow=fct_relevel(snow, "Normal"))) + coord_fixed(clip="off")+
  geom_point(aes(x=plot_x, y=plot_y, fill=snow), size=7*size_meter, shape=22, stroke=1)+
  scale_fill_manual("Snowmelt", values=c("grey","black"), guide=guide_legend(override.aes = list(size = 2*size_meter)))+
  new_scale_fill()+
  geom_point(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, fill=water4), size=2*size_meter, shape=22, stroke=1) + 
    geom_text(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, label=subplot), color="black") + 
  scale_fill_grey("Precipitation", start=1, end=0.2) +
  geom_text(aes(x=plot_x, y=plot_y, label=plot), color="white")+ theme_minimal() + 
  theme(legend.position="right", axis.title=element_blank(), axis.text = element_blank(), 
        axis.ticks = element_blank(), panel.background = element_blank(), panel.grid = element_blank())
```


# Snow melt vs. treatments

```{r hobo}
ggplot(hobo, aes(x=intensity, y=temp, color=hour(time))) +  geom_point(alpha=0.5, size=0.5) + geom_density_2d(color="black")+ 
  scale_x_log10(labels=function(x) format(x, scientific=F, big.mark=","), limits=c(10,1e6), breaks=10^(0:6)) +
  scale_color_gradientn(colours = rainbow(5))+
  geom_vline(xintercept = melt_threshold_sun) + 
  geom_hline(yintercept=melt_threshold_warm)+ geom_hline(yintercept=-melt_threshold_warm) +
  labs(x="Light intensity (lux)", y="Temperature (C)", color="Hour") +
  theme_dark() + theme(panel.grid.minor.x=element_blank())

ggplot(hobo %>% filter(hour(time)%in%5:17 & 
                  (year!="2019"&yday(time) %in% 110:130 | 
                   year=="2019"&yday(time) %in% 145:160)), 
       aes(x=intensity, y=temp, color=hour(time))) +   geom_point(alpha=0.7) +geom_density_2d(color="black")+
  scale_x_log10(labels=function(x) format(x, scientific=F, big.mark=","), limits=c(10,1e6), breaks=10^(0:6)) +
  scale_color_gradientn(colours = rainbow(5))+
  geom_vline(xintercept = melt_threshold_sun) + 
  geom_hline(yintercept=melt_threshold_warm)+ geom_hline(yintercept=-melt_threshold_warm) +
  labs(title="Temperature and light thresholds for snow melt", subtitle = "During the day during melt weeks",
       x="Light intensity (lux)", y="Temperature (C)", color="Hour") +
  theme_dark() + theme(panel.grid.minor.x=element_blank())

ggplot(hobo %>% filter(month(time)%in%4:6), aes(x=time, y=temp, color=plot)) + 
  facet_wrap(vars(year), scales="free_x", ncol=1) + 
  geom_hline(yintercept=melt_threshold_warm) + geom_hline(yintercept=-melt_threshold_warm) +geom_line() +
  geom_segment(data=meltdates, aes(x=warm_time, y=-10, xend=sun_time, yend=-15, color=plot), size=1) +
    geom_line(data=daily_CORBIL %>% filter(year!="2017", month(date)%in%4:6), aes(x=as.POSIXct(date), y=av_temp_2m_C), inherit.aes = F, color=brewer.pal(3, "Oranges")[3]) +
  labs(x="Date",y="Temperature (C)", color="Plot")
  
ggplot(hobo %>% filter(month(time)%in%4:6), aes(x=time, y=intensity, color=plot)) + 
  facet_wrap(vars(year), scales="free_x", ncol=1) + 
  geom_line() + geom_hline(yintercept=melt_threshold_sun) +
  geom_point(data=meltdates, aes(x=sun_time, y=-2e4, color=plot), size=1) +
  labs(x="Date",y="Light intensity (lux)", color="Plot")

meltdates %>% select(c("year","plot", "snow", ends_with("date"))) %>% pivot_longer(ends_with("date")) %>% arrange(name) %>% 
  ggplot(aes(y=name, x=value, color=plot, group=plot, linetype=snow)) + facet_wrap(vars(year), ncol=1) +
  geom_path(size=1) + #geom_text(aes(label=plot), hjust=1.5) +
  scale_linetype_manual(values=c(4,1)) + scale_x_continuous(minor_breaks=110:160)+
  labs(x="Day of year",y="Criteron", color="Plot", linetype="Snowmelt")
       
meltdates %>% ggplot(aes(x=melt_offset, fill=snow)) + geom_dotplot(method="histodot", binwidth=1, color="white") +
  facet_wrap(vars(year), ncol=1)+ scale_fill_manual(values=snow_pal) +
  scale_y_continuous(expand=expansion(0,0))+ scale_x_continuous(breaks=-10:4)+
  theme(panel.grid.minor = element_blank()) +
  labs(x="Offset from mean normal snow melt date (days)", y="Number of plots", fill="Snowmelt")
```

# Weather data

## Daily weather

```{r weather_daily}
ggplot(daily_all %>% filter(year(date) %in% 2018:2020), aes(y=EPA_NOAA_filled, x=yday(date))) + geom_col(aes(fill=ground_covered)) + 
  facet_wrap(vars(year(date)), ncol=1) + labs(x="Day of year", y="EPA/NOAA precipiation (mm)", fill="Snow cover")

daily_all %>% pivot_longer(unname(stations), names_to="station", values_to="precipitation_mm") %>% 
  filter(yday(date) >100, yday(date) < 240, year(date) %in% 2018:2020) %>%
  ggplot(aes(x=yday(date), y=precipitation_mm, color=station, shape=ground_covered)) +  
  geom_line() +  geom_point() + facet_wrap(vars(year), ncol=1) + theme(legend.position = "top", legend.direction="vertical") +
  labs(x="Day of year", y = "Precipitation (mm)", color="Station", shape="Snow cover")

precip_max <- 50
lower_continuous <- function (data, mapping, ...) {
    ggally_smooth(data = data, mapping = mapping, ..., method = "loess") + scale_shape_manual(values=c(1,2))+
    coord_cartesian(xlim=c(0,precip_max), ylim=c(0,precip_max)) + geom_abline(slope=1, intercept=0)+
    scale_x_continuous(breaks=seq(0,precip_max, by=precip_max/5))+scale_y_continuous(breaks=seq(0,precip_max, by=precip_max/5))
}
lower_combo <- function (data, mapping, ...) {
    ggally_facethist(data = data, mapping = mapping, binwidth=precip_max/50,...) +
    coord_cartesian(xlim=c(0,precip_max)) + scale_x_continuous(breaks=seq(0,precip_max, by=precip_max/5))
}
ggpairs(daily_all %>% select(all_of(unname(stations[-1])), ground_covered), mapping=aes(color=ground_covered,shape=ground_covered), lower=list(continuous=wrap(lower_continuous), combo=wrap(lower_combo)))

filter(daily_all) %>% ggplot(aes(y=EPA_RsrchMdw,x= NOAA_billy, color=ground_covered)) + geom_point(shape=1) + geom_smooth(method="lm", se=F) + geom_abline(intercept=0, slope=1) + labs(color="Snow cover")
```

## Monthly weather

```{r weather_monthly}
#recalculate Campbell and Wendlandt 2013 July average, 1989 - 2006 EPA station
(EPA_89_06_July_avg_mm <- monthly_all %>% filter(mo==7, yr >= 1989, yr <= 2006) %>% pull(EPA_NOAA_filled) %>% mean)

(water_addition_per_month_mm <- (14 / 4 / 2) * 31) #14 L applied to 4 m2 every 2 days = 1.75 mm daily precip * 31 days in July

monthly_all %>% pivot_longer(unname(stations), names_to="station", values_to="precipitation_mm") %>% 
ggplot(aes(x=mo*30.4-15.2, y=precipitation_mm, color=station)) + facet_wrap(vars(yr)) + #place points at month midpoint
  geom_rect(data=waterdates %>% pivot_wider(names_from=precip_treatments, values_from=c(day, date)) %>% mutate(yr=factor(year)), 
            aes(xmin=day_started, xmax=day_ended, ymin=0, ymax=water_addition_per_month_mm), inherit.aes=F,
            fill=water_pal["Addition"])+
  geom_line() +  geom_point()+
  geom_point(data=groundcover %>% filter(year>=2004) %>% mutate(yr=factor(year)), aes(x=first_0_cm_day,y=0), inherit.aes=F)+
  geom_point(data=groundcover %>% filter(year>=2005) %>% mutate(yr=factor(year-1)), aes(x=first_snow_day,y=0), inherit.aes=F)+
  labs(y="Monthly precipitation (mm)", x="Day of the year, with snowmelt and first snow dates", color="Station") +
  theme(legend.position="top") + coord_cartesian(ylim=c(0,200))

precip_max <- 250
ggpairs(monthly_all %>% mutate(season=factor(c("wntr","smmr"))[1+mo %in% 6:9]) %>% select(all_of(unname(stations[-1])), season), mapping=aes(color=season,shape=season), lower=list(continuous=wrap(lower_continuous), combo=wrap(lower_combo)))

filter(monthly_all) %>% ggplot(aes(y=EPA_RsrchMdw,x= NOAA_billy, color=ifelse(mo %in% 6:9,mo,"Oct - May"))) + geom_point() + geom_smooth(method="lm", se=F) + geom_abline(intercept=0, slope=1) + labs(color="Month")

monthly_all %>% filter(mo %in% 5:9) %>% 
  ggplot(aes(x=mo,  y=EPA_NOAA_filled)) + geom_smooth(span=0.5, color="black") +
  geom_line(aes(color=ifelse(yr>=2018, "study period (filled)", ifelse(yr>=2012,"filled from billy's NOAA data","original EPA data")), group=yr)) +
  geom_point(x=7, y=EPA_89_06_July_avg_mm)+ scale_y_continuous(limits=c(0,100))+
  geom_text(aes(label=ifelse(yr>=2018, yr, "")))  + 
  labs(y="Monthly precipitation (mm)", x="Month", title="EPA Research Meadow 1989-2020 summer precipitation", 
       subtitle="Black dot is mean July precipitation 1989-2006", color="Data filling") + theme(legend.position = "top")

monthly_all %>% filter(mo %in% 5:9) %>% ggplot(aes(x=yr, color=factor(mo), group=mo, shape=ifelse(yr>=2018, "study period (filled)", ifelse(yr>=2012,"filled from billy's NOAA data","original EPA data")), y=EPA_NOAA_filled)) +
  geom_vline(xintercept=2011.5) +geom_point() + geom_smooth(se=F, span=0.7) + 
  scale_shape_manual(values=c(1,2,19)) + scale_y_continuous(limits=c(0,100))+
  labs(x="Year",y="EPA Research Meadow monthly precipitation (mm)", color="Month", shape="Data filling") + 
  geom_segment(data=monthly_all %>% filter(mo %in% 5:9, yr <= 2007) %>% 
                 group_by(mo) %>% summarize(mean_precip=mean(EPA_NOAA_filled)), 
               aes(x=1989, xend=2006, y=mean_precip, yend=mean_precip, color=factor(mo)), inherit.aes=F, size=1) 
```

## Precipitation trends

```{r precip_trend}
daily_all %>% filter(year(date)>1989, year(date)<2021) %>%
  mutate(julian = yday(date), decade = factor(floor(year(date)/10)*10), decade = recode(decade, "2020"="2010")) %>%
  group_by(julian, decade) %>% 
  summarize(EPA_NOAA_filled = mean(EPA_NOAA_filled, na.rm=T), ground_covered=mean(as.integer(ground_covered), na.rm=T)-1) %>% 
  ggplot(aes(y=EPA_NOAA_filled, x=julian, color=decade)) + 
  geom_smooth(span=0.15, se=F)+  geom_point(shape=1, alpha=0.5) +   geom_line(aes(y=ground_covered*2), size=0.75) + 
  labs(x="Day of year", y="Mean EPA/NOAA precipiation (mm)", color="Decade", title="Snowmelt and monsoon timing by decade") +
    scale_color_brewer(palette = "Set1", direction = -1) + theme_minimal()

daily_all %>% filter(year(date)>1989, year(date)<2021, ground_covered=="smmr", EPA_NOAA_filled > 0) %>%   
  mutate(decade = factor(floor(year(date)/10)*10), decade = recode(decade, "2020"="2010")) %>% group_by(decade) %>% 
  ggplot(aes(x=EPA_NOAA_filled, color=decade)) + geom_density() +
    labs(x="EPA/NOAA precipiation (mm)", y="Proportion", color="Decade", title="Intensity spectrum of summer rainstorms > 0 mm") +
      scale_color_brewer(palette = "Set1", direction = -1) + 
  scale_x_continuous(breaks=scales::breaks_extended(10)) + theme_minimal()

summary_EPA_NOAA <- daily_all %>% filter(year(date)>1989, year(date)<2021, EPA_NOAA_filled > 2) %>% 
  mutate(drought_days = c(0,diff(date)), year=year(date)) %>% filter(ground_covered=="smmr") %>% 
  select(year, date, drought_days) %>% group_by(year) %>% summarize(max_drought_days_2mm = max(drought_days)) %>% 
  left_join(daily_all %>% filter(year(date)>1989, year(date)<2021, ground_covered=="smmr") %>% group_by(year=year(date)) %>% 
              summarize(rain_days_2mm = sum(EPA_NOAA_filled > 2, na.rm=T),
                        max_precip_mm = max(EPA_NOAA_filled, na.rm=T),
                        total_precip_mm = sum(EPA_NOAA_filled, na.rm=T)))

summary_names <- c(max_drought_days_2mm="Longest drought (< 2 mm precipitation)", 
                   rain_days_2mm = "Rain days (> 2 mm precipitation)",
                   max_precip_mm = "Maximum daily precipitation (mm)",
                   total_precip_mm = "Total precipitation (mm)")

ggplot(summary_EPA_NOAA %>% pivot_longer(!year), aes(y=value, x=year))+ 
  facet_wrap(vars(name), scales="free_y", labeller=as_labeller(summary_names)) +
  geom_smooth(method="lm", color="black", se=F) + geom_point() + labs(x="Year", y="") + 
  theme_bw() +theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            axis.title.y= element_blank(), strip.background = element_blank())
  

```

Long-term precipitation trends from Crested Butte station
```{r precip_trend_long}
ggplot(daily_NOAA, aes(x=date, y=NOAA_stations_name[id], color=prcp_mm)) + geom_tile() 
ggplot(daily_NOAA, aes(x=date, y=NOAA_stations_name[id], color=tmax_C)) + geom_tile()+ scale_color_viridis_c(option="inferno")

#weird long run of high precipitation in Sept 1927
daily_NOAA %>% filter(id=="USC00051959", year(date)>1909, year(date)!=1927) %>%
  mutate(julian = yday(date), decade = factor(floor(year(date)/10)*10), decade = recode(decade, "2020"="2010")) %>%
  group_by(julian, decade) %>% summarize(prcp_mm = mean(prcp_mm, na.rm=T)) %>% 
  ggplot(aes(y=prcp_mm, x=julian, color=decade)) + geom_smooth(span=0.15, se=F)+
  labs(x="Day of year", y="Mean Crested Butte precipiation (mm)", color="Decade", title="Snowmelt and monsoon timing by decade") +
  scale_color_viridis_d(option="inferno") + coord_cartesian(xlim=c(120,270)) + theme_dark()

daily_NOAA %>% filter(id=="USC00051959", year(date)>1909, year(date)!=1927,yday(date)>=120, yday(date)<= 270, prcp_mm > 0) %>%
  mutate(decade = factor(floor(year(date)/10)*10), decade = recode(decade, "2020"="2010")) %>% group_by(decade) %>% 
  ggplot(aes(x=prcp_mm, color=decade)) + geom_density() +
    labs(x="Crested Butte precipiation (mm)", y="Proportion", color="Decade", 
         title="Intensity spectrum of rainstorms > 0 mm \n(day 120-270, Crested Butte)") +
    scale_color_viridis_d(option="inferno") +  coord_cartesian(xlim=c(0,40)) +  theme_dark()

daily_NOAA %>% filter(id=="USC00051959", year(date)>1909, year(date)!=1927,prcp_mm > 2) %>%
  mutate(drought_days = c(0,diff(date)), year=year(date)) %>% filter(yday(date)>=120, yday(date)<= 270,) %>% 
  select(year, date, drought_days) %>% group_by(year) %>% summarize(max_drought_days = max(drought_days)) %>% 
  ggplot(aes(x= year, y=max_drought_days)) + geom_smooth(span=0.5) +geom_point() +
  labs(x="Year", y="Consecutive drought days with < 2 mm daily precipitation \n(day 120-270, Crested Butte)")

daily_NOAA %>% filter(id=="USC00051959", year(date)>1909, year(date)!=1927,yday(date)>=120, yday(date)<= 270, prcp_mm > 2) %>% 
  count(year=year(date)) %>% 
    ggplot(aes(x= year, y=n)) + geom_smooth(span=0.5) + geom_point() + 
  labs(x="Year", y="Days with > 2 mm daily precipitation (day 120-270, Crested Butte)")

daily_NOAA %>% filter(id=="USC00051959", year(date)>1909, year(date)!=1927, yday(date)>=120, yday(date)<= 270) %>% 
  group_by(year=year(date)) %>% summarize(total_mm=sum(prcp_mm, na.rm=T)) %>% 
    ggplot(aes(x= year, y=total_mm)) + geom_smooth(span=0.5) + geom_point() + 
  labs(x="Year", y="Total precipitation (mm, day 120-270, Crested Butte)")

daily_NOAA %>% filter(id=="USC00051959", year(date)>1909, yday(date)>=120, yday(date)<= 270) %>% 
  group_by(year=year(date)) %>% summarize(tmax_mean=mean(tmax_C, na.rm=T)) %>% 
    ggplot(aes(x= year, y=tmax_mean)) + geom_smooth(span=0.3) + geom_point() + 
  labs(x="Year", y="Average maximum temperature (C, day 120-270, Crested Butte)")

daily_NOAA %>% filter(id=="USC00051959", year(date)>1909, yday(date)>=120, yday(date)<= 270) %>% 
  group_by(year=year(date)) %>% summarize(tmin_mean=mean(tmin_C, na.rm=T)) %>% 
    ggplot(aes(x= year, y=tmin_mean)) + geom_smooth(span=0.3) + geom_point() + 
  labs(x="Year", y="Average minimum temperature (C, day 120-270, Crested Butte)")
```


## Snowmelt timing

```{r groundcover}
groundcover %>% mutate(first_snow_day=first_snow_day-365) %>% 
  pivot_longer(ends_with("day"), names_to="threshold", values_to="day") %>% drop_na(day) %>% select(-starts_with("first")) %>% 
  mutate(threshold = fct_reorder(factor(threshold),day, na.rm=T, .desc=T)) %>% 
  ggplot(aes(x=year, y=day, color=threshold)) +  
  annotate("rect", xmin=2017.5, xmax=2020.5, ymin=-Inf, ymax=Inf, fill="lightblue")+
  geom_line(aes(group=year), size=2)+ geom_smooth( method="lm", se=F) +geom_point() +
  scale_x_continuous(minor_breaks = 1976:2020, position = "top") + scale_y_continuous(n.breaks=10)+
  scale_color_grey(labels=function(x) str_to_sentence(str_replace_all(x, "_", " ")), start=0, end=0.85) +
  labs(x="Year", y="Day of year", color="Threshold") + theme_minimal()

cat("SD of first bare ground =", round(sd(groundcover$first_0_cm_day),1),"days over",range(groundcover$year))
summary(lm(first_0_cm_day ~ year, data=groundcover))

treatments %>% ggplot(aes(x=sun_date, shape=year, color=water, y=precip_est_mm))+geom_point(size=2)+
  geom_point(data=groundcover, aes(x=first_0_cm_day, color="Control", shape="1990-2017"), fill="white", size=2) +
  scale_color_manual(values=water_pal) + scale_shape_manual(values=c(21, 19, 17, 15))+ coord_cartesian(ylim=c(0,NA))+
  labs(y="Estimated summer precipiation with treatments (mm)", x="Snowmelt date", color="Precipitation", shape="Year",size="# Subplots") + 
  theme_minimal()
```


# Soil moisture

## Precipitation

```{r soil, fig.height=9, fig.width=9}
ggplot(sm, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
  geom_jitter(aes(shape=snow, color=water, y = VWC), height=0, width=0.3) + 
  geom_smooth(aes(linetype=snow, color=water, y = VWC), se=F, span=0.5) +
  geom_col(data=daily_all %>% filter(yr%in%2018:2020), aes(x=yday(date), y=EPA_NOAA_filled), 
           fill=brewer.pal(3, "Greens")[3]) +
  geom_line(data=daily_all %>% filter(yr%in%2018:2020), aes(x=yday(date), y=av_temp_2m_C), 
            color=brewer.pal(3, "Oranges")[3]) + 
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color=brewer.pal(3, "Purples")[3])+
  geom_line(data = snowcloth, aes(y=mean_snow_depth_cm/10, group=year))+
  geom_point(data = snowcloth, aes(y=mean_snow_depth_cm/10))+
  geom_line(data=groundcover %>% pivot_longer(ends_with("day"), names_to="threshold", values_to="day") %>%
              filter(threshold!="first_snow_day", year>2017) %>% 
              mutate(snow_depth_cm = as.integer(str_split_fixed(threshold,"_",n=4)[,2])),
            aes(x=day, y=snow_depth_cm/10), color="grey40")+
  geom_segment(data=waterdates %>% pivot_wider(names_from=precip_treatments, values_from=c(day,date)),
               aes(x=day_started, xend=day_ended, y=20, yend=20)) +
  geom_point(data=waterdates, aes(x=day,y=20)) +
  geom_point(data=ph, aes(y= 17*(open+buds)/(open+buds)), color="grey", shape=0)+
  geom_point(data=mt, aes(y= 18*corolla_length/corolla_length), color="purple", shape=0)+
  geom_point(data=mt, aes(y= 19*nectar_24_h_ul/nectar_24_h_ul), color="orange", shape=0)+
  scale_x_continuous(limits=c(85, 230), breaks=scales::breaks_extended(10), expand=expansion(0,0)) + scale_y_continuous(limits=c(-3, 20)) +
  labs(y="Soil VWC (%) / Daily precipitation (mm) / Mean temp. (C) / Snowpack (dm)", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_color_manual(values=water_pal) +
  scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(sm.subplot, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
      geom_col(data=daily_all %>% filter(yr%in%2018:2020), aes(x=yday(date), y=EPA_NOAA_filled), 
           fill=brewer.pal(3, "Greens")[2]) +
  geom_line(data=daily_precip_est, aes(y=VWC_Control_4*100), color=brewer.pal(3, "Greens")[2], size=1) +
  geom_point(aes(shape=snow, color=water, y = VWC)) + 
  geom_line(data=sm.subplot %>% group_by(water, snow, year, date) %>% summarize(VWC=mean(VWC, na.rm=T)), aes(linetype=snow, color=water, y = VWC)) +
    scale_color_manual(values=water_pal) +  scale_linetype_manual(values=c(2,1)) + 
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color="black", show.legend=F) +
  coord_cartesian(xlim=c(110, 230), ylim=c(0, round(max(sm.subplot$VWC, na.rm=T))), expand=F)+
    labs(x="Day of year", y="Mean soil moisture in subplot (%VWC) / Precipitation (mm)", color="Water", shape="Snowmelt", linetype="Snowmelt") + theme_minimal() + theme(legend.key.width = unit(2, "lines"))
```

## Treatments

```{r soil_treatments}
left_join(filter(sm.subplot20d, mo20d!="221 - 240"), meltdates) %>% 
  mutate(mo = paste0("Days ",mo20d)) %>% 
  ggplot(aes(x=sun_date, y=VWC, shape=year,color=water)) + 
  facet_wrap(vars(mo), nrow=1)+
  geom_point() + geom_smooth(se=F, method="lm") +
  scale_color_manual(values=water_pal) +
  labs(x="Snowmelt date", y="Mean soil moisture in subplot", color="Water", shape="Year")

sm.subplot %>% mutate(day=yday(date)) %>% group_by(year) %>% summarize(start=min(day), end=max(day)) %>% kable(caption = "Soil moisture collection dates")

# VWC split-plot analysis on subplot means (for the whole measurement period)
sm.mod <- lmer(VWC ~ year * snow * water + (1|snow:plot)+ (1|snow:plotid), data=sm.subplotyear)
anova(sm.mod) %>% kable(digits=3)
(sm.emm.snow <- summary(emmeans(sm.mod, ~ snow))) %>% kable(digits=3)
sm.emm.snow$emmean[1] - sm.emm.snow$emmean[2]
(sm.emm.water <- summary(emmeans(sm.mod, ~ water))) %>% kable(digits=3)
sm.emm.water$emmean[1] - sm.emm.water$emmean[2]
sm.emm.water$emmean[3] - sm.emm.water$emmean[2]

sm.mod.split.loc <- lmer(VWC ~ year * snow * water + location_y * water + (1|snow:plot) + (1|snow:plotid) + (1|date), data=mutate(sm, date=factor(date)))
anova(sm.mod.split.loc) %>% kable(digits=3)
emmeans(sm.mod.split.loc, ~ location_y * water, lmer.df = "satterthwaite") %>% plot

sm.mod.split <- lmer(VWC ~ year * snow * water + (1|snow:plot) + (1|snow:plotid) + (1|date), data=mutate(sm, date=factor(date)))
sjPlot::plot_model(sm.mod.split, type="re")

ggplot(sm.subplotyear, aes(x=water, color=snow, y=VWC)) +
  facet_wrap(vars(year)) + 
  geom_point(position=position_dodge(0.9)) + 
  #geom_pointrange(data=emmeans(sm.mod, ~ snow * water * year, lmer.df = "satterthwaite") %>% 
  #             summary %>% rename(VWC=emmean), shape=1, size=2, fatten=1.5,aes(ymin=VWC-SE, ymax=VWC+SE),
  #             position=position_dodge2(0.9)) + 
  geom_pointrange(data=emmeans(sm.mod.split, ~ snow * water * year, lmer.df = "satterthwaite") %>% 
               summary %>% rename(VWC=emmean), shape=3, aes(ymin=VWC-SE, ymax=VWC+SE),
             position=position_dodge2(0.9)) +
  scale_color_manual(values=snow_pal) + scale_x_discrete(guide=guide_axis(angle=90))

sm.mod.abs <- lmer(VWC ~ sun_date * water + (1|plot) + (1|plotid) + (1|date), data=mutate(sm, date=factor(date)))
anova(sm.mod.abs) %>% kable(digits=3)
sjPlot::plot_model(sm.mod.abs, type="int")
```

## Plant vs. subplot level
```{r sm_phys_compare}
ggplot(pt, aes(x=VWC.sm, y=VWC.plant, shape=year, group=year, color=as.integer(date_diff))) + geom_smooth(se=F) +geom_point() + geom_abline(intercept=0, slope=1) + 
  labs(x="Subplot VWC at closest date", y="VWC measured at plant for physiology", color="Date difference", shape="Year") +
  scale_color_viridis_c() + theme_minimal() + coord_fixed()

ggplot(pt, aes(x=VWC.sm.year, y=VWC.sm, shape=year, group=year, color=year)) + geom_smooth(se=F) +geom_point() + geom_abline(intercept=0, slope=1) + 
  labs(x="Mean subplot VWC across year", y="Subplot VWC at closest date", color="Year", shape="Year") +
  scale_color_manual(values=year_pal) + theme_minimal() + coord_fixed()

ggplot(pt, aes(x=VWC.sm.year, y=VWC.plant, shape=year, group=year, color=year)) + geom_smooth(se=F) +geom_point() + geom_abline(intercept=0, slope=1) + 
  labs(x="Mean subplot VWC across year", y="VWC measured at plant for physiology", color="Year", shape="Year") +
  scale_color_manual(values=year_pal) + theme_minimal() + coord_fixed()
```


# Timing of measurements

```{r timings}
plot_timings <- function(data) {
  timings_labels <- c(snowcloth="Snow cloths applied to early plots", 
                     meltdates="Mean snowmelt timing in early & unmanipulated plots", 
                     waterdates="Summer precipitation treatments applied",
                     sm="Soil moisture recorded",         ph="Inflorescence height recorded", 
                     mt="Floral morphology recorded",     nt="Nectar traits recorded", 
                     pt= "Physiology traits recorded",    lt="Vegetative traits recorded", 
                     sds="Flower collection period")
  
  data %>% mutate(variable=fct_reorder(variable, begin, na.rm=T, .desc=T),
                  drawline = ifelse(variable %in% c("lt","meltdates"), NA, TRUE)) %>% 
ggplot(aes(y=variable, color=year))+ 
  geom_linerange(aes(xmin=begin, xmax=end, linetype=drawline), 
                 position = position_dodge2(width=0.7, reverse = T), size=1.2, show.legend=FALSE) + 
  geom_text(data=data %>% drop_na(plots) %>% filter(year=="2019"),
            aes(x = end+7, label=paste(ifelse(nchar(plots)>1, "Plots","Plot"), str_replace(plots,","," & "))), 
            position = position_dodge2(width=0.8, reverse = T), size=4, hjust=0, show.legend=FALSE)+
  geom_point(aes(x=begin), position = position_dodge2(width=0.7, reverse = T), shape=15, size=2)+
  geom_point(aes(x=end), position = position_dodge2(width=0.7, reverse = T), shape=15, size=2) +
  scale_color_manual("Year", values=year_pal, guide=guide_legend(override.aes = list(size=5))) + 
  scale_y_discrete("", labels=timings_labels)+
  scale_x_continuous("Day of year", breaks=seq(80,240, by=20)) +
  theme_minimal() + theme(legend.position = "top")
}

plot_timings(filter(timings, !variable %in% c("sds","lt","pt")))
plot_timings(filter(timings, !variable %in% c("sds","mt","nt","ph")))

timings.long <- timings %>% select(-plots) %>% 
  pivot_longer(c(begin,end), names_to="time", values_to="day") 


timings.long %>% 
  left_join(timings.long %>% filter(variable=="meltdates", time=="end") %>% select(year, sun_date=day)) %>% 
  mutate(days_since_sun_date = day - sun_date) %>% 
  pivot_wider(id_cols = c(variable, year), names_from=time, names_prefix = "snowmelt_",
              values_from=days_since_sun_date, values_fn = list) %>% kable
```

## Sample sizes

```{r sample_sizes}
count_measurements <- function(df, tr) summarize(df, across(all_of(tr), ~ sum(!is.na(.x))))
bind_rows(n_measurements = count_measurements(bind_rows(mnps, lt, pt), alltraits),
          plant_year_round =   count_measurements(bind_rows(mnps.plantyr, lt.plantyr, pt.plantyr), alltraits), 
          plant_year = count_measurements(bind_rows(mnps, lt, pt) %>% group_by(year, plantid) %>% 
                                            summarize(across(all_of(alltraits), mean, na.rm=T)), alltraits) %>% 
            ungroup %>% summarize(across(all_of(alltraits), sum)),
          plant =       count_measurements(bind_rows(mnps, lt, pt) %>% group_by(plantid) %>% 
                                             summarize(across(all_of(alltraits), mean, na.rm=T)), alltraits),
          .id="sample_size") %>% column_to_rownames("sample_size") %>% 
  t %>% as.data.frame %>% rownames_to_column("trait") %>% as_tibble %>% 
  mutate(n_per_plant_year_round = n_measurements / plant_year_round,
         n_per_plant_year = n_measurements / plant_year,
         n_per_plant = n_measurements/ plant) %>% kable(digits=2)

#How many plants flowered twice?
with(cen.status, table(flowering_status_18 + flowering_status_19 + flowering_status_20))

ph %>% drop_na(height_cm) %>% group_by(year, plantid) %>% tally() %>% 
  ggplot(aes(x=n, fill=year)) + geom_histogram() + labs(title = "Phenology entries per plant", x="Entries",y="Count")+
  facet_wrap(vars(year), ncol=1) + scale_fill_manual(values=year_pal)
```


# Census

## Census status

The annual census at the beginning of the summer contains information on mortality, flowering status, and plant size (longest leaf and number of leaves for each rosette).

Visualize a plant's status in each census

```{r status}
ggplot(cen.size.long %>% filter(census!="2018r2"), aes(x=census, y=n_rosettes, group=plantid))+geom_path()+ geom_point() + scale_y_continuous(breaks=0:10)

ggplot(cen.status.long, aes(y=plantid, x=census, fill=status)) + geom_tile() +
  scale_fill_manual(values=brewer.pal(8, name="Set1")[c(1,3,5,2,7,6)])+
  theme(axis.text.y=element_blank(), axis.ticks = element_blank()) +
  labs(fill="Status", x="Census year",y="Plant ID")

ggplot(megatally.status.long, aes(y=plantid, x=dataset, fill=status)) + geom_tile() +
  scale_fill_manual(values=c(brewer.pal(5, name="Set1")[c(1,3,5,2)],brewer.pal(8, "Set2")[7:8]))+
  scale_x_discrete(position="top") +
  theme(axis.text.y=element_blank(), axis.ticks = element_blank(), axis.text.x = element_text(angle=90, size=14)) + 
  labs(fill="Status", x="Dataset and year",y="Plant ID")
```

## Status vs. treatment
```{r inventory}
cen.status.long %>% left_join(treatments) %>% drop_na(water, snow) %>% filter(status %in% status_ok, census !="2018r2") %>% 
  ggplot(aes(x=water, fill=status)) + facet_grid(snow ~ census, labeller=labeller(.cols=label_both)) + geom_bar(position="fill") +
  scale_fill_manual(values=status_pal) + labs(x="Precipitation", y="Proportion", fill="Status")
```

## Status transitions vs. treatment

What did plants of a given status in one year look like in the next?
```{r transitions}
with(cen.status, table(`2018`=paste(status_18), `2019`=paste(status_19)))

cen.status %>% mutate(year = "2018") %>% left_join(treatments) %>% filter(status_18 %in% status_ok[-3], status_19 %in% status_ok) %>% 
ggplot(aes(x=water, fill=status_19)) + facet_wrap(status_18 ~ snow, labeller=labeller(.cols=label_both)) +  
  labs(y="2019 Status", fill="2019 Status", x="Precipitation") +
  geom_bar(position="fill") + scale_fill_manual(values=status_pal) + theme_minimal()

with(cen.status, table(`2019`=paste(status_19), `2020`=paste(status_20)))

cen.status %>% mutate(year = "2019") %>% left_join(treatments) %>% filter(status_19 %in% status_ok[-3], status_20 %in% status_ok) %>% 
ggplot(aes(x=water, fill=status_20)) + facet_wrap(status_19 ~ snow, labeller=labeller(.cols=label_both)) +  
  labs(y="2020 Status", fill="2020 Status", x="Precipitation") +
  geom_bar(position="fill") + scale_fill_manual(values=status_pal) + theme_minimal()

#Model effect of treatments on survival and flowering of vegetative plants - both years combined
cen.status.veg1819 <- bind_rows(
  `2018`=cen.status %>% filter(status_18=="vegetative") %>% drop_na(alive_status_19, flowering_status_19) %>% mutate(year="2018") %>%
    left_join(treatments) %>% select(alive_status = alive_status_19, flowering_status = flowering_status_19, water, snow, plot, plotid),
  `2019`=cen.status %>% filter(status_19=="vegetative") %>% drop_na(alive_status_20, flowering_status_20) %>% mutate(year="2019") %>%
    left_join(treatments) %>% select(alive_status = alive_status_20, flowering_status = flowering_status_20, water, snow, plot, plotid), .id="year") %>% left_join(sm.subplotyear)
# could now also use cen.status.long %>% filter(status=="vegetative", census %in% c("2018","2019"))

library(glmmTMB)
options(contrasts = c("contr.sum", "contr.poly"))

(survival1819.mod <- glmmTMB(alive_status ~ year * water * snow + (1|snow:plot) + (1|snow:plotid), data=cen.status.veg1819, family=binomial))
car::Anova(survival1819.mod, type=3) %>% kable(digits=3)
emmeans(survival1819.mod, ~ year * water * snow) %>% as_tibble %>% 
  mutate(prob=plogis(emmean), uSE = plogis(emmean+SE), lSE = plogis(emmean-SE)) %>% 
  ggplot(aes(x=water, color=snow, y=prob))+ facet_wrap(vars(year))+
  geom_pointrange(aes(ymin=lSE, ymax=uSE), position=position_dodge2(0.9)) + 
  labs(x="Precipitation", y="Survival to next year", color="Snowmelt")+
  scale_y_continuous(limits=c(0,1)) + scale_color_manual(values=snow_pal) + theme_minimal()

(flowering1819.mod <- glmmTMB(flowering_status ~ year * water * snow + (1|snow:plot) + (1|snow:plotid), data=cen.status.veg1819, family=binomial))
car::Anova(flowering1819.mod, type=3) %>% kable(digits=3)
emmeans(flowering1819.mod, ~ year * water * snow) %>% as_tibble %>% 
  mutate(prob=plogis(emmean), uSE = plogis(emmean+SE), lSE = plogis(emmean-SE)) %>% 
  ggplot(aes(x=water, color=snow, y=prob))+ facet_wrap(vars(year))+
  geom_pointrange(aes(ymin=lSE, ymax=uSE), position=position_dodge2(0.9)) + 
  labs(x="Precipitation", y="Flowering the next year", color="Snowmelt")+
  scale_y_continuous(limits=c(0,1)) + scale_color_manual(values=snow_pal) + theme_minimal()
```

```{r transitions_sm}
library(glmmTMB)
options(contrasts = c("contr.sum", "contr.poly"))

(survival1819.mod.sm <- glmmTMB(alive_status ~ year * VWC + (1|plot) + (1|plotid), data=cen.status.veg1819, family=binomial))
car::Anova(survival1819.mod.sm, type=3) %>% kable(digits=3)
ggplot(cen.status.veg1819, aes(x=VWC, y=alive_status, color=year))+
  geom_point() + geom_smooth(method="glm", method.args = list(family=binomial))+
  labs(x="Soil moisture (VWC)", y="Survival to next year", color="Year")+
  scale_y_continuous(limits=c(0,1)) + scale_color_manual(values=year_pal) + theme_minimal()

(flowering1819.mod.sm <- glmmTMB(flowering_status ~ year * VWC + (1|plot) + (1|plotid), data=cen.status.veg1819, family=binomial))
car::Anova(flowering1819.mod.sm, type=3) %>% kable(digits=3)
ggplot(cen.status.veg1819, aes(x=VWC, y=flowering_status, color=year))+
  geom_point() + geom_smooth(method="glm", method.args = list(family=binomial))+
  labs(x="Soil moisture (VWC)", y="Flowering the next year", color="Year")+
  scale_y_continuous(limits=c(0,1)) + scale_color_manual(values=year_pal) + theme_minimal()
```


## Survival and flowering vs. vegetative plant size
```{r surv_size}
ggplot(filter(cen, status_19 %in% status_ok, status_18 == "vegetative"), aes(x=r1_longest_18, fill=status_19)) +
  facet_wrap(vars(status_19), ncol=1) + labs(x="2018 Rosette 1 Longest Leaf (mm)", fill="2019 status") +
  geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 85))
ggplot(filter(cen, status_20 %in% status_ok, status_19 == "vegetative"), aes(x=r1_longest_19, fill=status_20)) +
  facet_wrap(vars(status_20), ncol=1) + labs(x="2019 Rosette 1 Longest Leaf (mm)", fill="2020 status") +
   geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 85))

ggplot(filter(cen, status_19 %in% status_ok, status_18 == "vegetative"), aes(x=r1_leaves_18, fill=status_19)) +
  facet_wrap(vars(status_19), ncol=1) + labs(x="2018 Rosette 1 # Leaves", fill="2019 status") +
  geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 100))
ggplot(filter(cen, status_20 %in% status_ok, status_19 == "vegetative"), aes(x=r1_leaves_19, fill=status_20)) +
  facet_wrap(vars(status_20), ncol=1) + labs(x="2019 Rosette 1 # Leaves", fill="2020 status") +
  geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 100))
```

## Plant growth vs. treatments

Only includes plants that remained vegetative and did not die or flower.
```{r growth}
cen %>% drop_na(water_18, snow_18) %>% filter(status_18=="vegetative", status_19=="vegetative") %>% 
  ggplot(aes(x=r1_longest_18, y=r1_longest_19, color=water_18, shape=snow_18)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+coord_fixed(xlim=c(0,80), ylim=c(0,80)) +
  labs(x="2018 Rosette 1 Longest Leaf (mm)", y="2019 Rosette 1 Longest Leaf (mm)",  color="Precipitation", shape="Snowmelt")
cen %>% drop_na(water_19, snow_19) %>% filter(status_19=="vegetative", status_20=="vegetative") %>% 
  ggplot(aes(x=r1_longest_19, y=r1_longest_20, color=water_19, shape=snow_19)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+  coord_fixed(xlim=c(0,80), ylim=c(0,80)) +
    labs(x="2019 Rosette 1 Longest Leaf (mm)", y="2020 Rosette 1 Longest Leaf (mm)",  color="Precipitation", shape="Snowmelt")

cen %>% drop_na(water_18, snow_18) %>% filter(status_18=="vegetative", status_19=="vegetative") %>% 
  ggplot(aes(x=r1_leaves_18, y=r1_leaves_19, color=water_18, shape=snow_18)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+coord_fixed(xlim=c(0,100), ylim=c(0,100)) +
  labs(x="2018 Rosette 1 # Leaves", y="2019 Rosette 1 # Leaves",  color="Precipitation", shape="Snowmelt")
cen %>% drop_na(water_19, snow_19) %>% filter(status_19=="vegetative", status_20=="vegetative") %>% 
  ggplot(aes(x=r1_leaves_19, y=r1_leaves_20, color=water_19, shape=snow_19)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+  coord_fixed(xlim=c(0,100), ylim=c(0,100)) +
    labs(x="2019 Rosette 1 # Leaves", y="2020 Rosette 1 # Leaves",  color="Precipitation", shape="Snowmelt")
```

## Plant size vs. treatments
```{r plantsize}
ggplot(drop_na(cen18, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2018 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen18, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal)+ labs(x="Precipitation", y="2018 Rosette 1 # Leaves",  color="Snowmelt")

ggplot(drop_na(cen19, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2019 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen19, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal)+ labs(x="Precipitation", y="2019 Rosette 1 # Leaves",  color="Snowmelt")

ggplot(drop_na(cen20, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2020 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen20, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2020 Rosette 1 # Leaves",  color="Snowmelt")
```

# Leaf traits

```{r checkleaftraits, fig.height=8, fig.width=6}
with(lt, table(date, plot)) %>% kable

lt %>% group_by(date) %>% mutate(index=row_number(), date=factor(date)) %>% 
ggplot(aes(x=index, y=water_content, color=factor(date), group=factor(plot))) + 
  facet_wrap(vars(paste(year, round, sep=" round "), date)) +
  geom_hline(yintercept=0.6) + geom_hline(yintercept=0.85) + 
  geom_point() + geom_smooth(method="lm", color="black", se=F) + 
  scale_color_brewer(palette="Set1", direction=-1, guide=F) +
  labs(x="Measurement order", y="Water content", color="Plot")

lt %>% pivot_longer(c(trichomes, trichome_density, water_content, sla_wet, sla, leaf_area_cm2), names_to="trait") %>% 
  ggplot(aes(x=factor(date), y=value, color=factor(date))) + facet_wrap(vars(trait), scales="free_y") + geom_boxplot() +
  scale_x_discrete(guide=guide_axis(angle=90)) + labs(x="Date", y="", color="Date") + 
  scale_color_brewer(palette="Set1", direction = -1, guide=F) + theme(legend.position = "top")

ggplot(lt, aes(x=wet_weight_g, y=dry_weight_g, color=factor(date))) + geom_point() + geom_abline(slope=1, intercept = 0) +
  geom_abline(slope=1-median(lt$water_content, na.rm=T), intercept = 0) + geom_smooth(se=F) + 
  coord_cartesian(xlim=c(0,NA), ylim=c(0, NA)) + labs(x="Wet weight (g)", y = "Dry weight (g)", color="Date") +
    scale_color_brewer(palette="Set1", direction=-1) + theme(legend.position="top")

ggplot(lt, aes(x=sla_wet, y=sla, color=factor(date))) + geom_point() + geom_smooth(se=F) + geom_rug()+
  labs(x=paste("Wet", tolower(traitnames.units["sla"])), y = traitnames["sla"], color="Date") +
    scale_color_brewer(palette="Set1", direction=-1) + theme(legend.position="top")

ggplot(lt, aes(x=sla_wet, y=sla, color=factor(date))) + geom_point()

lt %>% pivot_longer(c(wet_weight_g, dry_weight_g, trichomes), names_to="trait") %>% 
  ggplot(aes(x=leaf_area_cm2, y=value, color=factor(date))) + facet_wrap(vars(trait), scales="free_y", ncol=1) + 
  geom_point(size=0.7) + geom_smooth(se=F) + labs(x="Leaf area (cm2)", y="", color="Date") + 
  scale_color_brewer(palette="Set1", direction=-1)

lt %>% pivot_longer(c(trichome_density, sla_wet, sla, water_content), names_to="trait") %>% 
  ggplot(aes(x=leaf_area_cm2, y=value, color=factor(date))) + facet_wrap(vars(trait), scales="free_y") + 
  geom_point(size=0.7) + geom_smooth(se=F) +   labs(x="Leaf area (cm2)", y="", color="Date") + 
  scale_color_brewer(palette="Set1", direction=-1) +
  theme(legend.position = "top")

library(ggfortify) 
lt %>%  mutate(id = paste0(str_sub(year(date),3,4), month(date), day(date), "-",plantid),
               id = ifelse(id=="20728-3C162" & trichomes==120, "20728-3C162B", id)) %>%  #two leaves collected
  select(id, wet_weight_g, dry_weight_g, leaf_area_cm2) %>% column_to_rownames("id") %>% drop_na() %>% 
  prcomp(scale=T) %>% 
  autoplot(data = lt %>% mutate(date=factor(date)) %>% drop_na(wet_weight_g, dry_weight_g, leaf_area_cm2),
           colour = "date", 
           loadings = TRUE, loadings.colour="black", loadings.label = TRUE, loadings.label.colour="black",
           shape=F, frame=T) + theme(legend.position = "top") +
  scale_color_brewer("Date", palette="Set1", direction=-1) + scale_fill_brewer("Date",palette="Set1", direction=-1)
```

# Physiology traits

```{r licor_all, fig.width=12, fig.height=12}
# plot all the LICOR variables by date
licor_date_pal <- sample(rainbow(nlevels(factor(licor$date))))
licor %>% pivot_longer(Photo:StableF) %>% mutate(index=row_number()) %>% filter(!name %in% c("Area","BLC_1","BLCond","StmRat")) %>% 
  ggplot(aes(x=index, y= value, color=factor(date), shape=year)) + facet_wrap(vars(name), scales="free_y", ncol=4) + geom_point() + scale_color_manual(values=licor_date_pal, guide=F) + theme_dark() + labs(x="Measurement", y="", shape="Year")
```

```{r licor}
licor %>% ggplot(aes(x=Cond, y= Photo, color=Ci>0, shape=year)) + geom_point()

licor %>% filter(Ci>0) %>%  ggplot(aes(x=Cond, y= Photo, color=factor(date), linetype=year, shape=year)) + geom_point() + geom_smooth(method="lm", se=F) + scale_linetype_manual(values=c(3,1,4)) + scale_color_manual(values=licor_date_pal, guide=F) + theme_dark()

licor %>% filter(Ci>0) %>% ggplot(aes(y=Photo/Cond, x=factor(date), color=year)) + geom_boxplot(size=1.5) + 
 scale_color_manual(values=year_pal) + scale_x_discrete(guide = guide_axis(angle = 90)) + theme_minimal()

licor %>% filter(Ci>0) %>% left_join(treatments)%>% ggplot(aes(y=Photo/Cond, x=factor(date), color=water)) + 
  geom_boxplot(position=position_dodge2(preserve="single")) + 
 scale_color_manual(values=water_pal) + scale_x_discrete(guide = guide_axis(angle = 90)) + theme_minimal()

pt %>% ggplot(aes(x=conductance, y= photosynthesis, color=factor(date), linetype=year, shape=year)) + geom_point() + geom_smooth(method="lm", se=F) + scale_linetype_manual(values=c(3,1,4)) + scale_color_manual(values=licor_date_pal, guide=F) + theme_dark()

pt %>% select(date, plantid, photosynthesis, conductance, WUE, notes) %>% 
  filter(is.na(conductance)) %>% mutate(conductance = photosynthesis/WUE) %>% kable
```
These conductance entries were blank, filled from WUE. There is another entry with conductance = 1.23 that was not blanked.

# Floral traits

## Floral trait correlations
```{r correlations}
mt %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width) %>% 
  ggpairs()

nt %>% select(nectar_24_h_ul, nectar_conc, nectar_sugar_24_h_mg) %>% 
  ggpairs()

ggplot(nt, aes(x=nectar_24_h_ul, y = nectar_conc)) + geom_point() + labs(x=traitnames[["nectar_conc"]], y=traitnames[["nectar_24_h_ul"]]) + geom_smooth(se=F)

mnps.plantyr %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width, nectar_24_h_ul, nectar_conc, nectar_sugar_24_h_mg, height_cm, flowers_est) %>% 
  ggcorr(hjust=0.85, layout.exp=2, label=T, label_round=2)

mnps.plantyr %>% select(all_of(seedtraits[c(1,7,2:6)])) %>% 
  ggcorr(hjust=0.85, layout.exp=2, label=T, label_round=2)
```

## Floral traits within season

No obvious effect of time within a season for corolla length/width or sepal width. The timing of trait measurements was different each year:

```{r morph_season}
ggplot(sm, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point(data=ph, aes(y= height_cm/2, color="Phenology (height)"))+
  geom_point(data=mt, aes(y= corolla_length, color="Morphology"))+
  geom_point(data=mt, aes(y= nectar_24_h_ul, color="Nectar"))+
  geom_point(aes(y = VWC, color="Soil moisture")) +
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color="grey40")+
  labs(x="Day of year", shape="Snow", linetype="Snow", y="", color="Measurement")+
  scale_color_manual(values=c("purple","orange","grey","black"))+
  scale_x_continuous(breaks=seq(120, 240, by=20))+
  scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(1.5, "lines"), axis.text.y=element_blank())

ggplot(data=mt, aes(x=yday(date), y= corolla_length, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y=traitnames[["corolla_length"]], x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()

ggplot(data=mt, aes(x=yday(date), y= nectar_24_h_ul, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y=traitnames[["nectar_24_h_ul"]], x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()

ggplot(data=nt, aes(x=yday(date), y= nectar_conc, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y=traitnames[["nectar_conc"]], x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()
```

## Selection on floral traits

```{r mt_selection, fig.height=8, fig.width=6}
mnps.plantyr %>% select(all_of(c(floraltraits, "seeds_est")), year) %>% 
  pivot_longer(all_of(floraltraits), names_to="trait") %>% 
  ggplot(aes(x=value, y=seeds_est, color=year)) + geom_point() + geom_smooth(method="lm",se=F) + 
  facet_wrap(vars(trait), scales="free_x", labeller=as_labeller(traitnames)) + 
  labs(x="Trait", y="Seed production", color="year")+
  theme_bw() + scale_color_manual(values=year_pal)

mt.mod.seeds <- map(set_names(floraltraits), 
                    ~ lm(seeds_rel ~ trait.z * year, 
                         data=mnps.plantyr %>% mutate(trait.z = mnps.plantyr[[.x]] / alltraits.sd[[.x]]) %>% 
                           group_by(year) %>% mutate(seeds_rel = seeds_est/mean(seeds_est, na.rm=T)))) # mean seeds in each yr

mt.mod.seeds.coefs <- map_dfr(mt.mod.seeds, tidy, .id="trait")
mt.mod.seeds.tests <- map(mt.mod.seeds, car::Anova, type=3) %>% map_dfr(tidy, .id="trait")
mt.mod.seeds.emm <- map_dfr(mt.mod.seeds, ~summary(emtrends(.x, ~1, var="trait.z")), .id="trait")

#get the the standardized selection differential S' (change in relative fitness / standardized trait value)
mt.mod.seeds.coefs %>% filter(term=="trait.z") %>% arrange(estimate)


#See if VWC affects relative fitness
ggplot(mnps.plantyr, aes(x=VWC, y=seeds_est, color=year)) + geom_point() + geom_smooth(method="lm",se=F) + 
  labs(x="VWC", y="Seed production", color="year")+ theme_bw() + scale_color_manual(values=year_pal)

seeds.mod.sm <- lmer(seeds_rel ~ VWC * year + (1|plotid), 
                     data=mnps.plantyr %>% group_by(year) %>% mutate(seeds_rel = seeds_est/mean(seeds_est, na.rm=T)))
anova(seeds.mod.sm, ddf="Ken")
emtrends(seeds.mod.sm, ~year, var="VWC")
```


# Models

## Plasticity: water and snowmelt treatments

This is a split-plot analysis, based on the original experimental design. The models use:

* *mnps.plantyr*, the plant means within each year
* *snow*, the snowmelt treatment (except for plot 2 in 2019)
* *water*, the water treatments with mock rainouts and controls combined
* *plot* as a random effect (subplots vary within this larger unit)
* *plotid* as a random effect (plants vary within this larger unit)

The models are specified as:

    lmer(trait ~ year * snow * water + (1|snow:plot) + (1|snow:plotid), data = mnps.plantyr)

### ANOVA tables

```{r splitplot}
lt.subplot.r2 <- lt.subplotround %>% filter(round=="2")
lt.plantyr.r2 <- lt.plantyr %>% filter(round=="2")

options(contrasts = c("contr.sum", "contr.poly"))
mod.split <- c(map(set_names(traits), 
                   ~ lmer(mnps.plantyr[[.x]]  ~ year * snow * water + (1|snow:plot) + (1|snow:plotid),
                          data=mnps.plantyr)),
               map(set_names(leaftraits), 
                   ~ lmer(lt.plantyr.r2[[.x]] ~ year * snow * water + (1|snow:plot) + (1|snow:plotid),
                          data=lt.plantyr.r2)),
               map(set_names(phystraits), 
                   ~ lmer(pt.plantyr[[.x]]    ~ year * snow * water + (1|snow:plot) + (1|snow:plotid) + (1|round),
                          data=pt.plantyr)))

mod.split.coefs <- map_dfr(mod.split, tidy, .id="trait")
mod.split.tests <- map(mod.split, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
mod.split.emm <- map_dfr(mod.split, ~ summary(emmeans(ref_grid(.x), ~ year * snow * water)), .id = "trait")
anova_table(mod.split.tests, mod.split) %>% kable()
re_table(mod.split.coefs) %>% kable(digits=3)
```

### Plots

```{r multi_splitplot, fig.width=4.5, fig.height=10}
plot_split_plot <- function(emm, data, traits.plot, geom = "point", plot.emm = TRUE, facets="year") {
  
  data.long <- data %>% select(c(all_of(traits.plot), all_of(facets), water, snow, sun_date, precip_est_mm)) %>% 
    pivot_longer(all_of(traits.plot), names_to="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal")) %>% 
           #value = (value - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)]) %>% 
    drop_na(value, water) %>% filter(trait %in% traits.plot)
  
  traitnames.multiline <- ifelse(nchar(traitnames.units) > 30, str_replace(traitnames.units, fixed("("),"\n("), traitnames.units)
  
  split_plot <- emm %>% filter(trait %in% traits.plot) %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal")) %>%  
           #emmean = (emmean - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)],
           #SE = SE/alltraits.sd[as.character(trait)]) %>% 
  ggplot(aes(x=water, color=snow)) +
    labs(x="Precipitation", y="Standardized trait", color="Snowmelt") + 
    scale_y_continuous(position="right") + scale_x_discrete(guide=guide_axis(angle=90)) + 
    #coord_fixed() + #maintains y scale across plots if figure width is constant
    scale_color_manual(values=c("grey","black"), guide=guide_legend(override.aes = list(shape=15, size=ifelse(plot.emm,5,1)))) + 
    theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            axis.title.y= element_blank(),
                            panel.border = element_rect(fill=NA, colour = "black"), 
                            panel.spacing=unit(0, "pt"), plot.margin = margin(0,0,0,0, "pt"), legend.position = "top") +
    switch(facets, 
           year = facet_grid(trait ~ year, scales="free_y", switch="y", 
                             labeller = as_labeller(c(traitnames.multiline, set_names(2018:2020)))),
           year.round = facet_grid(trait ~ year.round, scales="free_y", switch="y", 
                                   labeller = as_labeller(c(traitnames.multiline, set_names(levels(lt$year.round)))))) + 
    switch(geom,
           boxplot = geom_boxplot(data=data.long, aes(y=value), position=position_dodge(width=0.8), show.legend=!plot.emm,
                                  fatten=ifelse(plot.emm, NULL, 1), outlier.size=0.5),
           violin =   geom_violin(data=data.long, aes(y=value), position=position_dodge(width=0.8), show.legend=F),
           point =     geom_point(data=data.long, aes(y=value), position=position_dodge(width=0.8), shape=3))
  
  if(plot.emm) { return(split_plot + 
                          geom_linerange(aes(ymin=emmean-SE, ymax=emmean+SE,), position=position_dodge(width=0.8), size=1, show.legend=F) +
                          geom_point(aes(y=emmean), position=position_dodge(width=0.8), shape="-", size=16))
  } else return(split_plot + coord_cartesian(ylim=c(0,NA)))
}

# grid.arrange( #Morphological traits - boxplot of raw data
#   plot_split_plot(mt.mod.split.emm, mnps, floraltraits[1:4], geom="boxplot", plot.emm=F), 
#   plot_split_plot(mt.mod.split.emm, mnps, floraltraits[5:8], geom="boxplot", plot.emm=F)+ guides(color=F), nrow=1)

#Morphological traits - plot of EMMs and subplot means
  plot_split_plot(mod.split.emm, mnps.subplot, floraltraits[1:4])
  plot_split_plot(mod.split.emm, mnps.subplot, floraltraits[5:8])

# grid.arrange( #Vegetative and physiological traits - plot of raw data
#   plot_split_plot(lt.mod.split.emm, lt %>% filter(round==2), leaftraits, geom="boxplot", plot.emm=F), #second round only
#   plot_split_plot(pt.mod.split.emm, pt, phystraits, geom="boxplot", plot.emm=F) + guides(color=F), nrow=1)

#Vegetative and physiological traits - plot of EMMs and subplot means
  plot_split_plot(mod.split.emm, lt.subplotround %>% filter(round==2), leaftraits) #second round only
  plot_split_plot(mod.split.emm, pt.subplot, phystraits)
```

### Compensation of water addition for early snowmelt

```{r compensation}
mnps.plantyr <- mnps.plantyr %>% mutate(snowwater= paste0(snow, water))
lt.plantyr.r2 <- lt.plantyr.r2 %>% mutate(snowwater= paste0(snow, water))
pt.plantyr <- pt.plantyr %>% mutate(snowwater= paste0(snow, water))
library(multcomp)#careful, this loads MASS, which overrides dplyr::select
options(contrasts = c("contr.sum", "contr.poly"))
mod.split.glht <- c(map(set_names(traits), 
                   ~ lmer(mnps.plantyr[[.x]]  ~ snowwater * year + (1|snow:plot) + (1|snow:plotid),
                          data=mnps.plantyr)),
               map(set_names(leaftraits), 
                   ~ lmer(lt.plantyr.r2[[.x]] ~ snowwater * year + (1|snow:plot) + (1|snow:plotid),
                          data=lt.plantyr.r2)),
               map(set_names(phystraits), 
                   ~ lmer(pt.plantyr[[.x]]    ~ snowwater * year + (1|snow:plot) + (1|snow:plotid) + (1|round),
                          data=pt.plantyr))) %>% 
  #compare early snowmelt + addition vs. normal snowmelt + control
  map(glht, linfct = mcp(snowwater = c("EarlyAddition - NormalControl = 0"))) %>% 
  map_dfr(~ tibble(EarlyAddition_minus_NormalControl = coef(.x), p=summary(.x)$test$pvalues), .id="trait") 

mod.split.glht %>% kable(digits=3)
detach("package:multcomp", unload = TRUE); detach("package:TH.data", unload = TRUE)
```

### Compare treatment levels
```{r water_glht}
#This ignores any interactions with water (water:snow and water:year)
mod.split.tests.water <- map(mod.split, multcomp::glht, 
                             linfct = multcomp::mcp(water = c("Addition - Control = 0", "Reduction - Control = 0"))) %>% 
  map_dfr(tidy, .id="trait")

mod.split.tests.water %>% select(trait, contrast, adj.p.value) %>% pivot_wider(names_from=contrast, values_from=adj.p.value) %>% kable(digits=3)

map_dfr(mod.split, ~ summary(emmeans(ref_grid(.x), ~ snow)), .id = "trait")
```

### Split plot repeated measures 

#### Leaf traits

```{r lt_repeated}
lt %>% group_by(plantid, year.round) %>% tally() %>% ungroup %>% 
  pivot_wider(names_from=year.round, values_from=n, values_fill=0) %>% 
  rowwise() %>% mutate(tot = sum(c_across(!plantid))) %>% 
ggplot(aes(tot)) + geom_histogram(binwidth=1, color="black", fill="white") + 
  labs(x="Total samples", y="Number of plants", title="Leaf trait sampling") + theme_minimal()

with(lt, table(plantid, year.round)) %>% heatmap(scale="none", Colv = NA, col=c("white",status_pal["vegetative"],"yellow"))
```

```{r lt_connected, fig.height=11, fig.width=11}
lt %>% pivot_longer(all_of(leaftraits), names_to = "trait") %>% 
ggplot(aes(y=value, x=factor(round),color=water, fill=water)) + 
  facet_grid(trait ~year, scales="free_y", switch="y",labeller=as_labeller(c(traitnames.units, set_names(2018:2020)))) + 
  geom_violin(position=position_identity(), alpha=0.1)+ #, linetype=snow
  annotate("rect", xmin=1, xmax=2, ymin=-Inf, ymax=Inf, fill="white", color=NA) + 
  geom_point() + geom_path(aes(group=plantid))+
  scale_color_manual(values=water_pal) + scale_fill_manual(values=water_pal) + #scale_linetype_manual(values=c(2,1)) +
  labs(x="Round", y="", color="Precipitation", fill="Precipitation") + scale_y_continuous(position="right") + 
  theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"), legend.position = "top")
lt %>% pivot_longer(all_of(leaftraits), names_to = "trait") %>% 
ggplot(aes(y=value, x=factor(round),color=snow, fill=snow)) + 
  facet_grid(trait ~year, scales="free_y", switch="y",labeller=as_labeller(c(traitnames.units, set_names(2018:2020)))) + 
  geom_violin(position=position_identity(), alpha=0.1)+ 
  annotate("rect", xmin=1, xmax=2, ymin=-Inf, ymax=Inf, fill="white", color=NA) + 
  geom_point() + geom_path(aes(group=plantid))+
  scale_color_manual(values=snow_pal) + scale_fill_manual(values=snow_pal) + 
  labs(x="Round", y="", color="Snowmelt", fill="Snowmelt") + scale_y_continuous(position="right") + 
  theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"), legend.position = "top")

lt %>% pivot_longer(all_of(c(leaftraits, "sla_wet")), names_to = "trait") %>% arrange(date) %>% 
  ggplot(aes(y=value, x=year.round, color=water)) + 
  facet_wrap(vars(trait), scales="free_y",
             labeller=as_labeller(c(traitnames.units, set_names(paste("Wet" ,tolower(traitnames.units["sla"])),"sla_wet")))) + 
  geom_point() + geom_path(aes(group=plantid)) + scale_color_manual(values=water_pal) + 
  labs(x="Year and round", y="", color="Precipitation") + theme_minimal() + theme(legend.position = "top")
```

```{r lt_plantcorr, fig.height=11, fig.width=7.5}
lt.water_content <- lt %>% select(year.round, plantid, water, snow, water_content) %>% arrange(year.round) %>% 
  pivot_wider(names_from="year.round", values_from="water_content", values_fn=function(x) mean(x, na.rm=T))

plot_rm_change <- function(data, round_x, round_y, range_xy) {
  ggplot(data, aes_(x=as.name(round_x), y=as.name(round_y), color=as.name("water"))) + 
    geom_abline(intercept=0, slope=1) + geom_smooth(se=F, span=0.9) + geom_point(size=1)  +  geom_rug() +
    scale_color_manual(values=water_pal, guide=F) + coord_fixed(xlim=range_xy, ylim=range_xy) + theme_minimal()
}

do.call(grid.arrange, map2(c(levels(lt$year.round)[-6],"2018.1"), c(levels(lt$year.round)[-1],"2019.1"), plot_rm_change, data=lt.water_content, range_xy=range(lt$water_content, na.rm=T)))

#plot(lt.water_content %>% select(-c(plantid,water,snow)))
```


```{r lt_repeated_split, fig.height=8, fig.width=8}
lt.mod.split.rm <- map(set_names(leaftraits), 
                  ~ lmer(lt[[.x]] ~  year * round * water * snow + (1|snow:plot) + (1|snow:plotid) + (1|plantid), data=lt))
lt.mod.split.rm.coefs <- map_dfr(lt.mod.split.rm, tidy, .id="trait")
lt.mod.split.rm.tests <- map(lt.mod.split.rm, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
anova_table(lt.mod.split.rm.tests, lt.mod.split.rm) %>% kable()
lt.mod.split.rm.emm <- map_dfr(lt.mod.split.rm, ~ summary(emmeans(.x, ~  year * round * water * snow)), .id="trait") %>% 
  mutate(year.round=paste(year,round,sep="."))
plot_split_plot(lt.mod.split.rm.emm, lt.subplotround, leaftraits, facets="year.round") 

```

#### Physiology

```{r pt_repeated}
pt %>% group_by(plantid, year) %>% tally() %>% ungroup %>% 
  pivot_wider(names_from=year, values_from=n, values_fill=0) %>% 
  rowwise() %>% mutate(tot = sum(c_across(!plantid))) %>% 
ggplot(aes(tot)) + geom_histogram(binwidth=1, color="black", fill="white") + 
  labs(x="Total samples", y="Number of plants", title="Physiology sampling") + theme_minimal()

with(pt, table(plantid, year)) %>% heatmap(scale="none", Colv = NA, col=c("white",status_pal["vegetative"],"yellow"))

pt.mod.split.rm <- map(set_names(phystraits), 
                  ~ lmer(pt[[.x]] ~  year * snow * water + (1|snow:plot) + (1|snow:plotid) + (1|plantid) + (1|round), data=pt))
pt.mod.split.rm.coefs <- map_dfr(pt.mod.split.rm, tidy, .id="trait")
pt.mod.split.rm.tests <- map(pt.mod.split.rm, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
anova_table(pt.mod.split.rm.tests, pt.mod.split.rm) %>% kable()
```


## Plasticity: absolute snowmelt date and estimated summer precipitation

### ANOVA tables

```{r meltdate}
lt.subplot.r2 <- lt.subplotround %>% filter(round=="2")
lt.plantyr.r2 <- lt.plantyr %>% filter(round=="2")

options(contrasts = c("contr.sum", "contr.poly"))
mod.abs <- c(map(set_names(traits), 
                 ~ lmer(mnps.plantyr[[.x]]  ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), data=mnps.plantyr)),
             map(set_names(leaftraits), 
                 ~ lmer(lt.plantyr.r2[[.x]] ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), data=lt.plantyr.r2)),
             map(set_names(phystraits), 
                 ~ lmer(pt.plantyr[[.x]]    ~ sun_date * precip_est_mm + (1|plot) + (1|plotid) + (1|round),
                        data=pt.plantyr)))

mod.abs.coefs <- map_dfr(mod.abs, tidy, .id="trait")
mod.abs.tests <- map(mod.abs, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
anova_table(mod.abs.tests, mod.abs) %>% kable()
re_table(mod.abs.coefs) %>% kable(digits=3)
```

### Plots

```{r multi_meltdate, fig.width=9, fig.height=11}
plot_abs <- function(mod, data, traits.plot, facets = "trait") {
  data.long <- data %>% select(all_of(traits.plot), water, snow, year, sun_date, precip_est_mm) %>% 
    pivot_longer(all_of(traits.plot), names_to="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal"),
                      value = (value - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)]) %>% 
    drop_na(value, water) %>% filter(trait %in% traits.plot)
  
  grid_emmeans <- function(mods, use.rounds = F) {
    grid.points <- list(sun_date=range(treatments$sun_date), precip_est_mm=seq(25,150, by=25))
    if(use.rounds)  map_dfr(mods, ~ summary(emmeans(.x, ~ precip_est_mm * sun_date * round, at=grid.points)), .id="trait")
    else            map_dfr(mods, ~ summary(emmeans(.x, ~ precip_est_mm * sun_date,         at=grid.points)), .id="trait")
  }
  emm.grid <- mod[traits.plot] %>% grid_emmeans(use.rounds= facets=="trait.round") %>% 
    mutate(trait = fct_relevel(trait, traits.plot),
           emmean = (emmean - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)])
  
  traitnames.multiline <- ifelse(nchar(traitnames) > 30, str_replace(traitnames, fixed("("),"\n("), traitnames)
  
  abs_plot <- ggplot(emm.grid, aes(x=sun_date, y=emmean, color=precip_est_mm, group=precip_est_mm)) + 
    geom_point(data = data.long, aes(y=value, shape=year), size=2) + geom_line() + 
    scale_color_gradientn(colors=rev(water_pal), 
                          values=rev(c(1, (treatments %>% group_by(water) %>% summarize(across(precip_est_mm, mean)) %>%
                                             pull(precip_est_mm) %>% scales::rescale(from=c(25,150)))[2],0)),
                          breaks=seq(25,150, by=25)) + 
    labs(x="Snowmelt date", y="Standardized trait", color="Summer\nprecipitation\n(mm)", shape="Year") + 
    scale_y_continuous(position="right")+
    theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black")) +
    switch(facets, 
           trait =     facet_wrap(vars(trait), scales="fixed",labeller = as_labeller(traitnames.multiline)),
           trait.round = facet_grid(trait ~ round, scales="free_y", switch="y", 
                                   labeller = as_labeller(c(traitnames.multiline, 
                                                            set_names(paste("Round",1:2), 1:2)))))
    
  return(abs_plot)
}

# plot_abs(mt.mod.abs, mnps, floraltraits) #Morphological traits - raw data
plot_abs(mod.abs, mnps.subplot, floraltraits) + #Morphological traits - subplot means
  scale_y_continuous(limits=c(-1.7, 2.8), breaks=c(-1:2))

# grid.arrange(plot_abs(lt.mod.abs, lt, leaftraits), 
#              plot_abs(pt.mod.abs, pt, phystraits) + guides(color=F, shape=F), ncol=1) #raw data
grid.arrange(plot_abs(mod.abs, lt.subplot.r2, leaftraits)+ 
               scale_y_continuous(limits=c(-2.6, 1.8), breaks=c(-2:1)),
             plot_abs(mod.abs, pt.subplot, phystraits)+ guides(color=F, shape=F) +  
               scale_y_continuous(limits=c(-2.6, 1.8), breaks=c(-2:1)), ncol=1) #subplot means

```

### Slope plots and compensation

Compensation points calculated at average value of other predictor - only use for traits where there are effects of both predictors but no interaction
```{r compensation_meltdate_avg}
mod.abs.emm.avg <- map_dfr(set_names(c("sun_date", "precip_est_mm")), 
                   ~ mod.abs %>% map(emtrends, var = .x, ~1) %>% 
                     map(as_tibble) %>% set_names(alltraits) %>% bind_rows(.id="trait") %>% 
                     rename_with(.cols=ends_with("trend"), ~ "trend"), .id="predictor") %>% 
  mutate(predictor=fct_relevel(predictor,"sun_date"), 
         sd.plantyr = alltraits.sd[trait],
         trend = trend / sd.plantyr, SE = SE / sd.plantyr)

mod.abs.emm.avg.compensate <- mod.abs.emm.avg %>% 
  pivot_wider(names_from=predictor, values_from=c(trend, SE), id_cols=c(trait)) %>% 
  mutate(predictor = "compensation",
         precip_mm_per_snow_day = trend_sun_date / trend_precip_est_mm, 
         SE = abs(precip_mm_per_snow_day) * # propagate error in SEs using formula for quotients
           sqrt((SE_precip_est_mm / trend_precip_est_mm)^2 + (SE_sun_date / trend_sun_date)^2),
         prop_mean_total_precip_per_snow_day = precip_mm_per_snow_day / mean(groundcover$precip_est_mm, na.rm=T))
# divide compensation by mean summer precip 1990 - 2020 (mm)
mod.abs.emm.avg.compensate %>% select(-predictor) %>% kable(digits=3)
```

Compensation points calculated at four corners of the ranges of environemtnal variables
```{r compensation_meltdate, fig.width=9}
#get slopes with emtrends at range of other predictor, and scale slopes by SD of the plant means
predictor.ranges <- list(sun_date=c(115,160) , precip_est_mm=c(25, 150)) #a bit narrower than the experiment
mod.abs.emm <- map_dfr(set_names(c("sun_date", "precip_est_mm")), 
                   ~ mod.abs %>% 
                     map(emtrends, var = .x, 
                         at = predictor.ranges) %>% #[ifelse(.x=="sun_date","precip_est_mm","sun_date")]
                     map(as_tibble) %>% set_names(alltraits) %>% bind_rows(.id="trait") %>% 
                     rename_with(.cols=ends_with("trend"), ~ "trend"), .id="predictor") %>% 
  mutate(predictor=fct_relevel(predictor,"sun_date"), 
         sd.plantyr = alltraits.sd[trait],
         trend = trend / sd.plantyr, SE = SE / sd.plantyr)
  
mod.abs.emm.plot.sun_date <- 
  mod.abs.emm %>% filter(predictor=="sun_date", sun_date == predictor.ranges$sun_date[1]) %>% 
  ggplot(aes(x=fct_relevel(trait, rev(alltraits)), y=trend, color=factor(precip_est_mm))) +  
  facet_wrap(vars(predictor), nrow=1, scales="free_x", 
             labeller=labeller(predictor=c(sun_date="Snowmelt date"))) + 
  geom_hline(yintercept=0) +  
  geom_pointrange(aes(ymin=trend - SE, ymax = trend + SE), size=0.4, position=position_dodge2(0.2)) + 
  labs(y="Slope of standardized trait vs. predictor", x="Trait", color="Estimated summer precip (mm)") + 
  scale_color_manual(values= unname(water_pal[c(3,1)])) +
  scale_x_discrete(labels = traitnames) + coord_flip() +  theme_bw() + theme(legend.position = "top")

mod.abs.emm.plot.precip_est_mm <- 
  mod.abs.emm %>% filter(predictor=="precip_est_mm", precip_est_mm == predictor.ranges$precip_est_mm[1]) %>% 
  ggplot(aes(x=fct_relevel(trait, rev(alltraits)), y=trend, color=factor(sun_date))) +  
  facet_wrap(vars(predictor), nrow=1, scales="free_x", 
             labeller=labeller(predictor=c(precip_est_mm="Estimated summer precip (mm)"))) + 
  geom_hline(yintercept=0) +  
  geom_pointrange(aes(ymin=trend - SE, ymax = trend + SE), size=0.4, position=position_dodge2(0.2)) + 
  labs(y="Slope of standardized trait vs. predictor", x="", color="Snowmelt date") + 
  scale_color_manual(values= unname(snow_pal)) +
  scale_x_discrete(labels = traitnames) + coord_flip() +  theme_bw() + 
    theme(legend.position = "top",axis.text.y = element_blank(), axis.ticks.y = element_blank())


#calculate how much summer precip is required to compensate for 1 day earlier snowmelt
#mt.exclude.compensate <- mt.mod.abs.tests %>% filter(term %in% c("precip_est_mm", "sun_date"), p.value > 0.05) %>% pull(trait) %>% unique  #if P> 0.05 for effect of summer precip or snowmelt
exclude.compensate<- c("")#c("height_cm", #snowmelt/precip slope too big and P > 0.05 for both
                      #    "sepal_width",#sepal_width has interaction with sun_date,
                       # "seeds_per_flower", "seeds_per_fruit", "seeds_est", "open")
mod.abs.emm.compensate <- mod.abs.emm %>% 
  pivot_wider(names_from=predictor, values_from=c(trend, SE), id_cols=c(trait, sun_date, precip_est_mm)) %>% 
  mutate(predictor = "compensation",
         precip_mm_per_snow_day = trend_sun_date / trend_precip_est_mm, 
         SE = abs(precip_mm_per_snow_day) * # propagate error in SEs using formula for quotients
           sqrt((SE_precip_est_mm / trend_precip_est_mm)^2 + (SE_sun_date / trend_sun_date)^2),
         exclude = trait %in% exclude.compensate,
         precip_mm_per_snow_day = ifelse(exclude, NA, precip_mm_per_snow_day), SE = ifelse(exclude, NA, SE))

mod.abs.emm.compensate.plot <- mod.abs.emm.compensate %>% 
    ggplot(aes(x=fct_relevel(trait, rev(alltraits)), y=precip_mm_per_snow_day, 
               color=paste0(sun_date, ", ", precip_est_mm))) +  
  facet_wrap(vars(predictor)) +   geom_hline(yintercept=0) +   
  geom_pointrange(aes(ymin=precip_mm_per_snow_day - SE, ymax = precip_mm_per_snow_day + SE), size=0.4,
                  position=position_dodge2(0.4)) + 
  labs(y="Summer precipitation (mm) / day earlier snowmelt", x="", color="Snowmelt, Precip") +
    scale_x_discrete(labels = traitnames) +   coord_flip(ylim = c(-5,17)) +   
  theme_bw() + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), legend.position="top")
                                       
grid.arrange(mod.abs.emm.plot.sun_date, mod.abs.emm.plot.precip_est_mm, mod.abs.emm.compensate.plot, nrow=1, widths=c(1,0.5,0.8))
```


### Absolute snowmelt repeated measures 

```{r abs_repeated}
lt.mod.abs.rm <- map(set_names(leaftraits), 
                  ~ lmer(lt[[.x]] ~ sun_date * precip_est_mm * round + (1|plot) + (1|plotid) + (1|plantid), data=lt))
lt.mod.abs.rm.coefs <- map_dfr(lt.mod.abs.rm, tidy, .id="trait")
lt.mod.abs.rm.tests <- map(lt.mod.abs.rm, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
anova_table(lt.mod.abs.rm.tests, lt.mod.abs.rm) %>% kable()
plot_abs(lt.mod.abs.rm, lt.subplotround, leaftraits, facets="trait.round")

pt.mod.abs.rm <- map(set_names(phystraits), 
                  ~ lmer(pt[[.x]] ~ sun_date * precip_est_mm + (1|plot) + (1|plotid) + (1|plantid) + (1|round), data=pt))
pt.mod.abs.rm.coefs <- map_dfr(pt.mod.abs.rm, tidy, .id="trait")
pt.mod.abs.rm.tests <- map(pt.mod.abs.rm, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
anova_table(pt.mod.abs.rm.tests, pt.mod.abs.rm) %>% kable()
plot_abs(pt.mod.abs.rm, pt.subplot, phystraits)
```

### Lagged effects of snowmelt

```{r meltdate_lag}
mnps.plantyr <- mnps.plantyr %>% left_join(meltdates %>% select(year, plot, sun_date_lag=sun_date) %>% 
  bind_rows(groundcover %>% select(year, sun_date_lag=first_0_cm_day) %>% filter(year == 2017) %>%
              mutate(year=as.factor(year)) %>% left_join(tibble(year="2017", plot=as.character(1:6))))%>% 
  mutate(year = as.character(as.integer(as.character(year)) + 1)))

mod.abs.lag <- c(map(set_names(traits), 
                 ~ lmer(mnps.plantyr[[.x]]  ~ sun_date_lag * precip_est_mm + (1|plot) + (1|plotid), data=mnps.plantyr)))
mod.abs.lag.tests <- map(mod.abs.lag, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
anova_table(mod.abs.lag.tests, mod.abs.lag) %>% kable()

ggplot(mnps.plantyr %>% distinct(year, sun_date, sun_date_lag), aes(x=sun_date, y=sun_date_lag, color=year)) + geom_point()

ggplot(mnps.plantyr, aes(x=sun_date_lag, y=flowers_est, color=year)) + geom_point() + geom_smooth(method="lm") + scale_color_manual(values=year_pal) + theme_minimal() + labs(x="Melt date the previous year", y=traitnames["flowers_est"], color="Year")
```


## Plasticity: soil moisture

Both snowmelt timing and summer precipitation affect soil moisture and its timing. This analysis regresses the floral traits against the mean soil moisture in a subplot. Slopes may differ among years. The models use:

* *mt.plantyr*, the trait data averaged by plant within year
* *VWC*, the summer mean of the approximately weekly mean volumetric water content in a subplot. This may need to change to a subset of the summer because it includes soil moisture taken after traits were measured. See section on within-season variation for timing of measurements in each year.
* *plotid* (subplot) as a random effect (plants vary within this larger unit)

The models are specified as:

    lmer(trait ~ year * VWC + (1|plotid), data = mt.plantyr)

### ANOVAs

```{r sm}
lt.subplot.r2 <- lt.subplotround %>% filter(round=="2")
lt.plantyr.r2 <- lt.plantyr %>% filter(round=="2")

pt.plantyr <- pt.plantyr %>% mutate(VWC = ifelse(is.na(VWC.plant), VWC.sm, VWC.plant)) #phys traits had the VWC measured next to the plant in 2019 and 2020 only - substitute VWC in subplot on closest date
pt <- pt %>% mutate(VWC = ifelse(is.na(VWC.plant), VWC.sm, VWC.plant)) 

options(contrasts = c("contr.sum", "contr.poly"))
mod.sm <- c(map(set_names(traits), 
                  ~ lmer(mnps.plantyr[[.x]]  ~ year * VWC + (1|plotid), data=mnps.plantyr)),
            map(set_names(leaftraits), 
                  ~ lmer(lt.plantyr.r2[[.x]] ~ year * VWC + (1|plotid), data=lt.plantyr.r2)),
            map(set_names(phystraits), 
                  ~ lmer(pt.plantyr[[.x]]    ~ year * VWC + I(VWC^2) + (1|plotid) + (1|round), data=pt.plantyr)))
#significant effect of quadratic term for sepal_width, photosynthesis, conductance, but not year * VWC^2

mod.sm.coefs <- map_dfr(mod.sm, tidy, .id="trait")
mod.sm.tests <- map(mod.sm, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
anova_table(mod.sm.tests, mod.sm) %>% kable()
re_table(mod.sm.coefs) %>% kable(digits=3)
```

### Plots

```{r emm_sm, fig.height=8, fig.width=8}
mod.sm.emm <-          map_dfr(mod.sm, ~ summary(emtrends(.x, ~ year, var = "VWC")), .id="trait") %>% 
  mutate(trait.SD = alltraits.sd[trait], VWC.trend.SD = VWC.trend/trait.SD, SE.SD = SE/trait.SD, .after="VWC.trend")
#mod.sm.emm.allyears <- map_dfr(mod.sm, ~ summary(emtrends(.x, ~ 1   , var = "VWC")), .id="trait") %>% 
#  mutate(trait.SD = alltraits.sd[trait], VWC.trend.SD = VWC.trend/trait.SD, SE.SD = SE/trait.SD, .after="VWC.trend")

#library(googlesheets4)
#bind_rows(mod.sm.emm, mod.sm.emm.allyears %>% rename(year=`1`)) %>% 
#  mutate(predictor="VWC", .before=1) %>% mutate(traitname=traitnames[trait], .before="trait") %>% 
#  write_sheet(ss=filter(datasheets, name=="Maxfield Results"), sheet="slopes_VWC")
                      # left_join(mod.sm.emm, mnps.subplot %>% group_by(year) %>% select(traits) %>%
                      #   summarize(across(traits, ~ sum(!is.na(.x)))) %>%
                      #   pivot_longer(traits, names_to="trait", values_to="n"))

mod.sm.emm %>% #filter(trait %in% floraltraits) %>% 
  ggplot(aes(x=fct_relevel(trait, rev(alltraits)), y=VWC.trend.SD, color=year)) +  
  geom_hline(yintercept=0) + geom_pointrange(aes(ymin=VWC.trend.SD - SE.SD, ymax = VWC.trend.SD + SE.SD), position=position_dodge(width=0.3), size=0.4) + 
  labs(y="Slope of standardized trait vs. mean annual soil moisture", x="Trait", color="Year") +
  theme_classic() + scale_x_discrete(labels = traitnames) + coord_flip() + scale_color_manual(values=year_pal)
```

```{r multi_sm, fig.width=8, fig.height=10}
plot_sm <- function(mod, data, traits.plot) {
  data.long <- data %>% select(all_of(traits.plot), water, snow, year, sun_date, precip_est_mm, VWC) %>% 
    pivot_longer(all_of(traits.plot), names_to="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal"),
                      value = (value - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)]) %>% 
    drop_na(value, water) %>% filter(trait %in% traits.plot)
  
  VWC.range <- range(data$VWC, na.rm=T)
  emm.grid <- mod[traits.plot] %>% 
    map_dfr(~ emmeans(.x, ~ year*VWC, at=list(VWC=seq(from=floor(VWC.range[1]*10)/10, to=ceiling(VWC.range[2]*10)/10, by=0.1))) %>% as_tibble(), .id="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot),
           emmean = (emmean - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)]) %>% 
    left_join(data %>% drop_na(VWC) %>% group_by(year) %>% summarize(VWC.min = min(VWC), VWC.max=max(VWC))) %>%
    group_by(year) %>% filter(VWC >= VWC.min-0.1, VWC <= VWC.max+0.1)
  
  traitnames.multiline <- ifelse(nchar(traitnames) > 30, str_replace(traitnames, fixed("("),"\n("), traitnames)
  
  sm_plot <- ggplot(emm.grid, aes(x=VWC, y=emmean, color=year, group=year)) + 
    facet_wrap(vars(trait), scales="fixed",labeller = as_labeller(traitnames.multiline))+
    geom_point(data = data.long, aes(y=value, color=year), size=1) +  
    geom_line(size=2) + 
    scale_color_manual(values=year_pal) + 
    labs(x=ifelse("WUE" %in% traits.plot, "Soil moisture next to plant or in subplot on closest date (%VWC)", 
                  "Summer mean of soil moisture in subplot (%VWC)"), y="", color="Year") + 
    coord_fixed() + 
    theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"))
  return(sm_plot)
}

# plot_sm(mod.sm, mnps, floraltraits) #Morphological traits - raw data
plot_sm(mod.sm, mnps.plantyr, floraltraits) + scale_y_continuous(limits=c(-4.3,5), breaks=-4:4) #Morphological traits - plant means

# grid.arrange(plot_sm(mod.sm, lt, leaftraits), 
#              plot_sm(mod.sm, pt, phystraits) + guides(color=F, shape=F), ncol=1) #raw data
grid.arrange(plot_sm(mod.sm, lt.plantyr.r2, leaftraits),
             plot_sm(mod.sm, pt.plantyr,    phystraits)+ guides(color=F, shape=F) + coord_fixed(ratio=3), ncol=1) #plant means
```

## Mortality/flowering selection vs. treatments

```{r selection}
step_selection_mod <- function(trait, response, treatment, set, family="binomial", level=0.05) {
  re <- ifelse(treatment =="snow", "snow:plot", "plot")
  t.plantyr <- list(lt = lt.plantyr.r2, pt = pt.plantyr)[[set]] %>% 
    rename("trait"=trait)
  options(contrasts = c("contr.sum", "contr.poly"))
  mod <- glmmTMB(formula(paste(response, "~ trait * year * ", treatment," + (1|",re,")")), 
                            data=t.plantyr, family=family)
  
  droporder <- c(paste0("trait:year:", treatment),
                 paste0("trait:", treatment),
                 paste0("year:", treatment),
                 paste0("trait:year"))
  for(term_test in droporder) {
    pval <- car::Anova(mod) %>% tidy %>% filter(term==term_test) %>% pull(p.value)
    if(pval < level) return(mod)
    else mod <- update(mod, formula(paste0(". ~ . -", term_test)))
  }
  return(mod)
}

mod.selection <- 
  tibble(response = c("survived","flowered"), treatment=c("water","snow"), set = c("lt", "pt")) %>%
  tidyr::expand(response, treatment, set) %>% 
  full_join(tibble(trait=c(leaftraits, phystraits), 
                   set = c(rep("lt", length(leaftraits)), rep("pt", length(phystraits))))) %>% 
  mutate(model = pmap(., step_selection_mod),
         coefs = map(model, tidy),
         tests = map(model, ~ tidy(car::Anova(.x, type=3))))
```

```{r anova_selection, results='asis'}
mod.selection %>% group_by(response, treatment) %>% 
  group_walk(function(.x,.y) {cat(paste("Effect of trait on whether a plant",.y$response,"under different",.y$treatment,"conditions")); 
    Anova_table(.x %>% unnest(tests)) %>% kable %>% print})

mod.selection %>% unnest(coefs) %>% filter(effect=="ran_pars") %>% 
  select(trait, group, estimate, response, treatment) %>% 
  pivot_wider(names_from=group, values_from=estimate) %>% kable(digits=3)
```

```{r plot_selection}
# plot_selection <- function(trait, response, treatment, set) {
# list(lt = lt.plantyr.r2, pt = pt.plantyr)[[set]] %>% 
# ggplot(aes_string(y=response, x=trait, color=treatment, linetype="year")) + 
#   geom_point() + geom_smooth(se=F, method="glm", method.args = list(family=binomial)) +
#   scale_color_manual(values=c(water_pal, snow_pal))
# }
# mod.selection <- mod.selection %>% 
#   mutate(plot = pmap((.) %>% select(trait, response, treatment, set), plot_selection))
# do.call(grid.arrange, mod.selection %>% filter(response=="flowered", treatment=="water") %>% pull(plot))


plot_selection_facets <- function(response, treatment, set) {
  traits.plot <- list(lt=leaftraits, pt=phystraits)[[set]]
  p <- list(lt = lt.plantyr.r2, pt = pt.plantyr)[[set]] %>% 
    pivot_longer(all_of(traits.plot),names_to="trait") %>% 
    filter(year!="2020") %>% mutate(year=fct_drop(year)) %>% 
    ggplot(aes_string(y=response, x="value", color=treatment, linetype="year")) + 
    facet_wrap(vars(trait), scales="free_x")+
    geom_point() + geom_smooth(se=F, method="glm", method.args = list(family=binomial)) +
    scale_color_manual(values=c(water_pal, snow_pal)) + theme(legend.position="top") +
    labs(linetype="Year",x="",color=c(water="Precipitation",snow="Snowmelt")[treatment]) +
    theme_minimal() + theme(text=element_text(color="black"), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"), 
                            panel.spacing=unit(0, "pt"), plot.margin = margin(0,0,0,0, "pt"), 
                            legend.position = "top")
  if(set =="pt") return(p + guides(linetype=F, color=F))
  else return(p)
}

mod.selection.plots <- mod.selection %>% group_by(response, treatment, set) %>% summarize(.groups="drop") %>%  
  mutate(plots = pmap(., plot_selection_facets))

mod.selection.plots %>% group_by(response, treatment) %>% group_walk( ~ do.call(grid.arrange, .x$plots))
```

## Selection vs. absolute snowmelt, precip

```{r selection_abs, results='asis'}
pt.plantyr <- pt.plantyr %>% mutate(VWC = ifelse(is.na(VWC.plant), VWC.sm, VWC.plant)) #phys traits had the VWC measured next to the plant in 2019 and 2020 only - substitute VWC in subplot on closest date

selection_mod_abs <- function(trait, response, set) {
  t.plantyr <- list(lt = lt.plantyr.r2, pt = pt.plantyr)[[set]] %>% rename("trait"=trait)
  options(contrasts = c("contr.sum", "contr.poly"))
  glmmTMB(formula(paste(response, "~ trait * sun_date * precip_est_mm + (1|plot)")), data=t.plantyr, family="binomial")
}

mod.selection.abs <- 
  tibble(response = c("survived","flowered"), set = c("lt", "pt")) %>%
  tidyr::expand(response, set) %>% 
  full_join(tibble(trait=c(leaftraits, phystraits), 
                   set = c(rep("lt", length(leaftraits)), rep("pt", length(phystraits))))) %>% 
  mutate(model = pmap(., selection_mod_abs),
         coefs = map(model, tidy),
         tests = map(model, ~ tidy(car::Anova(.x, type=3))))

mod.selection.abs  %>% group_by(response) %>% 
  group_walk(function(.x,.y) {cat(paste("Effect of trait on whether a plant",.y$response,"under different absolute snowmelt / summer precip conditions")); 
    Anova_table(.x %>% unnest(tests)) %>% kable %>% print})

#sjPlot::plot_model((mod.selection.abs %>% filter(trait=="conductance", response=="flowered") %>% pull(model))[[1]])
```

## Selection vs. soil moisture

```{r selection_sm, results='asis'}
pt.plantyr <- pt.plantyr %>% mutate(VWC = ifelse(is.na(VWC.plant), VWC.sm, VWC.plant)) #phys traits had the VWC measured next to the plant in 2019 and 2020 only - substitute VWC in subplot on closest date

selection_mod_sm <- function(trait, response, set) {
  t.plantyr <- list(lt = lt.plantyr.r2, pt = pt.plantyr)[[set]] %>% rename("trait"=trait)
  options(contrasts = c("contr.sum", "contr.poly"))
  glmmTMB(formula(paste(response, "~ trait * year * VWC + (1|plot)")), data=t.plantyr, family="binomial")
}

mod.selection.sm <- 
  tibble(response = c("survived","flowered"), set = c("lt", "pt")) %>%
  tidyr::expand(response, set) %>% 
  full_join(tibble(trait=c(leaftraits, phystraits), 
                   set = c(rep("lt", length(leaftraits)), rep("pt", length(phystraits))))) %>% 
  mutate(model = pmap(., selection_mod_sm),
         coefs = map(model, tidy),
         tests = map(model, ~ tidy(car::Anova(.x, type=3))))

mod.selection.sm  %>% group_by(response) %>% 
  group_walk(function(.x,.y) {cat(paste("Effect of trait on whether a plant",.y$response,"under different VWC conditions")); 
    Anova_table(.x %>% unnest(tests)) %>% kable %>% print})
```

```{r summaryfigures, fig.width=7, fig.height=3, eval=F}
pt.plantyr %>% 
    pivot_longer(all_of(c("conductance","WUE")),names_to="trait") %>% 
    filter(year!="2020") %>% mutate(year=fct_drop(year)) %>% 
    ggplot(aes_string(y="survived", x="value", color="water", linetype="year")) + 
    facet_wrap(vars(trait), scales="free_x", labeller=as_labeller(traitnames.units))+
    geom_point() + geom_smooth(se=F, method="glm", method.args = list(family=binomial)) +
    scale_color_manual(values=c(water_pal, snow_pal)) + theme(legend.position="top") +
    labs(linetype="Year",x="",color="Precipitation", y="Survival") +
    theme_minimal() + theme(text=element_text(color="black"), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"), 
                            panel.spacing=unit(8, "pt"), plot.margin = margin(0,0,0,0, "pt"), 
                            legend.position = "top") 
ggsave(filename="../surv_water_pt.png", width=7, height=3)

plot_abs(mod.abs, mnps.subplot, floraltraits[1:6]) + #Morphological traits - subplot means
    scale_y_continuous(limits=c(-1.7, 2.8), breaks=c(-1:2))
ggsave(filename="../snow_precip_mt.png", width=7, height=4)
```

## Multivariate response
```{r multresp}
mt.morph <- mt[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","VWC")] %>% drop_na(1:5) #drop sepal_width
mt.pca <- prcomp(mt.morph[,-(6:7)], scale.=TRUE)

mt.morph.pca <- mt.morph %>%
  bind_cols(as.data.frame(mt.pca$x)) %>% 
  tibble::rowid_to_column("index") %>% 
  mutate(x_0=0,y_0=0,
         x_1=-corolla_width/2,y_1=corolla_length,
         x_2= corolla_width/2,y_2=corolla_length) %>% 
  pivot_longer(c(starts_with("x"),starts_with("y")), names_to=c(".value", "point"), names_sep="_")

scl <- 100
zoom <- 5.5
mt.pca.vars <- as.data.frame(mt.pca$rotation) %>% rownames_to_column("variable")
mt.pca.labs <- paste0("PC", 1:2, " (", round(100*mt.pca$sdev[1:2]^2/sum(mt.pca$sdev^2), 0), "% explained var.)") 
ggplot(mt.morph.pca) + theme_classic() +
  geom_polygon(aes(x= x/scl+PC1, y= y/scl+PC2, group=index, color=VWC, fill=VWC)) +
  #geom_point(aes(x=PC1, y=PC2, size=sepal_width/2), color="darkgreen") + 
  geom_point(aes(x=PC1, y=PC2 + anther_min/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + anther_max/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + style_length/scl), color="grey60", size=0.5, shape=4) +
  geom_segment(data=mt.pca.vars,aes(x=0, y=0, xend=PC1*zoom, yend=PC2*zoom), 
               arrow = arrow(length = unit(1/2, "picas")), color="black") +
  geom_text(data =mt.pca.vars, aes(label = gsub("_"," ",variable), x = PC1*zoom*1.4, y = PC2*zoom*1.4,
                                   angle =  (180/pi) * atan(PC2/PC1), hjust = 0), color="black")+
  scale_size_identity() + coord_fixed() + 
  scale_fill_viridis(direction=-1, option="plasma") + scale_color_viridis(direction=-1, option="plasma") + 
  labs(x=mt.pca.labs[1], y=mt.pca.labs[2])
ggsave("images/mt_morph_pca_noSW.pdf", height=15, width=20, device=cairo_pdf)

ggplot(mt, aes(x=corolla_length, y=corolla_width, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_smooth()

ggplot(mt, aes(x=corolla_length, y=style_length, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_abline(slope=1, intercept=0)

ggplot(mt, aes(y=style_length / corolla_length, x=VWC, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_abline(slope=0, intercept=1)+ geom_smooth()

ggplot(mt, aes(x=corolla_length, ymax = anther_max, ymin=anther_min, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_linerange() + theme_minimal()+ geom_abline(slope=1, intercept=0)

ggplot(mt, aes(y=anther_min / corolla_length, x=VWC, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_smooth()
```

```{r multresp_VWC, fig.height=7, fig.width=8}
mt.morph.plantyr <- mt.plantyr[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","VWC")] %>% drop_na() 
options(contrasts = c("contr.treatment", "contr.poly"))
mt.rda <- rda(scale(select(mt.morph.plantyr, !c(year, VWC))) ~ VWC * year, data=mt.morph.plantyr)
mt.rda
anova(mt.rda, by="term") %>% tidy %>% kable(digits=3)
#autoplot(mt.rda)

mt.rda.df <- fortify(update(mt.rda, ~ VWC + year))#drop interaction
ggplot(mapping=aes(x=RDA1, y=RDA2, label=Label)) + 
  geom_point(data=mt.rda.df %>% filter(Score=="sites") %>% bind_cols(mt.morph.plantyr), aes(color=VWC))+
  geom_segment(data=mt.rda.df %>% filter(Label=="VWC"), aes(x=0,y=0,xend=RDA1*3, yend=RDA2*3), 
               arrow=arrow(length=unit(0.5, "lines")), size=1.5)+
  geom_text(   data=mt.rda.df %>% filter(Label=="VWC"), aes(x=RDA1*3.5, y=RDA2*3.5, angle=atan(RDA2/RDA1)*180/pi))+
  geom_segment(data=mt.rda.df %>% filter(Score=="species"), aes(x=0,y=0,xend=RDA1, yend=RDA2), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(   data=mt.rda.df %>% filter(Score=="species"), aes(x=RDA1*1.9, y=RDA2*1.9, angle=atan(RDA2/RDA1)*180/pi,
    label = c(traitnames, c("anther_max"="Max stamen length", "anther_min"="Min stamen length"))[as.character(Label)] %>% str_remove(fixed(" (mm)"))), size=3)+
  geom_text(data=mt.rda.df %>% filter(Score=="centroids"), aes(label=str_remove(Label, "year")), fontface=2, size=6)+
  scale_color_viridis(direction=-1, option="plasma") + coord_fixed() + 
  scale_x_continuous(expand=expansion(mult=.16), breaks=c(-2,0,2))+   theme_minimal()
```


```{r multresp_discrete}
mt.morph.wm <- mt[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","precip_est_mm", "sun_date")] %>% drop_na(1:6) #and nectar (19,20) too
mt.cap.wm <- capscale(scale(mt.morph.wm[,-(6:8)]) ~ precip_est_mm * sun_date, data=mt.morph.wm)
mt.cap.wm
anova(mt.cap.wm, by="term") %>% tidy %>% kable
#plot CAP without interactions
autoplot(capscale(scale(mt.morph.wm[,-(6:8)]) ~ precip_est_mm + sun_date, data=mt.morph.wm)) + guides(color=F, shape=F)
```


# Phenology
```{r check_phenology}
walk(list(ph18.raw, ph19.raw, ph20.raw), 
     function(ph.raw) {
  y = ph.raw$year[1]
  p <- ph.raw %>% count(plot, subplot, plotid, plantid, julian) %>% add_count(plantid, name="n_plantid") %>% 
  mutate(plantid = fct_reorder(plantid, as.integer(factor(paste(plotid, n_plantid))),.desc=T)) %>% 
  ggplot(aes(julian, plantid, fill=factor(n)))+
  geom_tile()+ scale_fill_viridis(option="inferno", discrete=T) + 
  geom_point(aes(color=subplot, x=median(as.integer(julian))))+ scale_color_manual(values=sample(rainbow(10)))+
  geom_point(aes(color=plot, x=median(as.integer(julian))), position=position_nudge(x=-0.3))+ 
  theme_classic() + theme(panel.background = element_rect(fill="grey")) + guides(x=guide_axis(angle = 90)) +
  labs(x="Day of year", y="Plant (sorted by subplot)", fill="Observations", color="Plot", title=paste(y,"phenology"))
  print(p)})
```

```{r phenology_versus_total}
ggplot(mnps.plantyr, aes(x=flowers_est, y=open)) + geom_point() + facet_wrap(vars(year)) + geom_smooth(method="lm")
lm(open ~ flowers_est * year, data = mnps.plantyr) %>% emtrends(~year, var="flowers_est") %>% tidy
```


```{r phenology}
ggplot(ph %>% drop_na(height_cm), aes(x=julian, y=height_cm, shape=snow, color=water))+ geom_point() + geom_path(aes(group=plantid))+
  facet_wrap(vars(year), ncol=1)+
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Plant height", 
       x="Day of year", color="Precipitation", shape="Snow") +
  scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_color_manual(values=water_pal) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at(c("open","buds"), mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=open, shape=snow, linetype=snow, color=water))+ geom_point() + 
  facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Mean total flowers", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
    scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at("flowering", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=flowering, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Proportion flowering", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at("eggs", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=eggs, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Mean total eggs", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
    scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at("has_egg", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=has_egg, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Proportion of plants with eggs", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
    scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

```

## Floral display size
### 2020
```{r display_size}
ggplot(ph.plantid, aes(x=height_cm, y=open, color=water)) + geom_point() + geom_smooth(method="lm", se=F) + scale_color_manual(values=water_pal) + facet_wrap(vars(year), nrow=1) +
  labs(x="Mean inflorescence height (cm)", y="Mean number of open flowers", color="Precipitation")

ggplot(ph.plantid, aes(x=water, color=snow, y = height_cm)) +
    labs(x="Precipitation", y="Mean inflorescence height (cm)", color="Snowmelt") +
    geom_boxplot() + scale_color_manual(values=snow_pal) + facet_wrap(vars(year), nrow=1)
ggplot(ph.plantid, aes(x=VWC, y=height_cm, color=year)) + geom_point() + geom_smooth(se=F) +
    scale_color_manual(values=year_pal) + labs(x="Mean VWC", y="Mean inflorescence height (cm)") 
    
ggplot(ph.plantid, aes(x=water, color=snow, y = open)) +
    labs(x="Precipitation", y="Mean number of open flowers", color="Snowmelt") +
    geom_boxplot() + scale_color_manual(values=snow_pal) + facet_wrap(vars(year), nrow=1)
ggplot(ph.plantid, aes(x=VWC, y=open, color=year)) + geom_point() + geom_smooth(se=F)+
  scale_color_manual(values=year_pal) + labs(x="Mean VWC", y="Mean number of open flowers")
```

# Seeds

```{r seeds}
ggplot(sds %>% drop_na(plot), aes(color=water, x=fruits, y=seeds)) + geom_point() + geom_smooth(se=F) + 
  scale_color_manual(values=water_pal) + facet_wrap(vars(year), nrow=1) + theme_minimal() +
  labs(x="Counted fruits", y="Counted seeds", color="Precipitation")

for(tr in seedtraits) {
  print(ggplot(sds %>% drop_na(plot), aes_string(tr, x="water", color="snow")) + geom_boxplot() + 
    scale_color_manual(values=snow_pal) + facet_wrap(vars(year), nrow=1) + 
    scale_y_continuous(trans=ifelse(tr %in% c("seeds_est", "fruits_nonaborted"),"sqrt","identity"))+
    labs(x="Precipitation", y=traitnames[[tr]], color="Precipitation") +
        guides(x=guide_axis(angle = 90)))
  
  print(ggplot(sds, aes_string(tr, x="VWC", color="year")) + geom_point() + geom_smooth(se=F) + 
   scale_color_manual(values=year_pal) +
    scale_y_continuous(trans=ifelse(tr %in% c("seeds_est", "fruits_nonaborted"),"sqrt","identity"))+
    labs(x="Summer mean of soil moisture in subplot (VWC)", y=traitnames[[tr]], color="Year"))
}
```

## Fruiting phenology (2020)

```{r seed_phenology}
ggplot(sds20.date.zeroed, aes(x=julian, y=fruits_nonaborted, color=water, shape=snow, linetype=snow))+geom_jitter(width=2, height=0) + stat_summary(fun="mean", geom="line") + scale_y_sqrt() + scale_color_manual(values=water_pal)+
  labs(y="Nonaborted fruits", x="Day of year", color="Precipitation", shape="Snow", linetype="Snow")+ scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(sds20.date.zeroed, aes(x=julian, y=seeds, color=water, shape=snow, linetype=snow))+geom_jitter(width=2, height=0) + stat_summary(fun="mean", geom="line") + scale_y_sqrt() + scale_color_manual(values=water_pal)+
  labs(y="Seeds", x="Day of year", color="Precipitation", shape="Snow", linetype="Snow")+ scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(sds20.date.zeroed, aes(x=julian, y=prop_infested, color=water, shape=snow, linetype=snow))+geom_jitter(width=2, height=0) + stat_summary(fun="mean", geom="line") + scale_color_manual(values=water_pal)+
  labs(y="Proportion of nonaborted fruits infested", x="Day of year", color="Precipitation", shape="Snow", linetype="Snow")+ scale_linetype_manual(values=c(2,1)) +
  scale_y_continuous(labels = scales::percent)+ theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

sds20.date.zeroed %>% group_by(snow, water, julian) %>% summarize_at("fruiting", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=fruiting, shape=snow, linetype=snow, color=water))+ geom_point() +
  geom_path(aes(group=paste(snow, water))) +
  labs(y="Proportion fruiting (nonaborted fruits)", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))
```

