---
title: "Maxfield Meadow Ipomopsis Selection"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---


```{r setup, message=FALSE}
library(tidyverse)
library(emmeans)
library(broom)
library(lme4)
library(lmerTest)
library(GGally)
library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, 
                      fig.path = "images_selection/", dev="svglite", dev.args=list(fix_text_size=FALSE), fig.height=6, fig.width=6)
options(digits=3) # for kables

load("data/maxfield_data.rda") #load trait measurements, etc
mnps.plantyr <- mnps.plantyr %>% mutate(prop_uninfested = 1 - prop_infested)
#TODO eggs_per_flower has some high values when the eggs were averaged over 1 observation but the flowers were averaged over multiple weeks. need to calculate this before averaging

fitnesstraits <- c("seeds_est", "seeds_per_flower", "prop_uninfested", "eggs_per_flower")
fitnessnames <- c(traitnames, prop_uninfested = "Prop. uneaten fruits",
                  eggs_per_flower="Fly eggs per flower")[fitnesstraits]
morphtraits <- c("corolla_length", "corolla_width","style_length","anther_max","anther_min","sepal_width","nectar_24_h_ul", "nectar_conc")#floral traits less height and flower number, plus anthers
morphnames <- c(traitnames, anther_max="Max stamen length", anther_min="Min stamen length")[morphtraits]
```

# Fitness vs. treatments and soil moisture

## Seed traits

```{r seeds_treat, fig.height=12, fig.width=5}
sds %>% pivot_longer(all_of(seedtraits), names_to = "trait") %>% 
  ggplot(aes(y=value, x=water, color=snow)) + 
  facet_grid(trait~year, scales="free_y", labeller=as_labeller(c(set_names(2018:2020), traitnames))) +
  geom_boxplot() + scale_color_manual(values=snow_pal) +  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+
  labs(x="Precipitation", color="Snow") + 
  theme_bw() + theme(axis.title.y=element_blank()) + guides(x=guide_axis(angle = 90))

sds %>% pivot_longer(all_of(seedtraits), names_to = "trait") %>% 
  ggplot(aes(y=value, x=VWC, color=year)) + 
  facet_grid(trait~year, scales="free_y", labeller=as_labeller(c(set_names(2018:2020), traitnames))) +
  geom_point(size=0.5) + geom_smooth(se=F) + scale_color_manual(values=year_pal) +
  labs(x="Summer mean of soil moisture in subplot (VWC)", color="Year") + 
  theme_bw() + theme(axis.title.y=element_blank()) + guides(year="none")
```

## Fitness metrics

```{r fitness_treat, fig.height=8, fig.width=5}
mnps.plantyr %>% pivot_longer(all_of(fitnesstraits), names_to = "trait") %>% 
  ggplot(aes(y=value, x=water, color=snow)) + 
  facet_grid(trait~year, scales="free_y", labeller=as_labeller(c(set_names(2018:2020), fitnessnames))) +
  geom_boxplot() + scale_color_manual(values=snow_pal) +  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+
  labs(x="Precipitation", color="Snow") + 
  theme_bw() + theme(axis.title.y=element_blank()) + guides(x=guide_axis(angle = 90))

mnps.plantyr %>% pivot_longer(all_of(fitnesstraits), names_to = "trait") %>% 
  ggplot(aes(y=value, x=VWC, color=year)) + 
  facet_grid(trait~year, scales="free_y", labeller=as_labeller(c(set_names(2018:2020), fitnessnames))) +
  geom_point(size=0.5) + geom_smooth(se=F) + scale_color_manual(values=year_pal) +
  labs(x="Summer mean of soil moisture in subplot (VWC)", color="Year") + 
  theme_bw() + theme(axis.title.y=element_blank()) + guides(year="none")

# seeds.mod.sm <- lmer(seeds_rel ~ VWC * year + (1|plotid), 
#                      data=mnps.plantyr %>% group_by(year) %>% mutate(seeds_rel = seeds_est/mean(seeds_est, na.rm=T)))
# anova(seeds.mod.sm, ddf="Ken") %>% kable()
# emtrends(seeds.mod.sm, ~year, var="VWC") %>% kable()
```

## Fitness correlations

```{r fitness_corr}
mnps.plantyr %>% select(all_of(fitnesstraits)) %>% 
  ggpairs(mapping=aes(color=mnps.plantyr$year), lower=list(continuous="smooth_loess")) + 
  scale_color_manual(values=year_pal)+ scale_fill_manual(values=year_pal)

ggplot(sds %>% drop_na(plot), aes(color=water, x=fruits, y=seeds)) + geom_point() + geom_smooth(se=F) + 
  scale_color_manual(values=water_pal) + facet_wrap(vars(year), nrow=1) + theme_minimal() +
  labs(x="Counted fruits", y="Counted seeds", color="Precipitation")
```

# Floral trait correlations

```{r correlations}
mt %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width) %>% 
  ggpairs()

nt %>% select(nectar_24_h_ul, nectar_conc, nectar_sugar_24_h_mg) %>% 
  ggpairs()

ggplot(nt, aes(x=nectar_24_h_ul, y = nectar_conc)) + geom_point() + labs(y=traitnames[["nectar_conc"]], x=traitnames[["nectar_24_h_ul"]]) + geom_smooth(se=F)

mnps.plantyr %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width, nectar_24_h_ul, nectar_conc, nectar_sugar_24_h_mg, height_cm, flowers_est) %>% 
  ggcorr(hjust=0.85, layout.exp=2, label=T, label_round=2)

mnps.plantyr %>% select(all_of(seedtraits[c(1,7,2:6)])) %>% 
  ggcorr(hjust=0.85, layout.exp=2, label=T, label_round=2)
```

# Selection on individual traits

```{r mt_selection, fig.height=6, fig.width=10}
walk(fitnesstraits, ~ 
       print(mnps.plantyr %>% 
               mutate(prop_uninfested = if_else(fruits_nonaborted > 4, prop_uninfested, NA)) %>% 
               pivot_longer(all_of(floraltraits)) %>% 
               ggplot(aes_string(y=.x, x="value", color="year")) + 
               geom_point(size=0.5) + geom_smooth(method="lm", se=F) + 
               facet_wrap(vars(name), scales="free_x", labeller=as_labeller(traitnames), nrow=2) + 
               labs(y=fitnessnames[.x], color="Year")+
               theme_bw() + scale_color_manual(values=year_pal) + theme(axis.title.x=element_blank()))) 

walk(fitnesstraits, ~ 
       print(mnps.plantyr %>% 
               mutate(prop_uninfested = if_else(fruits_nonaborted > 4, prop_uninfested, NA)) %>% 
               pivot_longer(all_of(floraltraits)) %>%
               ggplot(aes(x=value, color=paste(water,snow))) + #shape=year, linetype=year,
               geom_point(aes_string(y=.x), size=0.5) + geom_smooth(aes_string(y=.x), method="lm",se=F) + 
               facet_wrap(vars(name), scales="free_x", labeller=as_labeller(traitnames), nrow=2) + 
               labs(y=fitnessnames[.x], 
                    color="Precipitation and snowmelt", linetype="Year", shape="Year")+
               theme_bw()  + scale_color_manual(values=water_snow_pal) +
               theme(axis.title.x=element_blank())))
```

```{r mt_indiv_mod}
mt.mod.indiv <- expand_grid(trait=floraltraits, fitnesstrait=fitnesstraits, yr=2018:2020) %>% 
  filter(!(yr==2018 & trait=="sepal_width")) %>% 
  mutate(model = pmap(., function(trait, fitnesstrait, yr) {
    lm(relfitness ~ trait.z, 
       data=mnps.plantyr %>% mutate(trait.z = mnps.plantyr[[trait]] / alltraits.sd[[trait]]) %>% 
         rename("fitness" = fitnesstrait) %>% 
         filter(year==yr) %>% mutate(relfitness = fitness/mean(fitness, na.rm=T)))}), # mean fitness in each yr
    coef = map(model, tidy))
    #test = map(model, ~tidy(car::Anova(.x), type=3)),
    #emm = map(model, ~summary(emtrends(.x, ~1, var="trait.z")))) 

#get the the standardized selection differential S' (change in relative fitness / standardized trait value)
mt.mod.indiv %>% select(trait, fitnesstrait, yr, coef) %>% unnest(coef) %>% filter(term=="trait.z") %>% 
  ggplot(aes(x=traitnames[trait], color=factor(yr), y=estimate, ymin=estimate-std.error, ymax=estimate+std.error)) + 
  facet_wrap(vars(fitnesstrait), labeller = as_labeller(fitnessnames)) + 
  geom_hline(yintercept=0) + geom_pointrange() + coord_flip() + scale_color_manual(values=year_pal) + 
  labs(y="Standardized selection differential S'", color="Year") + theme_bw() + theme(axis.title.y=element_blank())
```

# Selection gradients from PCA

Following method of [Chong et al. 2018](https://doi.org/10.1002/evl3.63) for reconstructing selection estimates from estimates of selection on PCs: Beta = E * A.

* PCA can't handle missing values so they are removed, but this biases the data and reduces sample size - multiple imputation?

## PCA of each flower
```{r pca}
mt.morph <- mt[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","VWC")] %>% drop_na(1:5) #drop sepal_width
mt.pca <- prcomp(mt.morph[,-(6:7)], scale.=TRUE)

mt.morph.pca <- mt.morph %>%
  bind_cols(as.data.frame(mt.pca$x)) %>% 
  tibble::rowid_to_column("index") %>% 
  mutate(x_0=0,y_0=0,
         x_1=-corolla_width/2,y_1=corolla_length,
         x_2= corolla_width/2,y_2=corolla_length) %>% 
  pivot_longer(c(starts_with("x"),starts_with("y")), names_to=c(".value", "point"), names_sep="_")

scl <- 100
zoom <- 5.5
mt.pca.vars <- as.data.frame(mt.pca$rotation) %>% rownames_to_column("variable")
mt.pca.labs <- paste0("PC", 1:2, " (", round(100*mt.pca$sdev[1:2]^2/sum(mt.pca$sdev^2), 0), "% explained var.)") 
ggplot(mt.morph.pca) + theme_classic() +
  geom_polygon(aes(x= x/scl+PC1, y= y/scl+PC2, group=index, color=VWC, fill=VWC)) +
  #geom_point(aes(x=PC1, y=PC2, size=sepal_width/2), color="darkgreen") + 
  geom_point(aes(x=PC1, y=PC2 + anther_min/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + anther_max/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + style_length/scl), color="grey60", size=0.5, shape=4) +
  geom_segment(data=mt.pca.vars,aes(x=0, y=0, xend=PC1*zoom, yend=PC2*zoom), 
               arrow = arrow(length = unit(1/2, "picas")), color="black") +
  geom_text(data =mt.pca.vars, aes(label = gsub("_"," ",variable), x = PC1*zoom*1.4, y = PC2*zoom*1.4,
                                   angle =  (180/pi) * atan(PC2/PC1), hjust = 0), color="black")+
  scale_size_identity() + coord_fixed() + 
  scale_fill_viridis_c(direction=-1, option="plasma") + scale_color_viridis_c(direction=-1, option="plasma") + 
  labs(x=mt.pca.labs[1], y=mt.pca.labs[2])
ggsave("images_selection/mt_morph_pca_noSW.svg", height=15, width=20)

zoom<-2
ggplot(mt.morph.pca) + geom_point(aes(x=PC1, y=PC2, color=corolla_width/corolla_length)) +
  geom_segment(data=mt.pca.vars,aes(x=0, y=0, xend=PC1*zoom, yend=PC2*zoom), 
               arrow = arrow(length = unit(1/2, "picas")), color="white") +
  geom_text(data =mt.pca.vars, aes(label = gsub("_"," ",variable), x = PC1*zoom*1.1, y = PC2*zoom*1.1,
                                   angle =  (180/pi) * atan(PC2/PC1), hjust = 0), color="white")+
  scale_color_viridis_c(option = "plasma")+ theme_dark()+
  theme(panel.grid=element_blank()) + labs(color="width/length") #+ coord_fixed()
```

## Gradients from averaged traits

```{r gradients, fig.width=8, fig.height=6}
morphtraits.pca <- morphtraits[-6]  #remove sepal_width (not measured in 2018)
#Standardize traits by mean and fitness by mean within each year
mnps.plantyr.std <- mnps.plantyr %>% select(all_of(morphtraits.pca),all_of(fitnesstraits), any_of(names(treatments))) %>%
  mutate(across(all_of(morphtraits.pca), ~ (.x-mean(.x, na.rm=T))/sd(.x, na.rm=T))) %>% 
  drop_na(all_of(morphtraits.pca)) 

# Run PCA on traits
mnps.PCA <- mnps.plantyr.std %>% select(all_of(morphtraits.pca)) %>% prcomp() #TODO check if these need to be scaled (scale. arg)
summary(mnps.PCA) # PCs 1 - 4 explain ~90% of the variation
mnps.PCA.E <- mnps.PCA$rotation[,1:4]
mnps.plantyr.std.PCA <- bind_cols(mnps.plantyr.std, as.data.frame(mnps.PCA$x))

calculate_beta <- function(E, slopes, resid.df) {
  data.frame(B = E %*% slopes$A,
             SE = sqrt(E ^ 2 %*% slopes$SE ^ 2)) %>% 
    mutate(p.value = 2*pt(abs(B)/SE, df=resid.df, lower.tail=F)) %>% #calculate t statistic from difference in means and SE, then look up one-tailed p in the upper tail, double to get a 2-tailed test
    rownames_to_column("trait")
}

selection_pca <- function(year, water=names(water_pal), snow=names(snow_pal), fitnesstrait = "seeds_est") {
  mnps.std.PCA <- mnps.plantyr.std.PCA %>% 
    filter(year == .env$year, water %in% .env$water, snow %in% .env$snow) %>%
    select(all_of(fitnesstrait), any_of(names(treatments)), starts_with("PC")) %>%
    rename("fitness" = fitnesstrait) %>%
    group_by(year) %>% mutate(relfitness = fitness/mean(fitness, na.rm=T), .keep="unused") %>% ungroup %>%
    drop_na(relfitness) %>%
    select(where(~sum(is.na(.))==0))

  # Regress relative fitness on PCs
  mnps.PCA.regression <- lm(relfitness ~ PC1 + PC2 + PC3 + PC4, data = mnps.std.PCA)
  mnps.PCA.slopes <- tidy(mnps.PCA.regression) %>% filter(term != "(Intercept)") %>% rename(A=estimate, SE=std.error)
  
  # Convert selection gradients on PCs to gradients on traits
  calculate_beta(mnps.PCA.E, mnps.PCA.slopes, aov(mnps.PCA.regression)$df.residual)
}

plot_beta <- function(betas, colorby="year") {
  bs <- betas
  if(colorby=="water_snow")  bs <- bs %>%  mutate(water_snow = paste(water, snow))
  ggplot(bs, aes(y=morphnames[trait], x = B, xmin=B-SE, xmax=B+SE,alpha=p.value<0.05)) + 
    geom_vline(xintercept=0)+ 
    geom_pointrange(mapping=aes_string(color=colorby), position=position_dodge(width=0.5)) + 
    labs(x = "Selection gradient \U03B2 estimated from gradients on PCs", y = "Trait",
         color=list(year="Year", water="Precipitation", snow="Snowmelt", 
                    water_snow = "Precipitation and snowmelt")[[colorby]]) +
    scale_color_manual(values=list(year=year_pal, water=water_pal, snow=snow_pal,
                                   water_snow=water_snow_pal)[[colorby]]) +
    theme_bw()
}

mnps.beta.yr <- expand_grid(year=unique(mnps.plantyr$year), fitnesstrait=fitnesstraits)  %>% 
 mutate(betas = pmap(., selection_pca)) %>% unnest(betas)
plot_beta(mnps.beta.yr)+ facet_wrap(vars(fitnesstrait), nrow=1, labeller=as_labeller(fitnessnames))

mnps.beta.water <- expand_grid(year=unique(mnps.plantyr$year), water=names(water_pal), fitnesstrait=fitnesstraits)  %>% 
 mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

mnps.beta.snow <- expand_grid(year=unique(mnps.plantyr$year), snow=names(snow_pal), fitnesstrait=fitnesstraits)  %>% 
 mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

mnps.beta.water.snow <- expand_grid(year=unique(mnps.plantyr$year), water=names(water_pal), snow=names(snow_pal),  fitnesstrait=fitnesstraits)  %>% 
  mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

walk(fitnesstraits, ~print(plot_beta(filter(mnps.beta.water.snow, fitnesstrait==.x), colorby="water_snow") + 
                             facet_wrap(vars(year), nrow=1) + ggtitle(fitnessnames[.x])))
```

# Adaptive plasticity


```{r adaptive, fig.height=5, fig.width=7}
adaptive.plasticity.data <- mnps.plantyr %>% pivot_longer(all_of(morphtraits)) %>% drop_na(value) %>% group_by(name, year) %>% nest() %>% 
  mutate(model = map(data, ~lm(value ~ snow + water, data=.)), # + (1|plotid) + (1|snow:plot)
         emm = map(model, ~ summary(emmeans(.x, ~ snow + water))%>% filter(water=="Control", snow=="Normal")), 
         test =  map(model, ~tidy(multcomp:::summary.glht(multcomp::glht(.x, 
                           linfct = multcomp::mcp(water = c("Addition - Control = 0",
                                                            "Reduction - Control = 0",
                                                            "Reduction - Addition = 0"),
                                                  snow = c("Early - Normal = 0"))), multcomp::adjusted(type="none"))))) %>%#adjust later 
  select(year, name, emm, test) %>% unnest(c(emm, test)) %>% 
  mutate(estimate=estimate/emmean, std.error=std.error/emmean, #divide estimated difference in means by emm in Normal Control subplots
         contrast = str_replace(str_remove_all(contrast, "[a-z\\s]"),"-","_"),
         newenv = recode(contrast, "A_C"="Addition", "R_C"="Reduction","E_N"="Early"),
         multiplecomps = contrast %in% c("A_C","R_C")) %>% #these two reuse the same control data
  group_by(multiplecomps, name, year) %>% #two comparisons to Control made within each volatile and year 
  mutate(adj.p.value = ifelse(multiplecomps, p.adjust(p.value, method="fdr"), p.value)) %>% 
  ungroup %>%
  select(year, name, newenv, adj.p.value, estimate, std.error) %>% 
  left_join(bind_rows(rename(mnps.beta.water, newenv=water),
                    rename(mnps.beta.snow, newenv=snow)) %>% 
            filter(fitnesstrait!="eggs_per_flower") %>% rename(name=trait)) %>% 
  drop_na(fitnesstrait) #TODO investigate NAs in fitnesstrait - why not computed in mnps.beta.water for some sepal_widths?

envnames <- c(Addition="additional precipitation",
             Reduction="reduced precipitation",
             Early="early snowmelt")

library(ggpubr)
adaptive.plasticity.data %>% filter(newenv!="R_A") %>% 
  mutate(fitnesstrait=factor(fitnesstrait, levels = rev(fitnesstraits))) %>% 
  group_by(newenv) %>% nest() %>% 
  pwalk(function(newenv, data) { 
    print(ggplot(data, aes(x=estimate, y=B, color=morphnames[name])) + 
            labs(x=paste("Trait change with",envnames[newenv]), 
                 y=paste("Selection gradient \U03B2 under",envnames[newenv]), color="Trait") + 
            facet_grid(rows=vars(year), cols=vars(fitnesstrait), scales="fixed", 
                       labeller=as_labeller(c(fitnessnames, set_names(levels(treatments$year)))))+
            geom_vline(xintercept=0) + geom_hline(yintercept=0) +
            annotate("rect", xmin = 0, xmax = Inf, ymin = 0, ymax = -Inf, fill= "black", alpha=0.1) + 
            annotate("rect", xmin = 0, xmax = -Inf, ymin = Inf, ymax = 0, fill= "black", alpha=0.1) +
            geom_errorbar(aes(ymin=B-SE, ymax=B+SE)) + 
            geom_errorbarh(aes(xmin=estimate-std.error, xmax=estimate+std.error)) +
            geom_point(size=2) + 
            scale_color_brewer(palette = "Dark2")+
            scale_x_continuous(labels=scales::percent, limits=c(NA,NA), breaks=c(-4:4)/4) + 
            theme_bw() + theme(axis.text = element_text(size=7)))})

ggplot(adaptive.plasticity.data %>% filter(newenv!="R_A") %>% drop_na(fitnesstrait),
       aes(x=estimate, xmin=estimate-std.error, xmax=estimate+std.error, y=morphnames[name], color=year, alpha=adj.p.value < 0.05)) +  
  facet_wrap(vars(newenv)) +
  geom_vline(xintercept = 0) + geom_pointrange() + scale_color_manual(values=year_pal) + 
  theme_bw() + theme(panel.grid.major.y=element_blank(), axis.title.y=element_blank()) +
  labs(x="Plastic change in new environment compared to control", color="Year")
```

# Long-term selection

```{r mt_selection_long, fig.height=6, fig.width=10}
#from /MyDocs/MEGA/Documents/RMBL 2013/Project/Data/flrtraits_master3_2013.xls
#TODO get raw weather data from RMBLWeatherData_2010.xls (1975-2010)
pg <- read_csv("data/PG_flrtraits_2013.csv") %>% 
  mutate(type2 = case_match(type, c("AA","agg")~"agg", "TT"~"ten", .default="hyb"))
floraltraits.pg <- c(floraltraits[c(1,2,3,5,8)], "redness")
traitnames.pg <- c(traitnames, redness="Redness")
alltraits.sd.pg <- map_dbl(floraltraits.pg, ~sd(pg[[.x]], na.rm=T)) %>% set_names(floraltraits.pg)

#TODO 2005 and 2010 have less than 10 plants!
pg %>% count(year,type2) %>% pivot_wider(names_from=type2, values_from=n, values_fill = 0) %>%
  mutate(n = agg+hyb+ten) %>% kable(caption = "number of plants of each species")

pg %>% pivot_longer(all_of(c(floraltraits.pg, "seeds_est"))) %>% 
  ggplot(aes(y=value, x=meltday, color=type2)) + 
  facet_wrap(vars(name), scales="free_y", labeller=as_labeller(traitnames.pg), nrow=2) +
  geom_point() + geom_smooth(method="lm", se=F) + 
  theme_bw() + theme(axis.title.y=element_blank())

pg %>% pivot_longer(all_of(floraltraits.pg)) %>% 
  ggplot(aes(y=seeds_est, x=value, color=factor(year))) + 
  geom_point(size=0.5) + geom_smooth(method="lm", se=F) + 
  facet_wrap(vars(name), scales="free_x", labeller=as_labeller(traitnames.pg), nrow=2) + 
  labs(y=fitnessnames["seeds_est"], color="Year")+
  theme_bw() + theme(axis.title.x=element_blank())

pg.climate <- pg %>% distinct(year, snowfall, octmay, junjul, meltday)
with(pg.climate, octmay - snowfall) #almost the same - units? maybe mm water?
ggplot(pg.climate, aes(x=snowfall, y=meltday)) + geom_point() + geom_smooth(method="lm")
```

## Selection in each year

```{r mt_indiv_mod_long}
mt.mod.indiv.pg <- pg %>% pivot_longer(all_of(floraltraits.pg), names_to="trait") %>% 
  drop_na(value) %>% distinct(trait, yr=year) %>% 
  mutate(fitnesstrait="seeds_est") %>% 
  mutate(model = pmap(., function(trait, fitnesstrait, yr) {
    lm(relfitness ~ trait.z, 
       data=pg %>% mutate(trait.z = pg[[trait]] / alltraits.sd.pg[[trait]]) %>% 
         rename("fitness" = fitnesstrait) %>% 
         filter(year==yr) %>% mutate(relfitness = fitness/mean(fitness, na.rm=T)))}), # mean fitness in each yr
    coef = map(model, tidy))
    #test = map(model, ~tidy(car::Anova(.x), type=3)),
    #emm = map(model, ~summary(emtrends(.x, ~1, var="trait.z")))) 

#get the the standardized selection differential S' (change in relative fitness / standardized trait value)
mt.mod.indiv.slopes.pg <- mt.mod.indiv.pg %>% select(trait, fitnesstrait, year=yr, coef) %>% 
  unnest(coef) %>% filter(term=="trait.z") %>% left_join(pg.climate)

ggplot(mt.mod.indiv.slopes.pg, aes(x=traitnames.pg[trait], color=factor(year), y=estimate, ymin=estimate-std.error, ymax=estimate+std.error)) + 
  facet_wrap(vars(fitnesstrait), labeller = as_labeller(fitnessnames)) + 
  geom_hline(yintercept=0) + geom_pointrange() + coord_flip() +  
  labs(y="Standardized selection differential S'", color="Year") + 
  theme_bw() + theme(axis.title.y=element_blank())

#TODO the corolla length plot doesn't quite match Fig 3B in Campbell & Powers 2015
ggplot(mt.mod.indiv.slopes.pg , aes(x=meltday, y=estimate, ymin=estimate-std.error, ymax=estimate+std.error)) +
    facet_wrap(vars(trait), labeller = as_labeller(traitnames.pg)) + 
  geom_smooth(method="lm", color="grey70", alpha=0.1) + 
  geom_hline(yintercept=0) + geom_linerange(color="grey70") + geom_text(aes(label=str_sub(year,3,4))) + 
  labs(y="Standardized selection differential S'", x="Snowmelt date") + theme_bw()

mt.mod.indiv.slopes.pg %>% group_by(trait) %>%  nest() %>% 
  mutate(model = map(data, ~lm(estimate ~ meltday, data=.)),
         test = map(model, ~tidy(anova(.)))) %>% 
  select(-data, -model) %>% unnest(test) %>% filter(term=="meltday") %>% kable(caption="selection differential vs. meltdate")
```

# Selection by snowmelt date

```{r mt_indiv_snow, fig.width=14, fig.height=10}
datasetnames <- c(max="Maxfield Meadow", pg="Poverty Gulch")
max.pg.std <- bind_rows(.id="dataset", pg=filter(pg, type2!="ten"), 
                        max=mutate(mnps.plantyr, site="max", year=as.numeric(year), meltday=sun_date)) %>%
  pivot_longer(any_of(c(floraltraits.pg, floraltraits))) %>% 
  select(dataset, year, site, plant, type, type2, seeds_est, meltday, name, value) %>% 
  group_by(site, year, name) %>% drop_na(value, seeds_est) %>%
  mutate(relfitness = seeds_est/mean(seeds_est), #relative fitness in each year for the plants with that trait measured
         stdtrait = (value-mean(value))/sd(value)) #mean-standardized trait

meltbins <- 4
ggplot(max.pg.std, aes(y=relfitness, x=stdtrait, color=cut(meltday,meltbins))) + 
  facet_grid(dataset ~ name, labeller = as_labeller(c(traitnames.pg, datasetnames))) + 
  geom_point(size=0.5) + geom_smooth(method="lm", se=F, linewidth=2) +
  coord_cartesian(xlim=c(-3,3), ylim=c(0, 5))+#some outliers not shown
  scale_color_brewer(palette="Oranges", direction=-1) + scale_y_continuous(expand=expansion(c(0,0.05)))+
  labs(x="Mean-centered and SD-scaled trait", y="Relative fitness (total seeds)", color = "Snowmelt date") +
  theme_dark() + theme(legend.position = "top")
```

```{r mt_indiv_snow_mod}
ggplot(max.pg.std, aes(x=meltday, fill=cut(meltday, meltbins))) + geom_histogram(binwidth=1)+
  facet_wrap(vars(dataset), labeller = as_labeller(datasetnames)) + 
  scale_fill_brewer(palette="Oranges", direction=-1, guide="none") + theme_dark() + 
  labs(x="Snowmelt date", y="Number of observations")

max.pg.std.mod <- max.pg.std %>% group_by(dataset, name) %>% nest() %>%
  mutate(model = map(data, ~lm(relfitness ~ stdtrait * meltday, data=.)),
         coef = map(model, ~tidy(.x))) %>%
  select(-data, -model) %>% unnest(coef) %>% filter(term!="(Intercept)")

ggplot(max.pg.std.mod, aes(y=traitnames.pg[name], x=estimate, shape=p.value<0.05, color=datasetnames[dataset])) + 
  facet_wrap(vars(term), scales = "free_x") +
  geom_vline(xintercept = 0)+ scale_shape_manual(values=c(1,19), guide="none")+ 
  scale_color_brewer("", palette = "Set1")+
  geom_pointrange(aes(xmin=estimate-std.error, xmax=estimate+std.error)) + 
  theme_bw() + theme(legend.position = "top", axis.title.y=element_blank()) 
```

