--- 
title: "Maxfield Meadow Ipomopsis Volatiles"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>


```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)
library(vegan)
library(viridis)
library(ggvegan)

library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images_volatiles2020/", fig.height=8, fig.width=8)
options(digits=4) # for kables

load("data/maxfield_data.rda")
load("data/volatiles/maxfield_volatiles.rda")
```

# Sample size

```{r sampsize}
with(vt, table(year, type)) %>% kable
with(vt %>% filter(type=="floral"), table(year, water, snow))
```

# Soil moisture

```{r sm}
ggplot(vt %>% drop_na(water), aes(x=factor(date), y=VWC, color=water)) + facet_wrap(vars(year), ncol = 1, scales="free_x") +
  geom_boxplot() + scale_color_manual(values=water_pal) + scale_x_discrete(guide = guide_axis(angle=90)) + 
  labs(x="Date", y="Soil moisture (%VWC)", color="Precipitation")
```

# Multivariate analyses

```{r samplesizes}
volsmeta <- maxfmeta %>% mutate(year=factor(year(sampledate))) %>% 
  drop_na(plotid) %>% #cut out ambients
  filter(year %in% c("2018","2019","2020")) %>% droplevels() %>% left_join(treatments)
rownames(volsmeta) <- volsmeta$index
vols <- maxf[volsmeta$index,]

with(maxfmeta, table(type))
with(volsmeta, table(year, plotid))
with(volsmeta, table(water, snow, year))

volsmeta %>% drop_na(sampledate) %>% group_by(sampledate) %>% tally() %>% 
  ggplot(aes(x=yday(sampledate), y=n)) + geom_col() + facet_wrap(vars(year(sampledate)), ncol=1) +
  labs(x="VOC sampling date (day of year)", y="Floral samples")
# TODO fix early 2020 date? and check singletons in 2018, 2019
# TODO there are some duplicate file names in maxfmeta, especially sampled 190805
```

```{r filtering}
vols.cut <- vols[,colSums(vols)>1e7]

ipochems <- read_csv("data/volatiles/chemsf_ipo.csv")
ipogood <- ipochems %>% filter(verdict %in% c("bis14", "fvoc", "keep","keep_leaf","merge")) %>% pull(name)
vols.good <- vols[,intersect(colnames(vols),ipogood)]
```


## Heatmap

```{r heatmap}
library(viridis)
library(pheatmap)
log10(sum(vols.cut$`(1R)-2,6,6-Trimethylbicyclo[3.1.1]hept-2-ene`))
pheatmap(as.matrix(t(vols[,colSums(vols)>3e8]))^(1/4), color=mako(512),
         annotation_col=volsmeta %>% select(year))

pheatmap(as.matrix(t(vols.good))^(1/4), color=mako(512),
         annotation_col=volsmeta %>% select(year))
```

## Boxplots

```{r boxplots, fig.width=10, fig.height=10}
vols.good.top <- vols.good[,colSums(decostand(vols.good,"pa"))>81]#occur in more than N samples
bind_cols(volsmeta, vols.good.top) %>%  pivot_longer(-all_of(colnames(volsmeta))) %>% 
  ggplot(aes(x = year, color = water, y = value)) + geom_boxplot() + facet_wrap(vars(name), scales="free_y") + 
  scale_y_sqrt() + scale_color_manual(values=water_pal) + labs(x="Year", y ="Emissions (ng/hr)") + guides(color="none")
```

## Ordinations

### Absolute amounts

```{r hmb_ords}
#Nmds 
(scents.mds <- metaMDS(sqrt(scents.dat), autotransform=FALSE)) #had noshare = FALSE
plot(scents.mds,type="t",display=c("species"))
plot(scents.mds,type="t",display=c("sites"))

#PerMANOVA using adonis function in vegan package
(scents.permanova <- adonis(sqrt(scents.dat) ~ water * year, planttype.dat))

#Canonical analysis of principal coordinates (CAP) allows for other distance measures
(scents.cap <-capscale(sqrt(scents.dat) ~ year * water, planttype.dat, dist="bray"))
anova(scents.cap, by="term")

scents.cap <-capscale(sqrt(scents.dat) ~ year + water, planttype.dat, dist="bray")
plot(scents.cap, type="n", xlim=c(-2,2))
points(scents.cap, display = "sites", col=water_pal[planttype.dat$water], pch=as.integer(planttype.dat$year))
text(scents.cap, display="cn", labels=c("2018","2019", names(water_pal)), col=c("black","black",water_pal), font=2)

plot(scents.cap, type="n", xlim=c(-1,3), ylim=c(-0.5,0.5))
text(scents.cap, display="sp", cex=0.7)
```


### Relative amounts

```{r hmb_ords_rel}
#Nmds 
(scents.mds <- metaMDS(decostand(scents.dat, "hellinger"), autotransform=FALSE)) #had noshare = FALSE
plot(scents.mds,type="t",display=c("species"))
plot(scents.mds,type="t",display=c("sites"))

#PerMANOVA using adonis function in vegan package
(scents.permanova <- adonis(decostand(scents.dat, "hellinger") ~ water * year, planttype.dat))

#Canonical analysis of principal coordinates (CAP) allows for other distance measures
(scents.cap <-capscale(sqrt(scents.dat) ~ year * water, planttype.dat, dist="bray"))
anova(scents.cap, by="term")

scents.cap <-capscale(decostand(scents.dat, "hellinger") ~ year + water, planttype.dat, dist="bray")
plot(scents.cap, type="n", xlim=c(-2,2))
points(scents.cap, display = "sites", col=water_pal[planttype.dat$water], pch=as.integer(planttype.dat$year))
text(scents.cap, display="cn", labels=c("2018","2019", names(water_pal)), col=c("black","black",water_pal), font=2)

plot(scents.cap, type="n", xlim=c(-1.5,1.5), ylim=c(-0.5,0.5))
text(scents.cap, display="sp", cex=0.7)
```

# Analysis

Code from applying_standards_2019-NOV18.R

## Load data

```{r hmb}
library(reshape2)

maxcomp <- read.table("data/volatiles/Max.2018.2019_combined_USE_NOV08plantsreduced.tsv",T,"\t",stringsAsFactors=F)

# don't need to remove air samples because they are already removed
#maxcomp <- as.data.frame(t(compounds4[grep("Air_",row.names(compounds4),invert=T),]))

# convert all numbers to numeric: There are a lot of NAs
maxcomp[,-1] <- data.frame(apply(maxcomp[,-1],2,as.numeric))

names(maxcomp)[1] <- "sample"

# make long-format:
maxcomp2 <- reshape2::melt(maxcomp, variable.name = "sample", value.name = "amount")
names(maxcomp2)[2] <- "chem_name"

####now trying to do some data exploration with volatiles####

#import sites 
sites<-read.table("data/volatiles/maxfield.sitematch.NOV08.tsv",T,"\t")

# melt for analysis and plotting:
maxcomp3 <- merge(maxcomp2, sites, by = 'sample', sort = F)

#write.table(maxcomp3, file = "maxcomp3_2019-NOV17.csv", col.names = TRUE, row.names = T, sep = ",")

## remove offending compounds:i.e the one that I cannot find "what" it is. 
# maxcomp3 <- dplyr::filter(maxcomp3, chem_name != "2-Trimethylsilyl-oxy-2-4-trimethylsilyl-oxy-phenyl-ethanamine")

# log transform amount:
maxcomp3$log_amount <- sapply(maxcomp3$amount, function(x) log(x+0.01, 10) + 2)


```

## Histograms

```{r histograms}
# let's plot some histograms:
# first: raw data prior to transformation
# assign facet to data groups above zero:
ggplot(maxcomp3[maxcomp3$amount > 0, ], aes(x = amount)) + 
  geom_histogram(bins = 100) +
  theme_bw() +
  xlab("untransformed data (zeros not shown)")
  
ggplot(maxcomp3[maxcomp3$log_amount > 0, ], aes(x = log_amount)) + 
  geom_histogram(bins = 100) +
  theme_bw() +
  xlab("log-transformed data")


# save raw plots together:
#library(cowplot)
#plot_grid(raw_1,raw_2) %>% ggsave(filename = "raw_plots_v1.pdf", width = 6.5, height = 3.25)

```

## Clean data

```{r data_cleaning}
# re-casting, using log scale for amount as value var:
maxcomp4 <- dcast(maxcomp3, precip + sample + snow.removal + plot_ID ~ chem_name, value.var="log_amount", fun.aggregate = function(x) log(sum(10^(x - 2)), 10) + 2)

#write.table(maxcomp3, file = "maxcomp_3.csv", col.names = TRUE, row.names = T, sep = ",")

# define vector of compound_names
compound_names <- paste0(levels(maxcomp3$chem_name))

# find rows whose compounds sum to zero:
maxcomp4$compound_sum <- rowSums(maxcomp4[,names(maxcomp4) %in% compound_names])

samples_without_data <- paste0(maxcomp4$sample[maxcomp4$compound_sum==0])
maxcomp4 <- droplevels(maxcomp4[!maxcomp4$sample %in% samples_without_data, ])

# generate composite treatment by combining control conditions:
maxcomp4$treatment[grepl('add ', maxcomp4$precip)] <- 'add_H2O'
maxcomp4$treatment[grepl('rainout', maxcomp4$precip)] <- 'rainout'
maxcomp4$treatment[grepl('mock', maxcomp4$precip)] <- 'Control'
maxcomp4$treatment[grepl('Control', maxcomp4$precip)] <- 'Control'

maxcomp4$year[grepl('2018', maxcomp4$sample)] <- '2018'
maxcomp4$year[grepl('2019', maxcomp4$sample)] <- '2019'

# generate response matrix for all, 2018, and 2019:

Y18 <- as.matrix(maxcomp4 %>% dplyr::filter(year == '2018') %>% dplyr::select(compound_names))
Y19 <- as.matrix(maxcomp4 %>% dplyr::filter(year == '2019') %>% dplyr::select(compound_names))
YAA <- as.matrix(maxcomp4 %>% dplyr::select(compound_names))

if (!dim(YAA)[1] == (dim(Y18)[1] + dim(Y19)[1])) {
  # scream
}

# now remove compounds with zero abundances within Y18 and Y19
Y18 <- Y18[, colSums(Y18) > 0]
Y19 <- Y19[, colSums(Y19) > 0]

maxcomp4$treatment <- factor(maxcomp4$treatment)
```

## MANOVA

### Precipitation

```{r manova}
# MANOVA on treatment
manova.scents.18 <- manova(Y18 ~ maxcomp4$treatment[maxcomp4$year == '2018'])
manova.scents.19 <- manova(Y19 ~ maxcomp4$treatment[maxcomp4$year == '2019'])
manova.scents.YA <- manova(YAA ~ maxcomp4$treatment)

# combined year analysis
summary(manova.scents.YA, test = "Wilks")

# Df  Wilks approx F num Df den Df Pr(>F)
# maxcomp4$treatment   2 0.7257   1.0384     72    430 0.4002
# Residuals          250  

# 2018 alone
# combined year analysis
summary(manova.scents.18, test = "Wilks")

# Df   Wilks approx F num Df den Df Pr(>F)
# maxcomp4$treatment[maxcomp4$year == "2018"]   2 0.63655  0.92906     60    220 0.6235
# Residuals                                   139 

# 2019 alone
# combined year analysis
summary(manova.scents.19, test = "Wilks")

# Df   Wilks approx F num Df den Df Pr(>F)
# maxcomp4$treatment[maxcomp4$year == "2019"]   2 0.49212   1.0706     62    156 0.3622
# Residuals                                   108  
```

### Snowmelt

```{r snowmelt}
## SNOW MELT
# combined year analysis
# MANOVA on treatment
manova.snowremoval.18 <- manova(Y18 ~ maxcomp4$snow.removal[maxcomp4$year == '2018'])
manova.snowremoval.19 <- manova(Y19 ~ maxcomp4$snow.removal[maxcomp4$year == '2019'])
manova.snowremoval.YA <- manova(YAA ~ maxcomp4$snow.removal)

summary(manova.snowremoval.YA, test = "Wilks")
# Df   Wilks approx F num Df den Df Pr(>F)
# maxcomp4$snow.removal   1 0.85602   1.0091     36    216 0.4616
# Residuals             251    

summary(manova.snowremoval.18, test = "Wilks")
# Df   Wilks approx F num Df den Df Pr(>F)
# maxcomp4$snow.removal[maxcomp4$year == "2018"]   1 0.73786   1.3145     30    111 0.1549
# Residuals                                      140 

summary(manova.snowremoval.19, test = "Wilks")
# Df   Wilks approx F num Df den Df Pr(>F)
# maxcomp4$snow.removal[maxcomp4$year == "2019"]   1 0.69374    1.125     31     79 0.3307
# Residuals      
```

## CANDISC

### Precipitation

```{r candisc}
library(candisc)
scents.can.YA <- candisc(manova.scents.YA, data = maxcomp4, scores = TRUE)
scents.can.18 <- candisc(manova.scents.18, data = maxcomp4[maxcomp4$year == '2018', ], scores = TRUE)
scents.can.19 <- candisc(manova.scents.19, data = maxcomp4[maxcomp4$year == '2019', ], scores = TRUE)

#pdf(file = "canda_v1_2019-NOV-19_YA.pdf",width=8,height=4)
#par(mfrow=c(1,2))
plot(scents.can.YA, col=c("white","white","white"), conf = 0.95,lwd=0.5,var.cex=0.7,
           cex.axis=0.75,var.col="gray40",bty="n",alpha=0.5)
plot(scents.can.YA, col=c("gray60","darkorange2","dodgerblue"), pch=c(19,19,19), conf = 0.95,
           var.labels = NULL, var.lwd=0,var.col = NULL,
           cex.axis=0.75,bty="n")
#dev.off()

print("2018")
# pdf(file = "canda_v1_2019-NOV-19_2018.pdf",width=8,height=4)
# par(mfrow=c(1,2))
plot(scents.can.18, col=c("white","white","white"), conf = 0.95,lwd=0.5,var.cex=0.7,
            cex.axis=0.75,var.col="gray40",bty="n",alpha=0.5)
plot(scents.can.18, col=c("gray60","darkorange2","dodgerblue"), pch=c(19,19,19), conf = 0.95,
            var.labels = NULL, var.lwd=0,var.col = NULL,
            cex.axis=0.75,bty="n")
# dev.off()
# 
print("2019")
# pdf(file = "canda_v1_2019-NOV-19_2019.pdf",width=8,height=4)
# par(mfrow=c(1,2))
plot(scents.can.19, col=c("white","white","white"), conf = 0.95,lwd=0.5,var.cex=0.7,
           cex.axis=0.75,var.col="gray40",bty="n",alpha=0.5)
plot(scents.can.19, col=c("gray60","darkorange2","dodgerblue"), pch=c(19,19,19), conf = 0.95,
           var.labels = NULL, var.lwd=0,var.col = NULL,
            cex.axis=0.75,bty="n")
# dev.off()

```

```{r heatmap_gplots, eval=FALSE}
library(gplots)
library(RColorBrewer) 

# define colors:
cols <- c("gray60","darkorange2","dodgerblue")[as.numeric(maxcomp4$treatment)]
hmcol <- colorRampPalette(brewer.pal(9, "GnBu"))(100)
hclustfun <- function(X) hclust(X,method="median")

heatmap.2(t(scents.dat), labCol=maxcomp4$treatment,
          trace="none", 
          ColSideColors=cols, 
          col=hmcol,
          distfun = dist,
          hclustfun = hclust)
```

## CANDISC scores
```{r candisc2}

#scents.can1 <- candisc(manova.scents1, data = maxcomp4, scores = TRUE)
#scents.can
#scents.can1

#Scores are the standardized response variables times the raw canonical coefficients
#coef(scents.can.YA, type="raw")
#scents.can["scores"]

coef_test <- cbind(coef(scents.can.YA, type="raw"),
                   coef(scents.can.YA, type="structure"))

# correlation between raw and structure coefficients from scents.can
plot(coef_test[,1],coef_test[,3])
plot(coef_test[,2],coef_test[,4])

```

