--- 
title: "Maxfield Meadow Ipomopsis Experiment"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>

# Purpose
Analyze selection on and plastic responses of floral and vegetative traits of *Ipomopsis aggregata* to manipulations of snowmelt and summer precipitation. Determine whether plasticity is adaptive (improves survival and reproduction). Part of a [larger project](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1654655) to understand the potential for evolutionary rescue in alpine plants under climate change.

# Treatments

```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)

library(lme4)
library(lmerTest)
library(emmeans)
library(broom)
library(vegan)

library(RColorBrewer)
library(viridis)
library(GGally)

library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images/")

signif.num <- function(x) {
    as.character(symnum(x, corr = FALSE, na = FALSE, legend = FALSE,
           cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
           symbols = c("***", "**", "*", ".", " ")))
}

load("data/maxfield_data.rda")
```

# Snow melt vs. treatments

```{r hobo}
ggplot(hobo, aes(x=intensity, y=temp, color=hour(time))) +  geom_point(alpha=0.5, size=0.5) + geom_density_2d(color="black")+ 
  scale_x_log10(labels=function(x) format(x, scientific=F, big.mark=","), limits=c(10,1e6), breaks=10^(0:6)) +
  scale_color_gradientn(colours = rainbow(5))+
  geom_vline(xintercept = melt_threshold_sun) + 
  geom_hline(yintercept=melt_threshold_warm)+ geom_hline(yintercept=-melt_threshold_warm) +
  labs(x="Light intensity (lux)", y="Temperature (C)", color="Hour") +
  theme_dark() + theme(panel.grid.minor.x=element_blank())

ggplot(hobo %>% filter(hour(time)%in%5:17 & 
                  (year!="2019"&yday(time) %in% 110:130 | 
                   year=="2019"&yday(time) %in% 145:160)), 
       aes(x=intensity, y=temp, color=hour(time))) +   geom_point(alpha=0.7) +geom_density_2d(color="black")+
  scale_x_log10(labels=function(x) format(x, scientific=F, big.mark=","), limits=c(10,1e6), breaks=10^(0:6)) +
  scale_color_gradientn(colours = rainbow(5))+
  geom_vline(xintercept = melt_threshold_sun) + 
  geom_hline(yintercept=melt_threshold_warm)+ geom_hline(yintercept=-melt_threshold_warm) +
  labs(title="Temperature and light thresholds for snow melt", subtitle = "During the day during melt weeks",
       x="Light intensity (lux)", y="Temperature (C)", color="Hour") +
  theme_dark() + theme(panel.grid.minor.x=element_blank())

ggplot(hobo %>% filter(month(time)%in%4:6), aes(x=time, y=temp, color=plot)) + 
  facet_wrap(vars(year), scales="free_x", ncol=1) + 
  geom_hline(yintercept=melt_threshold_warm) + geom_hline(yintercept=-melt_threshold_warm) +geom_line() +
  geom_segment(data=meltdates, aes(x=warm_time, y=-10, xend=sun_time, yend=-15, color=plot), size=1) +
    geom_line(data=daily_temp %>% filter(year!="2017", month(date)%in%4:6), aes(x=as.POSIXct(date), y=av_temp_2m_C), inherit.aes = F, color=brewer.pal(3, "Oranges")[3]) +
  labs(x="Date",y="Temperature (C)", color="Plot")
  
ggplot(hobo %>% filter(month(time)%in%4:6), aes(x=time, y=intensity, color=plot)) + 
  facet_wrap(vars(year), scales="free_x", ncol=1) + 
  geom_line() + geom_hline(yintercept=melt_threshold_sun) +
  geom_segment(data=meltdates, aes(x=sun_time, y=-2e4, xend=melt_time, yend=-4e4, color=plot), size=1) +
  labs(x="Date",y="Light intensity (lux)", color="Plot")

meltdates %>% select(c("year","plot", "snow", ends_with("date"))) %>% pivot_longer(ends_with("date")) %>% arrange(name) %>% 
  ggplot(aes(y=name, x=value, color=plot, group=plot, linetype=snow)) + facet_wrap(vars(year), ncol=1) +
  geom_path(size=1) + #geom_text(aes(label=plot), hjust=1.5) +
  scale_linetype_manual(values=c(4,1)) + scale_x_continuous(minor_breaks=110:160)+
  labs(x="Day of year",y="Criteron", color="Plot", linetype="Snowmelt")
       
meltdates %>% ggplot(aes(x=melt_offset, fill=snow)) + geom_dotplot(method="histodot", binwidth=1, color="white") +
  facet_wrap(vars(year), ncol=1)+ scale_fill_manual(values=snow_pal) +
  scale_y_continuous(expand=expansion(0,0))+ scale_x_continuous(breaks=-10:4)+
  theme(panel.grid.minor = element_blank()) +
  labs(x="Offset from mean normal snow melt date (days)", y="Number of plots", fill="Snowmelt")

left_join(sm.subplotmonth, meltdates) %>% 
  mutate(mo = paste0("Month ",mo,", ",month.abb[mo])) %>% 
  ggplot(aes(x=sun_date, y=VWC, shape=year,color=water)) + 
  facet_wrap(vars(mo), nrow=1)+
  geom_point() + geom_smooth(se=F, method="lm") +
  scale_color_manual(values=water_pal) +
  labs(x="Snowmelt date", y="Mean soil moisture in subplot", color="Water", shape="Year")
```

# Weather data
```{r weather}
ggplot(daily_precip, aes(y=precip_mm, x=julian)) + geom_col(fill=brewer.pal(3, "Greens")[3]) + facet_wrap(vars(year), ncol=1) + scale_y_continuous(limits=c(0,125)) + labs(x="Day of year", y="Precipiation (mm)") + theme_minimal()
```

# Soil moisture vs. treatments

```{r soil, fig.height=12, fig.width=12}
ggplot(sm, aes(x=yday(date), shape=snow, linetype=snow, color=water, y = VWC)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_jitter(height=0, width=0.3) + geom_smooth(se=F, span=0.5) +
  geom_col(data=daily_precip %>% filter(year!="2017"), aes(x=julian, y=precip_mm), inherit.aes = F, fill=brewer.pal(3, "Greens")[3]) + 
  geom_line(data=daily_temp %>% filter(year!="2017"), aes(x=julian, y=av_temp_2m_C), inherit.aes = F, color=brewer.pal(3, "Oranges")[3]) + 
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), inherit.aes=F, color=brewer.pal(3, "Purples")[3])+
  scale_x_continuous(limits=c(110, 230), breaks=scales::breaks_extended(10), expand=expansion(0,0)) + scale_y_continuous(limits=c(-3, 20)) +
  labs(y="Volumetric water content (%) / Daily precipitation (mm) / Mean temperature (C)", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))
```

# Census

The annual census at the beginning of the summer contains information on mortality, flowering status, and plant size (longest leaf and number of leaves for each rosette).

## Load census data

Tally the number of each status combination for each year. 

```{r load_census}
nrf_count <- bind_rows(
  allowed_nrf %>% filter(year==2019) %>% left_join(cen19 %>% mutate(nrf=paste(notes, rosettes, flowering)) %>% group_by(nrf) %>% tally()),
  allowed_nrf %>% filter(year==2020) %>% left_join(cen20 %>% mutate(nrf=paste(notes, rosettes, flowering)) %>% group_by(nrf) %>% tally()))

kable(nrf_count %>% select(-nrf) %>% arrange(year, status, desc(n)))
```

## Duplicates
When there is more than one record for a plant tag in the census, which duplicate should be kept? 
Currently resolved by the following rules:

1. sort by status (keep flowering > vegetative > recheck > dead_nf > tagnf)
2. sort by date within status (if statuses are the same, keep the observation with the newest date)
3. keep the first row and discard the other duplicates

This is automatic, but it may be better to resolve duplicates manually in some cases. The duplicates are written out to [spreadsheets](https://docs.google.com/spreadsheets/d/1vuQXBF8fCOjw-mYFLBk7Pm_iRUzSAZ8QmBCSJtt6IhA/edit#gid=367829026)

```{r dupes}
#write out the duplicates
#cen19 %>% group_by(id) %>% filter(n()>1) %>% arrange(id, status, desc(date)) %>% 
#  write_sheet(ss=filter(datasheets, name=="Maxfield Results"), sheet="2019duplicates")
#cen20 %>% group_by(id) %>% filter(n()>1) %>% arrange(id, status, desc(date)) %>% 
#  write_sheet(ss=filter(datasheets, name=="Maxfield Results"), sheet="2020duplicates")

#distinct() keeps the first row after sorting, discards the other rows
cen19 <- cen19 %>% arrange(id, status, desc(date)) %>% distinct(id, .keep_all = TRUE)
cen20 <- cen20 %>% arrange(id, status, desc(date)) %>% distinct(id, .keep_all = TRUE)
```

## Status vs. treatment
```{r inventory}
ggplot(drop_na(cen19, water, snow) %>% filter(status %in% status_ok), aes(x=water, fill=status)) + facet_wrap(vars(snow), labeller=labeller(.cols=label_both)) + geom_bar(position="fill") + scale_fill_manual(values=status_pal) + labs(title="2019 status", x="Precipitation", y="Proportion", fill="Status")
ggplot(drop_na(cen20, water, snow) %>% filter(status %in% status_ok), aes(x=water, fill=status)) + facet_wrap(vars(snow), labeller=labeller(.cols=label_both)) + geom_bar(position="fill") + scale_fill_manual(values=status_pal) + labs(title="2020 status", x="Precipitation", y="Proportion", fill="Status")
```

## Join 2019 and 2020 datasets

Check these tags missing in each year to make sure tags were not mis-read. When tags were not found in the other year, the part of the row for that year is "NA".

Recorded in 2019 but not in 2020:
```{r compare_years19}
sort(setdiff(cen19[cen19$notes != "tagnf",]$id, cen20$id))
```

Recorded in 2020 but not in 2019:
```{r compare_years20}
sort(setdiff(cen20[cen20$notes != "tagnf",]$id, cen19$id))
```

## Status transitions

What did plants of a given status in 2019 look like now in 2020?
```{r transitions}
with(cen, table(`2019`=paste(status_19), `2020`=paste(status)))

ggplot(filter(cen, status_19 %in% status_ok, status %in% status_ok), aes(x=status_19, fill=status)) + geom_bar(position="fill") + scale_fill_manual(values=status_pal) + labs(fill="2020 Status", y="Proportion", x="2019 Status")
```

## Survival and flowering vs. plant size
```{r surv_size}
ggplot(filter(cen, status %in% status_ok), aes(x=r1_longest_19, fill=status)) + facet_wrap(vars(status), ncol=1) +
  geom_density() + scale_fill_manual(values=status_pal) + labs(x="2019 Rosette 1 Longest Leaf (mm)", fill="2020 status")
ggplot(filter(cen, status %in% status_ok), aes(x=r1_leaves_19, fill=status)) + facet_wrap(vars(status), ncol=1) +
  geom_density() + scale_fill_manual(values=status_pal) + labs(x="2019 Rosette 1 # Leaves", fill="2020 status")
```

## Plant growth vs. treatments

Only includes plants that remained vegetative and did not die or flower.
```{r growth}
ggplot(drop_na(cen, water, snow), aes(x=r1_longest_19, y=r1_longest, color=water, shape=snow)) + geom_point() + geom_abline(intercept=0, slope=1) + scale_color_manual(values=water_pal) + theme_minimal()+ labs(x="2019 Rosette 1 Longest Leaf (mm)", y="2020 Rosette 1 Longest Leaf (mm)",  color="Precipitation", shape="Snowmelt")
ggplot(drop_na(cen, water, snow), aes(x=r1_leaves_19, y=r1_leaves, color=water, shape=snow)) + geom_point() + geom_abline(intercept=0, slope=1 )+ scale_color_manual(values=water_pal) + theme_minimal()+ labs(x="2019 Rosette 1 # Leaves", y="2020 Rosette 1 # Leaves",  color="Precipitation", shape="Snowmelt")
```

## Plant size vs. treatments
```{r plantsize}
ggplot(drop_na(cen19, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2019 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen19, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal)+ labs(x="Precipitation", y="2019 Rosette 1 # Leaves",  color="Snowmelt")

ggplot(drop_na(cen20, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2020 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen20, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2020 Rosette 1 # Leaves",  color="Snowmelt")
```

# Leaf traits

```{r checkleaftraits}
ggplot(lt, aes(x=wet_weight_g, y=dry_weight_g)) + geom_point() + geom_abline(slope=1, intercept = 0) + geom_abline(slope=1-median(lt$water_content), intercept = 0)
ggplot(lt, aes(x=leaf_area_cm2, y=dry_weight_g)) + geom_point() + geom_smooth(method="lm")
ggplot(lt, aes(x=leaf_area_cm2, y=wet_weight_g)) + geom_point() + geom_smooth(method="lm")
```


```{r leaftraits}
ggplot(lt, aes(x=water, color=snow, y = sla)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="Specific Leaf Area (cm2/g)", color="Snowmelt") + facet_wrap(vars(year, round), nrow=1) + scale_y_continuous(limits=c(0,300))

ggplot(lt, aes(x=sla, y=water_content, color=water, shape=snow, linetype=snow)) + geom_point() +  scale_color_manual(values=water_pal) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(lt, aes(x=water, color=snow, y = water_content)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="Water content", color="Snowmelt") + facet_wrap(vars(year, round), nrow=1)

ggplot(lt, aes(x=water, color=snow, y = trichome_density)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="Trichome density (/cm2)", color="Snowmelt") + facet_wrap(vars(year, round), nrow=1)
```

# Floral traits

## Floral morphology and nectar

### Floral trait correlations
```{r correlations}
traits <- c("corolla_length", "corolla_width", "style_length", "sepal_width", "nectar_24_h_ul", "nectar_conc","nectar_sugar_24_h_mg") #exclude anthers
traitnames <- setNames(c("Corolla Length (mm)", "Corolla Width (mm)", "Style Length (mm)", "Sepal Width (mm)", "Nectar Volume (uL/day)", "Nectar Concentration (% sucrose by mass)", "Nectar Sucrose (mg/day)"), traits)

ggpairs(mt %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width))

ggplot(mt, aes(x=nectar_24_h_ul, y = nectar_conc)) + geom_point() + labs(x="Nectar volume (uL / day)", y="Nectar concentration (% sucrose by mass)") + geom_smooth(se=F)
```

### Floral plasticity: water treatments and snowmelt treatments

This is a split-plot analysis, based on the original experimental design. The models use:

* *mt.subplot*, the 24 * 3 subplot means within each year
* *snow*, the snowmelt treatment (except for plot 2 in 2019)
* *water*, the water treatments with mock rainouts and controls combined
* *plot* as a random effect (subplots vary within this larger unit)

The models are specified as:

    lmer(trait ~ year * snow * water + (1|plot), data = mt.subplot)

* The models for nectar volume and total sugar are singular (variance at plot level near 0)
* Removing the interactions with year does not make those two models not singular
* The model fits are mostly singular when a random effect of *plotid* (subplot) is added for repeated measures among years

```{r morphology_split}
options(contrasts = c("contr.sum", "contr.poly"))
mt.mod <- mt.mod.test <- mod.emm <- mt.mod.plot <- list()
for(tr in traits) {
  mt.mod[[tr]] <- lmer(mt.subplot[[tr]] ~ year * snow * water + (1|plot), data=mt.subplot)
  mt.mod.test[[tr]] <- anova(mt.mod[[tr]])
  mod.emm[[tr]] <- summary(emmeans(ref_grid(mt.mod[[tr]]), ~ year + snow * water))
  
  mt.mod.plot[[tr]] <- 
    ggplot(mod.emm[[tr]], aes(x=water, color=snow, y = emmean)) + facet_wrap(vars(year), nrow=1) +
        labs(x="Precipitation", y=traitnames[[tr]], color="Snowmelt", shape="Year") +
        geom_point(data=mt.subplot, aes_string(x="water", color="snow", y = tr), inherit.aes = F, position=position_dodge(0.7)) + 
        geom_point(position=position_dodge(0.7), size=4, shape=0) + 
        geom_linerange(aes(ymin=emmean-SE, ymax=emmean+SE), position=position_dodge2(0.6)) +
        guides(x=guide_axis(angle = 90)) + scale_color_manual(values=snow_pal)
  print(mt.mod.plot[[tr]])
  cat(paste(tr,"\n"))
  print(mt.mod.test[[tr]])
}
mt.mod.tests <- map_dfr(mt.mod.test, tidy, .id="trait")
mt.mod.tests %>% select(c("trait", "term", "p.value")) %>% mutate(p.value = signif.num(p.value)) %>% 
  pivot_wider(names_from=trait, values_from=p.value) %>% 
  bind_rows(lapply(mt.mod, function(x) ifelse(isSingular(x),"Y","N"))) %>% mutate(term = replace_na(term,"Singular?")) %>% kable
```

### Floral plasticity: water treatments and snowmelt offset

The differences in snowmelt dates vary among years and plots, so this analysis uses a melt date offset. The models use:

* *mt.subplot*, the 24 * 3 subplot means within each year
* *water*, the water treatments with mock rainouts and controls combined
* *melt_offset*, the number of days between the snowmelt date (when the light entering the HOBO logger exceeds 10,000 lux) and the mean snowmelt date for the normal plots in that year (according to *snow*)
* *plotid* (subplot) as a random effect (plants vary within this larger unit)

The models are specified as: 

    lmer(trait ~ year * water * melt_offset + (1|plotid), data = mt.subplot)

```{r morphology_model}
options(contrasts = c("contr.sum", "contr.poly"))
mt.mod.raw <- mt.mod.plantyr <- mt.mod.subplot <- mt.mod.subplot.test <- mt.mod.plot <- list()
for(tr in traits) {
  #mt.mod.raw[[tr]]     <- lmer(mt[[tr]]         ~ year * water * melt_offset + (1|plotid) + (1|plantid), data=mt)
  #mt.mod.plantyr[[tr]] <- lmer(mt.plantyr[[tr]] ~ year * water * melt_offset + (1|plotid), data=mt.plantyr)
  mt.mod.subplot[[tr]] <- lmer(mt.subplot[[tr]] ~ year * water * melt_offset + (1|plot), data=mt.subplot)
  mt.subplot[[paste0("fit.",tr)]] <- NA
  mt.subplot[[paste0("fit.",tr)]][!is.na(mt.subplot[[tr]])]<- predict(mt.mod.subplot[[tr]], re.form=NA)
  mt.mod.subplot.test[[tr]] <- anova(mt.mod.subplot[[tr]])
  
  mt.mod.plot[[tr]] <- 
    ggplot(mt.subplot, aes_string(x="melt_offset", color="water", y = tr)) + geom_point() +
    geom_line(data=mt.subplot %>% drop_na(paste0("fit.",tr)), aes_string(y=paste0("fit.",tr)), size=2)+
     scale_color_manual(values=water_pal)+ facet_wrap(vars(year))+
        labs(x="Offset from mean normal snow melt date (days)", y=traitnames[[tr]], color="Precipitation")
  print(mt.mod.plot[[tr]])
}
mt.mod.subplot.tests <- map_dfr(mt.mod.subplot.test, tidy, .id="trait")
mt.mod.subplot.tests %>% select(c("trait", "term", "p.value")) %>% mutate(p.value = signif.num(p.value)) %>% 
  pivot_wider(names_from=trait, values_from=p.value) %>% 
  bind_rows(lapply(mt.mod.subplot, function(x) ifelse(isSingular(x),"Y","N"))) %>% mutate(term = replace_na(term,"Singular?")) %>% kable
```

### Floral plasticity: soil moisture

Both snowmelt timing and summer precipitation affect soil moisture and its timing. This analysis regresses the floral traits against the mean soil moisture in a subplot. Slopes may differ among years. The models use:

* *mt.plantyr*, the trait data averaged by plant within year
* *VWC*, the summer mean of the approximately weekly mean volumetric water content in a subplot. This may need to change to a subset of the summer because it includes soil moisture taken after traits were measured. See section on within-season variation for timing of measurements in each year.
* *plotid* (subplot) as a random effect (plants vary within this larger unit)

The models are specified as:

    lmer(trait ~ year * VWC + (1|plotid), data = mt.plantyr)

* The model for total sugar is singular (variance at subplot level near 0)

```{r morphology_sm}
options(contrasts = c("contr.sum", "contr.poly"))
mt.sm.mod <- mt.sm.mod.test <- mt.sm.mod.raw <- coef.VWC <- emm.VWC <- mt.sm.mod.plot <-list()
for(tr in traits) {
  #mt.sm.mod.raw[[tr]] <- lmer(mt[[tr]]    ~ year * VWC + (1|plot) + (1|plantid), data=mt)
  mt.sm.mod[[tr]] <-lmer(mt.plantyr[[tr]] ~ year * VWC + (1|plotid), data=mt.plantyr)
  mt.plantyr[[paste0("fit.sm.",tr)]] <- NA
  mt.plantyr[[paste0("fit.sm.",tr)]][!is.na(mt.plantyr[[tr]])]<- predict(mt.sm.mod[[tr]], re.form=NA)
  mt.sm.mod.test[[tr]] <- anova(mt.sm.mod[[tr]])
  
  #Z-score the data first to get standardized slopes
  mt.sm.mod[[tr]] <- lmer(scale(mt.plantyr[[tr]]) ~ year * VWC + (1|plotid), data=mt.plantyr)
  coef.VWC[[tr]] <- summary(mt.sm.mod[[tr]])$coefficients["VWC",]
  emm.VWC[[tr]] <- emtrends(mt.sm.mod[[tr]],  ~ year, var = "VWC")

   mt.sm.mod.plot[[tr]] <- 
    ggplot(mt.plantyr, aes_string(x="VWC", color="year", y = tr)) + geom_point() +
    geom_line(data=mt.plantyr %>% drop_na(paste0("fit.sm.",tr)), aes_string(y=paste0("fit.sm.",tr)), size=2)+
     scale_color_manual(values=year_pal)+
        labs(x="Summer mean of soil moisture in subplot (VWC)", y=traitnames[[tr]], color="Year")
  print(mt.sm.mod.plot[[tr]])
}
mt.sm.mod.tests <- map_dfr(mt.sm.mod.test, tidy, .id="trait")
mt.sm.mod.tests %>% select(c("trait", "term", "p.value")) %>% mutate(p.value = signif.num(p.value)) %>% 
  pivot_wider(names_from=trait, values_from=p.value) %>% 
    bind_rows(lapply(mt.sm.mod, function(x) ifelse(isSingular(x),"Y","N"))) %>% mutate(term = replace_na(term,"Singular?")) %>% kable

coef.VWC.df <- as.data.frame(t(as.data.frame(coef.VWC))) %>% rownames_to_column("trait")
emm.VWC.ls <- lapply(emm.VWC, as.data.frame)
emm.VWC.ls$sepal_width[3,] <- NA
emm.VWC.df  <- bind_rows(emm.VWC.ls, .id="trait")

ggplot(emm.VWC.df %>% drop_na(), aes(x=trait, y=VWC.trend, color=year)) +  
  geom_hline(yintercept=0) + geom_pointrange(aes(ymin=VWC.trend - SE, ymax = VWC.trend + SE), position=position_dodge(width=0.3), size=0.4) + 
  labs(y="Slope of z-scored trait vs. mean annual soil moisture", x="Floral Trait", color="Year") +
  theme_classic() + scale_x_discrete(labels = traitnames) + coord_flip() + scale_color_manual(values=year_pal)
```

### Multivariate response
```{r multresp}
mt.morph <- mt[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","VWC")] %>% drop_na(1:5) #drop sepal_width
mt.pca <- prcomp(mt.morph[,-(6:7)], scale.=TRUE)

mt.morph.pca <- mt.morph %>%
  bind_cols(as.data.frame(mt.pca$x)) %>% 
  tibble::rowid_to_column("index") %>% 
  mutate(x_0=0,y_0=0,
         x_1=-corolla_width/2,y_1=corolla_length,
         x_2= corolla_width/2,y_2=corolla_length) %>% 
  pivot_longer(c(starts_with("x"),starts_with("y")), names_to=c(".value", "point"), names_sep="_")

scl <- 100
zoom <- 5.5
mt.pca.vars <- as.data.frame(mt.pca$rotation) %>% rownames_to_column("variable")
mt.pca.labs <- paste0("PC", 1:2, " (", round(100*mt.pca$sdev[1:2]^2/sum(mt.pca$sdev^2), 0), "% explained var.)") 
ggplot(mt.morph.pca) + theme_classic() +
  geom_polygon(aes(x= x/scl+PC1, y= y/scl+PC2, group=index, color=VWC, fill=VWC)) +
  #geom_point(aes(x=PC1, y=PC2, size=sepal_width/2), color="darkgreen") + 
  geom_point(aes(x=PC1, y=PC2 + anther_min/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + anther_max/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + style_length/scl), color="grey60", size=0.5, shape=4) +
  geom_segment(data=mt.pca.vars,aes(x=0, y=0, xend=PC1*zoom, yend=PC2*zoom), 
               arrow = arrow(length = unit(1/2, "picas")), color="black") +
  geom_text(data =mt.pca.vars, aes(label = gsub("_"," ",variable), x = PC1*zoom*1.4, y = PC2*zoom*1.4,
                                   angle =  (180/pi) * atan(PC2/PC1), hjust = 0), color="black")+
  scale_size_identity() + coord_fixed() + 
  scale_fill_viridis(direction=-1, option="plasma") + scale_color_viridis(direction=-1, option="plasma") + 
  labs(x=mt.pca.labs[1], y=mt.pca.labs[2])
ggsave("images/mt_morph_pca_noSW.pdf", height=15, width=20, device=cairo_pdf)

ggplot(mt, aes(x=corolla_width, y=corolla_length, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal()
ggplot(mt, aes(y=style_length / corolla_length, x=VWC, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_smooth()

options(contrasts = c("contr.treatment", "contr.poly"))
mt.cap <- capscale(scale(mt.morph[,-(6:7)]) ~ VWC + year, data=mt.morph)
mt.cap
anova(mt.cap, by="term") %>% tidy %>% kable
library(ggvegan)
autoplot(mt.cap) + guides(color=F, shape=F)
```

```{r multresp_discrete}
mt.morph.wm <- mt[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","water","melt_offset")] %>% drop_na(1:6) #and nectar (19,20) too
mt.cap.wm <- capscale(scale(mt.morph.wm[,-(6:8)]) ~ year * melt_offset * water, data=mt.morph.wm)
mt.cap.wm
anova(mt.cap.wm, by="term") %>% tidy %>% kable
#plot CAP without interactions
autoplot(capscale(scale(mt.morph.wm[,-(6:8)]) ~ year + melt_offset + water, data=mt.morph.wm)) + guides(color=F, shape=F)
```

### Floral traits within season

No obvious effect of time within a season for corolla length/width or sepal width. The timing of trait measurements was different each year:

```{r morph_season}
ggplot(sm, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
  #geom_vline(xintercept=175) + geom_vline(xintercept=183) +
  geom_point(data=mt, aes(y= corolla_length), color="purple")+
    geom_point(data=mt, aes(y= nectar_24_h_ul), color="orange")+
  geom_point(aes(y = VWC), color="black") +
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color="grey40")+
  labs(title="Timing of measurements", subtitle="VWC (black) / Morphology (purple) / Nectar (orange)", x="Day of year", shape="Snow", linetype="Snow")+
  scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(data=mt, aes(x=yday(date), y= corolla_length, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y="Corolla Length (mm)", x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()

ggplot(data=mt, aes(x=yday(date), y= nectar_24_h_ul, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y="Nectar amount (48 hr, mm)", x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()

ggplot(data=nt, aes(x=yday(date), y= nectar_conc, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y="Nectar concentration", x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()
```

# Phenology
```{r phenology}
ggplot(ph20, aes(x=julian, y=height_cm, shape=snow, color=water))+ geom_point() + geom_path(aes(group=plantid))+
  labs(y="Plant height", 
       x="Day of year", color="Precipitation", shape="Snow") +
  scale_color_manual(values=water_pal) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

#ggplot(ph20, aes(x=julian, y=open+buds, shape=snow, linetype=snow, color=water))+ geom_point() + geom_path(aes(group=plantid)) +
#  labs(y="Total flowers and buds", 
#       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
#  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
#  theme(legend.key.size = unit(2, "lines"))

# ggplot(ph20, aes(x=julian, y=eggs, shape=snow, linetype=snow, color=water))+ geom_point() + geom_path(aes(group=plantid)) +
#   labs(y="Total eggs", 
#        x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
#   scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
#   theme(legend.key.size = unit(2, "lines"))

ph <-bind_rows(list("2020"= ph20,
                    "2018"= ph18 %>% mutate(plant=as.character(plant))), .id="year_df") %>% mutate(julian=as.integer(as.character(julian)))

ph %>% group_by(year_df, snow, water, julian) %>% summarize_at(c("open","buds"), mean) %>% 
  ggplot(aes(x=julian, y=open, shape=snow, linetype=snow, color=water))+ geom_point() + 
  facet_wrap(vars(year_df), nrow=2)+
  geom_path(aes(group=paste(snow, water))) +
  labs(y="Mean total flowers", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year_df, snow, water, julian) %>% summarize_at("flowering", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=flowering, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year_df), nrow=2)+
  geom_path(aes(group=paste(snow, water))) +
  labs(y="Proportion flowering", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year_df, snow, water, julian) %>% summarize_at("eggs", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=eggs, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year_df), nrow=2)+
  geom_path(aes(group=paste(snow, water))) +
  labs(y="Mean total eggs", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year_df, snow, water, julian) %>% summarize_at("has_egg", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=has_egg, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year_df), nrow=2)+
  geom_path(aes(group=paste(snow, water))) +
  labs(y="Proportion of plants with eggs", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

```

## Floral display size
### 2020
```{r display_size}
ph20.plantid <- ph20 %>% left_join(sm.subplotyear) %>% group_by(plotid, plantid, snow, water) %>% summarize_at(c("open","height_cm", "VWC"), mean, na.rm=T)

ggplot(ph20.plantid, aes(x=height_cm, y=open, color=water)) + geom_point() + geom_smooth(method="lm", se=F) + scale_color_manual(values=water_pal) +
  labs(x="Mean inflorescence height (cm)", y="Mean number of open flowers", color="Precipitation")

ggplot(ph20.plantid, aes(x=water, color=snow, y = height_cm)) +
    labs(x="Precipitation", y="Mean inflorescence height (cm)", color="Snowmelt") +
    geom_boxplot() + scale_color_manual(values=snow_pal) 
ggplot(ph20.plantid, aes(x=VWC, y=height_cm)) + geom_point() + geom_smooth() +
    labs(x="Mean VWC", y="Mean inflorescence height (cm)")
    
ggplot(ph20.plantid, aes(x=water, color=snow, y = open)) +
    labs(x="Precipitation", y="Mean number of open flowers", color="Snowmelt") +
    geom_boxplot() + scale_color_manual(values=snow_pal) 
ggplot(ph20.plantid, aes(x=VWC, y=open)) + geom_point() + geom_smooth()+
    labs(x="Mean VWC", y="Mean number of open flowers")
```

# Seeds
## 2018
```{r seeds, eval=FALSE}
ggplot(sds18, aes(color=water, fruits, seeds_total)) + geom_point() + geom_smooth(se=F) + scale_color_manual(values=water_pal)

ggplot(sds18, aes(color=water, water, seeds_total)) + geom_boxplot() + scale_color_manual(values=water_pal)
ggplot(sds18, aes(color=water, water, fruits)) + geom_boxplot() + scale_color_manual(values=water_pal)
ggplot(sds18, aes(color=water, water, seeds_total/fruits)) + geom_boxplot() + scale_color_manual(values=water_pal)
ggplot(sds18, aes(color=water, water, fruits_fly_no_seeds)) + geom_boxplot() + scale_color_manual(values=water_pal)
ggplot(sds18, aes(color=water, water, fruits_fly_no_seeds/fruits)) + geom_boxplot() + scale_color_manual(values=water_pal)
```

