---
title: "Maxfield Meadow Ipomopsis Selection"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---


```{r setup, message=FALSE}
library(tidyverse)
library(emmeans)
library(broom)
library(lme4)
library(lmerTest)
library(GGally)
library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, 
                      fig.path = "images_selection/", dev="svglite", dev.args=list(fix_text_size=FALSE), fig.height=6, fig.width=6)
options(digits=3) # for kables

load("data/maxfield_data.rda") #load trait measurements, etc
mnps.plantyr <- mnps.plantyr %>% mutate(prop_uninfested = 1 - prop_infested)
#TODO eggs_per_flower has some high values when the eggs were averaged over 1 observation but the flowers were averaged over multiple weeks. need to calculate this before averaging

fitnesstraits <- c("seeds_est", "seeds_per_flower", "prop_uninfested", "eggs_per_flower")
fitnessnames <- c(traitnames, prop_uninfested = "Prop. uneaten fruits",
                  eggs_per_flower="Fly eggs per flower")[fitnesstraits]
morphtraits <- c("corolla_length", "corolla_width","style_length","anther_max","anther_min","sepal_width","nectar_24_h_ul", "nectar_conc")#floral traits less height and flower number, plus anthers
morphnames <- c(traitnames, anther_max="Max stamen length", anther_min="Min stamen length")[morphtraits]
```

# Fitness vs. treatments and soil moisture

## Seed traits

```{r seeds_treat, fig.height=12, fig.width=5}
sds %>% pivot_longer(all_of(seedtraits), names_to = "trait") %>% 
  ggplot(aes(y=value, x=water, color=snow)) + 
  facet_grid(trait~year, scales="free_y", labeller=as_labeller(c(set_names(2018:2020), traitnames))) +
  geom_boxplot() + scale_color_manual(values=snow_pal) +  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+
  labs(x="Precipitation", color="Snow") + 
  theme_bw() + theme(axis.title.y=element_blank()) + guides(x=guide_axis(angle = 90))

sds %>% pivot_longer(all_of(seedtraits), names_to = "trait") %>% 
  ggplot(aes(y=value, x=VWC, color=year)) + 
  facet_grid(trait~year, scales="free_y", labeller=as_labeller(c(set_names(2018:2020), traitnames))) +
  geom_point(size=0.5) + geom_smooth(se=F) + scale_color_manual(values=year_pal) +
  labs(x="Summer mean of soil moisture in subplot (VWC)", color="Year") + 
  theme_bw() + theme(axis.title.y=element_blank()) + guides(year="none")
```

## Fitness metrics

```{r fitness_treat, fig.height=8, fig.width=5}
mnps.plantyr %>% pivot_longer(all_of(fitnesstraits), names_to = "trait") %>% 
  ggplot(aes(y=value, x=water, color=snow)) + 
  facet_grid(trait~year, scales="free_y", labeller=as_labeller(c(set_names(2018:2020), fitnessnames))) +
  geom_boxplot() + scale_color_manual(values=snow_pal) +  scale_y_continuous(expand=expansion(mult=c(0,0.05)))+
  labs(x="Precipitation", color="Snow") + 
  theme_bw() + theme(axis.title.y=element_blank()) + guides(x=guide_axis(angle = 90))

mnps.plantyr %>% pivot_longer(all_of(fitnesstraits), names_to = "trait") %>% 
  ggplot(aes(y=value, x=VWC, color=year)) + 
  facet_grid(trait~year, scales="free_y", labeller=as_labeller(c(set_names(2018:2020), fitnessnames))) +
  geom_point(size=0.5) + geom_smooth(se=F) + scale_color_manual(values=year_pal) +
  labs(x="Summer mean of soil moisture in subplot (VWC)", color="Year") + 
  theme_bw() + theme(axis.title.y=element_blank()) + guides(year="none")

# seeds.mod.sm <- lmer(seeds_rel ~ VWC * year + (1|plotid), 
#                      data=mnps.plantyr %>% group_by(year) %>% mutate(seeds_rel = seeds_est/mean(seeds_est, na.rm=T)))
# anova(seeds.mod.sm, ddf="Ken") %>% kable()
# emtrends(seeds.mod.sm, ~year, var="VWC") %>% kable()
```

## Fitness correlations

```{r fitness_corr}
mnps.plantyr %>% select(all_of(fitnesstraits)) %>% 
  ggpairs(mapping=aes(color=mnps.plantyr$year), lower=list(continuous="smooth_loess")) + 
  scale_color_manual(values=year_pal)+ scale_fill_manual(values=year_pal)

ggplot(sds %>% drop_na(plot), aes(color=water, x=fruits, y=seeds)) + geom_point() + geom_smooth(se=F) + 
  scale_color_manual(values=water_pal) + facet_wrap(vars(year), nrow=1) + theme_minimal() +
  labs(x="Counted fruits", y="Counted seeds", color="Precipitation")
```

# Floral trait correlations

```{r correlations}
mt %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width) %>% 
  ggpairs()

nt %>% select(nectar_24_h_ul, nectar_conc, nectar_sugar_24_h_mg) %>% 
  ggpairs()

ggplot(nt, aes(x=nectar_24_h_ul, y = nectar_conc)) + geom_point() + labs(x=traitnames[["nectar_conc"]], y=traitnames[["nectar_24_h_ul"]]) + geom_smooth(se=F)

mnps.plantyr %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width, nectar_24_h_ul, nectar_conc, nectar_sugar_24_h_mg, height_cm, flowers_est) %>% 
  ggcorr(hjust=0.85, layout.exp=2, label=T, label_round=2)

mnps.plantyr %>% select(all_of(seedtraits[c(1,7,2:6)])) %>% 
  ggcorr(hjust=0.85, layout.exp=2, label=T, label_round=2)
```

# Selection on individual traits

```{r mt_selection, fig.height=8, fig.width=6}
walk(fitnesstraits, ~ 
       print(mnps.plantyr %>% 
               mutate(prop_uninfested = if_else(fruits_nonaborted > 4, prop_uninfested, NA)) %>% 
               pivot_longer(all_of(floraltraits)) %>% 
               ggplot(aes_string(y=.x, x="value", color="year")) + 
               geom_point() + geom_smooth(method="lm", se=F) + 
               facet_wrap(vars(name), scales="free_x", labeller=as_labeller(traitnames)) + 
               labs(y=fitnessnames[.x], color="Year")+
               theme_bw() + scale_color_manual(values=year_pal) + theme(axis.title.x=element_blank()))) 

walk(fitnesstraits, ~ 
       print(mnps.plantyr %>% 
               mutate(prop_uninfested = if_else(fruits_nonaborted > 4, prop_uninfested, NA)) %>% 
               pivot_longer(all_of(floraltraits)) %>%
               ggplot(aes(x=value, color=paste(water,snow), linetype=year, shape=year)) + 
               geom_point(aes_string(y=.x)) + geom_smooth(aes_string(y=.x), method="lm",se=F) + 
               facet_wrap(vars(name), scales="free_x") + 
               labs(x="Emissions (ng/hr/flower)", y=fitnessnames[.x], 
                    color="Precipitation and snowmelt", linetype="Year", shape="Year")+
               theme_bw() + scale_x_sqrt() + scale_color_manual(values=water_snow_pal)))
```

```{r mt_indiv_mod}
mt.mod.indiv <- expand_grid(trait=floraltraits, fitnesstrait=fitnesstraits, yr=2018:2020) %>% 
  filter(!(yr==2018 & trait=="sepal_width")) %>% 
  mutate(model = pmap(., function(trait, fitnesstrait, yr) {
    lm(relfitness ~ trait.z, 
       data=mnps.plantyr %>% mutate(trait.z = mnps.plantyr[[trait]] / alltraits.sd[[trait]]) %>% 
         rename("fitness" = fitnesstrait) %>% 
         filter(year==yr) %>% mutate(relfitness = fitness/mean(fitness, na.rm=T)))}), # mean fitness in each yr
    coef = map(model, tidy))
    #test = map(model, ~tidy(car::Anova(.x), type=3)),
    #emm = map(model, ~summary(emtrends(.x, ~1, var="trait.z")))) 

#get the the standardized selection differential S' (change in relative fitness / standardized trait value)
mt.mod.indiv %>% select(trait, fitnesstrait, yr, coef) %>% unnest(coef) %>% filter(term=="trait.z") %>% 
  ggplot(aes(x=traitnames[trait], color=factor(yr), y=estimate, ymin=estimate-std.error, ymax=estimate+std.error)) + 
  facet_wrap(vars(fitnesstrait), labeller = as_labeller(fitnessnames)) + 
  geom_hline(yintercept=0) + geom_pointrange() + coord_flip() + scale_color_manual(values=year_pal) + 
  labs(y="Standardized selection differential S'", color="Year") + theme_bw() + theme(axis.title.y=element_blank())
```

# Selection gradients from PCA

Following method of [Chong et al. 2018](https://doi.org/10.1002/evl3.63) for reconstructing selection estimates from estimates of selection on PCs: Beta = E * A.

* PCA can't handle missing values so they are removed, but this biases the data and reduces sample size - multiple imputation?

```{r gradients, fig.width=8, fig.height=6}
morphtraits.pca <- morphtraits[-6]  #remove sepal_width (not measured in 2018)
#Standardize traits by mean and fitness by mean within each year
mnps.plantyr.std <- mnps.plantyr %>% select(all_of(morphtraits.pca),all_of(fitnesstraits), any_of(names(treatments))) %>%
  mutate(across(all_of(morphtraits.pca), ~ (.x-mean(.x, na.rm=T))/sd(.x, na.rm=T))) %>% 
  drop_na(all_of(morphtraits.pca)) 

# Run PCA on traits
mnps.PCA <- mnps.plantyr.std %>% select(all_of(morphtraits.pca)) %>% prcomp() 
summary(mnps.PCA) # PCs 1 - 5 explain 90% of the variation
mnps.PCA.E <- mnps.PCA$rotation[,1:5]
mnps.plantyr.std.PCA <- bind_cols(mnps.plantyr.std, as.data.frame(mnps.PCA$x))

calculate_beta <- function(E, slopes) {
  data.frame(B = E %*% slopes$A,
             SE = sqrt(E ^ 2 %*% slopes$SE ^ 2)) %>% 
    mutate(p.value = pt(abs(B)/SE, df=1, lower.tail=F)) %>% #calculate t statistic from difference in means and SE, then look up one-tailed p with 1 df in the upper tail
    rownames_to_column("trait")
}

selection_pca <- function(year, water=names(water_pal), snow=names(snow_pal), fitnesstrait = "seeds_est") {
  mnps.std.PCA <- mnps.plantyr.std.PCA %>% 
    filter(year == .env$year, water %in% .env$water, snow %in% .env$snow) %>%
    select(all_of(fitnesstrait), any_of(names(treatments)), starts_with("PC")) %>%
    rename("fitness" = fitnesstrait) %>%
    group_by(year) %>% mutate(relfitness = fitness/mean(fitness, na.rm=T), .keep="unused") %>% ungroup %>%
    drop_na(relfitness) %>%
    select(where(~sum(is.na(.))==0))

  # Regress relative fitness on PCs
  mnps.PCA.regression <- lm(relfitness ~ PC1 + PC2 + PC3 + PC4 + PC5, data = mnps.std.PCA)
  mnps.PCA.slopes <- tidy(mnps.PCA.regression) %>% filter(term != "(Intercept)") %>% rename(A=estimate, SE=std.error)
  
  # Convert selection gradients on PCs to gradients on traits
  calculate_beta(mnps.PCA.E, mnps.PCA.slopes)
}

plot_beta <- function(betas, colorby="year") {
  bs <- betas
  if(colorby=="water_snow")  bs <- bs %>%  mutate(water_snow = paste(water, snow))
  ggplot(bs, aes(y=morphnames[trait], x = B, xmin=B-SE, xmax=B+SE,)) + #alpha=p.value<0.05
    geom_vline(xintercept=0)+ 
    geom_pointrange(mapping=aes_string(color=colorby), position=position_dodge(width=0.5)) + 
    labs(x = "Selection gradient \U03B2 estimated from gradients on PCs", y = "Trait",
         color=list(year="Year", water="Precipitation", snow="Snowmelt", 
                    water_snow = "Precipitation and snowmelt")[[colorby]]) +
    scale_color_manual(values=list(year=year_pal, water=water_pal, snow=snow_pal,
                                   water_snow=water_snow_pal)[[colorby]]) +
    theme_bw()
}

mnps.beta.yr <- expand_grid(year=unique(mnps.plantyr$year), fitnesstrait=fitnesstraits)  %>% 
 mutate(betas = pmap(., selection_pca)) %>% unnest(betas)
plot_beta(mnps.beta.yr)+ facet_wrap(vars(fitnesstrait), nrow=1, labeller=as_labeller(fitnessnames))

mnps.beta.water <- expand_grid(year=unique(mnps.plantyr$year), water=names(water_pal), fitnesstrait=fitnesstraits)  %>% 
 mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

mnps.beta.snow <- expand_grid(year=unique(mnps.plantyr$year), snow=names(snow_pal), fitnesstrait=fitnesstraits)  %>% 
 mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

mnps.beta.water.snow <- expand_grid(year=unique(mnps.plantyr$year), water=names(water_pal), snow=names(snow_pal),  fitnesstrait=fitnesstraits)  %>% 
  mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

walk(fitnesstraits, ~print(plot_beta(filter(mnps.beta.water.snow, fitnesstrait==.x), colorby="water_snow") + 
                             facet_wrap(vars(year), nrow=1) + ggtitle(fitnessnames[.x])))
```

