--- 
title: "Code for \"Earlier snowmelt and reduced summer precipitation alter floral traits important to pollination\""
author: "John M. Powers, Heather M. Briggs, Rachel G. Dickson, Xinyu Li, Diane R. Campbell"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
    theme: cosmo
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>

```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)

library(lme4)
library(lmerTest)
library(MuMIn)
library(glmmTMB)
library(emmeans)
library(broom)
library(broom.mixed)
library(vegan)

library(RColorBrewer)
library(viridis)
library(gridExtra)
library(GGally)
library(ggvegan)
library(ggfortify) 
library(ggnewscale)
library(ggpattern)
library(ggpmisc)

library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images_cut/")
options(digits=4) # for kables

load("data/maxfield_data.rda")
load("data/daily_all.rda")

options(contrasts = c("contr.sum", "contr.poly")) #needed for Type III SS

anova_table <- function(df, mod) { #for lmer models and lmerTest::anova
  df %>% select(trait, term, p.value) %>% 
    pivot_wider(names_from=term, values_from=p.value) %>% 
    mutate(R2m = map_dbl(mod, ~ MuMIn::r.squaredGLMM(.x)[[1,"R2m"]]))
}
select <- dplyr::select
snow_pal_grey <- c("grey50","black")
```

# Metadata

```{r writeout}
alldata <- list("snowmelt"=groundcover %>% select(source, year, contains("cm_day"), precip_est_mm),
                "treatments"=treatments %>% select(year, plotid, water, water4, snow, sun_date,precip_est_mm),
                "soil_moisture"=sm.subplotyear %>% select(year, plotid, VWC),
                "traits"=mnps.plantyr %>% select(year, plotid, plant, any_of(floraltraits)) %>% 
                  arrange(year, plotid) %>% drop_na(plant) %>% filter(plant!="NA"))

purrr::walk(names(alldata), ~write_tsv(alldata[[.]], paste0("data_cut/",., ".tsv"), na=""))
```

The following four datasets are available in tab-delimited (TSV) format under the 'data' folder.

## Snowmelt

:Records of observed or inferred snowmelt timing from 1935 - 2021 in Gothic, CO, USA.

|Column|Description|
|:- |:------ |
|source|data source (billy: [billy barr’s plot observations](http://www.gothicwx.org/ground-cover.html) in Gothic, CO; 	streamflow: estimates from amount of peak runoff in the East River 	reported in [Anderson et al. 2012 *Plant 	Physiology*](https://doi.org/10.1104/pp.112.206219))|
|year|year|
|first_100_cm_day|day of the year the the snowpack first dropped below 100 cm|
|first_50_cm_day|day of the year the the snowpack first dropped below 50 cm|
|first_0_cm_day|day of the year the the snowpack first melted completely (snowmelt 	date)|
|precip_est_mm|summer precipitation from snowmelt date to day 214 (see Methods S1)|

## Treatments

:The precipitation and snowmelt treatments in the split-plot design, with snowmelt dates for each plot and precipitation totals for each subplot.

|Column|Description|
|:- |:------ |
|year|year|
|plotid|plot number and subplot letter|
|water|precipitation treatment, with mock rainouts combined with controls|
|water4|precipitation treatment, with mock rainouts separate from controls|
|snow|snowmelt treatment (in 2019, plot 2 was covered but coded as normal)|
|sun_date|snowmelt date in each plot, determined by sunlight threshold|
|precip_est_mm|estimated summer precipitation in each subplot (see Methods S1)|

## Soil moisture

:Summer averages of soil moisture in each subplot.

|Column|Description|
|:- |:------ |
|year|year|
|plotid|plot number and subplot letter|
|VWC|% volumetric water content, averaged within a subplot for each date and then averaged across each summer|

## Traits

:Summer averages of traits for each flowering plant.

|Column|Description|
|:- |:------ |
|year|year|
|plotid|plot number and subplot letter|
|plant|unique plant identifier|
|corolla_length|length of the corolla (mm)|
|style_length|length of the style (mm)|
|corolla_width|width of the corolla at the opening (mm)|
|sepal_width|width of one sepal at its widest point (mm)|
|nectar_24_h_ul|nectar production (µL/day)|
|nectar_conc|nectar sucrose concentration (% by mass)|
|height_cm|inflorescence height (cm)|
|flowers_est|total number of flowers produced during the growing season|

# Sample sizes

```{r sample_sizes}
years_flowering <- with(cen.status, table(flowering_status_18 + flowering_status_19 + flowering_status_20))

count_measurements <- function(df, tr) summarize(df, across(all_of(tr), ~ sum(!is.na(.x))))

bind_rows(n_measurements = count_measurements(bind_rows(mnps, lt, pt), floraltraits),
          plant_year = count_measurements(bind_rows(mnps, lt, pt) %>% group_by(year, plantid) %>% 
                                            summarize(across(all_of(floraltraits), mean, na.rm=T)), floraltraits) %>% 
            ungroup %>% summarize(across(all_of(floraltraits), sum)),
          plant =       count_measurements(bind_rows(mnps, lt, pt) %>% group_by(plantid) %>% 
                                             summarize(across(all_of(floraltraits), mean, na.rm=T)), floraltraits),
          .id="sample_size") %>% column_to_rownames("sample_size") %>% 
  t %>% as.data.frame %>% rownames_to_column("trait") %>% as_tibble %>% 
  mutate(n_per_plant_year = n_measurements / plant_year,
         plant_per_subplot_per_year = plant_year / 24 / 3) %>% 
  kable(digits=2, col.names=c("Trait", "Total measurements", "Total plant-years", 
        "Total unique plants measured", "Measurements / plant-year", "Plants / subplot / year"),
        caption=paste0("The number of unique plants is slightly less than the plant-years because ", 
                       round(100*years_flowering[["2"]]/(years_flowering[["1"]]+years_flowering[["2"]]),2),
                       "% of plants flowered in 2 years."))

```

# Trait correlations

```{r correlations}
mnps.plantyr %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width, nectar_24_h_ul, nectar_conc, nectar_sugar_24_h_mg, height_cm, flowers_est) %>% 
  ggcorr(hjust=0.85, layout.exp=2, label=T, label_round=2)
```


# Figure 1

```{r map, fig.height=4, fig.width=6.5}
size_meter <- 0.11
subplot_offset <- size_meter * 1.5
ggplot(treatments_map %>% mutate(snow=fct_relevel(snow, "Normal"))) + coord_fixed(clip="off")+
  geom_tile(aes(x=plot_x, y=plot_y, fill=snow), 
            width=7*size_meter, height=7*size_meter, color="black", size=0.5) +
  scale_fill_manual("Snowmelt", values=snow_pal_grey)+
  geom_tile_pattern(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, pattern_angle=water4), 
                    fill="white", pattern_fill="black", pattern_density=0.01, pattern_spacing=0.01,
                    height=2*size_meter, width=2*size_meter, color="black", size=0.5) + 
  scale_pattern_angle_manual("Precipitation", values=c(0,45,135,90))+
  geom_text(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, label=subplot), color="black") + 
  geom_text(aes(x=plot_x, y=plot_y, label=plot), color="white")+ theme_minimal() + 
  theme(legend.position="right", axis.title=element_blank(), axis.text = element_blank(), 
        axis.ticks = element_blank(), panel.background = element_blank(), panel.grid = element_blank())
```

# Figure 2

```{r hist_climate, fig.height=4, fig.width=6.5}
treatments %>% ggplot(aes(x=sun_date, shape=year, color=water, y=precip_est_mm))+geom_point(size=2)+
  geom_point(data=groundcover, aes(x=first_0_cm_day, color="Control", shape="1990-2017"), fill="white", size=2) +
  scale_color_manual(values=water_pal) + scale_shape_manual(values=c(21, 19, 17, 15))+ 
  scale_y_continuous(limits=c(0,NA), expand=expansion(mult = c(0,0.05)))+
  labs(y="Summer precipiation (mm)", x="Snowmelt date", color="Precipitation", shape="Year") + 
  theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"))
```

# Replicated split-plot models

## Table 1

```{r splitplot}
options(contrasts = c("contr.sum", "contr.poly"))
mod.split <- map(set_names(floraltraits), 
                   ~ lmer(mnps.plantyr[[.x]]  ~ year * snow * water + (1|snow:plot) + (1|snow:plotid),
                          data=mnps.plantyr))

mod.split.coefs <- map_dfr(mod.split, tidy, .id="trait")
mod.split.tests <- map(mod.split, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
mod.split.emm <- map_dfr(mod.split, ~ summary(emmeans(ref_grid(.x), ~ year * snow * water)), .id = "trait")
mod.split.tests.water <- map(mod.split, multcomp::glht, 
                             linfct = multcomp::mcp(water = c("Addition - Control = 0", "Reduction - Control = 0"))) %>% 
  map_dfr(tidy, .id="trait")

anova_table(mod.split.tests, mod.split) %>% 
  left_join(mod.split.tests.water %>% select(trait, contrast, adj.p.value) %>% 
              pivot_wider(names_from=contrast, values_from=adj.p.value)) %>% kable()
```

## Figure 3

```{r multi_splitplot, fig.width=9, fig.height=10}
plot_split_plot <- function(emm, data, traits.plot, geom = "point", plot.emm = TRUE, facets="year") {
  
  data.long <- data %>% select(c(all_of(traits.plot), all_of(facets), water, snow)) %>% 
    pivot_longer(all_of(traits.plot), names_to="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal")) %>% 
    drop_na(value, water) %>% filter(trait %in% traits.plot)
  
  traitnames.multiline <- ifelse(nchar(traitnames.units) > 30, str_replace(traitnames.units, fixed("("),"\n("), traitnames.units)
  
  split_plot <- emm %>% filter(trait %in% traits.plot) %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal")) %>%  
  ggplot(aes(x=water, color=snow)) +
    labs(x="Precipitation", y="Standardized trait", color="Snowmelt") + 
    scale_y_continuous(position="right") + scale_x_discrete(guide=guide_axis(angle=90)) + 
    scale_color_manual(values=snow_pal_grey, guide=guide_legend(override.aes = list(shape=15, size=ifelse(plot.emm,5,1)))) + 
    theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            axis.title.y= element_blank(),
                            panel.grid.major.x = element_blank(), panel.grid.minor.x=element_blank(), panel.grid.minor.y = element_blank(),
                            panel.border = element_rect(fill=NA, colour = "black"), 
                            panel.spacing=unit(0, "pt"), plot.margin = margin(0,0,0,0, "pt"), legend.position = "top") +
    switch(facets, 
           year = facet_grid(trait ~ year, scales="free_y", switch="y", 
                             labeller = as_labeller(c(traitnames.multiline, set_names(2018:2020)))),
           year.round = facet_grid(trait ~ year.round, scales="free_y", switch="y", 
                                   labeller = as_labeller(c(traitnames.multiline, set_names(levels(lt$year.round)))))) + 
    switch(geom,
           boxplot = geom_boxplot(data=data.long, aes(y=value), position=position_dodge(width=0.8), show.legend=!plot.emm,
                                  fatten=ifelse(plot.emm, NULL, 1), outlier.size=0.5),
           violin =   geom_violin(data=data.long, aes(y=value), position=position_dodge(width=0.8), show.legend=F),
           point =     geom_point(data=data.long, aes(y=value), position=position_dodge(width=0.8), shape=3))
  
  if(plot.emm) { return(split_plot + 
                          geom_linerange(aes(ymin=emmean-SE, ymax=emmean+SE,), position=position_dodge(width=0.8), size=1, show.legend=F) +
                          geom_point(aes(y=emmean), position=position_dodge(width=0.8), shape="-", size=16))
  } else return(split_plot)
}

#Morphological traits - plot of EMMs and subplot means
split_plot.floral <- grid.arrange(
  plot_split_plot(mod.split.emm, mnps.subplot, floraltraits[1:4]),
  plot_split_plot(mod.split.emm, mnps.subplot, floraltraits[5:8]) + 
    guides(color = guide_legend(override.aes = list(color="white", shape=15, size=5))) + 
    theme(legend.title = element_text(color = "white"), legend.text = element_text(color = "white")), nrow=1)
```

## Compensation

```{r compensation}
mnps.plantyr <- mnps.plantyr %>% mutate(snowwater= paste0(snow, water))
library(multcomp)#careful, this loads MASS, which overrides dplyr::select
options(contrasts = c("contr.sum", "contr.poly"))
mod.split.glht <- c(map(set_names(floraltraits), 
                   ~ lmer(mnps.plantyr[[.x]]  ~ snowwater * year + (1|snow:plot) + (1|snow:plotid),
                          data=mnps.plantyr))) %>% 
  #compare early snowmelt + addition vs. normal snowmelt + control
  map(glht, linfct = mcp(snowwater = c("EarlyAddition - NormalControl = 0"))) %>% 
  map_dfr(~ tibble(EarlyAddition_minus_NormalControl = coef(.x), p=summary(.x)$test$pvalues), .id="trait") 

mod.split.glht %>% kable(caption="Comparisons of traits in subplots with early snowmelt but water addition vs. subplots with normal snowmelt and precipiation.")
```

# Models with snowmelt date and total precipitation

## Table S2

### P-values

```{r meltdate}
options(contrasts = c("contr.sum", "contr.poly"))
mod.abs <- c(map(set_names(floraltraits), 
                 ~ lmer(mnps.plantyr[[.x]]  ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), data=mnps.plantyr)))

mod.abs.coefs <- map_dfr(mod.abs, tidy, .id="trait")
mod.abs.tests <- map(mod.abs, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
anova_table(mod.abs.tests, mod.abs) %>% kable()
```

### Regression slopes

```{r meltdate_slopes}
mod.abs.coefs %>% filter(effect=="fixed", term!="(Intercept)") %>% pivot_wider(id_cols=trait, names_from=term, values_from=c("estimate","std.error")) %>% kable()
```


## Figure 4

```{r multi_meltdate, fig.width=6, fig.height=10}
plot_abs <- function(mod, data, traits.plot, tests=NULL, facets = "trait") {
  data.long <- data %>% select(all_of(traits.plot), water, snow, year, sun_date, precip_est_mm) %>% 
    pivot_longer(all_of(traits.plot), names_to="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal"),
                      value = (value - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)]) %>% 
    drop_na(value, water) %>% filter(trait %in% traits.plot)
  
  precip.breaks <- seq(25,175, by=25)
  grid_emmeans <- function(mods, use.rounds = F) {
    grid.points <- list(sun_date=range(treatments$sun_date), precip_est_mm=precip.breaks)
    if(use.rounds)  map_dfr(mods, ~ summary(emmeans(.x, ~ precip_est_mm * sun_date * round, at=grid.points)), .id="trait")
    else            map_dfr(mods, ~ summary(emmeans(.x, ~ precip_est_mm * sun_date,         at=grid.points)), .id="trait")
  }
  emm.grid <- mod[traits.plot] %>% grid_emmeans(use.rounds= facets=="trait.round") %>% 
    mutate(trait = fct_relevel(trait, traits.plot),
           emmean = (emmean - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)])
  
  traitnames.multiline <- ifelse(nchar(traitnames) > 30, str_replace(traitnames, fixed("("),"\n("), traitnames)
  if(!is.null(tests)) {
    tests.sig <- tests %>% filter(p.value < 0.05) %>% 
          mutate(term.abbr = str_replace_all(term, c(sun_date="Sn", precip_est_mm="Pr",`:`="\U00D7"))) %>% 
          group_by(trait) %>% summarize(terms.sig = paste0("(",paste(term.abbr, collapse=", "),")")) %>% deframe
    traitnames.multiline[] <- paste(traitnames.multiline, replace_na(tests.sig[alltraits], ""))
  }
  
  abs_plot <- ggplot(emm.grid, aes(x=sun_date, y=emmean, color=precip_est_mm, group=precip_est_mm)) + 
    geom_point(data = data.long, aes(y=value, shape=year), size=2) + geom_line() + 
    scale_color_gradientn(colors=rev(water_pal), 
                          values=rev(c(1, (treatments %>% group_by(water) %>% summarize(across(precip_est_mm, mean)) %>%
                                             pull(precip_est_mm) %>% scales::rescale(from=range(precip.breaks)))[2],0)),
                          breaks=precip.breaks) + 
    labs(x="Snowmelt date", y="Standardized trait", color="Summer\nprecipitation\n(mm)", shape="Year") + 
    scale_y_continuous(position="right")+
    theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black")) +
    switch(facets, 
           trait =     facet_wrap(vars(trait), scales="fixed", dir = "v", ncol=2, labeller = as_labeller(traitnames.multiline)),
           trait.round = facet_grid(trait ~ round, scales="free_y", switch="y", 
                                   labeller = as_labeller(c(traitnames.multiline, 
                                                            set_names(paste("Round",1:2), 1:2)))))
    
  return(abs_plot)
}

plot_abs(mod.abs, mnps.subplot, floraltraits, mod.abs.tests) + #Morphological traits - subplot means
  scale_y_continuous(limits=c(-1.7, 2.8), breaks=c(-1:2))
```

## Compensation

```{r compensation_meltdate_avg}
mod.abs.emm.avg <- map_dfr(set_names(c("sun_date",  "precip_est_mm")), 
                   ~ mod.abs %>% map(emtrends, var = .x, ~1) %>% 
                     map(as_tibble) %>% set_names(floraltraits) %>% bind_rows(.id="trait") %>% 
                     rename_with(.cols=ends_with("trend"), ~ "trend"), .id="predictor") %>% 
  mutate(predictor=fct_relevel(predictor,"sun_date"), 
         sd.plantyr = alltraits.sd[trait],
         trend = trend / sd.plantyr, SE = SE / sd.plantyr)

(mod.abs.emm.avg.compensate <- mod.abs.emm.avg %>% 
  pivot_wider(names_from=predictor, values_from=c(trend, SE), id_cols=c(trait)) %>% 
  mutate(precip_mm_per_snow_day = trend_sun_date / trend_precip_est_mm, 
         SE = abs(precip_mm_per_snow_day) * # propagate error in SEs using formula for quotients
           sqrt((SE_precip_est_mm / trend_precip_est_mm)^2 + (SE_sun_date / trend_sun_date)^2),
         .keep="unused") %>% kable(caption="The amount of precipiation (mm) required to compensate for snowmelt occuring 1 day earlier."))
```

# Models with soil moisture

## Table S1

```{r soil_treatments}
# VWC split-plot analysis on subplot means (for the whole measurement period)
sm.mod <- lmer(VWC ~ year * snow * water + (1|snow:plot)+ (1|snow:plotid), data=sm.subplotyear)
anova(sm.mod) %>% kable()
```

### Soil moisture in each treatment

```{r soil_emms}
sm.mod %>% emmeans(~ snow)  %>% summary() %>% select(-c(df,lower.CL,upper.CL),mean_VWC=emmean) %>% kable()
sm.mod %>% emmeans(~ water) %>% summary() %>% select(-c(df,lower.CL,upper.CL),mean_VWC=emmean) %>% kable()
```

## Table S3

```{r sm}
options(contrasts = c("contr.sum", "contr.poly"))
mod.sm <- c(map(set_names(floraltraits), 
                  ~ lmer(mnps.plantyr[[.x]]  ~ year * VWC + (1|plotid), data=mnps.plantyr)))
mod.sm.coefs <- map_dfr(mod.sm, tidy, .id="trait")
mod.sm.tests <- map(mod.sm, anova, ddf="Ken") %>% map_dfr(tidy, .id="trait")
anova_table(mod.sm.tests, mod.sm) %>% kable()
```

## Figure 5

```{r multi_sm, fig.width=6, fig.height=10}
plot_sm <- function(mod, data, traits.plot, tests=NULL) {
  data.long <- data %>% select(all_of(traits.plot), water, snow, year, sun_date, precip_est_mm, VWC) %>% 
    pivot_longer(all_of(traits.plot), names_to="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal"),
                      value = (value - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)]) %>% 
    drop_na(value, water) %>% filter(trait %in% traits.plot)
  
  VWC.range <- range(data$VWC, na.rm=T)
  emm.grid <- mod[traits.plot] %>% 
    map_dfr(~ emmeans(.x, ~ year*VWC, at=list(VWC=seq(from=floor(VWC.range[1]*10)/10, to=ceiling(VWC.range[2]*10)/10, by=0.1))) %>% as_tibble(), .id="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot),
           emmean = (emmean - alltraits.mean[as.character(trait)])/alltraits.sd[as.character(trait)]) %>% 
    left_join(data %>% drop_na(VWC) %>% group_by(year) %>% summarize(VWC.min = min(VWC), VWC.max=max(VWC))) %>%
    group_by(year) %>% filter(VWC >= VWC.min-0.1, VWC <= VWC.max+0.1)
  
  traitnames.multiline <- ifelse(nchar(traitnames) > 30, str_replace(traitnames, fixed("("),"\n("), traitnames)
  if(!is.null(tests)) {
    tests.sig <- tests %>% filter(p.value < 0.05) %>% 
          mutate(term.abbr = str_replace_all(term, c(year="Yr", VWC="SM",`:`="\U00D7"))) %>% 
          group_by(trait) %>% summarize(terms.sig = paste0("(",paste(term.abbr, collapse=", "),")")) %>% deframe
    traitnames.multiline[] <- paste(traitnames.multiline, replace_na(tests.sig[alltraits], ""))
  }
  
  sm_plot <- ggplot(emm.grid, aes(x=VWC, y=emmean, color=year, group=year)) + 
    facet_wrap(vars(trait), scales="fixed", dir="v", ncol=ifelse(length(traits.plot)>3,2,4), labeller = as_labeller(traitnames.multiline))+
    geom_point(data = data.long, aes(y=value, color=year), size=1) +  
    geom_line(size=2) + 
    scale_color_manual(values=year_pal) + ylab("Standardized trait") + labs(color="Year") + 
    #coord_fixed() + 
    theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"))
  return(sm_plot)
}

plot_sm(mod.sm, mnps.plantyr, floraltraits, mod.sm.tests) + 
  scale_y_continuous(limits=c(-4.3,5)) + xlab("Summer mean of soil moisture in subplot (%VWC)")
```

# Figure 6

Drawn manually from results of script "maxfield_path_analysis.sas" executed in SAS 9.3.

# Figure S1

```{r timings}
  timings_labels <- c(snowcloth="Snow cloths applied to early plots", 
                     meltdates="Mean early and normal snowmelt timing", 
                     waterdates="Summer precipitation treatments applied",
                     sm="Soil moisture recorded",         ph="Inflorescence height recorded", 
                     mt="Floral morphology recorded",     nt="Nectar traits recorded", 
                     pt= "Physiology traits recorded",    lt="Vegetative traits recorded", 
                     sds="Flower collection period",      cen="Rosette size and survival recorded")

plot_timings <- function(data) {
  data %>% mutate(variable=fct_reorder(variable, begin, na.rm=T, .desc=T),
                  drawline = ifelse(variable %in% c("lt","meltdates"), NA, TRUE)) %>% 
ggplot(aes(y=variable, color=year))+ 
  geom_linerange(aes(xmin=begin, xmax=end, linetype=drawline), 
                 position = position_dodge2(width=0.7, reverse = T), size=1.2, show.legend=FALSE) + 
  geom_text(data=data %>% drop_na(plots) %>% filter(year=="2019"),
            aes(x = end+7, label=paste(ifelse(nchar(plots)>1, "Plots","Plot"), str_replace(plots,","," & "))), 
            position = position_dodge2(width=0.8, reverse = T), size=4, hjust=0, show.legend=FALSE)+
  geom_point(aes(x=begin), position = position_dodge2(width=0.7, reverse = T), shape=15, size=2)+
  geom_point(aes(x=end), position = position_dodge2(width=0.7, reverse = T), shape=15, size=2) +
  scale_color_manual("Year", values=year_pal, guide=guide_legend(override.aes = list(size=5))) + 
  scale_y_discrete("", labels=timings_labels)+
  scale_x_continuous("Day of year", breaks=seq(80,240, by=20)) +
  theme_minimal() + theme(legend.position = "top", text=element_text(size=14, color="black"), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"))
}

plot_timings(filter(timings, !variable %in% c("sds","lt","pt","cen")))
```

# Figure S2

```{r groundcover, fig.height=5, fig.width=7.5}
groundcover %>% mutate(first_snow_day=first_snow_day-365) %>% select(-first_snow_day) %>% 
  pivot_longer(ends_with("day"), names_to="threshold", values_to="day") %>% drop_na(day) %>% select(-starts_with("first")) %>% 
  mutate(threshold = fct_reorder(factor(threshold),day, na.rm=T, .desc=T)) %>% 
  ggplot(aes(x=year, y=day, color=threshold)) +  
  annotate("rect", xmin=2017.5, xmax=2020.5, ymin=-Inf, ymax=Inf, fill="lightblue")+
  geom_line(aes(group=year), size=1.5)+ geom_smooth(method="lm", se=F) +
  geom_point(aes(shape=source)) + scale_shape_manual("Source",values=c(19,17), labels=c("Observed", "Inferred from peak runoff")) +
  scale_x_continuous(breaks=seq(1940,2020, by=10), minor_breaks = scales::pretty_breaks(n = 100), position = "top") + scale_y_continuous(n.breaks=10)+
  scale_color_grey(labels=c(paste(c(0,50,100),"cm")), start=0, end=0.85) +
  labs(x="Year", y="Day of year", color="First day snowpack below") + theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),  panel.border = element_rect(fill=NA, colour = "black"))

meltdate.mod <- lm(first_0_cm_day ~ year, data=groundcover)
early.melt_offset <- treatments %>% filter(snow=="Early") %>% pull(melt_offset)
yearrange <- paste(range(groundcover$year), collapse=" - ")

meltdate.mod %>% tidy() %>% kable(caption=paste("Regression of snowmelt date vs. year from",yearrange))
cat("Snowmelt is advancing",-10*tidy(meltdate.mod)$estimate[2],"+-",10*tidy(meltdate.mod)$std.error[2], "days per decade")
cat("Early snow treatments melted", -round(mean(early.melt_offset),1), "+-", round(sd(early.melt_offset),1), "days earlier (mean +- SD), which is equivalent to", mean(early.melt_offset)/tidy(meltdate.mod)$estimate[2], "+-", (mean(early.melt_offset)/tidy(meltdate.mod)$estimate[2])*sqrt((sd(early.melt_offset)/mean(early.melt_offset))^2 + (tidy(meltdate.mod)$std.error[2]/tidy(meltdate.mod)$estimate[2])^2), "years of future warming")#propagate error from SD of treatment advance and SE of regression
```

# Figure S3

```{r soil, fig.height=9, fig.width=9}
max_VWC <- max(sm.subplot$VWC, na.rm=T)
ggplot(sm.subplot, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
  geom_segment(data=waterdates %>% pivot_wider(names_from=precip_treatments, values_from=c(day,date)),
               aes(x=day_started, xend=day_ended, y=max_VWC+0.5, yend=max_VWC+0.5)) +
  geom_col(data=daily_precip_est, aes(x=yday(date), y=EPA_NOAA_filled), fill=brewer.pal(7, "Set3")[7]) +
  geom_line(data=daily_precip_est, aes(y=VWC_Control_4*100), color=brewer.pal(7, "Set3")[6], size=1) +
  geom_point(aes(shape=snow, color=water, y = VWC)) + 
  geom_line(data=sm.subplot %>% group_by(water, snow, year, date) %>% summarize(VWC=mean(VWC, na.rm=T)), aes(linetype=snow, color=water, y = VWC)) +
  scale_color_manual(values=water_pal) +  scale_linetype_manual(values=c(2,1)) + 
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color="black", show.legend=F) +
  coord_cartesian(xlim=c(110, 230), ylim=c(0, max_VWC+1), expand=F)+
  labs(x="Day of year", y="Mean soil moisture in subplot (%VWC) / Precipitation (mm)", color="Precipitation", shape="Snowmelt", linetype="Snowmelt") + theme_minimal() + theme(legend.key.width = unit(2, "lines"), text=element_text(color="black", size=14), axis.text = element_text(color="black"), panel.border = element_rect(fill=NA, colour = "black"))
```


# Figure S4

```{r multresp_VWC, fig.height=7, fig.width=8}
mt.morph.plantyr <- mt.plantyr[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","VWC")] %>% drop_na() 
options(contrasts = c("contr.treatment", "contr.poly"))
mt.rda <- rda(scale(select(mt.morph.plantyr, !c(year, VWC))) ~ VWC * year, data=mt.morph.plantyr)

mt.rda.df <- fortify(update(mt.rda, ~ VWC + year))#drop interaction
ggplot(mapping=aes(x=RDA1, y=RDA2, label=Label)) + 
  geom_point(data=mt.rda.df %>% filter(Score=="sites") %>% bind_cols(mt.morph.plantyr), aes(color=VWC))+
  geom_segment(data=mt.rda.df %>% filter(Label=="VWC"), aes(x=0,y=0,xend=RDA1*3, yend=RDA2*3), 
               arrow=arrow(length=unit(0.5, "lines")), size=1.5)+
  geom_text(   data=mt.rda.df %>% filter(Label=="VWC"), aes(x=RDA1*3.7, y=RDA2*3.7, angle=atan(RDA2/RDA1)*180/pi), label="Soil moisture")+
  geom_segment(data=mt.rda.df %>% filter(Score=="species"), aes(x=0,y=0,xend=RDA1, yend=RDA2), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(   data=mt.rda.df %>% filter(Score=="species"), aes(x=RDA1*1.7, y=RDA2*1.7, angle=atan(RDA2/RDA1)*180/pi,
    label = c(traitnames, c("anther_max"="Max stamen length", "anther_min"="Min stamen length"))[as.character(Label)] %>% str_remove(fixed(" (mm)"))), size=3)+
  geom_text(data=mt.rda.df %>% filter(Score=="centroids"), aes(label=str_remove(Label, "year")), fontface=2, size=6)+
  scale_color_viridis("Soil\nmoisture\n(%VWC)", direction=-1, option="plasma") + coord_fixed() + 
  scale_x_continuous(expand=expansion(mult=.16), breaks=c(-2,0,2))+   theme_minimal() + 
  theme(panel.grid.minor=element_blank(), text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"))
```

## RDA summary and test

```{r rda_VWC}
mt.rda
anova(mt.rda, by="term")
```

# Figure S5

Drawn manually from results of script "maxfield_path_analysis.sas" executed in SAS 9.3.

# Figure S6

```{r precip_trend}
summary_EPA_NOAA <- daily_filled_7am %>% filter(year(date)>1989, year(date)<2021, EPA_NOAA_filled > 2) %>% 
  mutate(drought_days = c(0,diff(date)), year=year(date)) %>% filter(ground_covered=="smmr") %>% 
  select(year, date, drought_days) %>% group_by(year) %>% summarize(max_drought_days_2mm = max(drought_days)) %>% 
  left_join(daily_filled_7am  %>% filter(year(date)>1989, year(date)<2021, ground_covered=="smmr") %>% group_by(year=year(date)) %>% 
              summarize(snow_free_days = n(),
                        rain_days_2mm = sum(EPA_NOAA_filled > 2, na.rm=T),
                        prop_rain_days_2mm = rain_days_2mm / snow_free_days,
                        max_precip_mm = max(EPA_NOAA_filled, na.rm=T),
                        avg_precip_mm = mean(EPA_NOAA_filled, na.rm=T),
                        total_precip_mm = sum(EPA_NOAA_filled, na.rm=T)))

summary_names <- c(snow_free_days = "Snowmelt to first permanent snowfall (days)",
                   max_drought_days_2mm="Longest drought < 2 mm/day (days)", 
                   rain_days_2mm = "Days with > 2 mm precipitation",
                   prop_rain_days_2mm = "Proportion days > 2 mm precipitation",
                   max_precip_mm = "Maximum precipitation (mm/day)",
                   avg_precip_mm = "Average precipitation (mm/day)",
                   total_precip_mm = "Total precipitation (mm)")

summary_EPA_NOAA %>% pivot_longer(!year) %>% filter(!name %in% c("rain_days_2mm","max_precip_mm","total_precip_mm")) %>% 
  ggplot(aes(y=value, x=year))+ 
  facet_wrap(vars(name), scales="free_y", labeller=as_labeller(summary_names)) +
  annotate("rect", xmin=2017.5, xmax=2020.5, ymin=-Inf, ymax=Inf, fill="lightblue")+
  geom_smooth(method="lm", color="black", se=F) + geom_point(color="grey60") + labs(x="Year", y="") + 
  stat_poly_eq(formula = y ~ x, aes(label = paste(..eq.label.., ..rr.label.., ..p.value.label.., sep = "~~~")), parse = T) +
  theme_minimal() +theme(axis.title.y= element_blank(), text=element_text(color="black", size=14), 
                         axis.text = element_text(color="black"),  panel.border = element_rect(fill=NA, colour = "black"))
```

