--- 
title: "Maxfield Meadow Ipomopsis Experiment"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: inline
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>

# Purpose
Analyze selection on and plastic responses of floral and vegetative traits of *Ipomopsis aggregata* to manipulations of snowmelt and summer precipitation. Determine whether plasticity is adaptive (improves survival and reproduction). Part of a [larger project](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1654655) to understand the potential for evolutionary rescue in alpine plants under climate change.

# Treatments

```{r setup, message=FALSE, cache=FALSE}
library(tidyverse)
library(lubridate)
#library(googlesheets4)

library(lme4)
library(lmerTest)
library(MuMIn)
library(glmmTMB)
library(emmeans)
library(broom)
library(broom.mixed)
library(vegan)

library(RColorBrewer)
library(viridis)
library(gridExtra)
library(GGally)
library(ggvegan)
library(ggnewscale)

library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images/")

options(contrasts = c("contr.sum", "contr.poly")) #needed for Type III SS

print_anova_table <- function(df, mod) {
  df %>% select(c("trait", "term", "p.value")) %>% 
    mutate(p.value=p.value %>% map_chr(~ format(.x,scientific=-2, digits=2 , nsmall=2)) %>% 
             str_replace("e-0(\\d)"," x 10^-\\1^") %>% str_replace("e-(\\d+)"," x 10^-\\1^")) %>%
    pivot_wider(names_from=term, values_from=p.value) %>% 
    mutate(singular = map_lgl(mod, isSingular),
           R2m = map_dbl(mod, ~ MuMIn::r.squaredGLMM(.x)[[1,"R2m"]])) %>% 
    kable
}

load("data/maxfield_data.rda")
```

```{r map}
subplot_offset <- 1/6
size_meter <- 6
ggplot(treatments_map) + coord_fixed(clip="off")+
  geom_point(aes(x=plot_x, y=plot_y, fill=snow), size=7*size_meter, shape=22, stroke=1)+
  scale_fill_manual("Snowmelt", values=snow_pal, guide=guide_legend(override.aes = list(size = 2*size_meter)))+
  new_scale_fill()+
  geom_point(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, fill=water4), size=2*size_meter, shape=22, stroke=1) + 
    geom_text(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, label=subplot)) + 
  scale_fill_manual("Precipitation", values=water4_pal) +
  geom_text(aes(x=plot_x, y=plot_y, label=plot))+
  theme(legend.position="left", axis.title=element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.background = element_blank(), panel.grid = element_blank(), legend.box.background = element_blank())
```


# Snow melt vs. treatments

```{r hobo}
ggplot(hobo, aes(x=intensity, y=temp, color=hour(time))) +  geom_point(alpha=0.5, size=0.5) + geom_density_2d(color="black")+ 
  scale_x_log10(labels=function(x) format(x, scientific=F, big.mark=","), limits=c(10,1e6), breaks=10^(0:6)) +
  scale_color_gradientn(colours = rainbow(5))+
  geom_vline(xintercept = melt_threshold_sun) + 
  geom_hline(yintercept=melt_threshold_warm)+ geom_hline(yintercept=-melt_threshold_warm) +
  labs(x="Light intensity (lux)", y="Temperature (C)", color="Hour") +
  theme_dark() + theme(panel.grid.minor.x=element_blank())

ggplot(hobo %>% filter(hour(time)%in%5:17 & 
                  (year!="2019"&yday(time) %in% 110:130 | 
                   year=="2019"&yday(time) %in% 145:160)), 
       aes(x=intensity, y=temp, color=hour(time))) +   geom_point(alpha=0.7) +geom_density_2d(color="black")+
  scale_x_log10(labels=function(x) format(x, scientific=F, big.mark=","), limits=c(10,1e6), breaks=10^(0:6)) +
  scale_color_gradientn(colours = rainbow(5))+
  geom_vline(xintercept = melt_threshold_sun) + 
  geom_hline(yintercept=melt_threshold_warm)+ geom_hline(yintercept=-melt_threshold_warm) +
  labs(title="Temperature and light thresholds for snow melt", subtitle = "During the day during melt weeks",
       x="Light intensity (lux)", y="Temperature (C)", color="Hour") +
  theme_dark() + theme(panel.grid.minor.x=element_blank())

ggplot(hobo %>% filter(month(time)%in%4:6), aes(x=time, y=temp, color=plot)) + 
  facet_wrap(vars(year), scales="free_x", ncol=1) + 
  geom_hline(yintercept=melt_threshold_warm) + geom_hline(yintercept=-melt_threshold_warm) +geom_line() +
  geom_segment(data=meltdates, aes(x=warm_time, y=-10, xend=sun_time, yend=-15, color=plot), size=1) +
    geom_line(data=daily_CORBIL %>% filter(year!="2017", month(date)%in%4:6), aes(x=as.POSIXct(date), y=av_temp_2m_C), inherit.aes = F, color=brewer.pal(3, "Oranges")[3]) +
  labs(x="Date",y="Temperature (C)", color="Plot")
  
ggplot(hobo %>% filter(month(time)%in%4:6), aes(x=time, y=intensity, color=plot)) + 
  facet_wrap(vars(year), scales="free_x", ncol=1) + 
  geom_line() + geom_hline(yintercept=melt_threshold_sun) +
  geom_point(data=meltdates, aes(x=sun_time, y=-2e4, color=plot), size=1) +
  labs(x="Date",y="Light intensity (lux)", color="Plot")

meltdates %>% select(c("year","plot", "snow", ends_with("date"))) %>% pivot_longer(ends_with("date")) %>% arrange(name) %>% 
  ggplot(aes(y=name, x=value, color=plot, group=plot, linetype=snow)) + facet_wrap(vars(year), ncol=1) +
  geom_path(size=1) + #geom_text(aes(label=plot), hjust=1.5) +
  scale_linetype_manual(values=c(4,1)) + scale_x_continuous(minor_breaks=110:160)+
  labs(x="Day of year",y="Criteron", color="Plot", linetype="Snowmelt")
       
meltdates %>% ggplot(aes(x=melt_offset, fill=snow)) + geom_dotplot(method="histodot", binwidth=1, color="white") +
  facet_wrap(vars(year), ncol=1)+ scale_fill_manual(values=snow_pal) +
  scale_y_continuous(expand=expansion(0,0))+ scale_x_continuous(breaks=-10:4)+
  theme(panel.grid.minor = element_blank()) +
  labs(x="Offset from mean normal snow melt date (days)", y="Number of plots", fill="Snowmelt")
```

# Weather data

## Daily weather

```{r weather_daily}
ggplot(daily_all %>% filter(year(date) %in% 2018:2020), aes(y=EPA_NOAA_filled, x=yday(date))) + geom_col(aes(fill=ground_covered)) + 
  facet_wrap(vars(year(date)), ncol=1) + labs(x="Day of year", y="NOAA_billy precipiation (mm)", fill="Snow cover")

daily_all %>% pivot_longer(unname(stations), names_to="station", values_to="precipitation_mm") %>% 
  filter(yday(date) >100, yday(date) < 240, year(date) %in% 2018:2020) %>%
  ggplot(aes(x=yday(date), y=precipitation_mm, color=station, shape=ground_covered)) +  
  geom_line() +  geom_point() + facet_wrap(vars(year), ncol=1) + theme(legend.position = "top", legend.direction="vertical") +
  labs(x="Day of year", y = "Precipitation (mm)", color="Station", shape="Snow cover")

precip_max <- 50
lower_continuous <- function (data, mapping, ...) {
    ggally_smooth(data = data, mapping = mapping, ..., method = "loess") + scale_shape_manual(values=c(1,2))+
    coord_cartesian(xlim=c(0,precip_max), ylim=c(0,precip_max)) + geom_abline(slope=1, intercept=0)+
    scale_x_continuous(breaks=seq(0,precip_max, by=precip_max/5))+scale_y_continuous(breaks=seq(0,precip_max, by=precip_max/5))
}
lower_combo <- function (data, mapping, ...) {
    ggally_facethist(data = data, mapping = mapping, binwidth=precip_max/50,...) +
    coord_cartesian(xlim=c(0,precip_max)) + scale_x_continuous(breaks=seq(0,precip_max, by=precip_max/5))
}
ggpairs(daily_all %>% select(all_of(unname(stations[-1])), ground_covered), mapping=aes(color=ground_covered,shape=ground_covered), lower=list(continuous=wrap(lower_continuous), combo=wrap(lower_combo)))

filter(daily_all) %>% ggplot(aes(y=EPA_RsrchMdw,x= NOAA_billy, color=ground_covered)) + geom_point(shape=1) + geom_smooth(method="lm", se=F) + geom_abline(intercept=0, slope=1) + labs(color="Snow cover")
```

## Monthly weather

```{r weather_monthly}
#recalculate Campbell and Wendlandt 2013 July average, 1989 - 2006 EPA station
(EPA_89_06_July_avg_mm <- monthly_all %>% filter(mo==7, yr >= 1989, yr <= 2006) %>% pull(EPA_NOAA_filled) %>% mean)

(water_addition_per_month_mm <- (14 / 4 / 2) * 31) #14 L applied to 4 m2 every 2 days = 1.75 mm daily precip * 31 days in July

monthly_all %>% pivot_longer(unname(stations), names_to="station", values_to="precipitation_mm") %>% 
ggplot(aes(x=mo*30.4-15.2, y=precipitation_mm, color=station)) + facet_wrap(vars(yr)) + #place points at month midpoint
  geom_rect(data=waterdates %>% pivot_wider(names_from=precip_treatments, values_from=c(day, date)) %>% mutate(yr=factor(year)), 
            aes(xmin=day_started, xmax=day_ended, ymin=0, ymax=water_addition_per_month_mm), inherit.aes=F,
            fill=water_pal["Addition"])+
  geom_line() +  geom_point()+
  geom_point(data=groundcover %>% filter(year>=2004) %>% mutate(yr=factor(year)), aes(x=first_0_cm_day,y=0), inherit.aes=F)+
  geom_point(data=groundcover %>% filter(year>=2005) %>% mutate(yr=factor(year-1)), aes(x=first_snow_day,y=0), inherit.aes=F)+
  labs(y="Monthly precipitation (mm)", x="Day of the year, with snowmelt and first snow dates", color="Station") +
  theme(legend.position="top") + coord_cartesian(ylim=c(0,200))

precip_max <- 250
ggpairs(monthly_all %>% mutate(season=factor(c("wntr","smmr"))[1+mo %in% 6:9]) %>% select(all_of(unname(stations[-1])), season), mapping=aes(color=season,shape=season), lower=list(continuous=wrap(lower_continuous), combo=wrap(lower_combo)))

filter(monthly_all) %>% ggplot(aes(y=EPA_RsrchMdw,x= NOAA_billy, color=ifelse(mo %in% 6:9,mo,"Oct - May"))) + geom_point() + geom_smooth(method="lm", se=F) + geom_abline(intercept=0, slope=1) + labs(color="Month")

monthly_all %>% filter(mo %in% 5:9) %>% 
  ggplot(aes(x=mo,  y=EPA_NOAA_filled)) + geom_smooth(span=0.5, color="black") +
  geom_line(aes(color=ifelse(yr>=2018, "study period (filled)", ifelse(yr>=2012,"filled from billy's NOAA data","original EPA data")), group=yr)) +
  geom_point(x=7, y=EPA_89_06_July_avg_mm)+ scale_y_continuous(limits=c(0,100))+
  geom_text(aes(label=ifelse(yr>=2018, yr, "")))  + 
  labs(y="Monthly precipitation (mm)", x="Month", title="EPA Research Meadow 1989-2020 summer precipitation", 
       subtitle="Black dot is mean July precipitation 1989-2006", color="Data filling") + theme(legend.position = "top")

monthly_all %>% filter(mo %in% 5:9) %>% ggplot(aes(x=yr, color=factor(mo), group=mo, shape=ifelse(yr>=2018, "study period (filled)", ifelse(yr>=2012,"filled from billy's NOAA data","original EPA data")), y=EPA_NOAA_filled)) +
  geom_vline(xintercept=2011.5) +geom_point() + geom_smooth(se=F, span=0.7) + 
  scale_shape_manual(values=c(1,2,19)) + scale_y_continuous(limits=c(0,100))+
  labs(x="Year",y="EPA Research Meadow monthly precipitation (mm)", color="Month", shape="Data filling") + 
  geom_segment(data=monthly_all %>% filter(mo %in% 5:9, yr <= 2007) %>% 
                 group_by(mo) %>% summarize(mean_precip=mean(EPA_NOAA_filled)), 
               aes(x=1989, xend=2006, y=mean_precip, yend=mean_precip, color=factor(mo)), inherit.aes=F, size=1) 
```

## Snowmelt timing

```{r groundcover}
groundcover %>% mutate(first_snow_day=first_snow_day-365) %>% 
  pivot_longer(ends_with("day"), names_to="threshold", values_to="day") %>% drop_na(day) %>% select(-starts_with("first")) %>% 
  mutate(threshold = fct_reorder(factor(threshold),day, na.rm=T, .desc=T)) %>% 
  ggplot(aes(x=year, y=day, color=threshold)) +  
  annotate("rect", xmin=2017.5, xmax=2020.5, ymin=-Inf, ymax=Inf, fill="lightblue")+
  geom_line(aes(group=year), size=2)+ geom_smooth( method="lm", se=F) +geom_point() +
  scale_x_continuous(minor_breaks = 1976:2020, position = "top") + scale_y_continuous(n.breaks=10)+
  scale_color_grey(labels=function(x) str_to_sentence(str_replace_all(x, "_", " ")), start=0, end=0.85) +
  labs(x="Year", y="Day of year", color="Threshold") + theme_minimal()

cat("SD of first bare ground =", round(sd(groundcover$first_0_cm_day),1),"days over",range(groundcover$year))
summary(lm(first_0_cm_day ~ year, data=groundcover))

treatments %>% ggplot(aes(x=sun_date, shape=year, color=water, y=precip_est_mm))+geom_count()+
  geom_point(data=groundcover, aes(x=first_0_cm_day, color="Control", shape="1990-2017"), fill="white") +
  scale_color_manual(values=water_pal) + scale_shape_manual(values=c(21, 19, 17, 15))+ coord_cartesian(ylim=c(0,NA))+
  labs(y="Estimated summer precipiation with treatments (mm)", x="Snowmelt date", color="Precipitation", shape="Year",size="# Subplots") + 
  theme_minimal()
```


# Soil moisture vs. treatments

## Weather plots

```{r soil, fig.height=12, fig.width=12}
ggplot(sm, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
  geom_jitter(aes(shape=snow, color=water, y = VWC), height=0, width=0.3) + 
  geom_smooth(aes(linetype=snow, color=water, y = VWC), se=F, span=0.5) +
  geom_col(data=daily_all %>% filter(yr%in%2018:2020), aes(x=yday(date), y=EPA_NOAA_filled), 
           fill=brewer.pal(3, "Greens")[3]) +
  geom_line(data=daily_all %>% filter(yr%in%2018:2020), aes(x=yday(date), y=av_temp_2m_C), 
            color=brewer.pal(3, "Oranges")[3]) + 
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color=brewer.pal(3, "Purples")[3])+
  geom_line(data = snowcloth, aes(y=mean_snow_depth_cm/10, group=year))+
  geom_point(data = snowcloth, aes(y=mean_snow_depth_cm/10))+
  geom_line(data=groundcover %>% pivot_longer(ends_with("day"), names_to="threshold", values_to="day") %>%
              filter(threshold!="first_snow_day", year>2017) %>% 
              mutate(snow_depth_cm = as.integer(str_split_fixed(threshold,"_",n=4)[,2])),
            aes(x=day, y=snow_depth_cm/10), color="grey40")+
  geom_segment(data=waterdates %>% pivot_wider(names_from=precip_treatments, values_from=c(day,date)),
               aes(x=day_started, xend=day_ended, y=20, yend=20)) +
  geom_point(data=waterdates, aes(x=day,y=20)) +
  geom_point(data=ph, aes(y= 17*(open+buds)/(open+buds)), color="grey", shape=0)+
  geom_point(data=mt, aes(y= 18*corolla_length/corolla_length), color="purple", shape=0)+
  geom_point(data=mt, aes(y= 19*nectar_24_h_ul/nectar_24_h_ul), color="orange", shape=0)+
  scale_x_continuous(limits=c(85, 230), breaks=scales::breaks_extended(10), expand=expansion(0,0)) + scale_y_continuous(limits=c(-3, 20)) +
  labs(y="Soil VWC (%) / Daily precipitation (mm) / Mean temp. (C) / Snowpack (dm)", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_color_manual(values=water_pal) +
  scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(sm.subplot, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
      geom_col(data=daily_all %>% filter(yr%in%2018:2020), aes(x=yday(date), y=EPA_NOAA_filled), 
           fill=brewer.pal(3, "Greens")[2]) +
  geom_line(data=daily_precip_est, aes(y=VWC_Control_4*100), color=brewer.pal(3, "Greens")[2], size=1) +
  geom_point(aes(shape=snow, color=water, y = VWC)) + 
  geom_line(data=sm.subplot %>% group_by(water, snow, year, date) %>% summarize(VWC=mean(VWC, na.rm=T)), aes(linetype=snow, color=water, y = VWC)) +
    scale_color_manual(values=water_pal) +  scale_linetype_manual(values=c(2,1)) + 
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color="black", show.legend=F) +
  coord_cartesian(xlim=c(110, 230), ylim=c(0, round(max(sm.subplot$VWC, na.rm=T))), expand=F)+
    labs(x="Day of year", y="Mean soil VWC in subplot (%) / Precipitation (mm)", color="Water", shape="Snowmelt", linetype="Snowmelt") + theme_minimal() + theme(legend.key.width = unit(2, "lines"))
```

# Soil moisture vs. snowmelt date

```{r soil_treatments}
left_join(filter(sm.subplot20d, mo20d!="221 - 240"), meltdates) %>% 
  mutate(mo = paste0("Days ",mo20d)) %>% 
  ggplot(aes(x=sun_date, y=VWC, shape=year,color=water)) + 
  facet_wrap(vars(mo), nrow=1)+
  geom_point() + geom_smooth(se=F, method="lm") +
  scale_color_manual(values=water_pal) +
  labs(x="Snowmelt date", y="Mean soil moisture in subplot", color="Water", shape="Year")

sm.subplot %>% mutate(day=yday(date)) %>% group_by(year) %>% summarize(start=min(day), end=max(day)) %>% kable(caption = "Soil moisture collection dates")

# VWC split-plot analysis
sm.mod <- lmer(VWC ~ year * snow * water + (1|plot), data=sm.subplotyear)
anova(sm.mod) %>% kable(digits=3)
(sm.emm.snow <- as.data.frame(emmeans(sm.mod, ~ snow))) %>% kable(digits=3)
sm.emm.snow$emmean[1] - sm.emm.snow$emmean[2]
(sm.emm.water <- as.data.frame(emmeans(sm.mod, ~ water))) %>% kable(digits=3)
sm.emm.water$emmean[1] - sm.emm.water$emmean[2]
sm.emm.water$emmean[3] - sm.emm.water$emmean[2]
```

## Soil moisture at plant vs. subplot level
```{r sm_phys_compare}
ggplot(pt, aes(x=VWC.sm, y=VWC.plant, shape=year, group=year, color=as.integer(date_diff))) + geom_smooth(se=F) +geom_point() + geom_abline(intercept=0, slope=1) + 
  labs(x="Subplot VWC at closest date", y="VWC measured at plant for physiology", color="Date difference", shape="Year") +
  scale_color_viridis_c() + theme_minimal() + coord_fixed()

ggplot(pt, aes(x=VWC.sm.year, y=VWC.sm, shape=year, group=year, color=year)) + geom_smooth(se=F) +geom_point() + geom_abline(intercept=0, slope=1) + 
  labs(x="Mean subplot VWC across year", y="Subplot VWC at closest date", color="Year", shape="Year") +
  scale_color_manual(values=year_pal) + theme_minimal() + coord_fixed()

ggplot(pt, aes(x=VWC.sm.year, y=VWC.plant, shape=year, group=year, color=year)) + geom_smooth(se=F) +geom_point() + geom_abline(intercept=0, slope=1) + 
  labs(x="Mean subplot VWC across year", y="VWC measured at plant for physiology", color="Year", shape="Year") +
  scale_color_manual(values=year_pal) + theme_minimal() + coord_fixed()
```


# Timing of measurements

```{r timings}
ggplot(timings %>% filter(variable!="sds") %>% mutate(variable=fct_reorder(variable, begin, na.rm=T, .desc=T)), aes(y=variable, color=year))+ 
  geom_linerange(aes(xmin=begin, xmax=end), 
                 position = position_dodge2(width=0.7, reverse = T), size=1.2, show.legend=FALSE) + 
  #geom_text(aes(x=begin-10, label=begin), position = position_dodge(width=0.7), fontface="bold")+
  #geom_text(aes(x=end+10, label=end), position = position_dodge(width=0.7), fontface="bold") +
  geom_text(data=timings %>% drop_na(plots) %>% filter(year=="2019"),
            aes(x = end+7, label=paste(ifelse(nchar(plots)>1, "Plots","Plot"), str_replace(plots,","," & "))), 
            position = position_dodge2(width=0.8, reverse = T), size=4, hjust=0, show.legend=FALSE)+
  geom_point(aes(x=begin), position = position_dodge2(width=0.7, reverse = T), shape=15, size=2)+
  geom_point(aes(x=end), position = position_dodge2(width=0.7, reverse = T), shape=15, size=2) +
  scale_color_manual("Year", values=year_pal, guide=guide_legend(override.aes = list(size=5))) + 
  scale_y_discrete("", labels=c(snowcloth="Snow cloths applied to early plots", meltdates="Mean snowmelt timing (early plots - unmanipulated plots)", waterdates="Summer precipitation treatments applied",sm="Soil moisture recorded",ph="Inflorescence height recorded", mt="Floral morphology recorded", nt="Nectar traits recorded", sds="Flower collection period"))+
  scale_x_continuous("Day of year", breaks=seq(80,240, by=20)) +
  theme_minimal() + theme(legend.position = "top")
```

## Sample sizes

```{r sample_sizes}
cat("Morphology measurements per plant:", mt %>% drop_na(corolla_length) %>% group_by(year, plantid) %>% tally() %>% ungroup() %>% summarize(n=mean(n)) %>% pull(n) %>% round(2))
cat("Nectar measurements per plant:", mt %>% drop_na(nectar_24_h_ul) %>% group_by(year, plantid) %>% tally() %>% ungroup() %>% summarize(n=mean(n)) %>% pull(n) %>% round(2))
cat("Inflo height measurements per plant:")
ph %>% drop_na(height_cm) %>% group_by(year, plantid) %>% tally() %>% summarize(n=mean(n)) %>% kable
ph %>% drop_na(height_cm) %>% group_by(year, plantid) %>% tally() %>% 
  ggplot(aes(x=n, fill=year)) + geom_histogram() + labs(title = "Phenology entries per plant", x="Entries",y="Count")+
  facet_wrap(vars(year), ncol=1) + scale_fill_manual(values=year_pal)


#Total sample sizes
mt %>% drop_na(corolla_length) %>% group_by(year, plantid) %>% tally() %>% ungroup %>% tally() %>% pull(n)
mt %>% drop_na(corolla_length, anther_min) %>% group_by(year, plantid) %>% tally() %>% ungroup %>% tally() %>% pull(n)
mt %>% drop_na(nectar_24_h_ul) %>% group_by(year, plantid) %>% tally() %>% ungroup %>% tally() %>% pull(n)
sds %>% drop_na(flowers_est) %>% group_by(year, plantid) %>% tally() %>% ungroup %>% tally() %>% pull(n)
ph %>% drop_na(height_cm) %>% group_by(year, plantid) %>% tally() %>% ungroup %>% tally() %>% pull(n)
```


# Census

## Census status

The annual census at the beginning of the summer contains information on mortality, flowering status, and plant size (longest leaf and number of leaves for each rosette).

Visualize a plant's status in each census

```{r status}
ggplot(cen.size.long %>% filter(census!="2018r2"), aes(x=census, y=n_rosettes, group=plantid))+geom_path()+ geom_point() + scale_y_continuous(breaks=0:10)

ggplot(cen.status.long, aes(y=plantid, x=census, fill=status)) + geom_tile() +
  scale_fill_manual(values=brewer.pal(8, name="Set1")[c(1,3,5,2,7,6)])+
  theme(axis.text.y=element_blank(), axis.ticks = element_blank()) +
  labs(fill="Status", x="Census year",y="Plant ID")

ggplot(megatally.status.long, aes(y=plantid, x=dataset, fill=status)) + geom_tile() +
  scale_fill_manual(values=c(brewer.pal(5, name="Set1")[c(1,3,5,2)],brewer.pal(8, "Set2")[7:8]))+
  scale_x_discrete(position="top") +
  theme(axis.text.y=element_blank(), axis.ticks = element_blank(), axis.text.x = element_text(angle=90, size=14)) + 
  labs(fill="Status", x="Dataset and year",y="Plant ID")
```

## Status vs. treatment
```{r inventory}
cen.status.long %>% left_join(treatments) %>% drop_na(water, snow) %>% filter(status %in% status_ok, census !="2018r2") %>% 
  ggplot(aes(x=water, fill=status)) + facet_grid(snow ~ census, labeller=labeller(.cols=label_both)) + geom_bar(position="fill") +
  scale_fill_manual(values=status_pal) + labs(x="Precipitation", y="Proportion", fill="Status")
```

## Status transitions vs. treatment

What did plants of a given status in one year look like in the next?
```{r transitions}
with(cen.status, table(`2018`=paste(status_18), `2019`=paste(status_19)))

cen.status %>% mutate(year = "2018") %>% left_join(treatments) %>% filter(status_18 %in% status_ok[-3], status_19 %in% status_ok) %>% 
ggplot(aes(x=water, fill=status_19)) + facet_wrap(status_18 ~ snow, labeller=labeller(.cols=label_both)) +  
  labs(y="2019 Status", fill="2019 Status", x="Precipitation") +
  geom_bar(position="fill") + scale_fill_manual(values=status_pal) + theme_minimal()

with(cen.status, table(`2019`=paste(status_19), `2020`=paste(status_20)))

cen.status %>% mutate(year = "2019") %>% left_join(treatments) %>% filter(status_19 %in% status_ok[-3], status_20 %in% status_ok) %>% 
ggplot(aes(x=water, fill=status_20)) + facet_wrap(status_19 ~ snow, labeller=labeller(.cols=label_both)) +  
  labs(y="2020 Status", fill="2020 Status", x="Precipitation") +
  geom_bar(position="fill") + scale_fill_manual(values=status_pal) + theme_minimal()

#Model effect of treatments on survival and flowering of vegetative plants - both years combined
cen.status.veg1819 <- bind_rows(
  `2018`=cen.status %>% filter(status_18=="vegetative") %>% drop_na(alive_status_19, flowering_status_19) %>% mutate(year="2018") %>%
    left_join(treatments) %>% select(alive_status = alive_status_19, flowering_status = flowering_status_19, water, snow, plot, plotid),
  `2019`=cen.status %>% filter(status_19=="vegetative") %>% drop_na(alive_status_20, flowering_status_20) %>% mutate(year="2019") %>%
    left_join(treatments) %>% select(alive_status = alive_status_20, flowering_status = flowering_status_20, water, snow, plot, plotid), .id="year")
# could now also use cen.status.long %>% filter(status=="vegetative", census %in% c("2018","2019"))

library(glmmTMB)
options(contrasts = c("contr.sum", "contr.poly"))

(survival1819.mod <- glmmTMB(alive_status ~ year * water * snow + (1|plot) + (1|plotid), data=cen.status.veg1819, family=binomial))
car::Anova(survival1819.mod, type=3) %>% kable(digits=3)
emmeans(survival1819.mod, ~ year * water * snow) %>% as_tibble %>% 
  mutate(prob=plogis(emmean), uSE = plogis(emmean+SE), lSE = plogis(emmean-SE)) %>% 
  ggplot(aes(x=water, color=snow, y=prob))+ facet_wrap(vars(year))+
  geom_pointrange(aes(ymin=lSE, ymax=uSE), position=position_dodge2(0.9)) + 
  labs(x="Precipitation", y="Survival to next year", color="Snowmelt")+
  scale_y_continuous(limits=c(0,1)) + scale_color_manual(values=snow_pal) + theme_minimal()

(flowering1819.mod <- glmmTMB(flowering_status ~ year * water * snow + (1|plot) + (1|plotid), data=cen.status.veg1819, family=binomial))
car::Anova(flowering1819.mod, type=3) %>% kable(digits=3)
emmeans(flowering1819.mod, ~ year * water * snow) %>% as_tibble %>% 
  mutate(prob=plogis(emmean), uSE = plogis(emmean+SE), lSE = plogis(emmean-SE)) %>% 
  ggplot(aes(x=water, color=snow, y=prob))+ facet_wrap(vars(year))+
  geom_pointrange(aes(ymin=lSE, ymax=uSE), position=position_dodge2(0.9)) + 
  labs(x="Precipitation", y="Flowering the next year", color="Snowmelt")+
  scale_y_continuous(limits=c(0,1)) + scale_color_manual(values=snow_pal) + theme_minimal()
```

## Survival and flowering vs. vegetative plant size
```{r surv_size}
ggplot(filter(cen, status_19 %in% status_ok, status_18 == "vegetative"), aes(x=r1_longest_18, fill=status_19)) +
  facet_wrap(vars(status_19), ncol=1) + labs(x="2018 Rosette 1 Longest Leaf (mm)", fill="2019 status") +
  geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 85))
ggplot(filter(cen, status_20 %in% status_ok, status_19 == "vegetative"), aes(x=r1_longest_19, fill=status_20)) +
  facet_wrap(vars(status_20), ncol=1) + labs(x="2019 Rosette 1 Longest Leaf (mm)", fill="2020 status") +
   geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 85))

ggplot(filter(cen, status_19 %in% status_ok, status_18 == "vegetative"), aes(x=r1_leaves_18, fill=status_19)) +
  facet_wrap(vars(status_19), ncol=1) + labs(x="2018 Rosette 1 # Leaves", fill="2019 status") +
  geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 100))
ggplot(filter(cen, status_20 %in% status_ok, status_19 == "vegetative"), aes(x=r1_leaves_19, fill=status_20)) +
  facet_wrap(vars(status_20), ncol=1) + labs(x="2019 Rosette 1 # Leaves", fill="2020 status") +
  geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 100))
```

## Plant growth vs. treatments

Only includes plants that remained vegetative and did not die or flower.
```{r growth}
cen %>% drop_na(water_18, snow_18) %>% filter(status_18=="vegetative", status_19=="vegetative") %>% 
  ggplot(aes(x=r1_longest_18, y=r1_longest_19, color=water_18, shape=snow_18)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+coord_fixed(xlim=c(0,80), ylim=c(0,80)) +
  labs(x="2018 Rosette 1 Longest Leaf (mm)", y="2019 Rosette 1 Longest Leaf (mm)",  color="Precipitation", shape="Snowmelt")
cen %>% drop_na(water_19, snow_19) %>% filter(status_19=="vegetative", status_20=="vegetative") %>% 
  ggplot(aes(x=r1_longest_19, y=r1_longest_20, color=water_19, shape=snow_19)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+  coord_fixed(xlim=c(0,80), ylim=c(0,80)) +
    labs(x="2019 Rosette 1 Longest Leaf (mm)", y="2020 Rosette 1 Longest Leaf (mm)",  color="Precipitation", shape="Snowmelt")

cen %>% drop_na(water_18, snow_18) %>% filter(status_18=="vegetative", status_19=="vegetative") %>% 
  ggplot(aes(x=r1_leaves_18, y=r1_leaves_19, color=water_18, shape=snow_18)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+coord_fixed(xlim=c(0,100), ylim=c(0,100)) +
  labs(x="2018 Rosette 1 # Leaves", y="2019 Rosette 1 # Leaves",  color="Precipitation", shape="Snowmelt")
cen %>% drop_na(water_19, snow_19) %>% filter(status_19=="vegetative", status_20=="vegetative") %>% 
  ggplot(aes(x=r1_leaves_19, y=r1_leaves_20, color=water_19, shape=snow_19)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+  coord_fixed(xlim=c(0,100), ylim=c(0,100)) +
    labs(x="2019 Rosette 1 # Leaves", y="2020 Rosette 1 # Leaves",  color="Precipitation", shape="Snowmelt")
```

## Plant size vs. treatments
```{r plantsize}
ggplot(drop_na(cen18, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2018 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen18, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal)+ labs(x="Precipitation", y="2018 Rosette 1 # Leaves",  color="Snowmelt")

ggplot(drop_na(cen19, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2019 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen19, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal)+ labs(x="Precipitation", y="2019 Rosette 1 # Leaves",  color="Snowmelt")

ggplot(drop_na(cen20, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2020 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen20, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2020 Rosette 1 # Leaves",  color="Snowmelt")
```

# Leaf traits

```{r checkleaftraits, fig.height=8, fig.width=6}
with(lt, table(date, plot)) %>% kable

lt %>% group_by(date) %>% mutate(index=row_number(), date=factor(date)) %>% 
ggplot(aes(x=index, y=water_content, color=factor(date), group=factor(plot))) + 
  facet_wrap(vars(paste(year, round, sep=" round "), date)) +
  geom_hline(yintercept=0.6) + geom_hline(yintercept=0.85) + 
  geom_point() + geom_smooth(method="lm", color="black", se=F) + 
  scale_color_brewer(palette="Set1", direction=-1, guide=F) +
  labs(x="Measurement order", y="Water content", color="Plot")

lt %>% pivot_longer(c(trichomes, trichome_density, water_content, sla_wet, sla, leaf_area_cm2), names_to="trait") %>% 
  ggplot(aes(x=factor(date), y=value, color=factor(date))) + facet_wrap(vars(trait), scales="free_y") + geom_boxplot() +
  scale_x_discrete(guide=guide_axis(angle=90)) + labs(x="Date", y="", color="Date") + 
  scale_color_brewer(palette="Set1", direction = -1, guide=F) + theme(legend.position = "top")

ggplot(lt, aes(x=wet_weight_g, y=dry_weight_g, color=factor(date))) + geom_point() + geom_abline(slope=1, intercept = 0) +
  geom_abline(slope=1-median(lt$water_content, na.rm=T), intercept = 0) + geom_smooth(se=F) + 
  coord_cartesian(xlim=c(0,NA), ylim=c(0, NA)) + labs(x="Wet weight (g)", y = "Dry weight (g)", color="Date") +
    scale_color_brewer(palette="Set1", direction=-1) + theme(legend.position="top")

lt %>% pivot_longer(c(wet_weight_g, dry_weight_g, trichomes), names_to="trait") %>% 
  ggplot(aes(x=leaf_area_cm2, y=value, color=factor(date))) + facet_wrap(vars(trait), scales="free_y", ncol=1) + 
  geom_point(size=0.7) + geom_smooth(se=F) + labs(x="Leaf area (cm2)", y="", color="Date") + 
  scale_color_brewer(palette="Set1", direction=-1)

lt %>% pivot_longer(c(trichome_density, sla_wet, sla, water_content), names_to="trait") %>% 
  ggplot(aes(x=leaf_area_cm2, y=value, color=factor(date))) + facet_wrap(vars(trait), scales="free_y") + 
  geom_point(size=0.7) + geom_smooth(se=F) +   labs(x="Leaf area (cm2)", y="", color="Date") + 
  scale_color_brewer(palette="Set1", direction=-1) +
  theme(legend.position = "top")

library(ggfortify) 
lt %>%  mutate(id = paste0(str_sub(year(date),3,4), month(date), day(date), "-",plantid),
               id = ifelse(id=="20728-3C162" & trichomes==120, "20728-3C162B", id)) %>%  #two leaves collected
  select(id, wet_weight_g, dry_weight_g, leaf_area_cm2) %>% column_to_rownames("id") %>% drop_na() %>% 
  prcomp(scale=T) %>% 
  autoplot(data = lt %>% mutate(date=factor(date)) %>% drop_na(wet_weight_g, dry_weight_g, leaf_area_cm2),
           colour = "date", 
           loadings = TRUE, loadings.colour="black", loadings.label = TRUE, loadings.label.colour="black",
           shape=F, frame=T) + theme(legend.position = "top") +
  scale_color_brewer("Date", palette="Set1", direction=-1) + scale_fill_brewer("Date",palette="Set1", direction=-1)
```

# Physiology traits

```{r licor_all, fig.width=12, fig.height=12}
# plot all the LICOR variables by date
licor_date_pal <- sample(rainbow(nlevels(factor(licor$date))))
licor %>% pivot_longer(Photo:StableF) %>% mutate(index=row_number()) %>% filter(!name %in% c("Area","BLC_1","BLCond","StmRat")) %>% 
  ggplot(aes(x=index, y= value, color=factor(date), shape=year)) + facet_wrap(vars(name), scales="free_y", ncol=4) + geom_point() + scale_color_manual(values=licor_date_pal, guide=F) + theme_dark() + labs(x="Measurement", y="", shape="Year")
```

```{r licor}
licor %>% ggplot(aes(x=Cond, y= Photo, color=Ci>0, shape=year)) + geom_point()

licor %>% filter(Ci>0) %>%  ggplot(aes(x=Cond, y= Photo, color=factor(date), linetype=year, shape=year)) + geom_point() + geom_smooth(method="lm", se=F) + scale_linetype_manual(values=c(3,1,4)) + scale_color_manual(values=licor_date_pal, guide=F) + theme_dark()

licor %>% filter(Ci>0) %>% ggplot(aes(y=Photo/Cond, x=factor(date), color=year)) + geom_boxplot(size=1.5) + 
 scale_color_manual(values=year_pal) + scale_x_discrete(guide = guide_axis(angle = 90)) + theme_minimal()

licor %>% filter(Ci>0) %>% left_join(treatments)%>% ggplot(aes(y=Photo/Cond, x=factor(date), color=water)) + 
  geom_boxplot(position=position_dodge2(preserve="single")) + 
 scale_color_manual(values=water_pal) + scale_x_discrete(guide = guide_axis(angle = 90)) + theme_minimal()

pt %>% ggplot(aes(x=conductance, y= photosynthesis, color=factor(date), linetype=year, shape=year)) + geom_point() + geom_smooth(method="lm", se=F) + scale_linetype_manual(values=c(3,1,4)) + scale_color_manual(values=licor_date_pal, guide=F) + theme_dark()

pt %>% select(date, plantid, photosynthesis, conductance, WUE, notes) %>% 
  filter(is.na(conductance)) %>% mutate(conductance = photosynthesis/WUE) %>% kable
```
These conductance entries were blank, filled from WUE. There is another entry with conductance = 1.23 that was not blanked.

# Floral traits

## Floral trait correlations
```{r correlations}
mt %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width) %>% 
  ggpairs()

nt %>% select(nectar_24_h_ul, nectar_conc, nectar_sugar_24_h_mg) %>% 
  ggpairs()

ggplot(nt, aes(x=nectar_24_h_ul, y = nectar_conc)) + geom_point() + labs(x=traitnames[["nectar_conc"]], y=traitnames[["nectar_24_h_ul"]]) + geom_smooth(se=F)

mnps.plantyr %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width, nectar_24_h_ul, nectar_conc, nectar_sugar_24_h_mg, height_cm, flowers_est) %>% 
  ggcorr(hjust=0.85, layout.exp=2, label=T, label_round=2)

mnps.plantyr %>% select(all_of(seedtraits[c(1,7,2:6)])) %>% 
  ggcorr(hjust=0.85, layout.exp=2, label=T, label_round=2)
```

# Models

## Plasticity: water and snowmelt treatments

This is a split-plot analysis, based on the original experimental design. The models use:

* *mnps.plantyr*, the plant means within each year
* *snow*, the snowmelt treatment (except for plot 2 in 2019)
* *water*, the water treatments with mock rainouts and controls combined
* *plot* as a random effect (subplots vary within this larger unit)
* *plotid* as a random effect (plants vary within this larger unit)

The models are specified as:

    lmer(trait ~ year * snow * water + (1|plot) + (1|plotid) +, data = mnps.plantyr)

```{r splitplot}
lt.subplot.r2 <- lt.subplotround %>% filter(round=="2")
lt.plantyr.r2 <- lt.plantyr %>% filter(round=="2")

options(contrasts = c("contr.sum", "contr.poly"))
mt.mod.split <- map(set_names(traits), 
                    ~ lmer(mnps.plantyr[[.x]]  ~ year * snow * water + (1|plot) + (1|plotid), data=mnps.plantyr))
lt.mod.split <- map(set_names(leaftraits), 
                    ~ lmer(lt.plantyr.r2[[.x]] ~ year * snow * water + (1|plot) + (1|plotid), data=lt.plantyr.r2))
pt.mod.split <- map(set_names(phystraits), 
                    ~ lmer(pt.plantyr[[.x]]    ~ year * snow * water + (1|plot) + (1|plotid) + (1|round), data=pt.plantyr))

mt.mod.split.coefs <- map_dfr(mt.mod.split, tidy, .id="trait")
lt.mod.split.coefs <- map_dfr(lt.mod.split, tidy, .id="trait")
pt.mod.split.coefs <- map_dfr(pt.mod.split, tidy, .id="trait")

mt.mod.split.tests <- map(mt.mod.split, anova) %>% map_dfr(tidy, .id="trait")
lt.mod.split.tests <- map(lt.mod.split, anova) %>% map_dfr(tidy, .id="trait")
pt.mod.split.tests <- map(pt.mod.split, anova) %>% map_dfr(tidy, .id="trait")

mt.mod.split.emm <- map(mt.mod.split, ~ summary(emmeans(ref_grid(.x), ~ year * snow * water)))
lt.mod.split.emm <- map(lt.mod.split, ~ summary(emmeans(ref_grid(.x), ~ year * snow * water)))
pt.mod.split.emm <- map(pt.mod.split, ~ summary(emmeans(ref_grid(.x), ~ year * snow * water)))
```

### ANOVA tables

```{r anova_splitplot}
print_anova_table(mt.mod.split.tests, mt.mod.split)
print_anova_table(lt.mod.split.tests, lt.mod.split)
print_anova_table(pt.mod.split.tests, pt.mod.split)

```

```{r re_splitplot}
re_table <- function(coefs) {
  coefs %>% filter(effect=="ran_pars") %>% select(trait, group, estimate) %>% pivot_wider(names_from=group, values_from=estimate)
  }
map_dfr(list(mt.mod.split.coefs, lt.mod.split.coefs, pt.mod.split.coefs), re_table) %>% kable(digits=3)
```

### Plots

```{r multi_splitplot, fig.width=10, fig.height=11}
plot_split_plot <- function(emm, data, traits.plot, geom = "point", plot.emm = TRUE) {
  
  data.long <- data %>% select(all_of(traits.plot), water, snow, year, sun_date, precip_est_mm) %>% 
    pivot_longer(all_of(traits.plot), names_to="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal")) %>% 
    drop_na(value, water) %>% filter(trait %in% traits.plot)
  
  traitnames.multiline <- ifelse(nchar(traitnames) > 30, str_replace(traitnames, fixed("("),"\n("), traitnames)
  split_plot <- emm %>% bind_rows(.id="trait") %>% filter(trait %in% traits.plot) %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal")) %>% 
    ggplot(aes(x=water, color=snow)) + 
    facet_grid(trait ~ year, scales="free_y", switch="y",
               labeller = as_labeller(c(traitnames.multiline, set_names(2018:2020)))) + 
    labs(x="Precipitation", y="", color="Snowmelt") + 
    scale_y_continuous(position="right")+ scale_x_discrete(guide=guide_axis(angle=90)) + 
    scale_color_manual(values=snow_pal, guide=guide_legend(override.aes = list(shape=15, size=ifelse(plot.emm,5,1)))) + 
    theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"), 
                            panel.spacing=unit(0, "pt"), plot.margin = margin(0,0,0,0, "pt"), legend.position = "top") +
    switch(geom,
           boxplot = geom_boxplot(data=data.long, aes(y=value), position=position_dodge(width=0.8), show.legend=!plot.emm,
                                  fatten=ifelse(plot.emm, NULL, 1), outlier.size=0.5),
           violin =   geom_violin(data=data.long, aes(y=value), position=position_dodge(width=0.8), show.legend=F),
           point =     geom_point(data=data.long, aes(y=value), position=position_dodge(width=0.8), shape=3))
  
  if(plot.emm) { return(split_plot + 
           geom_linerange(aes(ymin=emmean-SE, ymax=emmean+SE,), position=position_dodge(width=0.8), size=1, show.legend=F) +
           geom_point(aes(y=emmean), position=position_dodge(width=0.8), shape="-", size=16))
  } else return(split_plot + coord_cartesian(ylim=c(0,NA)))
}

grid.arrange( #Morphological traits - boxplot of raw data
  plot_split_plot(mt.mod.split.emm, mnps, traits[1:4], geom="boxplot", plot.emm=F), 
  plot_split_plot(mt.mod.split.emm, mnps, traits[c(5,6,8,13)], geom="boxplot", plot.emm=F)+ guides(color=F), nrow=1)

grid.arrange( #Morphological traits - plot of EMMs and subplot means
  plot_split_plot(mt.mod.split.emm, mnps.subplot, traits[1:4]), 
  plot_split_plot(mt.mod.split.emm, mnps.subplot, traits[c(5,6,8,13)])+ guides(color=F), nrow=1)

grid.arrange( #Vegetative and physiological traits - plot of raw data
  plot_split_plot(lt.mod.split.emm, lt %>% filter(round==2), leaftraits, geom="boxplot", plot.emm=F), #second round only
  plot_split_plot(pt.mod.split.emm, pt, phystraits, geom="boxplot", plot.emm=F) + guides(color=F), nrow=1)

grid.arrange( #Vegetative and physiological traits - plot of EMMs and subplot means
  plot_split_plot(lt.mod.split.emm, lt.subplotround %>% filter(round==2), leaftraits), #second round only
  plot_split_plot(pt.mod.split.emm, pt.subplot, phystraits)+ guides(color=F), nrow=1)
```

### Compensation of water addition for early snowmelt

```{r compensation}
#compare early snowmelt + addition vs. normal snowmelt + control
emm.mod <- mt.mod.glht <- list()
mnps.plantyr$snowwater <- paste0(mnps.plantyr$snow, mnps.subplot$water)
library(multcomp)#careful, this loads MASS, which overrides dplyr::select
for(tr in traits) {
    #emm.mod[[tr]] <- emmeans(mt.mod[[tr]], ~ water * snow, var=tr)
    mt.mod.glht[[tr]] <- lmer(mnps.plantyr[[tr]] ~ snowwater * year+ (1|plot) + (1|plotid), data=mnps.plantyr) %>%
      glht(linfct = mcp(snowwater = c("EarlyAddition - NormalControl = 0")))
}
map_dfr(mt.mod.glht, ~ tibble(EarlyAddition_minus_NormalControl = coef(.x), p=summary(.x)$test$pvalues), .id="trait") %>% kable(digits=3)
detach("package:multcomp", unload = TRUE); detach("package:TH.data", unload = TRUE); detach("package:MASS", unload = TRUE)
```

## Plasticity: absolute snowmelt date and estimated summer precipitation

```{r meltdate}
lt.subplot.r2 <- lt.subplotround %>% filter(round=="2")
lt.plantyr.r2 <- lt.plantyr %>% filter(round=="2")

options(contrasts = c("contr.sum", "contr.poly"))
mt.mod.abs <- map(set_names(traits), 
                  ~ lmer(mnps.plantyr[[.x]]  ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), data=mnps.plantyr))
lt.mod.abs <- map(set_names(leaftraits), 
                  ~ lmer(lt.plantyr.r2[[.x]] ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), data=lt.plantyr.r2))
pt.mod.abs <- map(set_names(phystraits), 
                  ~ lmer(pt.plantyr[[.x]]    ~ sun_date * precip_est_mm + (1|plot) + (1|plotid) + (1|round), data=pt.plantyr))

mt.mod.abs.coefs <- map_dfr(mt.mod.abs, tidy, .id="trait")
lt.mod.abs.coefs <- map_dfr(lt.mod.abs, tidy, .id="trait")
pt.mod.abs.coefs <- map_dfr(pt.mod.abs, tidy, .id="trait")

mt.mod.abs.tests <- map(mt.mod.abs, anova) %>% map_dfr(tidy, .id="trait")
lt.mod.abs.tests <- map(lt.mod.abs, anova) %>% map_dfr(tidy, .id="trait")
pt.mod.abs.tests <- map(pt.mod.abs, anova) %>% map_dfr(tidy, .id="trait")
```

### ANOVA tables

```{r anova_meltdate}
print_anova_table(mt.mod.abs.tests, mt.mod.abs)
print_anova_table(lt.mod.abs.tests, lt.mod.abs)
print_anova_table(pt.mod.abs.tests, pt.mod.abs)
```

```{r re_meltdate}
map_dfr(list(mt.mod.abs.coefs, lt.mod.abs.coefs, pt.mod.abs.coefs), re_table) %>% kable(digits=3)
```

### Plots

```{r multi_meltdate, fig.width=9, fig.height=11}
plot_abs <- function(mod, data, traits.plot) {
  data.long <- data %>% select(all_of(traits.plot), water, snow, year, sun_date, precip_est_mm) %>% 
    pivot_longer(all_of(traits.plot), names_to="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal")) %>% 
    drop_na(value, water) %>% filter(trait %in% traits.plot)
  
  emm.grid <- mod[traits.plot] %>% 
    map_dfr(~ emmeans(.x, ~ precip_est_mm * sun_date, at=list(sun_date=range(treatments$sun_date), precip_est_mm=seq(25,150, by=25))) %>% as_tibble(), .id="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot))
  
  traitnames.multiline <- ifelse(nchar(traitnames) > 30, str_replace(traitnames, fixed("("),"\n("), traitnames)
  
  abs_plot <- ggplot(emm.grid, aes(x=sun_date, y=emmean, color=precip_est_mm, group=precip_est_mm)) + 
    facet_wrap(vars(trait), scales="free_y",labeller = as_labeller(traitnames.multiline))+
    geom_point(data = data.long, aes(y=value, shape=year), size=2) + geom_line() + 
    scale_color_gradientn(colors=rev(water_pal), 
                          values=rev(c(1, (treatments %>% group_by(water) %>% summarize(across(precip_est_mm, mean)) %>%
                                             pull(precip_est_mm) %>% scales::rescale(from=c(25,150)))[2],0)),
                          breaks=seq(25,150, by=25)) + 
    labs(x="Snowmelt date", y="", color="Summer\nprecipitation\n(mm)", shape="Year") + 
    #coord_cartesian(ylim=c(0,NA)) + scale_y_continuous(expand=expansion(mult=c(0,0.05)))+
    theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                            panel.border = element_rect(fill=NA, colour = "black"))
  return(abs_plot)
}

plot_abs(mt.mod.abs, mnps, traits[c(1:4,5,6,8,13)]) #Morphological traits - raw data
plot_abs(mt.mod.abs, mnps.subplot, traits[c(1:4,5,6,8,13)]) #Morphological traits - subplot means

grid.arrange(plot_abs(lt.mod.abs, lt, leaftraits), 
             plot_abs(pt.mod.abs, pt, phystraits) + guides(color=F, shape=F), ncol=1) #raw data
grid.arrange(plot_abs(lt.mod.abs, lt.subplot.r2, leaftraits),
             plot_abs(pt.mod.abs, pt.subplot, phystraits)+ guides(color=F, shape=F), ncol=1) #subplot means

```

### Slope plots and compensation

```{r compensation_meltdate}

# library(googlesheets4)
# write_sheet(left_join(emm.meltprecip.df %>% mutate(traitname=traitnames[trait], .before="trait"),
#                        mnps.plantyr %>% group_by(water) %>% select(traits) %>% 
#                          summarize(across(traits, ~ sum(!is.na(.x)))) %>% 
#                          pivot_longer(traits, names_to="trait", values_to="n"))
#              , ss=filter(datasheets, name=="Maxfield Results"), sheet="slopes_melt")

#get slopes with emtrends
mt.mod.abs.emm <- map(set_names(c("sun_date", "precip_est_mm")), ~ map(mt.mod.abs, emtrends, ~ 1, var = .x) %>% 
      map(as.data.frame) %>% set_names(traits) %>% bind_rows(.id="trait") %>% rename_with(.cols=ends_with("trend"), ~ "trend")) %>% 
  bind_rows(.id="predictor") %>% mutate(predictor=fct_relevel(predictor,"sun_date"), 
                                        sd.plantyr = map_dbl(trait, ~sd(mnps.plantyr[[.x]], na.rm=T)),
                                        trend = trend / sd.plantyr,SE = SE / sd.plantyr) %>% #Scale slope by SD of the plant means
    filter(trait %in% traits[c(1:8, 13)]) # traits for floral plasticty paper

mt.mod.abs.emm.plot <- ggplot(mt.mod.abs.emm, aes(x=fct_relevel(trait, rev(traits)), y=trend)) +  
  facet_wrap(vars(predictor), nrow=1, scales="free_x", 
             labeller=labeller(predictor=c(precip_est_mm="Estimated summer precip (mm)", sun_date="Snowmelt date"))) + 
  geom_hline(yintercept=0) +  geom_pointrange(aes(ymin=trend - SE, ymax = trend + SE), size=0.4) + 
  labs(y="Slope and SE of z-scored trait vs. predictor", x="Floral Trait") + 
  scale_x_discrete(labels = traitnames) + coord_flip() +  theme_bw()

#calculate how much summer precip is required to compensate for 1 day earlier snowmelt
mt.nonsignif.traits <- mt.mod.abs.tests %>% filter(p.value > 0.05) %>% pull(trait) %>% unique  #if P> 0.05 for effect of summer precip or snowmelt
mt.mod.abs.emm.compensate <- mt.mod.abs.emm %>% pivot_wider(names_from=predictor, values_from=c(trend, SE), id_cols=trait) %>% 
  mutate(predictor = "Compensation",
         precip_mm_per_snow_day = trend_sun_date / trend_precip_est_mm, 
         SE = abs(precip_mm_per_snow_day) * # propagate error in SEs using formula for quotients
           sqrt((SE_precip_est_mm / trend_precip_est_mm)^2 + (SE_sun_date / trend_sun_date)^2),
        keep = !trait %in% mt.nonsignif.traits & trait!="sepal_width",   #sepal_width has interaction with sun_date,
        precip_mm_per_snow_day = ifelse(keep, precip_mm_per_snow_day, NA), SE = ifelse(keep, SE, NA))

mt.mod.abs.emm.compensate.plot <- ggplot(mt.mod.abs.emm.compensate, aes(x=fct_relevel(trait, rev(traits)), y=precip_mm_per_snow_day)) +  
  facet_wrap(vars(predictor)) +   geom_hline(yintercept=0) +   
  geom_pointrange(aes(ymin=precip_mm_per_snow_day - SE, ymax = precip_mm_per_snow_day + SE), size=0.4) + 
  labs(y="Summer precipitation (mm) / day earlier snowmelt", x="") + scale_x_discrete(labels = traitnames) + 
  coord_flip() +   theme_bw() + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
                                       
grid.arrange(mt.mod.abs.emm.plot, mt.mod.abs.emm.compensate.plot, nrow=1, widths=c(3,1))
```

## Plasticity: soil moisture

Both snowmelt timing and summer precipitation affect soil moisture and its timing. This analysis regresses the floral traits against the mean soil moisture in a subplot. Slopes may differ among years. The models use:

* *mt.plantyr*, the trait data averaged by plant within year
* *VWC*, the summer mean of the approximately weekly mean volumetric water content in a subplot. This may need to change to a subset of the summer because it includes soil moisture taken after traits were measured. See section on within-season variation for timing of measurements in each year.
* *plotid* (subplot) as a random effect (plants vary within this larger unit)

The models are specified as:

    lmer(trait ~ year * VWC + (1|plotid), data = mt.plantyr)

* The model for total sugar is singular (variance at subplot level near 0)

```{r sm}
lt.subplot.r2 <- lt.subplotround %>% filter(round=="2")
lt.plantyr.r2 <- lt.plantyr %>% filter(round=="2")

pt.plantyr <- pt.plantyr %>% mutate(VWC = ifelse(is.na(VWC.plant), VWC.sm, VWC.plant)) #phys traits had the VWC measured next to the plant in 2019 and 2020 only - substitute VWC in subplot on closest date
pt <- pt %>% mutate(VWC = ifelse(is.na(VWC.plant), VWC.sm, VWC.plant)) 

options(contrasts = c("contr.sum", "contr.poly"))
mt.mod.sm <- map(set_names(traits), 
                  ~ lmer(mnps.plantyr[[.x]]  ~ year * VWC + (1|plotid), data=mnps.plantyr))
lt.mod.sm <- map(set_names(leaftraits), 
                  ~ lmer(lt.plantyr.r2[[.x]] ~ year * VWC + (1|plotid), data=lt.plantyr.r2))
pt.mod.sm <- map(set_names(phystraits), 
                  ~ lmer(pt.plantyr[[.x]]    ~ year * VWC + (1|plotid) + (1|round), data=pt.plantyr))

mt.mod.sm.coefs <- map_dfr(mt.mod.sm, tidy, .id="trait")
lt.mod.sm.coefs <- map_dfr(lt.mod.sm, tidy, .id="trait")
pt.mod.sm.coefs <- map_dfr(pt.mod.sm, tidy, .id="trait")

mt.mod.sm.tests <- map(mt.mod.sm, anova) %>% map_dfr(tidy, .id="trait")
lt.mod.sm.tests <- map(lt.mod.sm, anova) %>% map_dfr(tidy, .id="trait")
pt.mod.sm.tests <- map(pt.mod.sm, anova) %>% map_dfr(tidy, .id="trait")
```

### ANOVA tables

```{r anova_sm}
print_anova_table(mt.mod.sm.tests, mt.mod.sm)
print_anova_table(lt.mod.sm.tests, lt.mod.sm)
print_anova_table(pt.mod.sm.tests, pt.mod.sm)
```

```{r re_sm}
map_dfr(list(mt.mod.sm.coefs, lt.mod.sm.coefs, pt.mod.sm.coefs), re_table) %>% kable(digits=3)
```

### Plots


```{r morphology_sm}
options(contrasts = c("contr.sum", "contr.poly"))

mt.sm.mod <- mt.sm.mod.z <- mt.sm.mod.test <- mt.sm.mod.raw <- coef.VWC <- emm.VWC <- emm.VWC.allyears <- mt.sm.mod.plot <-list()
for(tr in traits) {
  #mt.sm.mod.raw[[tr]] <- lmer(mt[[tr]]    ~ year * VWC + (1|plot) + (1|plantid), data=mt)
  mt.sm.mod[[tr]] <-lmer(mnps.plantyr[[tr]] ~ year * VWC + (1|plotid), data=mnps.plantyr)
  mnps.plantyr[[paste0("fit.sm.",tr)]] <- NA
  mnps.plantyr[[paste0("fit.sm.",tr)]][!is.na(mnps.plantyr[[tr]])]<- predict(mt.sm.mod[[tr]], re.form=NA) #TODO why jumps?
  mt.sm.mod.test[[tr]] <- anova(mt.sm.mod[[tr]])
  
  #Z-score the data first to get standardized slopes
  mt.sm.mod.z[[tr]] <- lmer(scale(mnps.plantyr[[tr]]) ~ year * VWC + (1|plotid), data=mnps.plantyr)
  coef.VWC[[tr]] <- summary(mt.sm.mod.z[[tr]])$coefficients["VWC",]
  emm.VWC[[tr]] <- emtrends(mt.sm.mod.z[[tr]],  ~ year, var = "VWC")
  emm.VWC.allyears[[tr]] <- emtrends(mt.sm.mod.z[[tr]],  ~ 1, var = "VWC")

  #  mt.sm.mod.plot[[tr]] <- 
  #   ggplot(mnps.plantyr, aes_string(x="VWC", color="year", y = tr)) + geom_point() +
  #   geom_line(data=mnps.plantyr %>% drop_na(paste0("fit.sm.",tr)), aes_string(y=paste0("fit.sm.",tr)), size=2)+
  #    scale_color_manual(values=year_pal)+
  #       labs(x="Summer mean of soil moisture in subplot (VWC)", y=traitnames[[tr]], color="Year")
  # print(mt.sm.mod.plot[[tr]])
}
mt.sm.mod.tests <- map_dfr(mt.sm.mod.test, tidy, .id="trait")
mt.sm.mod.tests %>%  print_anova_table(mt.sm.mod)
coef.VWC.df <- as.data.frame(t(as.data.frame(coef.VWC))) %>% rownames_to_column("trait")
emm.VWC.ls <- map(emm.VWC, as.data.frame)
emm.VWC.ls$sepal_width[3,] <- NA
emm.VWC.df  <- bind_rows(emm.VWC.ls, .id="trait")
emm.VWC.allyears[c(1:3,5)] %>% map(as.data.frame) %>% bind_rows(.id="trait")

# library(googlesheets4)
# write_sheet(left_join(emm.VWC.df %>% mutate(predictor="VWC", .before=1) %>%
#                         mutate(traitname=traitnames[trait], .before="trait"),
#                       mnps.subplot %>% group_by(year) %>% select(traits) %>%
#                         summarize(across(traits, ~ sum(!is.na(.x)))) %>%
#                         pivot_longer(traits, names_to="trait", values_to="n"))
#             , ss=filter(datasheets, name=="Maxfield Results"), sheet="slopes_VWC")

emm.VWC.df %>% drop_na() %>% filter(trait %in% traits[c(1:8, 13)]) %>% 
  ggplot(aes(x=fct_relevel(trait, rev(traits)), y=VWC.trend, color=year)) +  
  geom_hline(yintercept=0) + geom_pointrange(aes(ymin=VWC.trend - SE, ymax = VWC.trend + SE), position=position_dodge(width=0.3), size=0.4) + 
  labs(y="Slope of z-scored trait vs. mean annual soil moisture", x="Floral Trait", color="Year") +
  theme_classic() + scale_x_discrete(labels = traitnames) + coord_flip() + scale_color_manual(values=year_pal)
```

```{r multi_sm, fig.width=9, fig.height=11}
plot_sm <- function(mod, data, traits.plot) {
  data.long <- data %>% select(all_of(traits.plot), water, snow, year, sun_date, precip_est_mm, VWC) %>% 
    pivot_longer(all_of(traits.plot), names_to="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal")) %>% 
    drop_na(value, water) %>% filter(trait %in% traits.plot)
  
  emm.grid <- mod[traits.plot] %>% 
    map_dfr(~ emmeans(.x, ~ year*VWC, at=list(VWC=range(data$VWC, na.rm=T))) %>% as_tibble(), .id="trait") %>% 
    mutate(trait = fct_relevel(trait, traits.plot))
  
  traitnames.multiline <- ifelse(nchar(traitnames) > 30, str_replace(traitnames, fixed("("),"\n("), traitnames)
  
  sm_plot <- ggplot(emm.grid, aes(x=VWC, y=emmean, color=year, group=year)) + 
  facet_wrap(vars(trait), scales="free_y",labeller = as_labeller(traitnames))+
  geom_point(data = data.long, aes(y=value, color=year)) +  
  geom_line(size=2) + 
  scale_color_manual(values=year_pal) + 
  labs(x=ifelse("WUE" %in% traits.plot, "Soil moisture next to plant or in subplot on closest date (%VWC)", 
                "Summer mean of soil moisture in subplot (%VWC)"), y="", color="Year") + 
  coord_cartesian(ylim=c(0,NA)) + scale_y_continuous(expand=expansion(mult=c(0,0.05)))+
  theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                          panel.border = element_rect(fill=NA, colour = "black"))
  return(sm_plot)
}

plot_sm(mt.mod.sm, mnps, traits[c(1:4,5,6,8,13)]) #Morphological traits - raw data
plot_sm(mt.mod.sm, mnps.plantyr, traits[c(1:4,5,6,8,13)]) #Morphological traits - plant means

grid.arrange(plot_sm(lt.mod.sm, lt, leaftraits), 
             plot_sm(pt.mod.sm, pt, phystraits) + guides(color=F, shape=F), ncol=1) #raw data
grid.arrange(plot_sm(lt.mod.sm, lt.plantyr.r2, leaftraits),
             plot_sm(pt.mod.sm, pt.plantyr, phystraits)+ guides(color=F, shape=F), ncol=1) #plant means
```

## Selection: water, survival/flowering

### Survival

```{r selection_surv}
options(contrasts = c("contr.sum", "contr.poly"))

lt.mod.surv <- map(set_names(leaftraits), 
                   ~ glmmTMB(survived ~ lt.plantyr.r2[[.x]] * year * water + (1|plot), 
                            data=lt.plantyr.r2, family=binomial))
pt.mod.surv <- map(set_names(phystraits), 
                   ~ glmmTMB(survived ~ pt.plantyr[[.x]] * year * water * snow + (1|plot), 
                            data=pt.plantyr, family=binomial))

lt.mod.surv.coefs <- map_dfr(lt.mod.surv, tidy, .id="trait")
pt.mod.surv.coefs <- map_dfr(pt.mod.surv, tidy, .id="trait")

lt.mod.surv.tests <- map(lt.mod.surv, car::Anova, type=3) %>% map_dfr(tidy, .id="trait")
pt.mod.surv.tests <- map(pt.mod.surv, car::Anova, type=3) %>% map_dfr(tidy, .id="trait")
```

```{r anova_surv}
print_Anova_table <- function(df, mod) {
  df %>% select(c("trait", "term", "p.value")) %>% 
    mutate(p.value=p.value %>% map_chr(~ format(.x,scientific=-2, digits=2 , nsmall=2)) %>% 
             str_replace("e-0(\\d)"," x 10^-\\1^") %>% str_replace("e-(\\d+)"," x 10^-\\1^")) %>%
    pivot_wider(names_from=trait, values_from=p.value) %>% 
    kable
}
print_Anova_table(lt.mod.surv.tests, lt.mod.surv)
print_Anova_table(pt.mod.surv.tests, pt.mod.surv)
map_dfr(list(lt.mod.surv.coefs, pt.mod.surv.coefs), re_table) %>% kable(digits=3)
```

### Flowering

```{r selection_flrd}
options(contrasts = c("contr.sum", "contr.poly"))

lt.mod.flrd <- map(set_names(leaftraits), 
                   ~ glmmTMB(flowered ~ lt.plantyr.r2[[.x]] * year * water + (1|plot) , 
                            data=lt.plantyr.r2, family=binomial))
pt.mod.flrd <- map(set_names(phystraits), 
                   ~ glmmTMB(flowered ~ pt.plantyr[[.x]] * year * water + (1|plot), 
                            data=pt.plantyr, family=binomial))

lt.mod.flrd.coefs <- map_dfr(lt.mod.flrd, tidy, .id="trait")
pt.mod.flrd.coefs <- map_dfr(pt.mod.flrd, tidy, .id="trait")

lt.mod.flrd.tests <- map(lt.mod.flrd, car::Anova, type=3) %>% map_dfr(tidy, .id="trait")
pt.mod.flrd.tests <- map(pt.mod.flrd, car::Anova, type=3) %>% map_dfr(tidy, .id="trait")
```

```{r anova_flrd}
print_Anova_table(lt.mod.flrd.tests, lt.mod.flrd)
print_Anova_table(pt.mod.flrd.tests, pt.mod.flrd)
map_dfr(list(lt.mod.flrd.coefs, pt.mod.flrd.coefs), re_table) %>% kable(digits=3)
```

## Selection: snow, survival/flowering

### Survival

```{r selection_surv_snow}
options(contrasts = c("contr.sum", "contr.poly"))

lt.mod.surv <- map(set_names(leaftraits), 
                   ~ glmmTMB(survived ~ lt.plantyr.r2[[.x]] * year * snow + (1|plot), 
                            data=lt.plantyr.r2, family=binomial))
pt.mod.surv <- map(set_names(phystraits), 
                   ~ glmmTMB(survived ~ pt.plantyr[[.x]] * year * snow * snow + (1|plot), 
                            data=pt.plantyr, family=binomial))

lt.mod.surv.coefs <- map_dfr(lt.mod.surv, tidy, .id="trait")
pt.mod.surv.coefs <- map_dfr(pt.mod.surv, tidy, .id="trait")

lt.mod.surv.tests <- map(lt.mod.surv, car::Anova, type=3) %>% map_dfr(tidy, .id="trait")
pt.mod.surv.tests <- map(pt.mod.surv, car::Anova, type=3) %>% map_dfr(tidy, .id="trait")
```

```{r anova_surv_snow}
print_Anova_table(lt.mod.surv.tests, lt.mod.surv)
print_Anova_table(pt.mod.surv.tests, pt.mod.surv)
map_dfr(list(lt.mod.surv.coefs, pt.mod.surv.coefs), re_table) %>% kable(digits=3)
```

### Flowering

```{r selection_flrd_snow}
options(contrasts = c("contr.sum", "contr.poly"))

lt.mod.flrd <- map(set_names(leaftraits), 
                   ~ glmmTMB(flowered ~ lt.plantyr.r2[[.x]] * year * snow + (1|plot) , 
                            data=lt.plantyr.r2, family=binomial))
pt.mod.flrd <- map(set_names(phystraits), 
                   ~ glmmTMB(flowered ~ pt.plantyr[[.x]] * year * snow + (1|plot), 
                            data=pt.plantyr, family=binomial))

lt.mod.flrd.coefs <- map_dfr(lt.mod.flrd, tidy, .id="trait")
pt.mod.flrd.coefs <- map_dfr(pt.mod.flrd, tidy, .id="trait")

lt.mod.flrd.tests <- map(lt.mod.flrd, car::Anova, type=3) %>% map_dfr(tidy, .id="trait")
pt.mod.flrd.tests <- map(pt.mod.flrd, car::Anova, type=3) %>% map_dfr(tidy, .id="trait")
```

```{r anova_flrd_snow}
print_Anova_table(lt.mod.flrd.tests, lt.mod.flrd)
print_Anova_table(pt.mod.flrd.tests, pt.mod.flrd)
map_dfr(list(lt.mod.flrd.coefs, pt.mod.flrd.coefs), re_table) %>% kable(digits=3)
```
## Multivariate response
```{r multresp}
mt.morph <- mt[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","VWC")] %>% drop_na(1:5) #drop sepal_width
mt.pca <- prcomp(mt.morph[,-(6:7)], scale.=TRUE)

mt.morph.pca <- mt.morph %>%
  bind_cols(as.data.frame(mt.pca$x)) %>% 
  tibble::rowid_to_column("index") %>% 
  mutate(x_0=0,y_0=0,
         x_1=-corolla_width/2,y_1=corolla_length,
         x_2= corolla_width/2,y_2=corolla_length) %>% 
  pivot_longer(c(starts_with("x"),starts_with("y")), names_to=c(".value", "point"), names_sep="_")

scl <- 100
zoom <- 5.5
mt.pca.vars <- as.data.frame(mt.pca$rotation) %>% rownames_to_column("variable")
mt.pca.labs <- paste0("PC", 1:2, " (", round(100*mt.pca$sdev[1:2]^2/sum(mt.pca$sdev^2), 0), "% explained var.)") 
ggplot(mt.morph.pca) + theme_classic() +
  geom_polygon(aes(x= x/scl+PC1, y= y/scl+PC2, group=index, color=VWC, fill=VWC)) +
  #geom_point(aes(x=PC1, y=PC2, size=sepal_width/2), color="darkgreen") + 
  geom_point(aes(x=PC1, y=PC2 + anther_min/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + anther_max/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + style_length/scl), color="grey60", size=0.5, shape=4) +
  geom_segment(data=mt.pca.vars,aes(x=0, y=0, xend=PC1*zoom, yend=PC2*zoom), 
               arrow = arrow(length = unit(1/2, "picas")), color="black") +
  geom_text(data =mt.pca.vars, aes(label = gsub("_"," ",variable), x = PC1*zoom*1.4, y = PC2*zoom*1.4,
                                   angle =  (180/pi) * atan(PC2/PC1), hjust = 0), color="black")+
  scale_size_identity() + coord_fixed() + 
  scale_fill_viridis(direction=-1, option="plasma") + scale_color_viridis(direction=-1, option="plasma") + 
  labs(x=mt.pca.labs[1], y=mt.pca.labs[2])
ggsave("images/mt_morph_pca_noSW.pdf", height=15, width=20, device=cairo_pdf)

ggplot(mt, aes(x=corolla_length, y=corolla_width, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_smooth()

ggplot(mt, aes(x=corolla_length, y=style_length, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_abline(slope=1, intercept=0)

ggplot(mt, aes(y=style_length / corolla_length, x=VWC, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_abline(slope=0, intercept=1)+ geom_smooth()

ggplot(mt, aes(x=corolla_length, ymax = anther_max, ymin=anther_min, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_linerange() + theme_minimal()+ geom_abline(slope=1, intercept=0)

ggplot(mt, aes(y=anther_min / corolla_length, x=VWC, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_smooth()
```

```{r multresp_VWC, fig.height=6, fig.width=7}
mt.morph.plantyr <- mt.plantyr[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","VWC")] %>% drop_na() 
options(contrasts = c("contr.treatment", "contr.poly"))
mt.rda <- rda(scale(select(mt.morph.plantyr, !c(year, VWC))) ~ VWC * year, data=mt.morph.plantyr)
mt.rda
anova(mt.rda, by="term") %>% tidy %>% kable(digits=3)
#autoplot(mt.rda)

mt.rda.df <- fortify(update(mt.rda, ~ VWC + year))#drop interaction
ggplot(mapping=aes(x=RDA1, y=RDA2, label=Label)) + 
  geom_point(data=mt.rda.df %>% filter(Score=="sites") %>% bind_cols(mt.morph.plantyr), aes(color=VWC))+
  geom_segment(data=mt.rda.df %>% filter(Label=="VWC"), aes(x=0,y=0,xend=RDA1*3, yend=RDA2*3), 
               arrow=arrow(length=unit(0.5, "lines")), size=1.5)+
  geom_text(   data=mt.rda.df %>% filter(Label=="VWC"), aes(x=RDA1*3.5, y=RDA2*3.5, angle=atan(RDA2/RDA1)*180/pi))+
  geom_segment(data=mt.rda.df %>% filter(Score=="species"), aes(x=0,y=0,xend=RDA1, yend=RDA2), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(   data=mt.rda.df %>% filter(Score=="species"), aes(x=RDA1*1.7, y=RDA2*1.7, angle=atan(RDA2/RDA1)*180/pi,
    label = c(traitnames, c("anther_max"="Max stamen length", "anther_min"="Min stamen length"))[as.character(Label)] %>% str_remove(fixed(" (mm)"))), size=3)+
  geom_text(data=mt.rda.df %>% filter(Score=="centroids"), aes(label=str_remove(Label, "year")), fontface=2, size=6)+
  scale_color_viridis(direction=-1, option="plasma") + coord_fixed() + 
  scale_x_continuous(expand=expansion(mult=.16), breaks=c(-2,0,2))+   theme_minimal()
```


```{r multresp_discrete}
mt.morph.wm <- mt[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","precip_est_mm", "sun_date")] %>% drop_na(1:6) #and nectar (19,20) too
mt.cap.wm <- capscale(scale(mt.morph.wm[,-(6:8)]) ~ precip_est_mm * sun_date, data=mt.morph.wm)
mt.cap.wm
anova(mt.cap.wm, by="term") %>% tidy %>% kable
#plot CAP without interactions
autoplot(capscale(scale(mt.morph.wm[,-(6:8)]) ~ precip_est_mm + sun_date, data=mt.morph.wm)) + guides(color=F, shape=F)
```

### Floral traits within season

No obvious effect of time within a season for corolla length/width or sepal width. The timing of trait measurements was different each year:

```{r morph_season}
ggplot(sm, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point(data=ph, aes(y= height_cm/2, color="Phenology (height)"))+
  geom_point(data=mt, aes(y= corolla_length, color="Morphology"))+
  geom_point(data=mt, aes(y= nectar_24_h_ul, color="Nectar"))+
  geom_point(aes(y = VWC, color="Soil moisture")) +
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color="grey40")+
  labs(x="Day of year", shape="Snow", linetype="Snow", y="", color="Measurement")+
  scale_color_manual(values=c("purple","orange","grey","black"))+
  scale_x_continuous(breaks=seq(120, 240, by=20))+
  scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(1.5, "lines"), axis.text.y=element_blank())

ggplot(data=mt, aes(x=yday(date), y= corolla_length, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y=traitnames[["corolla_length"]], x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()

ggplot(data=mt, aes(x=yday(date), y= nectar_24_h_ul, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y=traitnames[["nectar_24_h_ul"]], x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()

ggplot(data=nt, aes(x=yday(date), y= nectar_conc, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y=traitnames[["nectar_conc"]], x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()
```

# Phenology
```{r check_phenology}
walk(list(ph18.raw, ph19.raw, ph20.raw), 
     function(ph.raw) {
  y = ph.raw$year[1]
  p <- ph.raw %>% count(plot, subplot, plotid, plantid, julian) %>% add_count(plantid, name="n_plantid") %>% 
  mutate(plantid = fct_reorder(plantid, as.integer(factor(paste(plotid, n_plantid))),.desc=T)) %>% 
  ggplot(aes(julian, plantid, fill=factor(n)))+
  geom_tile()+ scale_fill_viridis(option="inferno", discrete=T) + 
  geom_point(aes(color=subplot, x=median(as.integer(julian))))+ scale_color_manual(values=sample(rainbow(10)))+
  geom_point(aes(color=plot, x=median(as.integer(julian))), position=position_nudge(x=-0.3))+ 
  theme_classic() + theme(panel.background = element_rect(fill="grey")) + guides(x=guide_axis(angle = 90)) +
  labs(x="Day of year", y="Plant (sorted by subplot)", fill="Observations", color="Plot", title=paste(y,"phenology"))
  print(p)})
```


```{r phenology}
ggplot(ph %>% drop_na(height_cm), aes(x=julian, y=height_cm, shape=snow, color=water))+ geom_point() + geom_path(aes(group=plantid))+
  facet_wrap(vars(year), ncol=1)+
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Plant height", 
       x="Day of year", color="Precipitation", shape="Snow") +
  scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_color_manual(values=water_pal) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at(c("open","buds"), mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=open, shape=snow, linetype=snow, color=water))+ geom_point() + 
  facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Mean total flowers", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
    scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at("flowering", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=flowering, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Proportion flowering", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at("eggs", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=eggs, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Mean total eggs", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
    scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at("has_egg", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=has_egg, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Proportion of plants with eggs", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
    scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

```

## Floral display size
### 2020
```{r display_size}
ggplot(ph.plantid, aes(x=height_cm, y=open, color=water)) + geom_point() + geom_smooth(method="lm", se=F) + scale_color_manual(values=water_pal) + facet_wrap(vars(year), nrow=1) +
  labs(x="Mean inflorescence height (cm)", y="Mean number of open flowers", color="Precipitation")

ggplot(ph.plantid, aes(x=water, color=snow, y = height_cm)) +
    labs(x="Precipitation", y="Mean inflorescence height (cm)", color="Snowmelt") +
    geom_boxplot() + scale_color_manual(values=snow_pal) + facet_wrap(vars(year), nrow=1)
ggplot(ph.plantid, aes(x=VWC, y=height_cm, color=year)) + geom_point() + geom_smooth(se=F) +
    scale_color_manual(values=year_pal) + labs(x="Mean VWC", y="Mean inflorescence height (cm)") 
    
ggplot(ph.plantid, aes(x=water, color=snow, y = open)) +
    labs(x="Precipitation", y="Mean number of open flowers", color="Snowmelt") +
    geom_boxplot() + scale_color_manual(values=snow_pal) + facet_wrap(vars(year), nrow=1)
ggplot(ph.plantid, aes(x=VWC, y=open, color=year)) + geom_point() + geom_smooth(se=F)+
  scale_color_manual(values=year_pal) + labs(x="Mean VWC", y="Mean number of open flowers")
```

# Seeds

```{r seeds}
ggplot(sds %>% drop_na(plot), aes(color=water, x=fruits, y=seeds)) + geom_point() + geom_smooth(se=F) + 
  scale_color_manual(values=water_pal) + facet_wrap(vars(year), nrow=1) + theme_minimal() +
  labs(x="Counted fruits", y="Counted seeds", color="Precipitation")

for(tr in seedtraits) {
  print(ggplot(sds %>% drop_na(plot), aes_string(tr, x="water", color="snow")) + geom_boxplot() + 
    scale_color_manual(values=snow_pal) + facet_wrap(vars(year), nrow=1) + 
    scale_y_continuous(trans=ifelse(tr %in% c("seeds_est", "fruits_nonaborted"),"sqrt","identity"))+
    labs(x="Precipitation", y=traitnames[[tr]], color="Precipitation") +
        guides(x=guide_axis(angle = 90)))
  
  print(ggplot(sds, aes_string(tr, x="VWC", color="year")) + geom_point() + geom_smooth(se=F) + 
   scale_color_manual(values=year_pal) +
    scale_y_continuous(trans=ifelse(tr %in% c("seeds_est", "fruits_nonaborted"),"sqrt","identity"))+
    labs(x="Summer mean of soil moisture in subplot (VWC)", y=traitnames[[tr]], color="Year"))
}
```

## Fruiting phenology (2020)

```{r seed_phenology}
ggplot(sds20.date.zeroed, aes(x=julian, y=fruits_nonaborted, color=water, shape=snow, linetype=snow))+geom_jitter(width=2, height=0) + stat_summary(fun="mean", geom="line") + scale_y_sqrt() + scale_color_manual(values=water_pal)+
  labs(y="Nonaborted fruits", x="Day of year", color="Precipitation", shape="Snow", linetype="Snow")+ scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(sds20.date.zeroed, aes(x=julian, y=seeds, color=water, shape=snow, linetype=snow))+geom_jitter(width=2, height=0) + stat_summary(fun="mean", geom="line") + scale_y_sqrt() + scale_color_manual(values=water_pal)+
  labs(y="Seeds", x="Day of year", color="Precipitation", shape="Snow", linetype="Snow")+ scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(sds20.date.zeroed, aes(x=julian, y=prop_infested, color=water, shape=snow, linetype=snow))+geom_jitter(width=2, height=0) + stat_summary(fun="mean", geom="line") + scale_color_manual(values=water_pal)+
  labs(y="Proportion of nonaborted fruits infested", x="Day of year", color="Precipitation", shape="Snow", linetype="Snow")+ scale_linetype_manual(values=c(2,1)) +
  scale_y_continuous(labels = scales::percent)+ theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

sds20.date.zeroed %>% group_by(snow, water, julian) %>% summarize_at("fruiting", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=fruiting, shape=snow, linetype=snow, color=water))+ geom_point() +
  geom_path(aes(group=paste(snow, water))) +
  labs(y="Proportion fruiting (nonaborted fruits)", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))
```

