--- 
title: "Maxfield Meadow Ipomopsis Volatiles"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>


```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)
library(vegan)
library(broom)
library(viridis)
library(ggvegan)
library(gridExtra)
library(dendsort)
library(pheatmap)
library(RColorBrewer)
library(DT)

library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images_volatiles2020/", fig.height=8, fig.width=8)
options(digits=4) # for kables

load("data/volatiles/maxfield_volatiles.rda")
load("data/maxfield_data.rda")

#check for duplicates
#View(vt %>% filter(!exclude %in% c("flowers", "incomplete", "resample")) %>% count(sampledate.file, plot.file, subplot.file, plant.file) %>% filter(n>1) %>% left_join(vt)) #exclude="resample"
#View(maxfmeta %>% count(sampledate,plantid) %>% filter(n>1) %>% left_join(maxfmeta))
#TODO fix these GCMS filename duplicates

volsmeta <- maxfmeta %>% mutate(sampledate = as.Date(sampledate), year = factor(year(sampledate))) %>% 
  drop_na(plotid) %>% #cut out ambients, leaving 459 chromatograms
  filter(year %in% c("2018","2019","2020")) %>% droplevels() %>% 
  mutate(plot=str_sub(plotid, 1,1), subplot=str_sub(plotid, 2,2),
         sequence.date = date(sequence.start),
         sequence.year = year(sequence.date)) %>% 
  rename(sampledate.file=sampledate, plot.file = plot, subplot.file=subplot, plant.file=plant) %>% 
  left_join(vt %>% select(-c(colnames(treatments),"type","plantid")) %>% 
              filter(!exclude %in% c("flowers", "incomplete", "resample"))) %>% 
  left_join(treatments) %>% #assume that the written metadata is correct not the info in the  .file name
  mutate(total_time = end - bag,
         total_time = replace_na(total_time, median(total_time, na.rm=T)),# fill in 45 minutes when missing metadata
         pump_id=as.character(pump_id))
volsmeta <- left_join(volsmeta, volsmeta %>% count(sequence.date, sequence.start) %>% group_by(sequence.date) %>% mutate(sequence.12 = paste(sequence.date, row_number()))) #add the batch number for that day (1 or 2)
rownames(volsmeta) <- volsmeta$index 
```

# Sample size

Calculated from two datasets that will eventually match: datasheets from sampling and GC-MS filenames.

```{r samplesizes, fig.width=10, fig.height=5}
bind_rows(sampling=vt %>% group_by(year, type) %>% tally(),
  filenames=maxfmeta %>% mutate(year=factor(year(sampledate))) %>% group_by(year, type) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year,type), values_from = n) %>% kable()

bind_rows(sampling=vt %>% filter(type=="floral") %>% group_by(year, water, snow) %>% tally(),
  filenames=volsmeta %>% mutate(year=factor(year(sampledate.file))) %>% group_by(year, water, snow) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year), values_from = n) %>% kable()

bind_rows(sampling = vt %>% filter(type=="floral") %>% group_by(sampledate.file) %>% tally(),
         filenames = volsmeta %>% drop_na(sampledate.file) %>% group_by(sampledate.file) %>% tally(), .id="source") %>% 
  ggplot(aes(x=yday(sampledate.file), y=n, fill=source)) + geom_col(position=position_dodge(preserve="single")) +
  facet_wrap(vars(year(sampledate.file)), ncol=1) +
  labs(x="VOC sampling date (day of year)", y="Floral samples")
```

## Matching datasheets and filenames

```{r matching}
# diagnostics for matching up vt (datasheets) and volsmeta (filenames)
volsmeta.samp <- volsmeta %>% mutate(samp = paste(sampledate.file, plot.file, subplot.file, plant.file)) %>% arrange(samp) %>% pull(samp)
vt.samp <- vt %>% mutate(samp = paste(sampledate.file, plot.file, subplot.file, plant.file)) %>% arrange(samp) %>% pull(samp)
#intersect(volsmeta.samp, vt.samp) #in both volsmeta and vt
cat("Have files but not datasheet entries:")
setdiff(volsmeta.samp, vt.samp)
cat("Have datasheet entries but no files:")
setdiff(vt.samp, volsmeta.samp)

vt.volsmeta <- full_join(vt %>% select(vt.n, sampledate.file, plot.file, subplot.file, plant.file) %>% mutate(in.vt=T),
                         volsmeta %>% select(FileName, sampledate.file, plot.file, subplot.file, plant.file) %>% mutate(in.volsmeta=T)) %>% 
  mutate(across(c(in.vt, in.volsmeta), ~replace_na(.,FALSE)))
vt.volsmeta.count <- vt.volsmeta  %>% count(sampledate.file, in.vt, in.volsmeta) %>% mutate(status=paste("datasheet",ifelse(in.vt,"yes","no"),"+","GCfile", ifelse(in.volsmeta,"yes","no")))
#View(vt.volsmeta.count %>% filter(!in.vt | !in.volsmeta) %>% arrange(sampledate.file))

ggplot(vt.volsmeta.count, aes(x=as.Date(sampledate.file),y=n, fill=status))+ geom_col() +
  facet_wrap(vars(year(sampledate.file)), scales="free_x", ncol=1) +
  scale_x_date(breaks=scales::date_breaks(width = "7 days"), minor_breaks = scales::date_breaks(width = "1 days")) +
  theme_minimal() + theme(legend.position="top") + scale_fill_manual(values=c("red","orange","darkgreen")) +
  labs(x="Sampling date (according to GCfile)", y="Samples", fill="Data status")
```


# Soil moisture

```{r sm}
ggplot(vt %>% drop_na(water), aes(x=factor(sampledate), y=VWC, color=water)) + facet_wrap(vars(year), ncol = 1, scales="free_x") +
  geom_boxplot() + scale_color_manual(values=water_pal) + scale_x_discrete(guide = guide_axis(angle=90)) + 
  labs(x="Date", y="Soil moisture (%VWC)", color="Precipitation")
```

# Multivariate analyses

## Filtering

Filtering criteria (verdict="fvoc" in the table below):

* Present in 10% of samples
* Maximum peak area is at least 4000000
* Passes t-test against ambient controls with FDR correction OR mean exceeds 4x mean of controls
* Not a known contaminant (Oxybenzone, Ethanol, 2,2'-oxybis-, Butylated Hydroxytoluene, Caprolactam, [1,1'-Bicyclopentyl]-2-one, L-Lactic acid, Pentadecane, 8-hexyl-, Decane, 2,8,8-trimethyl-, 2-Methyltetradecan-2-ol)

Note that these potential floral compounds did not meet the tests against ambient controls: 2-phenylacetaldehyde, 3-carene, D-limonene, camphene, benzaldehyde, a-thujene, sabinene, abietadiene, 2-ethylhexan-1-ol 

Plus (verdict other than "fvoc"): 

* bis14: compounds included in Bischoff et al. (2014)
* keep: kept since they are known floral volatiles
* keep_leaf: kept since they are known leaf volatiles
* merge: a compound that is a synonym of one of the floral volatiles (b-caryophyllene, b-ocimene, verbenone)

```{r filtering}
vols <- maxf[volsmeta$index,]
vols.cut <- vols[,colSums(vols)>3e8]

k <- 2 #split into this many k-means cluster per year
volsmeta$cluster <- 0
for(yr in unique(year(volsmeta$sequence.start))) {
  thisyr <- volsmeta$sequence.year==yr
  set.seed(1)
  volsmeta$cluster[thisyr] <- kmeans(decostand(vols.cut[thisyr,], method="log"), k, nstart=3)$cluster
}

ipochems <- read_csv("data/volatiles/chemsf_ipo.csv") %>% 
  filter(verdict %in% c("bis14", "fvoc", "keep","keep_leaf","merge"),
         name %in% colnames(vols)[colSums(vols)>0]) %>% #filter empty compounds
  left_join(read_csv("data/volatiles/regressions_2019_prexfer_filtered_slopes.csv")) #TODO get appropriate slopes for each year
ipogood <- ipochems$name
vols.good <- vols[,ipogood] %>% 
  sweep(2, ipochems$area_per_ng, FUN = '/') %>%  # convert areas to nanograms 
  sweep(1, as.numeric(volsmeta$total_time, units="hours"),  FUN = '/') #divide by equilibration + pumping time
nonzero <- rowSums(vols.good) > 0 #one sample ("55" on row 44) has no volatiles

ipochems %>% select(verdict, name, RT, mean, max, match, freq.floral, freq.ambient, ambient_ratio, filter_ambient_ttest, filters_passed, standard) %>%
  arrange(desc(freq.floral)) %>% mutate(across(starts_with("freq."), round, 3)) %>% 
  DT::datatable(caption="Volatile compounds to use for analysis")

vols.good.top <- vols.good[,colSums(decostand(vols.good,"pa"))>81]#occur in more than N samples
ipogood.top <- names(sort(colSums(vols.good.top), decreasing=T))
volsmeta.good.top <- bind_cols(volsmeta, vols.good.top) %>%  
  pivot_longer(-all_of(colnames(volsmeta))) %>% 
  mutate(name = fct_relevel(name, ipogood.top))

traits.good.top <- volsmeta.good.top %>% 
  left_join(mnps.plantyr %>% select(year, plotid, plant, all_of(traits))) %>% 
  mutate(name = fct_relevel(name, ipogood.top))

vols.drydown <- vols[,c("1,3-Octadiene","Benzyl alcohol", "(1R)-2,6,6-Trimethylbicyclo[3.1.1]hept-2-ene",".alpha.-Farnesene", "Bicyclo[3.1.1]heptane, 6,6-dimethyl-2-methylene-, (1S)-", ".beta.-Ocimene")]
volsmeta.drydown <- bind_cols(volsmeta, vols.drydown) %>%  pivot_longer(-all_of(colnames(volsmeta)))
```

## Heatmap

```{r heatmap, fig.width=16, fig.height=12}
annot_pal <- list(
  sequence.date = volsmeta %>% mutate(yr = sequence.year) %>% group_by(yr) %>% summarize(n = length(unique(sequence.date))) %>% 
    pull(n) %>% map(scales::hue_pal()) %>% unlist() %>% set_names(unique(volsmeta$sequence.date)),
  cluster = set_names(brewer.pal(8, "Set1")[1:length(unique(volsmeta$cluster))], unique(volsmeta$cluster)))

pheatmap(as.matrix(t(vols.cut))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),main="Unfiltered top volatiles",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors =  annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))

pheatmap(as.matrix(t(vols.good))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),main="Filtered volatiles",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors = annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))

pinene_correlates <- cor(sqrt(vols[,colSums(vols)>3e7])) %>% as.data.frame() %>% filter(`(1R)-2,6,6-Trimethylbicyclo[3.1.1]hept-2-ene`>0.15) %>% rownames()
pheatmap(as.matrix(t(vols[,pinene_correlates]))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),
                  annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors = annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0), main="Volatiles that correlate with alpha-pinene")
```

## Chromatograms by GCMS run date

### Univariate

```{r gcms_date, fig.height=8, fig.width=7}
plot_vols_rundate <- function(type, checkvols) {
volsmeta %>% mutate(yr=year(Desorb.Start.Time)) %>% 
  bind_cols(vols %>% select(checkvols)) %>% pivot_longer(cols=checkvols) %>% 
ggplot(aes(x=Desorb.Start.Time)) + facet_grid(name ~ yr, scales="free") + 
  geom_point(aes(y=value, color=name), alpha=0.3)  + 
  guides(color="none") + theme_bw() + labs(x="GCMS run time", y="Area", title=type)
}

plot_vols_rundate("Terpenes", c("(1R)-2,6,6-Trimethylbicyclo[3.1.1]hept-2-ene", ".beta.-Myrcene",".beta.-Ocimene","Linalool","Caryophyllene", "Longifolene"))

plot_vols_rundate("Benzenoids", c("Benzophenone","Methyl salicylate","Anisole"))

plot_vols_rundate("Green leaf volatiles", c("4-Hexen-1-ol, (Z)-", "3-Hexen-1-ol, (E)-"))
```

### Multivariate

```{r gcms_sample_date, fig.height=6, fig.width=10}
for(categories in list(c("water","snow"),"plot","observer","pump_id","sequence.12", "sampledate", c("sampledate", "sequence.12"), c("sampledate.observer", "sequence.12"))) {
  cluster_counts <- volsmeta %>% 
    mutate(sampledate = coalesce(sampledate, sampledate.file),
           sampledate.observer = paste(sampledate, observer)) %>% 
    count(!!!syms(c("sequence.year", categories, "cluster")))
  p <- ggplot(cluster_counts, aes(x=factor(paste(!!!syms(categories), sep=" x ")), 
                                  y=n, fill=factor(paste0(sequence.year," C",cluster)))) + 
    geom_col() + facet_wrap(vars(sequence.year), scales="free_x") +
    theme(axis.text.x = element_text(angle=90)) + labs(fill="Cluster", x=paste(categories, collapse=" x ")) +
    scale_fill_brewer(palette="Paired")
  print(p)
  
  if(length(categories)>1) {
    (p2 <- cluster_counts %>% filter(sequence.year !="2018") %>% mutate(cluster=factor(cluster), across(where(~ !is.numeric(.)),factor)) %>% 
     ggplot(aes_string(x=categories[[1]], y = categories[[2]], size="n", color="cluster")) + geom_point(shape=1, stroke=1) +
      facet_wrap(vars(sequence.year), scales="free") + theme_minimal()+ scale_size_area(max_size=15)+
      theme(axis.text.x = element_text(angle=90)) + scale_color_manual(values=c("darkgreen","magenta")))
    print(p2)
  }
}

ggplot(volsmeta, aes(x=Desorb.Start.Time, y=cluster))  + geom_line() + geom_point(aes(color=factor(mday(sampledate))))+
  facet_wrap(vars(sequence.year), scales="free_x", ncol=1) + scale_color_manual(values=sample(rainbow(30)))+
  scale_x_datetime(date_breaks="1 day", guide=guide_axis(angle=90)) + scale_y_discrete()
```


```{r gcms_date_ord}
for(yr in unique(year(volsmeta$sequence.start))) {
  print(yr)
  yrfilter <- year(volsmeta$sequence.start) == yr
  vols.cap <- capscale(sqrt(vols.cut[yrfilter,]) ~ factor(date(sequence.start)), data=volsmeta[yrfilter,])
  print(vols.cap)
  print(anova(vols.cap, by="term"))
  # vols.cap.points <-  fortify(vols.cap) %>% as_tibble()
  # vols.cap.samples <- vols.cap.points %>% filter(Score=="sites") %>% 
  #   left_join(volsmeta, by=c("Label"="index"))
  # vols.cap.chems <- vols.cap.points %>% filter(Score=="species")
  # p <- vols.cap.samples %>% 
  #   mutate(sequence.date = date(sequence.start)) %>% 
  #   filter(sequence.year == yr) %>% 
  #   ggplot(aes(x=CAP1, y=CAP2)) + theme_classic()+
  #   geom_point(aes(color=factor(sequence.date)), size=3) + scale_color_hue("Run date")
  # print(p)
}
```

## Ordinations

### NMDS

```{r nmds, fig.height=9, fig.width=6}
(vols.nmds <- metaMDS(sqrt(vols.cut), autotransform=FALSE, trace=F))
vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
  left_join(volsmeta, by=c("Label"="index"))
vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")

(scents.nmds <- metaMDS(sqrt(vols.good[nonzero,]), autotransform=FALSE, trace=F)) #had noshare = FALSE
nmds.points <-  fortify(scents.nmds) %>% as_tibble()
nmds.samples <- nmds.points %>% filter(Score=="sites") %>% left_join(volsmeta, by=c("Label"="index"))
nmds.chems <- nmds.points %>% filter(Score=="species")

vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() +
  geom_point(aes(color=factor(sequence.year))) + 
  labs(title="Unfiltered NMDS by year", color="Year")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  #geom_text(data=nmds.chems, aes(label=Label)) +
  geom_point(aes(color=factor(sequence.year)))+ 
  #geom_point(aes(color=water, shape=year)) + scale_color_manual("Precipitation", values=water_pal) +
  labs(title="Filtered NMDS by year", color="Year")
grid.arrange(vols.plot, vols.good.plot)

for(yr in unique(year(volsmeta$sequence.start))) {
  thisyr <- year(volsmeta$sequence.start)==yr
  (vols.nmds <- metaMDS(sqrt(vols.cut[thisyr,]), autotransform=FALSE, trace=F))
  vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
  vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
    left_join(volsmeta, by=c("Label"="index")) %>% filter(sequence.year == yr)
  vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")
  
  (scents.nmds <- metaMDS(sqrt(vols.good[nonzero & thisyr,]), autotransform=FALSE, trace=F))
  nmds.points <-  fortify(scents.nmds) %>% as_tibble()
  nmds.samples <- nmds.points %>% filter(Score=="sites") %>% 
    left_join(volsmeta, by=c("Label"="index")) %>% filter(sequence.year == yr)
  nmds.chems <- nmds.points %>% filter(Score=="species")
  
  vols.plot <- vols.nmds.samples %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Unfiltered NMDS run",yr))+
    geom_text(aes(color=factor(sequence.date), label=cluster), size=4) + scale_color_hue("Run date")
  vols.good.plot <- nmds.samples %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Filtered NMDS run",yr))+
        geom_text(aes(color=factor(sequence.date), label=cluster), size=4)  + scale_color_hue("Run date")
  grid.arrange(vols.plot, vols.good.plot)
}
```

```{r nmds_chems, fig.height=14, fig.width=12}
(vols.nmds <- metaMDS(sqrt(vols.cut), autotransform=FALSE, trace=F))
vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
  left_join(volsmeta, by=c("Label"="index"))
vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")

(scents.nmds <- metaMDS(sqrt(vols.good[nonzero,]), autotransform=FALSE, trace=F)) #had noshare = FALSE
nmds.points <-  fortify(scents.nmds) %>% as_tibble()
nmds.samples <- nmds.points %>% filter(Score=="sites") %>% left_join(volsmeta, by=c("Label"="index"))
nmds.chems <- nmds.points %>% filter(Score=="species")

vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() +
  geom_point(aes(color=factor(sequence.year))) + 
    geom_text(data=vols.nmds.chems, aes(label=Label)) +
  labs(title="Unfiltered NMDS by year", color="Year")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=factor(sequence.year)))+ 
    geom_text(data=nmds.chems, aes(label=Label)) +
  #geom_point(aes(color=water, shape=year)) + scale_color_manual("Precipitation", values=water_pal) +
  labs(title="Filtered NMDS by year", color="Year")
grid.arrange(vols.plot, vols.good.plot)


for(yr in unique(year(volsmeta$sequence.start))) {
  
  (vols.nmds <- metaMDS(sqrt(vols.cut[year(volsmeta$sequence.start)==yr,]), autotransform=FALSE, trace=F))
vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
  left_join(volsmeta, by=c("Label"="index"))
vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")

(scents.nmds <- metaMDS(sqrt(vols.good[nonzero & year(volsmeta$sequence.start)==yr,]), autotransform=FALSE, trace=F)) #had noshare = FALSE
nmds.points <-  fortify(scents.nmds) %>% as_tibble()
nmds.samples <- nmds.points %>% filter(Score=="sites") %>% left_join(volsmeta, by=c("Label"="index"))
nmds.chems <- nmds.points %>% filter(Score=="species")
  
  
  vols.plot <- vols.nmds.samples %>% 
    filter(sequence.year == yr) %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Unfiltered NMDS run",yr))+
    geom_point(aes(color=factor(sequence.date)), size=3) + scale_color_hue("Run date") +
          geom_text(data=vols.nmds.chems, aes(label=Label))
  vols.good.plot <- nmds.samples %>% 
    filter(sequence.year == yr) %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Filtered NMDS run",yr))+
    geom_point(aes(color=factor(sequence.date)), size=3) + scale_color_hue("Run date")+
          geom_text(data=nmds.chems, aes(label=Label))
  grid.arrange(vols.plot, vols.good.plot)
}
```

### CAP
```{r cap}
(scents.cap <- capscale(sqrt(vols.good) ~ year * water * snow, data=volsmeta))
anova(scents.cap, by="term")

scents.cap <- capscale(sqrt(vols.good) ~ year + water, data=volsmeta)
plot(scents.cap, type="n")
points(scents.cap, display = "sites", col=water_pal[volsmeta$water], pch=as.integer(volsmeta$year)+14)
text(scents.cap, display="cn", labels=c(2018:2020, names(water_pal)), col=c(rep("black",3),water_pal), font=2, cex=1.5)

plot(scents.cap, type="n")
text(scents.cap, display="species", col="black", cex=0.5)
```

# Univariate analyses

## Volatiles vs. treatments

```{r boxplots, fig.width=16, fig.height=10}
ggplot(volsmeta.good.top, aes(x = year, fill = water, y = value)) + geom_boxplot() + facet_wrap(vars(name), scales="free_y") + 
  scale_y_sqrt() + scale_fill_manual(values=water_pal) + labs(x="Year", y ="Emissions (ng/hr)", fill="Precipitation")

ggplot(volsmeta.drydown, aes(x = year, fill = water, y = value)) + geom_boxplot() + facet_wrap(vars(name), scales="free_y") + 
  scale_y_sqrt() + scale_fill_manual(values=water_pal) + labs(x="Year", y ="Emissions (ng/hr)", fill="Precipitation")

ggplot(volsmeta.good.top, aes(x = year, fill = snow, y = value)) + geom_boxplot() + facet_wrap(vars(name), scales="free_y") + 
  scale_y_sqrt() + scale_fill_manual(values=snow_pal) + labs(x="Year", y ="Emissions (ng/hr)", fill="Snowmelt")

ggplot(volsmeta.drydown, aes(x = year, fill = snow, y = value)) + geom_boxplot() + facet_wrap(vars(name), scales="free_y") + 
  scale_y_sqrt() + scale_fill_manual(values=snow_pal) + labs(x="Year", y ="Emissions (ng/hr)", fill="Snowmelt")
```

## Volatiles vs. soil moisture

```{r sm_vols, fig.width=16, fig.height=10}
ggplot(volsmeta.good.top, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") + 
  geom_point() + geom_smooth(se=F)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + scale_color_manual(values=year_pal) + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr)", color="Year")

ggplot(volsmeta.drydown, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") + 
  geom_point() + geom_smooth(se=F)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + scale_color_manual(values=year_pal) + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr)", color="Year")
```

# Selection analyses

## Plots of selection in treatments and years
```{r selection, fig.width=16, fig.height=10}
ggplot(traits.good.top, aes(x=value, y=seeds_est, color=year)) + 
  geom_point() + geom_smooth(method="loess",se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Seed production", color="Year")+
  theme_bw()+scale_color_manual(values=year_pal) +
  scale_x_sqrt()

ggplot(traits.good.top, aes(x=value, y=seeds_est, color=water, linetype=year)) + 
  geom_point() + geom_smooth(method="loess",se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Seed production", color="Precipitation", linetype="Year")+
  theme_bw() + scale_color_manual(values=water_pal)+
  scale_x_sqrt()

ggplot(traits.good.top, aes(x=value, y=seeds_est, color=snow, linetype=year)) + 
  geom_point() + geom_smooth(method="loess",se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y="Seed production", color="Snowmelt", linetype="Year")+
  theme_bw() + scale_color_manual(values=snow_pal)+
  scale_x_sqrt()
```

## Fly infestation vs. volatiles

```{r flies}
ggplot(traits.good.top, aes(x=value, y=prop_infested, color=year)) + 
  geom_point() + geom_smooth(method="loess",span=1,se=F) + 
  facet_wrap(vars(name), scales="free_x") + 
  labs(x="Emissions", y=traitnames.units["prop_infested"], color="Year")+
  theme_bw()+scale_color_manual(values=year_pal) +
  scale_x_sqrt()
```


## Selection gradients

Following method of [Chong et al. 2018](https://doi.org/10.1002/evl3.63) for reconstructing selection estimates from estimates of selection on PCs: Beta = E * A.

* PCA can't handle missing values so they are removed, but this biases the data and reduces sample size - multiple imputation?

```{r gradients}
calculate_beta <- function(E, slopes) {
  data.frame(B = E %*% slopes$A,
             SE = sqrt(E ^ 2 %*% slopes$SE ^ 2)) %>% rownames_to_column("trait")
}
plot_beta <- function(beta) {
  beta %>% mutate(trait = fct_reorder(trait, B)) %>% 
    ggplot(aes(y=trait, x = B, xmin=B-SE, xmax=B+SE)) + 
    geom_vline(xintercept=0)+ geom_pointrange() + 
    labs(x = "Selection gradient (beta) estimated from gradients on PCs", y = "Trait")
}

vols.std <- bind_cols(volsmeta, vols.good.top) %>% left_join(mnps.plantyr %>% select(year, plotid, plant, seeds_est)) %>% 
  select(all_of(ipogood.top),seeds_est, any_of(names(treatments))) %>% 
  mutate(across(all_of(ipogood.top), ~ (.x-mean(.x, na.rm=T))/sd(.x, na.rm=T))) %>% 
  group_by(year) %>% mutate(seeds_rel = seeds_est/mean(seeds_est, na.rm=T), .keep="unused") %>% ungroup %>% 
  drop_na(all_of(ipogood.top)) 

# Run PCA on traits
vols.PCA <- vols.std %>% select(ipogood.top) %>% prcomp() 
summary(vols.PCA) # PCs 1 - 5 explain 51% of the variation
vols.PCA.E <- vols.PCA$rotation[,1:5]
vols.std.PCA <- bind_cols(vols.std, as.data.frame(vols.PCA$x))

# Regress relative fitness on PCs
vols.PCA.regression <- lm(seeds_rel ~ PC1 + PC2 + PC3 + PC4 + PC5, data = vols.std.PCA)
vols.PCA.slopes <- tidy(vols.PCA.regression) %>% filter(term != "(Intercept)") %>% rename(A=estimate, SE=std.error)

# Convert selection gradients on PCs to gradients on traits
(vols.beta <- calculate_beta(vols.PCA.E, vols.PCA.slopes))
plot_beta(vols.beta)
```

# Volatiles vs. floral traits

```{r flrsize, fig.width=16, fig.height=10}
ggplot(traits.good.top, aes(x=corolla_length, y=value, color=year)) + geom_point() + geom_smooth(se=F) + 
  facet_wrap(vars(name), scales="free_y") + 
  labs(y="Emissions", x=traitnames.units["corolla_length"], color="Year")+
  theme_bw() + scale_color_manual(values=year_pal) +
  scale_y_sqrt()

ggplot(traits.good.top, aes(x=sepal_width, y=value, color=year)) + geom_point() + geom_smooth(se=F) + 
  facet_wrap(vars(name), scales="free_y") + 
  labs(y="Emissions", x=traitnames.units["sepal_width"], color="Year")+
  theme_bw() + scale_color_manual(values=year_pal) +
  scale_y_sqrt()

ggplot(traits.good.top, aes(x=nectar_24_h_ul, y=value, color=year)) + geom_point() + geom_smooth(se=F) + 
  facet_wrap(vars(name), scales="free_y") + 
  labs(y="Emissions", x=traitnames.units["nectar_24_h_ul"], color="Year")+
  theme_bw() + scale_color_manual(values=year_pal) +
  scale_y_sqrt()
```

