--- 
title: "Maxfield Meadow Ipomopsis Experiment"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>

# Purpose
Analyze selection on and plastic responses of floral and vegetative traits of *Ipomopsis aggregata* to manipulations of snowmelt and summer precipitation. Determine whether plasticity is adaptive (improves survival and reproduction). Part of a [larger project](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1654655) to understand the potential for evolutionary rescue in alpine plants under climate change.

# Treatments

```{r setup, message=FALSE, cache=FALSE}
library(tidyverse)
library(lubridate)
#library(googlesheets4)

library(lme4)
library(lmerTest)
library(MuMIn)
library(emmeans)
library(broom)
library(broom.mixed)
library(vegan)

library(RColorBrewer)
library(viridis)
library(gridExtra)
library(GGally)
library(ggvegan)
library(ggnewscale)

library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images/")

signif.num <- function(x) {
    as.character(symnum(x, corr = FALSE, na = FALSE, legend = FALSE,
           cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
           symbols = c("***", "**", "*", ".", " ")))
}
print_anova_table <- function(df, mod) {
  df %>% select(c("trait", "term", "p.value")) %>% 
    mutate(p.value=p.value %>% map_chr(~ format(.x,scientific=-2, digits=2 , nsmall=2)) %>% 
             str_replace("e-0(\\d)"," x 10^-\\1^") %>% str_replace("e-(\\d+)"," x 10^-\\1^")) %>%
    pivot_wider(names_from=term, values_from=p.value) %>% 
    mutate(singular = map_lgl(mod, isSingular),
           R2m = map_dbl(mod, ~ MuMIn::r.squaredGLMM(.x)[[1,"R2m"]])) %>% 
    kable
}

load("data/maxfield_data.rda")
```

```{r map}
subplot_offset <- 1/6
size_meter <- 6
ggplot(treatments_map) + coord_fixed(clip="off")+
  geom_point(aes(x=plot_x, y=plot_y, fill=snow), size=7*size_meter, shape=22, stroke=1)+
  scale_fill_manual("Snowmelt", values=snow_pal, guide=guide_legend(override.aes = list(size = 2*size_meter)))+
  new_scale_fill()+
  geom_point(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, fill=water4), size=2*size_meter, shape=22, stroke=1) + 
    geom_text(aes(x=plot_x+subplot_x*subplot_offset, y=plot_y+subplot_y*subplot_offset, label=subplot)) + 
  scale_fill_manual("Precipitation", values=water4_pal) +
  geom_text(aes(x=plot_x, y=plot_y, label=plot))+
  theme(legend.position="left", axis.title=element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.background = element_blank(), panel.grid = element_blank(), legend.box.background = element_blank())
```


# Snow melt vs. treatments

```{r hobo}
ggplot(hobo, aes(x=intensity, y=temp, color=hour(time))) +  geom_point(alpha=0.5, size=0.5) + geom_density_2d(color="black")+ 
  scale_x_log10(labels=function(x) format(x, scientific=F, big.mark=","), limits=c(10,1e6), breaks=10^(0:6)) +
  scale_color_gradientn(colours = rainbow(5))+
  geom_vline(xintercept = melt_threshold_sun) + 
  geom_hline(yintercept=melt_threshold_warm)+ geom_hline(yintercept=-melt_threshold_warm) +
  labs(x="Light intensity (lux)", y="Temperature (C)", color="Hour") +
  theme_dark() + theme(panel.grid.minor.x=element_blank())

ggplot(hobo %>% filter(hour(time)%in%5:17 & 
                  (year!="2019"&yday(time) %in% 110:130 | 
                   year=="2019"&yday(time) %in% 145:160)), 
       aes(x=intensity, y=temp, color=hour(time))) +   geom_point(alpha=0.7) +geom_density_2d(color="black")+
  scale_x_log10(labels=function(x) format(x, scientific=F, big.mark=","), limits=c(10,1e6), breaks=10^(0:6)) +
  scale_color_gradientn(colours = rainbow(5))+
  geom_vline(xintercept = melt_threshold_sun) + 
  geom_hline(yintercept=melt_threshold_warm)+ geom_hline(yintercept=-melt_threshold_warm) +
  labs(title="Temperature and light thresholds for snow melt", subtitle = "During the day during melt weeks",
       x="Light intensity (lux)", y="Temperature (C)", color="Hour") +
  theme_dark() + theme(panel.grid.minor.x=element_blank())

ggplot(hobo %>% filter(month(time)%in%4:6), aes(x=time, y=temp, color=plot)) + 
  facet_wrap(vars(year), scales="free_x", ncol=1) + 
  geom_hline(yintercept=melt_threshold_warm) + geom_hline(yintercept=-melt_threshold_warm) +geom_line() +
  geom_segment(data=meltdates, aes(x=warm_time, y=-10, xend=sun_time, yend=-15, color=plot), size=1) +
    geom_line(data=daily_CORBIL %>% filter(year!="2017", month(date)%in%4:6), aes(x=as.POSIXct(date), y=av_temp_2m_C), inherit.aes = F, color=brewer.pal(3, "Oranges")[3]) +
  labs(x="Date",y="Temperature (C)", color="Plot")
  
ggplot(hobo %>% filter(month(time)%in%4:6), aes(x=time, y=intensity, color=plot)) + 
  facet_wrap(vars(year), scales="free_x", ncol=1) + 
  geom_line() + geom_hline(yintercept=melt_threshold_sun) +
  geom_point(data=meltdates, aes(x=sun_time, y=-2e4, color=plot), size=1) +
  labs(x="Date",y="Light intensity (lux)", color="Plot")

meltdates %>% select(c("year","plot", "snow", ends_with("date"))) %>% pivot_longer(ends_with("date")) %>% arrange(name) %>% 
  ggplot(aes(y=name, x=value, color=plot, group=plot, linetype=snow)) + facet_wrap(vars(year), ncol=1) +
  geom_path(size=1) + #geom_text(aes(label=plot), hjust=1.5) +
  scale_linetype_manual(values=c(4,1)) + scale_x_continuous(minor_breaks=110:160)+
  labs(x="Day of year",y="Criteron", color="Plot", linetype="Snowmelt")
       
meltdates %>% ggplot(aes(x=melt_offset, fill=snow)) + geom_dotplot(method="histodot", binwidth=1, color="white") +
  facet_wrap(vars(year), ncol=1)+ scale_fill_manual(values=snow_pal) +
  scale_y_continuous(expand=expansion(0,0))+ scale_x_continuous(breaks=-10:4)+
  theme(panel.grid.minor = element_blank()) +
  labs(x="Offset from mean normal snow melt date (days)", y="Number of plots", fill="Snowmelt")
```

# Weather data

## Daily weather

```{r weather_daily}
ggplot(daily_all %>% filter(year(date) %in% 2018:2020), aes(y=EPA_NOAA_filled, x=yday(date))) + geom_col(aes(fill=ground_covered)) + 
  facet_wrap(vars(year(date)), ncol=1) + labs(x="Day of year", y="NOAA_billy precipiation (mm)", fill="Snow cover")

daily_all %>% pivot_longer(unname(stations), names_to="station", values_to="precipitation_mm") %>% 
  filter(yday(date) >100, yday(date) < 240, year(date) %in% 2018:2020) %>%
  ggplot(aes(x=yday(date), y=precipitation_mm, color=station, shape=ground_covered)) +  
  geom_line() +  geom_point() + facet_wrap(vars(year), ncol=1) + theme(legend.position = "top", legend.direction="vertical") +
  labs(x="Day of year", y = "Precipitation (mm)", color="Station", shape="Snow cover")

precip_max <- 50
lower_continuous <- function (data, mapping, ...) {
    ggally_smooth(data = data, mapping = mapping, ..., method = "loess") + scale_shape_manual(values=c(1,2))+
    coord_cartesian(xlim=c(0,precip_max), ylim=c(0,precip_max)) + geom_abline(slope=1, intercept=0)+
    scale_x_continuous(breaks=seq(0,precip_max, by=precip_max/5))+scale_y_continuous(breaks=seq(0,precip_max, by=precip_max/5))
}
lower_combo <- function (data, mapping, ...) {
    ggally_facethist(data = data, mapping = mapping, binwidth=precip_max/50,...) +
    coord_cartesian(xlim=c(0,precip_max)) + scale_x_continuous(breaks=seq(0,precip_max, by=precip_max/5))
}
ggpairs(daily_all %>% select(all_of(unname(stations[-1])), ground_covered), mapping=aes(color=ground_covered,shape=ground_covered), lower=list(continuous=wrap(lower_continuous), combo=wrap(lower_combo)))

filter(daily_all) %>% ggplot(aes(y=EPA_RsrchMdw,x= NOAA_billy, color=ground_covered)) + geom_point(shape=1) + geom_smooth(method="lm", se=F) + geom_abline(intercept=0, slope=1) + labs(color="Snow cover")
```

## Monthly weather

```{r weather_monthly}
#recalculate Campbell and Wendlandt 2013 July average, 1989 - 2006 EPA station
(EPA_89_06_July_avg_mm <- monthly_all %>% filter(mo==7, yr >= 1989, yr <= 2006) %>% pull(EPA_NOAA_filled) %>% mean)

(water_addition_per_month_mm <- (14 / 4 / 2) * 31) #14 L applied to 4 m2 every 2 days = 1.75 mm daily precip * 31 days in July

monthly_all %>% pivot_longer(unname(stations), names_to="station", values_to="precipitation_mm") %>% 
ggplot(aes(x=mo*30.4-15.2, y=precipitation_mm, color=station)) + facet_wrap(vars(yr)) + #place points at month midpoint
  geom_rect(data=waterdates %>% pivot_wider(names_from=precip_treatments, values_from=c(day, date)) %>% mutate(yr=factor(year)), 
            aes(xmin=day_started, xmax=day_ended, ymin=0, ymax=water_addition_per_month_mm), inherit.aes=F,
            fill=water_pal["Addition"])+
  geom_line() +  geom_point()+
  geom_point(data=groundcover %>% filter(year>=2004) %>% mutate(yr=factor(year)), aes(x=first_0_cm_day,y=0), inherit.aes=F)+
  geom_point(data=groundcover %>% filter(year>=2005) %>% mutate(yr=factor(year-1)), aes(x=first_snow_day,y=0), inherit.aes=F)+
  labs(y="Monthly precipitation (mm)", x="Day of the year, with snowmelt and first snow dates", color="Station") +
  theme(legend.position="top") + coord_cartesian(ylim=c(0,200))

precip_max <- 250
ggpairs(monthly_all %>% mutate(season=factor(c("wntr","smmr"))[1+mo %in% 6:9]) %>% select(all_of(unname(stations[-1])), season), mapping=aes(color=season,shape=season), lower=list(continuous=wrap(lower_continuous), combo=wrap(lower_combo)))

filter(monthly_all) %>% ggplot(aes(y=EPA_RsrchMdw,x= NOAA_billy, color=ifelse(mo %in% 6:9,mo,"Oct - May"))) + geom_point() + geom_smooth(method="lm", se=F) + geom_abline(intercept=0, slope=1) + labs(color="Month")

monthly_all %>% filter(mo %in% 5:9) %>% 
  ggplot(aes(x=mo,  y=EPA_NOAA_filled)) + geom_smooth(span=0.5, color="black") +
  geom_line(aes(color=ifelse(yr>=2018, "study period (filled)", ifelse(yr>=2012,"filled from billy's NOAA data","original EPA data")), group=yr)) +
  geom_point(x=7, y=EPA_89_06_July_avg_mm)+ scale_y_continuous(limits=c(0,100))+
  geom_text(aes(label=ifelse(yr>=2018, yr, "")))  + 
  labs(y="Monthly precipitation (mm)", x="Month", title="EPA Research Meadow 1989-2020 summer precipitation", 
       subtitle="Black dot is mean July precipitation 1989-2006", color="Data filling") + theme(legend.position = "top")

monthly_all %>% filter(mo %in% 5:9) %>% ggplot(aes(x=yr, color=factor(mo), group=mo, shape=ifelse(yr>=2018, "study period (filled)", ifelse(yr>=2012,"filled from billy's NOAA data","original EPA data")), y=EPA_NOAA_filled)) +
  geom_vline(xintercept=2011.5) +geom_point() + geom_smooth(se=F, span=0.7) + 
  scale_shape_manual(values=c(1,2,19)) + scale_y_continuous(limits=c(0,100))+
  labs(x="Year",y="EPA Research Meadow monthly precipitation (mm)", color="Month", shape="Data filling") + 
  geom_segment(data=monthly_all %>% filter(mo %in% 5:9, yr <= 2007) %>% 
                 group_by(mo) %>% summarize(mean_precip=mean(EPA_NOAA_filled)), 
               aes(x=1989, xend=2006, y=mean_precip, yend=mean_precip, color=factor(mo)), inherit.aes=F, size=1) 
```

## Snowmelt timing

```{r groundcover}
groundcover %>% mutate(first_snow_day=first_snow_day-365) %>% 
  pivot_longer(ends_with("day"), names_to="threshold", values_to="day") %>% drop_na(day) %>% select(-starts_with("first")) %>% 
  mutate(threshold = fct_reorder(factor(threshold),day, na.rm=T, .desc=T)) %>% 
  ggplot(aes(x=year, y=day, color=threshold)) +  
  annotate("rect", xmin=2017.5, xmax=2020.5, ymin=-Inf, ymax=Inf, fill="lightblue")+
  geom_line(aes(group=year), size=2)+ geom_smooth( method="lm", se=F) +geom_point() +
  scale_x_continuous(minor_breaks = 1976:2020, position = "top") + scale_y_continuous(n.breaks=10)+
  scale_color_grey(labels=function(x) str_to_sentence(str_replace_all(x, "_", " ")), start=0, end=0.85) +
  labs(x="Year", y="Day of year", color="Threshold") + theme_minimal()

cat("SD of first bare ground =", round(sd(groundcover$first_0_cm_day),1),"days over",range(groundcover$year))
summary(lm(first_0_cm_day ~ year, data=groundcover))

treatments %>% ggplot(aes(x=sun_date, shape=year, color=water, y=precip_est_mm))+geom_count()+
  geom_point(data=groundcover, aes(x=first_0_cm_day, color="Control", shape="1990-2017"), fill="white") +
  scale_color_manual(values=water_pal) + scale_shape_manual(values=c(21, 19, 17, 15))+ coord_cartesian(ylim=c(0,NA))+
  labs(y="Estimated summer precipiation with treatments (mm)", x="Snowmelt date", color="Precipitation", shape="Year",size="# Plots") + 
  theme_minimal()
```


# Soil moisture vs. treatments

## Weather plots

```{r soil, fig.height=12, fig.width=12}
ggplot(sm, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
  geom_jitter(aes(shape=snow, color=water, y = VWC), height=0, width=0.3) + 
  geom_smooth(aes(linetype=snow, color=water, y = VWC), se=F, span=0.5) +
  geom_col(data=daily_all %>% filter(yr%in%2018:2020), aes(x=yday(date), y=EPA_NOAA_filled), 
           fill=brewer.pal(3, "Greens")[3]) +
  geom_line(data=daily_all %>% filter(yr%in%2018:2020), aes(x=yday(date), y=av_temp_2m_C), 
            color=brewer.pal(3, "Oranges")[3]) + 
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color=brewer.pal(3, "Purples")[3])+
  geom_line(data = snowcloth, aes(y=mean_snow_depth_cm/10, group=year))+
  geom_point(data = snowcloth, aes(y=mean_snow_depth_cm/10))+
  geom_line(data=groundcover %>% pivot_longer(ends_with("day"), names_to="threshold", values_to="day") %>%
              filter(threshold!="first_snow_day", year>2017) %>% 
              mutate(snow_depth_cm = as.integer(str_split_fixed(threshold,"_",n=4)[,2])),
            aes(x=day, y=snow_depth_cm/10), color="grey40")+
  geom_segment(data=waterdates %>% pivot_wider(names_from=precip_treatments, values_from=c(day,date)),
               aes(x=day_started, xend=day_ended, y=20, yend=20)) +
  geom_point(data=waterdates, aes(x=day,y=20)) +
  geom_point(data=ph, aes(y= 17*(open+buds)/(open+buds)), color="grey", shape=0)+
  geom_point(data=mt, aes(y= 18*corolla_length/corolla_length), color="purple", shape=0)+
  geom_point(data=mt, aes(y= 19*nectar_24_h_ul/nectar_24_h_ul), color="orange", shape=0)+
  scale_x_continuous(limits=c(85, 230), breaks=scales::breaks_extended(10), expand=expansion(0,0)) + scale_y_continuous(limits=c(-3, 20)) +
  labs(y="Soil VWC (%) / Daily precipitation (mm) / Mean temp. (C) / Snowpack (dm)", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_color_manual(values=water_pal) +
  scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(sm.subplot, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
      geom_col(data=daily_all %>% filter(yr%in%2018:2020), aes(x=yday(date), y=EPA_NOAA_filled), 
           fill=brewer.pal(3, "Greens")[2]) +
  geom_line(data=daily_precip_est, aes(y=VWC_Control_4*100), color=brewer.pal(3, "Greens")[2], size=1) +
  geom_point(aes(shape=snow, color=water, y = VWC)) + 
  geom_line(data=sm.subplot %>% group_by(water, snow, year, date) %>% summarize(VWC=mean(VWC, na.rm=T)), aes(linetype=snow, color=water, y = VWC)) +
    scale_color_manual(values=water_pal) +  scale_linetype_manual(values=c(2,1)) + 
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color="black", show.legend=F) +
  coord_cartesian(xlim=c(110, 230), ylim=c(0, round(max(sm.subplot$VWC, na.rm=T))), expand=F)+
    labs(x="Day of year", y="Mean soil VWC in subplot (%) / Precipitation (mm)", color="Water", shape="Snowmelt", linetype="Snowmelt") + theme_minimal() + theme(legend.key.width = unit(2, "lines"))
```

# Soil moisture vs. snowmelt date

```{r soil_treatments}
left_join(filter(sm.subplot20d, mo20d!="221 - 240"), meltdates) %>% 
  mutate(mo = paste0("Days ",mo20d)) %>% 
  ggplot(aes(x=sun_date, y=VWC, shape=year,color=water)) + 
  facet_wrap(vars(mo), nrow=1)+
  geom_point() + geom_smooth(se=F, method="lm") +
  scale_color_manual(values=water_pal) +
  labs(x="Snowmelt date", y="Mean soil moisture in subplot", color="Water", shape="Year")

sm.subplot %>% mutate(day=yday(date)) %>% group_by(year) %>% summarize(start=min(day), end=max(day)) %>% kable(caption = "Soil moisture collection dates")

# VWC split-plot analysis
sm.mod <- lmer(VWC ~ year * snow * water + (1|plot), data=sm.subplotyear)
anova(sm.mod) %>% kable(digits=3)
(sm.emm.snow <- as.data.frame(emmeans(sm.mod, ~ snow))) %>% kable(digits=3)
sm.emm.snow$emmean[1] - sm.emm.snow$emmean[2]
(sm.emm.water <- as.data.frame(emmeans(sm.mod, ~ water))) %>% kable(digits=3)
sm.emm.water$emmean[1] - sm.emm.water$emmean[2]
sm.emm.water$emmean[3] - sm.emm.water$emmean[2]
```

## Timing of measurements

```{r timings}
ggplot(timings %>% filter(variable!="sds") %>% mutate(variable=fct_reorder(variable, begin, na.rm=T, .desc=T)), aes(y=variable, color=year))+ 
  geom_linerange(aes(xmin=begin, xmax=end), 
                 position = position_dodge2(width=0.7, reverse = T), size=1.2, show.legend=FALSE) + 
  #geom_text(aes(x=begin-10, label=begin), position = position_dodge(width=0.7), fontface="bold")+
  #geom_text(aes(x=end+10, label=end), position = position_dodge(width=0.7), fontface="bold") +
  geom_text(data=timings %>% drop_na(plots) %>% filter(year=="2019"),
            aes(x = end+7, label=paste(ifelse(nchar(plots)>1, "Plots","Plot"), str_replace(plots,","," & "))), 
            position = position_dodge2(width=0.8, reverse = T), size=4, hjust=0, show.legend=FALSE)+
  geom_point(aes(x=begin), position = position_dodge2(width=0.7, reverse = T), shape=15, size=2)+
  geom_point(aes(x=end), position = position_dodge2(width=0.7, reverse = T), shape=15, size=2) +
  scale_color_manual("Year", values=year_pal, guide=guide_legend(override.aes = list(size=5))) + 
  scale_y_discrete("", labels=c(snowcloth="Snow cloths applied to early plots", meltdates="Mean snowmelt timing (early plots - unmanipulated plots)", waterdates="Summer precipitation treatments applied",sm="Soil moisture recorded",ph="Inflorescence height recorded", mt="Floral morphology recorded", nt="Nectar traits recorded", sds="Flower collection period"))+
  scale_x_continuous("Day of year", breaks=seq(80,240, by=20)) +
  theme_minimal() + theme(legend.position = "top")
```

## Sample sizes

```{r sample_sizes}
cat("Morphology measurements per plant:", mt %>% drop_na(corolla_length) %>% group_by(year, plantid) %>% tally() %>% ungroup() %>% summarize(n=mean(n)) %>% pull(n) %>% round(2))
cat("Nectar measurements per plant:", mt %>% drop_na(nectar_24_h_ul) %>% group_by(year, plantid) %>% tally() %>% ungroup() %>% summarize(n=mean(n)) %>% pull(n) %>% round(2))
cat("Inflo height measurements per plant:")
ph %>% drop_na(height_cm) %>% group_by(year, plantid) %>% tally() %>% summarize(n=mean(n)) %>% kable
ph %>% drop_na(height_cm) %>% group_by(year, plantid) %>% tally() %>% 
  ggplot(aes(x=n, fill=year)) + geom_histogram() + labs(title = "Phenology entries per plant", x="Entries",y="Count")+
  facet_wrap(vars(year), ncol=1) + scale_fill_manual(values=year_pal)


#Total sample sizes
mt %>% drop_na(corolla_length) %>% group_by(year, plantid) %>% tally() %>% ungroup %>% tally() %>% pull(n)
mt %>% drop_na(nectar_24_h_ul) %>% group_by(year, plantid) %>% tally() %>% ungroup %>% tally() %>% pull(n)
sds %>% drop_na(flowers_est) %>% group_by(year, plantid) %>% tally() %>% ungroup %>% tally() %>% pull(n)
ph %>% drop_na(height_cm) %>% group_by(year, plantid) %>% tally() %>% ungroup %>% tally() %>% pull(n)
```


# Census

## Census status

The annual census at the beginning of the summer contains information on mortality, flowering status, and plant size (longest leaf and number of leaves for each rosette).

Visualize a plant's status in each census

```{r status}
ggplot(cen.size.long %>% filter(census!="2018r2"), aes(x=census, y=n_rosettes, group=plantid))+geom_path()+ geom_point() + scale_y_continuous(breaks=0:10)

ggplot(cen.status.long, aes(y=plantid, x=census, fill=status)) + geom_tile() +
  scale_fill_manual(values=brewer.pal(8, name="Set1")[c(1,3,5,2,7,6)])+
  theme(axis.text.y=element_blank(), axis.ticks = element_blank()) +
  labs(fill="Status", x="Census year",y="Plant ID")

ggplot(megatally.status.long, aes(y=plantid, x=dataset, fill=status)) + geom_tile() +
  scale_fill_manual(values=c(brewer.pal(5, name="Set1")[c(1,3,5,2)],brewer.pal(8, "Set2")[7:8]))+
  scale_x_discrete(position="top") +
  theme(axis.text.y=element_blank(), axis.ticks = element_blank(), axis.text.x = element_text(angle=90, size=14)) + 
  labs(fill="Status", x="Dataset and year",y="Plant ID")
```

## Status vs. treatment
```{r inventory}
cen.status.long %>% left_join(treatments) %>% drop_na(water, snow) %>% filter(status %in% status_ok, census !="2018r2") %>% 
  ggplot(aes(x=water, fill=status)) + facet_grid(snow ~ census, labeller=labeller(.cols=label_both)) + geom_bar(position="fill") +
  scale_fill_manual(values=status_pal) + labs(x="Precipitation", y="Proportion", fill="Status")
```

## Status transitions vs. treatment

What did plants of a given status in one year look like in the next?
```{r transitions}
with(cen.status, table(`2018`=paste(status_18), `2019`=paste(status_19)))

cen.status %>% mutate(year = "2018") %>% left_join(treatments) %>% filter(status_18 %in% status_ok[-3], status_19 %in% status_ok) %>% 
ggplot(aes(x=water, fill=status_19)) + facet_wrap(status_18 ~ snow, labeller=labeller(.cols=label_both)) +  
  labs(y="2019 Status", fill="2019 Status", x="Precipitation") +
  geom_bar(position="fill") + scale_fill_manual(values=status_pal) + theme_minimal()

with(cen.status, table(`2019`=paste(status_19), `2020`=paste(status_20)))

cen.status %>% mutate(year = "2019") %>% left_join(treatments) %>% filter(status_19 %in% status_ok[-3], status_20 %in% status_ok) %>% 
ggplot(aes(x=water, fill=status_20)) + facet_wrap(status_19 ~ snow, labeller=labeller(.cols=label_both)) +  
  labs(y="2020 Status", fill="2020 Status", x="Precipitation") +
  geom_bar(position="fill") + scale_fill_manual(values=status_pal) + theme_minimal()
```

## Survival and flowering vs. vegetative plant size
```{r surv_size}
ggplot(filter(cen, status_19 %in% status_ok, status_18 == "vegetative"), aes(x=r1_longest_18, fill=status_19)) +
  facet_wrap(vars(status_19), ncol=1) + labs(x="2018 Rosette 1 Longest Leaf (mm)", fill="2019 status") +
  geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 85))
ggplot(filter(cen, status_20 %in% status_ok, status_19 == "vegetative"), aes(x=r1_longest_19, fill=status_20)) +
  facet_wrap(vars(status_20), ncol=1) + labs(x="2019 Rosette 1 Longest Leaf (mm)", fill="2020 status") +
   geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 85))

ggplot(filter(cen, status_19 %in% status_ok, status_18 == "vegetative"), aes(x=r1_leaves_18, fill=status_19)) +
  facet_wrap(vars(status_19), ncol=1) + labs(x="2018 Rosette 1 # Leaves", fill="2019 status") +
  geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 100))
ggplot(filter(cen, status_20 %in% status_ok, status_19 == "vegetative"), aes(x=r1_leaves_19, fill=status_20)) +
  facet_wrap(vars(status_20), ncol=1) + labs(x="2019 Rosette 1 # Leaves", fill="2020 status") +
  geom_density() + scale_fill_manual(values=status_pal) + coord_cartesian(xlim=c(0, 100))
```

## Plant growth vs. treatments

Only includes plants that remained vegetative and did not die or flower.
```{r growth}
cen %>% drop_na(water_18, snow_18) %>% filter(status_18=="vegetative", status_19=="vegetative") %>% 
  ggplot(aes(x=r1_longest_18, y=r1_longest_19, color=water_18, shape=snow_18)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+coord_fixed(xlim=c(0,80), ylim=c(0,80)) +
  labs(x="2018 Rosette 1 Longest Leaf (mm)", y="2019 Rosette 1 Longest Leaf (mm)",  color="Precipitation", shape="Snowmelt")
cen %>% drop_na(water_19, snow_19) %>% filter(status_19=="vegetative", status_20=="vegetative") %>% 
  ggplot(aes(x=r1_longest_19, y=r1_longest_20, color=water_19, shape=snow_19)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+  coord_fixed(xlim=c(0,80), ylim=c(0,80)) +
    labs(x="2019 Rosette 1 Longest Leaf (mm)", y="2020 Rosette 1 Longest Leaf (mm)",  color="Precipitation", shape="Snowmelt")

cen %>% drop_na(water_18, snow_18) %>% filter(status_18=="vegetative", status_19=="vegetative") %>% 
  ggplot(aes(x=r1_leaves_18, y=r1_leaves_19, color=water_18, shape=snow_18)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+coord_fixed(xlim=c(0,100), ylim=c(0,100)) +
  labs(x="2018 Rosette 1 # Leaves", y="2019 Rosette 1 # Leaves",  color="Precipitation", shape="Snowmelt")
cen %>% drop_na(water_19, snow_19) %>% filter(status_19=="vegetative", status_20=="vegetative") %>% 
  ggplot(aes(x=r1_leaves_19, y=r1_leaves_20, color=water_19, shape=snow_19)) + geom_point() + geom_abline(intercept=0, slope=1) +
  scale_color_manual(values=water_pal) + theme_minimal()+  coord_fixed(xlim=c(0,100), ylim=c(0,100)) +
    labs(x="2019 Rosette 1 # Leaves", y="2020 Rosette 1 # Leaves",  color="Precipitation", shape="Snowmelt")
```

## Plant size vs. treatments
```{r plantsize}
ggplot(drop_na(cen18, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2019 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen18, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal)+ labs(x="Precipitation", y="2019 Rosette 1 # Leaves",  color="Snowmelt")

ggplot(drop_na(cen19, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2019 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen19, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal)+ labs(x="Precipitation", y="2019 Rosette 1 # Leaves",  color="Snowmelt")

ggplot(drop_na(cen20, water, snow), aes(x=water, color=snow, y=r1_longest)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2020 Rosette 1 Longest Leaf (mm)",  color="Snowmelt")
ggplot(drop_na(cen20, water, snow), aes(x=water, color=snow, y=r1_leaves)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="2020 Rosette 1 # Leaves",  color="Snowmelt")
```

# Leaf traits

```{r checkleaftraits}
ggplot(lt, aes(x=wet_weight_g, y=dry_weight_g)) + geom_point() + geom_abline(slope=1, intercept = 0) + geom_abline(slope=1-median(lt$water_content), intercept = 0)
ggplot(lt, aes(x=leaf_area_cm2, y=dry_weight_g)) + geom_point() + geom_smooth(method="lm")
ggplot(lt, aes(x=leaf_area_cm2, y=wet_weight_g)) + geom_point() + geom_smooth(method="lm")
```


```{r leaftraits}
ggplot(lt, aes(x=water, color=snow, y = sla)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="Specific Leaf Area (cm2/g)", color="Snowmelt") + facet_wrap(vars(year, round), nrow=1) + scale_y_continuous(limits=c(0,300))

ggplot(lt, aes(x=sla, y=water_content, color=water, shape=snow, linetype=snow)) + geom_point() +  scale_color_manual(values=water_pal) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(lt, aes(x=water, color=snow, y = water_content)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="Water content", color="Snowmelt") + facet_wrap(vars(year, round), nrow=1)

ggplot(lt, aes(x=water, color=snow, y = trichome_density)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="Trichome density (/cm2)", color="Snowmelt") + facet_wrap(vars(year, round), nrow=1)
```

# Physiology traits

```{r licor_all, fig.width=12, fig.height=12}
# plot all the LICOR variables by date
licor_date_pal <- sample(rainbow(nlevels(factor(licor$date))))
licor %>% pivot_longer(Photo:StableF) %>% mutate(index=row_number()) %>% filter(!name %in% c("Area","BLC_1","BLCond","StmRat")) %>% 
  ggplot(aes(x=index, y= value, color=factor(date), shape=year)) + facet_wrap(vars(name), scales="free_y", ncol=4) + geom_point() + scale_color_manual(values=licor_date_pal, guide=F) + theme_dark() + labs(x="Measurement", y="", shape="Year")
```

```{r licor}
licor %>% ggplot(aes(x=Cond, y= Photo, color=factor(date), linetype=year, shape=year)) + geom_point() + geom_smooth(method="lm", se=F) + scale_linetype_manual(values=c(3,1,4)) + scale_color_manual(values=licor_date_pal, guide=F) + theme_dark()

licor %>% ggplot(aes(y=Photo/Cond, x=factor(date), color=year)) + geom_boxplot(size=1.5) + 
  scale_y_continuous(limits=c(0,350)) + scale_color_manual(values=year_pal) + scale_x_discrete(guide = guide_axis(angle = 90)) + theme_minimal()

licor %>% left_join(treatments)%>% ggplot(aes(y=Photo/Cond, x=factor(date), color=water)) + 
  geom_boxplot(position=position_dodge2(preserve="single")) + 
  scale_y_continuous(limits=c(0,350)) + scale_color_manual(values=water_pal) + scale_x_discrete(guide = guide_axis(angle = 90)) + theme_minimal()
```

```{r wue}
ggplot(wue, aes(x=water, color=snow, y = photosynthesis)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="Photosynthesis", color="Snowmelt") + facet_wrap(vars(year), nrow=1)

photo.mod <- lmer(photosynthesis ~ year * snow * water + (1|plot), data=wue.subplot)
anova(photo.mod) %>% kable(digits=3)

ggplot(wue, aes(x=water, color=snow, y = conductance)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="Conductance", color="Snowmelt") + facet_wrap(vars(year), nrow=1)

cond.mod <- lmer(conductance ~ year * snow * water + (1|plot), data=wue.subplot)
anova(cond.mod) %>% kable(digits=3)

ggplot(wue, aes(x=water, color=snow, y = WUE)) + geom_boxplot() + scale_color_manual(values=snow_pal) + labs(x="Precipitation", y="Water use efficiency (photo/cond)", color="Snowmelt") + facet_wrap(vars(year), nrow=1)

WUE.mod <- lmer(WUE ~ year * snow * water + (1|plot), data=wue.subplot)
anova(WUE.mod) %>% kable(digits=3)
```


# Floral traits

## Floral morphology, nectar, phenology, seeds

### Floral trait correlations
```{r correlations}

ggpairs(mt %>% select(corolla_length,	corolla_width,	style_length,	anther_max,	anther_min,	sepal_width))

ggplot(nt, aes(x=nectar_24_h_ul, y = nectar_conc)) + geom_point() + labs(x=traitnames[["nectar_conc"]], y=traitnames[["nectar_24_h_ul"]]) + geom_smooth(se=F)
```

### Floral plasticity: water treatments and snowmelt treatments

This is a split-plot analysis, based on the original experimental design. The models use:

* *mnps.subplot*, the 24 * 3 subplot means within each year
* *snow*, the snowmelt treatment (except for plot 2 in 2019)
* *water*, the water treatments with mock rainouts and controls combined
* *plot* as a random effect (subplots vary within this larger unit)

The models are specified as:

    lmer(trait ~ year * snow * water + (1|plot), data = mnps.subplot)

* The models for nectar volume and total sugar are singular (variance at plot level near 0)
* Removing the interactions with year does not make those two models not singular
* The model fits are mostly singular when a random effect of *plotid* (subplot) is added for repeated measures among years

```{r morphology_split}
options(contrasts = c("contr.sum", "contr.poly"))

mt.mod <- mt.mod.test <- mod.emm <- mod.emm.snow <- mt.mod.plot <- list()
for(tr in traits) {
  mt.mod[[tr]] <- lmer(mnps.subplot[[tr]] ~ year * snow * water + (1|plot), data=mnps.subplot)
  mt.mod.test[[tr]] <- anova(mt.mod[[tr]])
  mod.emm[[tr]] <- summary(emmeans(ref_grid(mt.mod[[tr]]), ~ year + snow * water))
  mod.emm.snow[[tr]] <- summary(emmeans(ref_grid(mt.mod[[tr]]), ~ snow))
  
  # mt.mod.plot[[tr]] <- 
  #   ggplot(mod.emm[[tr]] %>% mutate(year=factor(year, levels=2018:2020)), aes(x=water, color=snow, y = emmean)) + facet_wrap(vars(year), nrow=1) +
  #       labs(x="Precipitation", y=traitnames[[tr]], color="Snowmelt") +
  #       geom_point(data=mnps.subplot, aes_string(x="water", color="snow", y = tr), inherit.aes = F, position=position_dodge(0.7)) + 
  #       geom_point(position=position_dodge(0.7), size=4, shape=0) + 
  #       geom_linerange(aes(ymin=emmean-SE, ymax=emmean+SE), position=position_dodge2(0.7)) +
  #       guides(x=guide_axis(angle = 90)) + scale_color_manual(values=snow_pal)
  # print(mt.mod.plot[[tr]])
  #cat(paste(tr,"\n"))
  #print(mt.mod.test[[tr]])
}

mt.mod.tests <- map_dfr(mt.mod.test, tidy, .id="trait")
mt.mod.tests %>%  print_anova_table(mt.mod)
```

```{r multi_splitplot, fig.width=9, fig.height=11}
traits.include <- traits[c(1:4,8,5:7,13)]

mnps.subplot.long <- mnps.subplot %>% select(all_of(traits.include), water, snow, year, VWC, sun_date, precip_est_mm) %>% 
  pivot_longer(all_of(traits.include), names_to="trait") %>% 
  mutate(trait = fct_relevel(trait, traits.include), snow=fct_relevel(snow, "Normal"))

mnps.plantyr.long <- mnps.plantyr %>% select(all_of(traits.include), water, snow, year, VWC, sun_date, precip_est_mm) %>% 
  pivot_longer(all_of(traits.include), names_to="trait") %>% 
  mutate(trait = fct_relevel(trait, traits.include), snow=fct_relevel(snow, "Normal"))

mod.emm.plots <- list()
for(i in 1:2) {
traits.plot <- list(traits.include[1:5], traits.include[6:9])[[i]]
mod.emm.plots[[i]] <- mod.emm %>% bind_rows(.id="trait") %>% filter(trait %in% traits.plot) %>% 
  mutate(trait = fct_relevel(trait, traits.plot), snow=fct_relevel(snow, "Normal")) %>% 
  ggplot(aes(x=water, color=snow)) + 
  facet_grid(trait ~ year, scales="free_y", labeller = as_labeller(c(traitnames, set_names(2018:2020))), switch="y") + 
  geom_violin(data=mnps.plantyr.long %>% drop_na(value, water) %>% filter(trait %in% traits.plot), aes(y=value), 
            position=position_dodge(width=0.8), show.legend=F) + 
  #geom_point(data=mnps.subplot.long %>% filter(trait %in% traits.plot) , aes(y=value),
  #           position=position_dodge(width=0.6), shape=4) +  
  geom_linerange(aes(ymin=emmean-SE, ymax=emmean+SE,), position=position_dodge(width=0.8), size=2, show.legend=F) +
  geom_point(aes(y=emmean), position=position_dodge(width=0.8), shape="-", size=16) +  
  labs(x="Precipitation", y="", color="Snowmelt") + 
  coord_cartesian(ylim=c(0,NA)) + scale_y_continuous(position="right", expand=expansion(mult=c(0,0.05)))+
  scale_x_discrete(guide=guide_axis(angle=90)) + 
  scale_color_manual(values=snow_pal, guide=guide_legend(override.aes = list(shape=15, size=5))) + 
  theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
        panel.border = element_rect(fill=NA, colour = "black"), panel.spacing=unit(12, "pt"),
        legend.position = "top")
  if(i !=2) mod.emm.plots[[i]] <- mod.emm.plots[[i]] + guides(color=F)
}

grid.arrange(mod.emm.plots[[1]], mod.emm.plots[[2]], nrow=1)
```

#### Compensation of water addition for early snowmelt

```{r compensation}
#compare early snowmelt + addition vs. normal snowmelt + control
emm.mod <- mt.mod.glht <- list()
mnps.subplot$snowwater <- paste0(mnps.subplot$snow, mnps.subplot$water)
library(multcomp)#careful, this loads MASS, which overrides dplyr::select
for(tr in traits) {
    #emm.mod[[tr]] <- emmeans(mt.mod[[tr]], ~ water * snow, var=tr)
    mt.mod.glht[[tr]] <- lmer(mnps.subplot[[tr]] ~ snowwater + year+ (1|plot), data=mnps.subplot) %>% # drop year??
      glht(linfct = mcp(snowwater = c("EarlyAddition - NormalControl = 0")))
}
map_dfr(mt.mod.glht, ~ tibble(EarlyAddition_minus_NormalControl = coef(.x), p=summary(.x)$test$pvalues), .id="trait") %>% kable(digits=3)
detach("package:multcomp", unload = TRUE); detach("package:TH.data", unload = TRUE); detach("package:MASS", unload = TRUE)
```


### Floral plasticity: water treatments and snowmelt offset

The differences in snowmelt dates vary among years and plots, so this analysis uses a melt date offset. The models use:

* *mnps.subplot*, the 24 * 3 subplot means within each year
* *water*, the water treatments with mock rainouts and controls combined
* *melt_offset*, the number of days between the snowmelt date (when the light entering the HOBO logger exceeds 10,000 lux) and the mean snowmelt date for the normal plots in that year (according to *snow*)
* *plotid* (subplot) as a random effect (plants vary within this larger unit)

The models are specified as: 

    lmer(trait ~ year * water * melt_offset + (1|plotid), data = mnps.subplot)

```{r morphology_meltoffset}
options(contrasts = c("contr.sum", "contr.poly"))

mt.mod.raw <- mt.mod.plantyr <- mt.mod.subplot <- mt.mod.subplot.test <- mt.mod.plot <- list()
for(tr in traits) {
  #mt.mod.raw[[tr]]     <- lmer(mt[[tr]]         ~ year * water * melt_offset + (1|plotid) + (1|plantid), data=mt)
  #mt.mod.plantyr[[tr]] <- lmer(mt.plantyr[[tr]] ~ year * water * melt_offset + (1|plotid), data=mt.plantyr)
  mt.mod.subplot[[tr]] <- lmer(mnps.subplot[[tr]] ~ year * water * melt_offset + (1|plot), data=mnps.subplot)
  mnps.subplot[[paste0("fit.",tr)]] <- NA
  mnps.subplot[[paste0("fit.",tr)]][!is.na(mnps.subplot[[tr]])]<- predict(mt.mod.subplot[[tr]], re.form=NA)
  mt.mod.subplot.test[[tr]] <- anova(mt.mod.subplot[[tr]])
  
  mt.mod.plot[[tr]] <- 
    ggplot(mnps.subplot, aes_string(x="melt_offset", color="water", y = tr)) + geom_point() +
    geom_line(data=mnps.subplot %>% drop_na(paste0("fit.",tr)), aes_string(y=paste0("fit.",tr)), size=2)+
     scale_color_manual(values=water_pal)+ facet_wrap(vars(year))+
        labs(x="Offset from mean normal snow melt date (days)", y=traitnames[[tr]], color="Precipitation")
  print(mt.mod.plot[[tr]])
}
mt.mod.subplot.tests <- map_dfr(mt.mod.subplot.test, tidy, .id="trait")
mt.mod.subplot.tests %>%  print_anova_table(mt.mod.subplot)
```

### Floral plasticity: absolute snowmelt date and estimated summer precipitation

```{r morphology_meltdate}
#dropped total summer precipitation (early_precip_mm and water:early_precip_mm) since it has n = 3 years
options(contrasts = c("contr.sum", "contr.poly"))

mt.mod.raw <- mt.mod.plantyr <- mt.mod.subplot <- mt.mod.subplot.full <- mt.mod.subplot.test <- mt.mod.plot <- coef.melt.precip <-emm.melt<- emm.precip <- list()
for(tr in traits) {
  #mt.mod.raw[[tr]]     <- lmer(mt[[tr]]         ~ sun_date * precip_est_mm + (1|plotid) + (1|plantid), data=mt)
  #mt.mod.plantyr[[tr]] <- lmer(mt.plantyr[[tr]] ~ sun_date * precip_est_mm + (1|plotid), data=mt.plantyr)
  mt.mod.subplot.full[[tr]] <- lmer(mnps.subplot[[tr]] ~ sun_date * precip_est_mm + (1|plot), data=mnps.subplot)
  #drop interaction if not significant
  if(mt.mod.subplot.full[[tr]] %>% tidy %>% filter(term=="sun_date:precip_est_mm") %>% pull(p.value) > 0.05) {
    mt.mod.subplot[[tr]] <- lmer(mnps.subplot[[tr]] ~ sun_date + precip_est_mm + (1|plot), data=mnps.subplot)
  } else {
    mt.mod.subplot[[tr]] <- mt.mod.subplot.full[[tr]]
  }
  mnps.subplot[[paste0("fit.",tr)]] <- NA
  mnps.subplot[[paste0("fit.",tr)]][!is.na(mnps.subplot[[tr]])]<- predict(mt.mod.subplot[[tr]], re.form=NA)
  mt.mod.subplot.test[[tr]] <- anova(mt.mod.subplot[[tr]])
  
  #Z-score the data first to get standardized slopes
  mt.mod.subplot[[tr]] <- lmer(scale(mnps.subplot[[tr]]) ~ sun_date + precip_est_mm + (1|plot), data=mnps.subplot)
  coef.melt.precip[[tr]] <- summary(mt.mod.subplot[[tr]])$coefficients[c("sun_date", "precip_est_mm"),]
  emm.melt[[tr]] <- emtrends(mt.mod.subplot[[tr]],  ~ 1, var = "sun_date")
  emm.precip[[tr]] <- emtrends(mt.mod.subplot[[tr]],  ~ 1, var = "precip_est_mm")
  
  # mt.mod.plot[[tr]] <-
  #   ggplot(mnps.subplot, aes_string(x="sun_date", color="water", y = tr, shape="year")) + geom_point() +
  #   geom_line(data=mnps.subplot %>% drop_na(paste0("fit.",tr)) %>% mutate(year.water = paste(year, water)), 
  #             aes_string(y=paste0("fit.",tr), group="year.water"), size=2)+
  #    scale_color_manual(values=water_pal)+
  #       labs(x="Snow melt date (day of year)", y=traitnames[[tr]], color="Precipitation", shape="Year")
  # print(mt.mod.plot[[tr]])
}
mt.mod.subplot.coefs <- map_dfr(mt.mod.subplot, tidy, .id="trait")
mt.mod.subplot.tests <- map_dfr(mt.mod.subplot.test, tidy, .id="trait")
mt.mod.subplot.tests %>%  print_anova_table(mt.mod.subplot)

emm.melt.df <- bind_rows(sun_date = emm.melt %>% map(as.data.frame) %>% bind_rows(.id="trait") %>% rename(trend=sun_date.trend)
          ,precip_est_mm = emm.precip %>% map(as.data.frame) %>% bind_rows(.id="trait") %>% rename(trend=precip_est_mm.trend)
          ,.id="predictor") %>% mutate(predictor=fct_relevel(predictor,"sun_date"))

# library(googlesheets4)
# write_sheet(left_join(emm.melt.df %>% mutate(traitname=traitnames[trait], .before="trait"),
#                        mnps.subplot %>% group_by(water) %>% select(traits) %>% 
#                          summarize(across(traits, ~ sum(!is.na(.x)))) %>% 
#                          pivot_longer(traits, names_to="trait", values_to="n"))
#              , ss=filter(datasheets, name=="Maxfield Results"), sheet="slopes_melt")

ggplot(emm.melt.df, aes(x=fct_relevel(trait, rev(traits)), y=trend)) +  
  facet_wrap(vars(predictor), nrow=1, scales="free_x", labeller=labeller(predictor=c(precip_est_mm="Estimated summer precip (mm)", sun_date="Snowmelt date"))) + 
  geom_hline(yintercept=0) + 
  geom_pointrange(aes(ymin=trend - SE, ymax = trend + SE), position=position_dodge(width=0.3), size=0.4) + 
  labs(y="Slope and SE of z-scored trait vs. predictor", x="Floral Trait", color="Water") + scale_x_discrete(labels = traitnames) + coord_flip() + scale_color_manual(values=water_pal) +
  theme(legend.position = "top")
```

```{r multi_meltdate, fig.width=9, fig.height=11}
traits.include <- traits[c(1,2,4,3,8,13,5:7)]

mt.mod.subplot.full[traits.include] %>% 
  map_dfr(~ emmeans(.x, ~ precip_est_mm * sun_date, at=list(sun_date=range(treatments$sun_date), precip_est_mm=seq(25,150, by=25))) %>% as_tibble(), .id="trait") %>% 
  mutate(trait = fct_relevel(trait, traits.include)) %>% 
  ggplot(aes(x=sun_date, y=emmean, color=precip_est_mm, group=precip_est_mm)) + 
  facet_wrap(vars(trait), scales="free_y",labeller = as_labeller(traitnames))+
  geom_point(data = mnps.plantyr.long, aes(y=value, shape=year)) +  geom_line() + 
  scale_color_gradientn(colors=rev(c(water_pal[1],water_pal, water_pal[3])), 
                        values=rev(c(1, treatments %>% group_by(water) %>% summarize(across(precip_est_mm, mean)) %>%
                                       pull(precip_est_mm) %>% scales::rescale(from=c(25,150)),0)),
                        breaks=seq(25,150, by=25)) + 
  labs(x="Snowmelt date", y="", color="Summer\nprecipitation\n(mm)", shape="Year") + 
  coord_cartesian(ylim=c(0,NA)) + scale_y_continuous(expand=expansion(mult=c(0,0.05)))+
  theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                          panel.border = element_rect(fill=NA, colour = "black"))
```

### Floral plasticity: soil moisture

Both snowmelt timing and summer precipitation affect soil moisture and its timing. This analysis regresses the floral traits against the mean soil moisture in a subplot. Slopes may differ among years. The models use:

* *mt.plantyr*, the trait data averaged by plant within year
* *VWC*, the summer mean of the approximately weekly mean volumetric water content in a subplot. This may need to change to a subset of the summer because it includes soil moisture taken after traits were measured. See section on within-season variation for timing of measurements in each year.
* *plotid* (subplot) as a random effect (plants vary within this larger unit)

The models are specified as:

    lmer(trait ~ year * VWC + (1|plotid), data = mt.plantyr)

* The model for total sugar is singular (variance at subplot level near 0)

```{r morphology_sm}
options(contrasts = c("contr.sum", "contr.poly"))

mt.sm.mod <- mt.sm.mod.z <- mt.sm.mod.test <- mt.sm.mod.raw <- coef.VWC <- emm.VWC <- emm.VWC.allyears <- mt.sm.mod.plot <-list()
for(tr in traits) {
  #mt.sm.mod.raw[[tr]] <- lmer(mt[[tr]]    ~ year * VWC + (1|plot) + (1|plantid), data=mt)
  mt.sm.mod[[tr]] <-lmer(mnps.plantyr[[tr]] ~ year * VWC + (1|plotid), data=mnps.plantyr)
  mnps.plantyr[[paste0("fit.sm.",tr)]] <- NA
  mnps.plantyr[[paste0("fit.sm.",tr)]][!is.na(mnps.plantyr[[tr]])]<- predict(mt.sm.mod[[tr]], re.form=NA) #TODO why jumps?
  mt.sm.mod.test[[tr]] <- anova(mt.sm.mod[[tr]])
  
  #Z-score the data first to get standardized slopes
  mt.sm.mod.z[[tr]] <- lmer(scale(mnps.plantyr[[tr]]) ~ year * VWC + (1|plotid), data=mnps.plantyr)
  coef.VWC[[tr]] <- summary(mt.sm.mod.z[[tr]])$coefficients["VWC",]
  emm.VWC[[tr]] <- emtrends(mt.sm.mod.z[[tr]],  ~ year, var = "VWC")
  emm.VWC.allyears[[tr]] <- emtrends(mt.sm.mod.z[[tr]],  ~ 1, var = "VWC")

  #  mt.sm.mod.plot[[tr]] <- 
  #   ggplot(mnps.plantyr, aes_string(x="VWC", color="year", y = tr)) + geom_point() +
  #   geom_line(data=mnps.plantyr %>% drop_na(paste0("fit.sm.",tr)), aes_string(y=paste0("fit.sm.",tr)), size=2)+
  #    scale_color_manual(values=year_pal)+
  #       labs(x="Summer mean of soil moisture in subplot (VWC)", y=traitnames[[tr]], color="Year")
  # print(mt.sm.mod.plot[[tr]])
}
mt.sm.mod.tests <- map_dfr(mt.sm.mod.test, tidy, .id="trait")
mt.sm.mod.tests %>%  print_anova_table(mt.sm.mod)
coef.VWC.df <- as.data.frame(t(as.data.frame(coef.VWC))) %>% rownames_to_column("trait")
emm.VWC.ls <- map(emm.VWC, as.data.frame)
emm.VWC.ls$sepal_width[3,] <- NA
emm.VWC.df  <- bind_rows(emm.VWC.ls, .id="trait")
emm.VWC.allyears[c(1:3,5)] %>% map(as.data.frame) %>% bind_rows(.id="trait")

# library(googlesheets4)
# write_sheet(left_join(emm.VWC.df %>% mutate(predictor="VWC", .before=1) %>%
#                         mutate(traitname=traitnames[trait], .before="trait"),
#                       mnps.subplot %>% group_by(year) %>% select(traits) %>%
#                         summarize(across(traits, ~ sum(!is.na(.x)))) %>%
#                         pivot_longer(traits, names_to="trait", values_to="n"))
#             , ss=filter(datasheets, name=="Maxfield Results"), sheet="slopes_VWC")

ggplot(emm.VWC.df %>% drop_na(), aes(x=trait, y=VWC.trend, color=year)) +  
  geom_hline(yintercept=0) + geom_pointrange(aes(ymin=VWC.trend - SE, ymax = VWC.trend + SE), position=position_dodge(width=0.3), size=0.4) + 
  labs(y="Slope of z-scored trait vs. mean annual soil moisture", x="Floral Trait", color="Year") +
  theme_classic() + scale_x_discrete(labels = traitnames) + coord_flip() + scale_color_manual(values=year_pal)
```

```{r multi_sm, fig.width=9, fig.height=11}
traits.include <- traits[c(1,2,4,3,8,13,5:7)]

mt.sm.mod[traits.include] %>% 
  map_dfr(~ emmeans(.x, ~ year*VWC, at=list(VWC=range(sm.subplotyear$VWC))) %>% as_tibble(), .id="trait") %>% 
  mutate(trait = fct_relevel(trait, traits.include)) %>% 
  ggplot(aes(x=VWC, y=emmean, color=year, group=year)) + 
  facet_wrap(vars(trait), scales="free_y",labeller = as_labeller(traitnames))+
  geom_point(data = mnps.plantyr.long, aes(y=value, color=year)) +  
  geom_line(size=2) + 
  scale_color_manual(values=year_pal) + 
  labs(x="Summer mean of soil moisture in subplot (%VWC)", y="", color="Year") + 
  coord_cartesian(ylim=c(0,NA)) + scale_y_continuous(expand=expansion(mult=c(0,0.05)))+
  theme_minimal() + theme(text=element_text(color="black", size=14), axis.text = element_text(color="black"),
                          panel.border = element_rect(fill=NA, colour = "black"))

```


### Traits for plasticity paper
```{r emmeans_filtered, fig.width=8, fig.height=5}
bind_rows(sun_date = emm.melt %>% map(as.data.frame) %>% bind_rows(.id="trait") %>% rename(trend=sun_date.trend),
          precip_est_mm = emm.precip %>% map(as.data.frame) %>% bind_rows(.id="trait") %>% rename(trend=precip_est_mm.trend),      .id="predictor") %>% mutate(predictor=fct_relevel(predictor,"sun_date")) %>%
  filter(trait %in% traits[c(1:8, 13)]) %>% 
  ggplot(aes(x=fct_relevel(trait, rev(traits)), y=trend)) +  
  facet_wrap(vars(predictor), nrow=1, scales="free_x", labeller=labeller(predictor=c(precip_est_mm="Summer precip (mm)", sun_date="Snowmelt date"))) + 
  geom_hline(yintercept=0) + 
  geom_pointrange(aes(ymin=trend - SE, ymax = trend + SE), position=position_dodge(width=0.3), size=0.4) + 
  labs(y="Slope of z-scored trait vs. predictor", x="Floral Trait", color="Water") + 
  scale_x_discrete(labels = traitnames) + coord_flip() +
  theme(legend.position = "top")

ggplot(emm.VWC.df %>% drop_na() %>% filter(trait %in% traits[c(1:8, 13)]), aes(x=fct_relevel(trait, rev(traits)), y=VWC.trend, color=year)) +  
  geom_hline(yintercept=0) + geom_pointrange(aes(ymin=VWC.trend - SE, ymax = VWC.trend + SE), position=position_dodge(width=0.3), size=0.4) + 
  labs(y="Slope of z-scored trait vs. mean annual soil moisture", x="Floral Trait", color="Year") +
  theme_classic() + scale_x_discrete(labels = traitnames) + coord_flip() + scale_color_manual(values=year_pal)
```

### Multivariate response
```{r multresp}
mt.morph <- mt[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","VWC")] %>% drop_na(1:5) #drop sepal_width
mt.pca <- prcomp(mt.morph[,-(6:7)], scale.=TRUE)

mt.morph.pca <- mt.morph %>%
  bind_cols(as.data.frame(mt.pca$x)) %>% 
  tibble::rowid_to_column("index") %>% 
  mutate(x_0=0,y_0=0,
         x_1=-corolla_width/2,y_1=corolla_length,
         x_2= corolla_width/2,y_2=corolla_length) %>% 
  pivot_longer(c(starts_with("x"),starts_with("y")), names_to=c(".value", "point"), names_sep="_")

scl <- 100
zoom <- 5.5
mt.pca.vars <- as.data.frame(mt.pca$rotation) %>% rownames_to_column("variable")
mt.pca.labs <- paste0("PC", 1:2, " (", round(100*mt.pca$sdev[1:2]^2/sum(mt.pca$sdev^2), 0), "% explained var.)") 
ggplot(mt.morph.pca) + theme_classic() +
  geom_polygon(aes(x= x/scl+PC1, y= y/scl+PC2, group=index, color=VWC, fill=VWC)) +
  #geom_point(aes(x=PC1, y=PC2, size=sepal_width/2), color="darkgreen") + 
  geom_point(aes(x=PC1, y=PC2 + anther_min/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + anther_max/scl), color="yellow", size=0.3) +
  geom_point(aes(x=PC1, y=PC2 + style_length/scl), color="grey60", size=0.5, shape=4) +
  geom_segment(data=mt.pca.vars,aes(x=0, y=0, xend=PC1*zoom, yend=PC2*zoom), 
               arrow = arrow(length = unit(1/2, "picas")), color="black") +
  geom_text(data =mt.pca.vars, aes(label = gsub("_"," ",variable), x = PC1*zoom*1.4, y = PC2*zoom*1.4,
                                   angle =  (180/pi) * atan(PC2/PC1), hjust = 0), color="black")+
  scale_size_identity() + coord_fixed() + 
  scale_fill_viridis(direction=-1, option="plasma") + scale_color_viridis(direction=-1, option="plasma") + 
  labs(x=mt.pca.labs[1], y=mt.pca.labs[2])
ggsave("images/mt_morph_pca_noSW.pdf", height=15, width=20, device=cairo_pdf)

ggplot(mt, aes(x=corolla_width, y=corolla_length, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal()
ggplot(mt, aes(y=style_length / corolla_length, x=VWC, color=VWC)) + scale_color_viridis(direction=-1, option="plasma") + geom_point() + theme_minimal() + geom_smooth()

options(contrasts = c("contr.treatment", "contr.poly"))
mt.cap <- capscale(scale(mt.morph[,-(6:7)]) ~ VWC + year, data=mt.morph)
mt.cap
anova(mt.cap, by="term") %>% tidy %>% kable

autoplot(mt.cap) + guides(color=F, shape=F)
```

```{r multresp_discrete}
mt.morph.wm <- mt[,c("corolla_length", "corolla_width","style_length","anther_max","anther_min","year","water","melt_offset")] %>% drop_na(1:6) #and nectar (19,20) too
mt.cap.wm <- capscale(scale(mt.morph.wm[,-(6:8)]) ~ year * melt_offset * water, data=mt.morph.wm)
mt.cap.wm
anova(mt.cap.wm, by="term") %>% tidy %>% kable
#plot CAP without interactions
autoplot(capscale(scale(mt.morph.wm[,-(6:8)]) ~ year + melt_offset + water, data=mt.morph.wm)) + guides(color=F, shape=F)
```

### Floral traits within season

No obvious effect of time within a season for corolla length/width or sepal width. The timing of trait measurements was different each year:

```{r morph_season}
ggplot(sm, aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point(data=ph, aes(y= height_cm/2, color="Phenology (height)"))+
  geom_point(data=mt, aes(y= corolla_length, color="Morphology"))+
  geom_point(data=mt, aes(y= nectar_24_h_ul, color="Nectar"))+
  geom_point(aes(y = VWC, color="Soil moisture")) +
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow), color="grey40")+
  labs(x="Day of year", shape="Snow", linetype="Snow", y="", color="Measurement")+
  scale_color_manual(values=c("purple","orange","grey","black"))+
  scale_x_continuous(breaks=seq(120, 240, by=20))+
  scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(1.5, "lines"), axis.text.y=element_blank())

ggplot(data=mt, aes(x=yday(date), y= corolla_length, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y=traitnames[["corolla_length"]], x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()

ggplot(data=mt, aes(x=yday(date), y= nectar_24_h_ul, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y=traitnames[["nectar_24_h_ul"]], x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()

ggplot(data=nt, aes(x=yday(date), y= nectar_conc, color=water)) + 
  facet_wrap(vars(year), ncol=1) +
  geom_point()+   geom_smooth(se=F)+ 
  labs(y=traitnames[["nectar_conc"]], x="Day of year")+
  scale_color_manual(values=water_pal) + theme_minimal()
```

# Phenology
```{r check_phenology}
walk(list(ph18.raw, ph19.raw, ph20.raw), 
     function(ph.raw) {
  y = ph.raw$year[1]
  p <- ph.raw %>% count(plot, subplot, plotid, plantid, julian) %>% add_count(plantid, name="n_plantid") %>% 
  mutate(plantid = fct_reorder(plantid, as.integer(factor(paste(plotid, n_plantid))),.desc=T)) %>% 
  ggplot(aes(julian, plantid, fill=factor(n)))+
  geom_tile()+ scale_fill_viridis(option="inferno", discrete=T) + 
  geom_point(aes(color=subplot, x=median(as.integer(julian))))+ scale_color_manual(values=sample(rainbow(10)))+
  geom_point(aes(color=plot, x=median(as.integer(julian))), position=position_nudge(x=-0.3))+ 
  theme_classic() + theme(panel.background = element_rect(fill="grey")) + guides(x=guide_axis(angle = 90)) +
  labs(x="Day of year", y="Plant (sorted by subplot)", fill="Observations", color="Plot", title=paste(y,"phenology"))
  print(p)})
```


```{r phenology}
ggplot(ph %>% drop_na(height_cm), aes(x=julian, y=height_cm, shape=snow, color=water))+ geom_point() + geom_path(aes(group=plantid))+
  facet_wrap(vars(year), ncol=1)+
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Plant height", 
       x="Day of year", color="Precipitation", shape="Snow") +
  scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_color_manual(values=water_pal) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at(c("open","buds"), mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=open, shape=snow, linetype=snow, color=water))+ geom_point() + 
  facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
  geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Mean total flowers", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
    scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at("flowering", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=flowering, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Proportion flowering", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at("eggs", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=eggs, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Mean total eggs", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
    scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ph %>% group_by(year, snow, water, julian) %>% summarize_at("has_egg", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=has_egg, shape=snow, linetype=snow, color=water))+ geom_point() + 
    facet_wrap(vars(year), ncol=1)+
  geom_path(aes(group=paste(snow, water))) +
    geom_vline(data=meltdates, aes(xintercept=sun_date, linetype=snow))+
  labs(y="Proportion of plants with eggs", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
    scale_x_continuous(breaks=seq(120,240, by=20))+
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

```

## Floral display size
### 2020
```{r display_size}
ggplot(ph.plantid, aes(x=height_cm, y=open, color=water)) + geom_point() + geom_smooth(method="lm", se=F) + scale_color_manual(values=water_pal) + facet_wrap(vars(year), nrow=1) +
  labs(x="Mean inflorescence height (cm)", y="Mean number of open flowers", color="Precipitation")

ggplot(ph.plantid, aes(x=water, color=snow, y = height_cm)) +
    labs(x="Precipitation", y="Mean inflorescence height (cm)", color="Snowmelt") +
    geom_boxplot() + scale_color_manual(values=snow_pal) + facet_wrap(vars(year), nrow=1)
ggplot(ph.plantid, aes(x=VWC, y=height_cm, color=year)) + geom_point() + geom_smooth(se=F) +
    scale_color_manual(values=year_pal) + labs(x="Mean VWC", y="Mean inflorescence height (cm)") 
    
ggplot(ph.plantid, aes(x=water, color=snow, y = open)) +
    labs(x="Precipitation", y="Mean number of open flowers", color="Snowmelt") +
    geom_boxplot() + scale_color_manual(values=snow_pal) + facet_wrap(vars(year), nrow=1)
ggplot(ph.plantid, aes(x=VWC, y=open, color=year)) + geom_point() + geom_smooth(se=F)+
  scale_color_manual(values=year_pal) + labs(x="Mean VWC", y="Mean number of open flowers")
```

# Seeds

```{r seeds}
ggplot(sds %>% drop_na(plot), aes(color=water, x=fruits, y=seeds)) + geom_point() + geom_smooth(se=F) + 
  scale_color_manual(values=water_pal) + facet_wrap(vars(year), nrow=1) + theme_minimal() +
  labs(x="Counted fruits", y="Counted seeds", color="Precipitation")

for(tr in seedtraits) {
  print(ggplot(sds %>% drop_na(plot), aes_string(tr, x="water", color="snow")) + geom_boxplot() + 
    scale_color_manual(values=snow_pal) + facet_wrap(vars(year), nrow=1) + 
    scale_y_continuous(trans=ifelse(tr %in% c("seeds_est", "fruits_nonaborted"),"sqrt","identity"))+
    labs(x="Precipitation", y=seedtraitnames[[tr]], color="Precipitation") +
        guides(x=guide_axis(angle = 90)))
  
  print(ggplot(sds, aes_string(tr, x="VWC", color="year")) + geom_point() + geom_smooth(se=F) + 
   scale_color_manual(values=year_pal) +
    scale_y_continuous(trans=ifelse(tr %in% c("seeds_est", "fruits_nonaborted"),"sqrt","identity"))+
    labs(x="Summer mean of soil moisture in subplot (VWC)", y=seedtraitnames[[tr]], color="Year"))
}
```

## Fruiting phenology (2020)

```{r seed_phenology}
ggplot(sds20.date.zeroed, aes(x=julian, y=fruits_nonaborted, color=water, shape=snow, linetype=snow))+geom_jitter(width=2, height=0) + stat_summary(fun="mean", geom="line") + scale_y_sqrt() + scale_color_manual(values=water_pal)+
  labs(y="Nonaborted fruits", x="Day of year", color="Precipitation", shape="Snow", linetype="Snow")+ scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(sds20.date.zeroed, aes(x=julian, y=seeds, color=water, shape=snow, linetype=snow))+geom_jitter(width=2, height=0) + stat_summary(fun="mean", geom="line") + scale_y_sqrt() + scale_color_manual(values=water_pal)+
  labs(y="Seeds", x="Day of year", color="Precipitation", shape="Snow", linetype="Snow")+ scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

ggplot(sds20.date.zeroed, aes(x=julian, y=prop_infested, color=water, shape=snow, linetype=snow))+geom_jitter(width=2, height=0) + stat_summary(fun="mean", geom="line") + scale_color_manual(values=water_pal)+
  labs(y="Proportion of nonaborted fruits infested", x="Day of year", color="Precipitation", shape="Snow", linetype="Snow")+ scale_linetype_manual(values=c(2,1)) +
  scale_y_continuous(labels = scales::percent)+ theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))

sds20.date.zeroed %>% group_by(snow, water, julian) %>% summarize_at("fruiting", mean, na.rm=T) %>% 
  ggplot(aes(x=julian, y=fruiting, shape=snow, linetype=snow, color=water))+ geom_point() +
  geom_path(aes(group=paste(snow, water))) +
  labs(y="Proportion fruiting (nonaborted fruits)", 
       x="Day of year", color="Precipitation", shape="Snow", linetype="Snow") +
  scale_y_continuous(labels = scales::percent)+
  scale_color_manual(values=water_pal) + scale_linetype_manual(values=c(2,1)) + theme_minimal() + 
  theme(legend.key.size = unit(2, "lines"))
```

