---
title: "Maxfield Meadow Ipomopsis Volatiles"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>


```{r setup, message=FALSE}
library(tidyverse)
library(lubridate)
library(vegan)
library(broom)
library(emmeans)
library(viridis)
library(ggvegan)
library(gridExtra)
library(dendsort)
library(pheatmap)
library(RColorBrewer)
library(DT)
library(lme4)
library(lmerTest)
library(bouquet)
library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, 
                      fig.path = "images_volatiles2020/", dev="svglite", dev.args=list(fix_text_size=FALSE), fig.height=8, fig.width=8)
options(digits=4) # for kables
```

```{r loadscents}
load("data/maxfield_data.rda") #load trait measurements, etc
load("data/volatiles/maxfield_volatiles_lessdupes.rda") # from maxfield_gc.R, loads maxfmeta and maxf (wide GC data)

#check for duplicates in vt and maxfmeta
#vt %>% filter(!exclude %in% c("flowers", "incomplete", "resample")) %>% count(sampledate.file, plot.file, subplot.file, plant.file) %>% filter(n>1)  #exclude="resample" #no duplicates in vt
#I think these plantid duplicates are all just two samples taken on the same day, except maybe 3D858
# maxfmeta %>% count(sampledate,plantid) %>% filter(n>1)
# maxfmeta %>% count(sampledate,plantid, other) %>% filter(n>1)# other column has info on vial, run position, etc

volsmeta <- maxfmeta %>% mutate(sampledate = as.Date(sampledate), year = factor(year(sampledate))) %>% 
  drop_na(plotid) %>% #cut out ambients, leaving 459 chromatograms
  filter(year %in% c("2018","2019","2020")) %>% droplevels() %>% 
  mutate(plot=str_sub(plotid, 1,1), subplot=str_sub(plotid, 2,2),
         sequence.date = date(sequence.start),
         sequence.year = year(sequence.date)) %>% 
  rename(sampledate.file=sampledate, plot.file = plot, subplot.file=subplot, plant.file=plant) %>% 
  left_join(vt %>% select(-c(colnames(treatments),"type","plantid")) %>% 
              filter(!exclude %in% c("flowers", "incomplete", "resample"))) %>% 
  left_join(treatments) %>% #assume that the written metadata is correct not the info in the  .file name
  mutate(total_time = end - bag,
         total_time = replace_na(total_time, median(total_time, na.rm=T)),# fill in 45 minutes when missing metadata
         pump_id=as.character(pump_id))
volsmeta <- left_join(volsmeta, volsmeta %>% count(sequence.date, sequence.start) %>% group_by(sequence.date) %>% mutate(sequence.12 = paste(sequence.date, row_number()))) #add the batch number for that day (1 or 2)
rownames(volsmeta) <- volsmeta$index 
```


# Sample size

Calculated from two datasets that will eventually match: datasheets from sampling and GC-MS filenames.

```{r samplesizes, fig.width=10, fig.height=5}
bind_rows(sampling=vt %>% group_by(year, type) %>% tally(),
  filenames=maxfmeta %>% mutate(year=factor(year(sampledate))) %>% group_by(year, type) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year,type), values_from = n) %>% kable()

bind_rows(sampling=vt %>% filter(type=="floral") %>% group_by(year, water, snow) %>% tally(),
  filenames=volsmeta %>% mutate(year=factor(year(sampledate.file))) %>% group_by(year, water, snow) %>% tally(), .id="source") %>% 
  pivot_wider(names_from=c(year), values_from = n) %>% kable()

bind_rows(sampling = vt %>% filter(type=="floral") %>% group_by(sampledate.file) %>% tally(),
         filenames = volsmeta %>% drop_na(sampledate.file) %>% group_by(sampledate.file) %>% tally(), .id="source") %>% 
  ggplot(aes(x=yday(sampledate.file), y=n, fill=source)) + geom_col(position=position_dodge(preserve="single")) +
  facet_wrap(vars(year(sampledate.file)), ncol=1) +
  labs(x="VOC sampling date (day of year)", y="Floral samples")
```

## Matching datasheets and filenames

```{r matching}
# diagnostics for matching up vt (datasheets) and volsmeta (filenames)
volsmeta.samp <- volsmeta %>% mutate(samp = paste(sampledate.file, plot.file, subplot.file, plant.file)) %>% arrange(samp) %>% pull(samp)
vt.samp <- vt %>% mutate(samp = paste(sampledate.file, plot.file, subplot.file, plant.file)) %>% arrange(samp) %>% pull(samp)
#intersect(volsmeta.samp, vt.samp) #in both volsmeta and vt
cat("Have files but not datasheet entries:")
setdiff(volsmeta.samp, vt.samp)
cat("Have datasheet entries but no files:")
setdiff(vt.samp, volsmeta.samp)

vt.volsmeta <- full_join(vt %>% select(vt.n, sampledate.file, plot.file, subplot.file, plant.file) %>% mutate(in.vt=T),
                         volsmeta %>% select(FileName, sampledate.file, plot.file, subplot.file, plant.file) %>% mutate(in.volsmeta=T)) %>% 
  mutate(across(c(in.vt, in.volsmeta), ~replace_na(.,FALSE)))
vt.volsmeta.count <- vt.volsmeta  %>% count(sampledate.file, in.vt, in.volsmeta) %>% mutate(status=paste("datasheet",ifelse(in.vt,"yes","no"),"+","GCfile", ifelse(in.volsmeta,"yes","no")))
#View(vt.volsmeta.count %>% filter(!in.vt | !in.volsmeta) %>% arrange(sampledate.file))

ggplot(vt.volsmeta.count, aes(x=as.Date(sampledate.file),y=n, fill=status))+ geom_col() +
  facet_wrap(vars(year(sampledate.file)), scales="free_x", ncol=1) +
  scale_x_date(breaks=scales::date_breaks(width = "7 days"), minor_breaks = scales::date_breaks(width = "1 days")) +
  theme_minimal() + theme(legend.position="top") + scale_fill_manual(values=c("red","orange","darkgreen")) +
  labs(x="Sampling date (according to GCfile)", y="Samples", fill="Data status")

ggplot(volsmeta %>% filter(year %in% c(2018, 2020) | sequence.year==year), aes(x=sequence.date - sampledate)) + 
  geom_histogram() + facet_wrap(vars(year), ncol=1, scales="free_x") + theme_minimal()
```


# Soil moisture

```{r sm}
ggplot(vt %>% drop_na(water) %>% mutate(date=sampledate), aes(x=yday(date))) + 
  facet_wrap(vars(year), ncol = 1) +
  scale_x_continuous(limits=c(160, 230)) +
  geom_col(data=daily_precip_est, aes(y=EPA_NOAA_filled), fill=brewer.pal(7, "Set3")[7]) +
  geom_point(aes(y=VWC, color=water), size=0.5) + 
  geom_line(aes(y=VWC, color=water),
            data=vt %>% drop_na(water, VWC) %>% group_by(year, water, date=sampledate) %>% summarize(VWC=mean(VWC)), size=1) + 
  geom_line(aes(y=VWC, color=water),
            data=sm %>% group_by(year, water, date) %>% summarize(VWC=mean(VWC)), size=0.5, alpha=0.5) + 
  scale_color_manual(values=water_pal) + 
  labs(x="Day of year", y="Soil moisture (%VWC) / Daily precipitation (mm)", color="Precipitation") + theme_bw()
```

# Ambient controls
```{r ambients}
vt.ambi %>% filter(type=="ambient") %>% mutate(year=year(sampledate)) %>% 
ggplot(aes(x=yday(sampledate))) + geom_bar() + facet_wrap(vars(year), ncol=1) + 
  labs(x="Day of year", y="Ambient samples") + scale_fill_manual(values=water_snow_pal)+  theme_minimal() 

print("dates missing ambient controls")
setdiff(vt.ambi %>% filter(type=="floral") %>% pull(sampledate) %>% unique, 
        vt.ambi %>% filter(type=="ambient") %>% pull(sampledate) %>% unique) %>% as.Date(origin=ymd("1970-01-01"))

vt.ambi %>% count(type) #TODO check if all of these were run and used for filtering
volsmeta %>% count(year, type)
volsmeta %>% count(type)

ggplot(vt.ambi %>% filter(type=="ambient"), aes(x=hour(bag), fill=year)) + 
  geom_histogram(show.legend = F, binwidth=1) + facet_wrap(vars(year), ncol=1) + 
  scale_fill_manual(values=year_pal) + theme_minimal() + 
  labs(x="Ambient control bagged (hr)", caption= paste(format(c(min(volsmeta$bag, na.rm=T), max(volsmeta$bag, na.rm=T)), "%H:%M"), collapse=" - "))
```


# GCMS data checks

## Bouquet

```{r bouquetprep}
# load short names and standard regessions
ipochems <- read_csv("data/volatiles/Ipo volatile compounds - chemsf_ipo.csv") %>% 
  select(name, shortname, standard, verdict) %>%  filter(verdict != "") %>% 
  mutate(standard = fct_recode(standard, "Methyl_salicylate"="Benzaldehyde"), # benzaldehyde regressions not reliable
         class = fct_recode(standard, Aliphatic="Hexenol", Benzenoid="Methyl_salicylate", Benzenoid="Indole", 
                            Sesquiterpene="Caryophyllene", Monoterpene="alpha_Pinene", Monoterpene="Linalool")) %>%  
  left_join(read_csv("data/volatiles/regressions_181921_filtered_slopes.csv") %>% 
              pivot_wider(id_cols=standard, names_from="year", names_prefix="area_per_ng", values_from=area_per_ng)) 
class_pal <- set_names(c("#BC0060","#027B8C","#E56E00","#86D400"), levels(ipochems$class))[-2]#no more

#shorten chemical names and merge compounds with multiple names
shortnames <- ipochems %>% select(name, shortname) %>% filter(shortname!="") %>% deframe()
#shortnames[shortnames %in% shortnames[duplicated(shortnames)]]

greekify <- function(names) {
  names %>% 
    str_replace("^a-","\U03B1-") %>% str_replace("-a-","-\U03B1-") %>% 
    str_replace("^b-","\U03B2-") %>% str_replace("-b-","-\U03B2-") %>% 
    str_replace("^g-","\U03B3-") %>% str_replace("-g-","-\U03B3-")
}

metadata <- maxfmeta %>% 
  mutate(sampledate = as.Date(sampledate), year = year(sampledate)) %>% 
  filter(year != 2021) %>% mutate(year=factor(year)) %>% #drop samples collected in 2021
  as.data.frame() %>% #tibbles break bouquet's add_count_freqs
  select(-sample) %>% #name conflict with bouquet
  load_metadata(date="Date", sample="FileName", type="type")#group="year"

load("data/volatiles/maxfield210827.Rdata") #load long maxf.data
longdata <- maxf.data %>% 
  filter(batch != "heather_maxfield190813_new.txt", #duplicates - subset of heather_maxfield190813_new001.txt
         Filename %in% metadata$sample, #match list in metadata
         Name != "") %>% #drop unidentified peaks
  mutate(Name = droplevels(recode(Name, !!!shortnames))) %>% 
  load_longdata(sample="Filename", RT="Ret.Time", name="Name", area="Area", match = "SI", maxmatch=100)
remove(maxf.data)

vol.all <- make_sampletable(longdata, metadata)
chems <- make_chemtable(longdata, metadata)
```

```{r bouquet}
chemsf <- chems %>%
  filter_RT(2, 17) %>% 
  filter_match(0.8) %>% 
  filter_freq(0.1) %>% 
  filter_contaminant(cont.list = c("Butanenitrile, 2-methyl-","2-Oxepanone, 7-methyl-", #RT order
                                   "Ethylbenzene", "Hexyl chloroformate", "L-Lactic acid",
                                   "1,5-Dimethyl-6-oxa-bicyclo[3.1.0]hexane","Cyclopentanone, 2-ethyl-",
                                   "trisiloxane, 1,1,1,5,5,5-hexamethyl-3-[(trimethylsilyl)oxy]-",
                                   "Cyclobutane, 1,2-bis(1-methylethenyl)-, trans-",
                                   "N-.alpha.,N-.omega.-Di-cbz-L-arginine", "3-Ethyl-3-methylheptane",
                                   "Trichloroacetic acid, 6-ethyl-3-octyl ester", "11-Methyldodecanol",
                                   "8-Hexadecenal, 14-methyl-, (Z)-",
                                   "Tritetracontane","Carbonic acid, decyl undecyl ester",
                                   "Eicosane, 2,4-dimethyl-", "1-Decanol, 2-octyl-")) %>%
  filter_area(min_maximum = 1e5) %>% #not used for final filtering
  filter_ambient_ratio(vol.all, metadata, ratio = 4) %>%
  combine_filters() 

chemsf$filter_final <- with(chemsf, filter_RT=="OK" & filter_match=="OK" & 
                              filter_freq.floral=="OK" & filter_contaminant=="OK")

#now run t-tests after rarity filtering
#don't filter by ambient ratio first, see paper on problems: https://doi.org/10.1186/1471-2105-11-450
chemsf <- bind_rows(chemsf %>% filter(filter_final) %>% 
  filter_ambient_ttest(prune_sampletable(vol.all, chemsf, metadata), 
                       metadata, alpha = 0.05, adjust = "fdr"), #not used for final filtering
  chemsf %>% filter(!filter_final)) %>% 
  mutate(filter_ambient_ttest=fct_na_value_to_level(filter_ambient_ttest, "notchecked"),
         filter_final = filter_final & filter_ambient_ratio =="OK") %>% 
  arrange(match(name, colnames(vol.all)))

bouquet::plot_filters(chemsf, option="rarity", yrange=3.5)
bouquet::plot_filters(chemsf, option="ambient", yrange=3.5)
bouquet::plot_filters(chemsf, option="volcano", yrange=2.5, alpha_excluded = 0.5)
bouquet::plot_filters(chemsf, option="prop")
```

```{r quant}
#vols <- maxf[volsmeta$index,]
vols <- vol.all[volsmeta$FileName,]
vols.cut <- vols[,colSums(vols)>3e8]
rownames(vols.cut) <- volsmeta$index
vol <- vols[volsmeta$FileName, chemsf$name[chemsf$filter_final]]#don't cut ambients yet
rownames(vol) <- volsmeta$index

#load quantitative integrations
quant <- read_tsv("../ipomopsis-temp/data/volatiles/quant.tsv") %>%
  mutate(Name = recode(Name, !!!shortnames)) %>% 
  pivot_wider(names_from = "Name", values_from="Area") %>% as.data.frame()

quant.meta <- quant %>% select(batch, Dirname, Filename) %>% 
  mutate(isblank = str_detect(Filename, "lank"),
         Year = as.integer(str_match(Filename, "(201[7-9]|202[1-2])[_.]")[,2]),
         GCOrder = as.integer(str_match(Filename, "_([0-9][0-9]).qgd")[,2]),
         RunDate = str_remove(str_extract(Filename, paste0("_[0-9]{2,4}",Year)), "_[0]*"),
         RunDate = ymd(paste(str_sub(RunDate, -4,-1), str_sub(RunDate,1,1), str_sub(RunDate, 2, -5))))

rownames(quant) <- quant$Filename
quant[,c("batch","Dirname","Filename")] <- NULL
quant.all <- quant[rownames(vols),]

setdiff(colnames(quant.all), colnames(vol)) #TODO include any of these from temp paper?
setdiff(colnames(vol), colnames(quant.all)) #TODO set up quant integrations for these

quant.slopes <- set_names(rep(0,ncol(quant.all)), colnames(quant.all))
for(compvol in colnames(quant.all)) {
  nozeros <- bind_cols(select(vols,qual=compvol), select(quant.all,quant=compvol)) %>% 
    filter(!(qual==0 | quant == 0))
  quant.slopes[compvol] = coef(lm(quant ~ 0 + qual, data=nozeros))["qual"]
  # quant_wins <- sum(vols[,compvol] == 0 & quant.all[,compvol] > 0, na.rm=T)
  # qual_wins <-  sum(quant.all[,compvol] == 0 & vols[,compvol] > 0, na.rm=T)
  # plot(x=sqrt(vols[,compvol]), y=sqrt(quant.all[,compvol]),
  #      main=compvol, sub = paste0("quant:qual ",quant_wins,":",qual_wins,
  #                                 " slope ",round(quant.slopes[compvol],3)),
  #      xlab="qual",ylab="quant")
  # abline(a=0, b=sqrt(quant.slopes[compvol]))
}
write_tsv(enframe(quant.slopes) %>% left_join(enframe(shortnames, name="Name", value="name")),
          "data/volatiles/quant_slopes.tsv")

quant.scaled <- sweep(quant.all, 2, quant.slopes, FUN = '/')

quant.scaled.vol <- quant.scaled[volsmeta$FileName, 
                                 colnames(quant.scaled) %in% 
                                   intersect(colnames(quant.all), colnames(vol))] 

#swap in scaled quant integrations
#plot(unlist(vol[,colnames(quant.scaled.vol)]), unlist(quant.scaled.vol))
vol[,colnames(quant.scaled.vol)] <- quant.scaled.vol
```

```{r retention, fig.width=12}
longdata.RT <- longdata %>% filter(name %in% chemsf$name[chemsf$filter_final]) %>% 
  left_join(select(metadata, sample, year)) %>%
  group_by(name) %>%  mutate(n=n(), RT.sd=sd(RT)) %>% ungroup() %>% 
  mutate(name=fct_reorder(name, RT))

RT.year <- longdata.RT %>% group_by(name, year) %>% summarize(RT=median(RT)) %>% pivot_wider(names_from=year, values_from=RT) 
RT.year %>% filter(name %in% setdiff(colnames(vol), colnames(quant.all))) %>% arrange(`2018`)


RT.tol <- 0.17  # 10 seconds before or after in quantitative params
ggplot(longdata.RT, aes(x=RT, y=name,color=RT.sd)) + geom_point(position=position_jitter(width=0, height=0.5)) + 
  scale_color_viridis_c(option="magma") + theme_dark()+
  geom_text(aes(x=4.5, label=n), color="grey80") +
  stat_summary(fun="median", fun.max=~median(.x)+RT.tol, fun.min=~median(.x)-RT.tol, geom="pointrange", color="white", size=0.5, shape=15)

ggplot(longdata.RT,  aes(y=RT, x=factor(year), color=RT.sd)) + 
  facet_wrap(vars(name), scales="free_y") + geom_boxplot() + scale_color_viridis_c(option="magma") + theme_dark() +
  geom_pointrange(data=RT.year, aes(y=mean1821, ymin=mean1821-0.1, ymax=mean1821+0.1, x="2019"), color="blue")

#RT fixes
longdata.butanoate <- longdata %>% filter(str_detect(name , "\\)-hex-3-enyl\\] (2-methylpropa|buta)noate")) %>% 
    left_join(select(metadata, sample, year)) 
ggplot(longdata.butanoate, aes(x=RT, fill=factor(year))) + geom_histogram(binwidth=0.02) + 
  facet_wrap(vars(name), ncol=1) + scale_x_continuous(limits=c(NA,NA))
longdata.butanoate %>% mutate(EZ=ifelse(RT>11, "Z","E")) %>% group_by(EZ, year) %>% 
  summarize(RT=median(RT)) %>% kable(digits=2)
```

```{r diverse, eval=FALSE}
tointegrate <- c("hexan-1-ol","sabinene","3-carene","verbenone","[(Z)-hex-3-enyl] 2-methylbutanoate","2-butyloctan-1-ol","b-caryophyllene oxide")
best <- list()
i <- 1
vol.left <- vol[,tointegrate]^(1/4) # halfway between areas and presence-absence
while(ncol(vol.left)>0) {
  print(dim(vol.left))
  best[[i]] <- which.max(rowSums(vol.left))
  vol.left <- vol.left[ , !decostand(vol.left[best[[i]],], method="pa"), drop=FALSE]
  i <- i+1
}
best.vol <- as.data.frame(t(vol[unlist(best),tointegrate]))
colnames(best.vol) <- metadata[colnames(best.vol),"sample"]
best.vol
```

## Filtering

This section is outdated, see bouquet filters above.

OLD Filtering criteria (verdict="fvoc" in the table below):

* Present in 10% of samples
* Maximum peak area is at least 4000000
* Passes t-test against ambient controls with FDR correction OR mean exceeds 4x mean of controls
* Not a known contaminant (Oxybenzone, Ethanol, 2,2'-oxybis-, Butylated Hydroxytoluene, Caprolactam, [1,1'-Bicyclopentyl]-2-one, L-Lactic acid, Pentadecane, 8-hexyl-, Decane, 2,8,8-trimethyl-, 2-Methyltetradecan-2-ol)

Note that these potential floral compounds did not meet the tests against ambient controls: 2-phenylacetaldehyde, 3-carene, D-limonene, camphene, benzaldehyde, a-thujene, sabinene, abietadiene, 2-ethylhexan-1-ol 

Plus (verdict other than "fvoc"): 

* bis14: compounds included in Bischoff et al. (2014)
* keep: kept since they are known floral volatiles
* keep_leaf: kept since they are known leaf volatiles
* merge: a compound that is a synonym of one of the floral volatiles (b-caryophyllene, b-ocimene, verbenone)

```{r filtering}
# ipochems <- read_csv("data/volatiles/Ipo volatile compounds - chemsf_ipo.csv") %>% 
#   filter(verdict %in% c("bis14", "fvoc", "keep","keep_leaf","merge"),
#          name %in% colnames(vols)[colSums(vols)>0], name != "Anisole") %>% #filter empty compounds
#   mutate(standard = fct_recode(standard, "Methyl_salicylate"="Benzaldehyde"), # benzaldehyde regressions not reliable
#          class = fct_recode(standard, Aliphatic="Hexenol", Benzenoid="Methyl_salicylate", Benzenoid="Indole", 
#                             Sesquiterpene="Caryophyllene", Monoterpene="alpha_Pinene", Monoterpene="Linalool")) %>%  
#   left_join(read_csv("data/volatiles/regressions_181921_filtered_slopes.csv") %>% 
#               pivot_wider(id_cols=standard, names_from="year", names_prefix="area_per_ng", values_from=area_per_ng)) 
# class_pal <- set_names(c("#BC0060","#027B8C","#E56E00","#86D400"), levels(ipochems$class))
# 
# ipochems %>% select(verdict, name, Ret.Time, mean, max, match, freq.floral, freq.ambient, ambient_ratio, filter_ambient_ttest, filters_passed, standard) %>%
#   arrange(desc(freq.floral)) %>% mutate(across(starts_with("freq."), round, 3)) %>% 
#   DT::datatable(caption="Volatile compounds to use for analysis")

#ipogood <- ipochems$name
ipogood <- colnames(vol)
#vols.good <- vols[,ipogood] %>% 
vols.good <- vol %>% 
  #sweep(2, ipochems$area_per_ng, FUN = '/') %>%  # convert areas to nanograms 
  sweep(1, as.numeric(volsmeta$total_time, units="hours"),  FUN = '/') #divide by equilibration + pumping time
ipochemsf <- ipochems[match(colnames(vols.good), ipochems$shortname),]
for(yr in unique(volsmeta$sequence.year)) {
  thisyr <- volsmeta$sequence.year==yr
  #TODO these slopes are leading to most of the year effects. try using one year for now. 
  vols.good[thisyr,] <- sweep(vols.good[thisyr,], 2, pull(ipochemsf, paste0("area_per_ng","2019")), FUN = '/')# convert areas to nanograms 
}

# merge peaks identified as multiple compounds
# give volatiles shorter names too
# vols.good <- left_join(ipochems %>% select(name, shortname), 
#          vols.good %>% t %>% as.data.frame %>% rownames_to_column("name")) %>% 
#   select(-name) %>% group_by(shortname) %>% 
#   summarize(across(everything(), sum)) %>% 
#   column_to_rownames("shortname") %>% t %>% as.data.frame

ipochemsf <- ipochemsf %>% #group_by(name=shortname, verdict, standard, class) %>%
  #summarize(.groups = "drop") %>%
  left_join(tibble(shortname = colnames(vols.good), # the following values are new, computed on the filtered dataset
                            freq.floral = map_dbl(as.data.frame(vols.good>0), mean),
                            mean.floral = map_dbl(vols.good, mean),
                            max.floral =  map_dbl(vols.good, max))) %>% 
  select(-name) %>% rename(name=shortname) %>%
  mutate(class=droplevels(class))#no benzenoids
rownames(ipochemsf) <- ipochemsf$name
#ipogood == ipochemsf$name

k <- 2 #split into this many k-means cluster per year
volsmeta$cluster <- 0
for(yr in unique(year(volsmeta$sequence.start))) {
  thisyr <- volsmeta$sequence.year==yr
  chems.include <- setdiff(colnames(vols.cut), ipogood) #drop alpha-pinene so this is all nonfloral
  set.seed(1)
  volsmeta$cluster[thisyr] <- kmeans(decostand(vols.cut[thisyr,chems.include], method="log"), k, nstart=3)$cluster
}

volsmeta <- volsmeta %>% 
  mutate(year.cluster = paste0(year, if_else(sequence.year == 2018, "", paste0(".",cluster))),
         total = rowSums(vols.good),
         sampledate = coalesce(sampledate, sampledate.file))#get sampling date from filename when no metadata

# Cut contaminated samples
#nonzero <-rowSums(vols.good) > 0 #one sample ("55" on row 44) has no volatiles 
# set.seed(1) # ensures the contaminated samples are in cluster.good = 1
# volsmeta$cluster.good <-factor(kmeans((vols.good)^(1/3), centers=3, nstart=3)$cluster)
# set.seed(1) # ensures the contaminated samples are in cluster.cut = 1
# volsmeta$cluster.cut <-factor(kmeans((vols.cut)^(1/3), centers=3, nstart=3)$cluster)

# keep.samples <- nonzero#unname(nonzero & !(volsmeta$year.cluster %in% c("2019.1", "2020.1")))
# volsmeta <- volsmeta[keep.samples,]
# vols <- vols[keep.samples,]
# vols.cut <- vols.cut[keep.samples,]
# vols.good <- vols.good[keep.samples,]

minsamples <- nrow(vol) * 0.15 #15% of samples (was 70 samples)
vols.good.top <- vols.good[,colSums(decostand(vols.good,"pa"))> minsamples]#occur in more than minsamples
ipogood.top <- names(sort(colSums(vols.good.top), decreasing=T))
volsmeta.good.top <- bind_cols(volsmeta, vols.good.top) %>%  
  pivot_longer(-all_of(colnames(volsmeta))) %>% 
  mutate(name = fct_relevel(name, ipogood.top))

mnps.plantyr.traits <- mnps.plantyr %>% 
  select(year, plotid, plant, all_of(traits), eggs_per_flower) %>%
  mutate(prop_uninfested = 1 - prop_infested)

traits.good.top <- volsmeta.good.top %>% 
  left_join(mnps.plantyr.traits) %>% 
  mutate(name = fct_relevel(name, ipogood.top))

#vols.drydown <- vols.good[,c("1,3-octadiene","benzyl alcohol", "a-pinene","a-farnesene", "b-pinene", "b-ocimene")]
#volsmeta.drydown <- bind_cols(volsmeta, vols.drydown) %>%  pivot_longer(-all_of(colnames(volsmeta)))

vols.class <- vols.good %>% t() %>% as.data.frame() %>% rownames_to_column("name") %>% 
  left_join(select(ipochemsf, name=name, class)) %>% 
  group_by(class) %>% summarize(across(-name, sum)) %>% column_to_rownames("class") %>%  t() %>% as.data.frame()
volsmeta.class <- bind_cols(volsmeta, vols.class) %>%  
  pivot_longer(-all_of(colnames(volsmeta)))
traits.class    <- volsmeta.class %>% left_join(mnps.plantyr.traits)
volsmeta.traits <- volsmeta %>%       left_join(mnps.plantyr.traits) #wide format
```


## Compound relative amounts

```{r relamounts}
vols.good %>% decostand(method = "tot") %>% summarize(across(everything(), mean)) %>% 
  t() %>% as.data.frame() %>% rownames_to_column("name") %>% rename(mean_percent = V1) %>% 
  arrange(desc(mean_percent)) %>% kable(caption = "Mean relative amounts of each volatile")
```


## Sample size after filtering

```{r samplesizes_filtered, fig.width=10, fig.height=5}
volsmeta %>% count(year, name="samples") %>% 
  left_join(volsmeta %>% count(year, plantid) %>% count(year, name="plants")) %>% 
  kable(caption = "Samples and plants by year")
volsmeta %>% count(year, water, snow) %>% 
  pivot_wider(names_from=c(year), values_from = n) %>% 
  kable(caption = "Samples by year and treatment")
volsmeta %>% count(year, sampledate) %>% 
  kable(caption = "Samples by date")
volsmeta %>% count(year, sampledate) %>% count(year) %>% 
  kable(caption="Number of sampling dates")

volsmeta %>% count(year, plantid) %>% ggplot(aes(x=n, fill=year)) + facet_wrap(vars(year), ncol=1) +geom_histogram(binwidth=1) +
  xlab("samples per plant") + scale_fill_manual(values=year_pal, guide="none") + theme_minimal()
```

## Time of day

```{r sample_times}
ggplot(volsmeta, aes(x=bag, fill=year)) + geom_histogram(show.legend = F) + facet_wrap(vars(year), ncol=1) + 
  scale_fill_manual(values=year_pal) + theme_minimal() + 
  labs(x="Flower bagged", caption= paste(format(c(min(volsmeta$bag, na.rm=T), max(volsmeta$bag, na.rm=T)), "%H:%M"), collapse=" - "))
ggplot(volsmeta, aes(x=bag+total_time, fill=year)) + geom_histogram(show.legend = F) + facet_wrap(vars(year), ncol=1) + 
  scale_fill_manual(values=year_pal) + theme_minimal() + 
  labs(x="Pump stopped", caption= paste(format(c(min(volsmeta$bag+volsmeta$total_time, na.rm=T), max(volsmeta$bag+volsmeta$total_time, na.rm=T)), "%H:%M"), collapse=" - "))

volsmeta %>% count(beforenoon=hour(bag+total_time) <13)

ggplot(volsmeta, aes(x=end-bag)) + geom_histogram() + labs(x="Equilibration and pumping time (min)")
```

## Distributed among treatments

```{r random_sampling}
ggplot(volsmeta, aes(x=yday(sampledate), fill=paste(water, snow))) + geom_bar() + facet_wrap(vars(year), ncol=1) + 
  labs(fill="Precipitation, snowmelt", x="Day of year", y="Samples") + scale_fill_manual(values=water_snow_pal)+ theme_minimal() 

list(water=chisq.test(with(vt, table(sampledate, water))),
     snow=chisq.test(with(vt, table(sampledate, snow))),
     water_snow=chisq.test(with(vt, table(sampledate, paste(water, snow))))) %>% 
  map_dfr(tidy, .id="treatment") %>% kable(caption="Test of randomness of treatments on each sampling date")
```

## Flowers and buds

```{r buds}
volsmeta %>% count(flowers)
volsmeta %>% count(buds>0)
cat(paste("Buds included in", round(sum(volsmeta$buds>0, na.rm=T)/sum(volsmeta$buds>=0, na.rm=T)*100), "% of samples"))
```

## Heatmap

```{r heatmap, dev="png", dev.args=list(),fig.width=16, fig.height=12}
annot_pal <- list(
  sequence.date = volsmeta %>% mutate(yr = sequence.year) %>% group_by(yr) %>% summarize(n = length(unique(sequence.date))) %>% 
    pull(n) %>% map(scales::hue_pal()) %>% unlist() %>% set_names(unique(volsmeta$sequence.date)),
  cluster = set_names(brewer.pal(8, "Set1")[1:length(unique(volsmeta$cluster))], unique(volsmeta$cluster)))

pheatmap(as.matrix(t(vols.cut))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),main="Unfiltered top volatiles",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors =  annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))

pheatmap(as.matrix(t(vols.good))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),main="Filtered volatiles",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors = annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))

pinene_correlates <- cor(sqrt(vols[,colSums(vols)>1e7])) %>% as.data.frame() %>%
  filter(`a-pinene`>0.2) %>% rownames()
pheatmap(as.matrix(t(vols[,pinene_correlates]))^(1/4), 
         cluster_cols=F, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         scale="none", color=mako(512),
                  annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors = annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0), main="Volatiles that correlate with alpha-pinene")

pheatmap(as.matrix(t(vols.class))^(1/4), 
         cluster_cols=F, cluster_rows=F, show_colnames=F,
         scale="none", color=mako(512),main="Filtered volatiles classes",
         annotation_col=volsmeta %>% select(sequence.date, cluster) %>% mutate(across(everything(), factor)),
         annotation_colors =  annot_pal,
         gaps_col = which(diff(year(volsmeta$sequence.date))>0))
```

## Chromatograms by GCMS run date

### Univariate

```{r gcms_date, fig.height=8, fig.width=7, eval=FALSE}
plot_vols_rundate <- function(type, checkvols) {
volsmeta %>% mutate(yr=year(Desorb.Start.Time)) %>% 
  bind_cols(vols %>% select(checkvols)) %>% pivot_longer(cols=checkvols) %>% 
ggplot(aes(x=Desorb.Start.Time)) + facet_grid(name ~ yr, scales="free") + 
  geom_point(aes(y=value, color=name), alpha=0.3)  + 
  guides(color="none") + theme_bw() + labs(x="GCMS run time", y="Area", title=type)
}

plot_vols_rundate("Terpenes", c("a-pinene", "b-myrcene","(E)-b-ocimene","b-caryophyllene","linalool","longifolene"))

plot_vols_rundate("Benzenoids", c("benzophenone","methyl salicylate","anisole"))

plot_vols_rundate("Green leaf volatiles", c("(E)-hex-3-en-1-ol"))
```

### Multivariate

```{r gcms_sample_date, fig.height=6, fig.width=10, eval=FALSE}
for(categories in list(c("water","snow"),"plot","observer","pump_id","sequence.12", "sampledate", c("sampledate", "sequence.12"), c("sampledate.observer", "sequence.12"))) {
  cluster_counts <- volsmeta %>% 
    mutate(sampledate = coalesce(sampledate, sampledate.file),
           sampledate.observer = paste(sampledate, observer)) %>% 
    count(!!!syms(c("sequence.year", categories, "cluster")))
  p <- ggplot(cluster_counts, aes(x=factor(paste(!!!syms(categories), sep=" x ")), 
                                  y=n, fill=factor(paste0(sequence.year," C",cluster)))) + 
    geom_col() + facet_wrap(vars(sequence.year), scales="free_x") +
    theme(axis.text.x = element_text(angle=90)) + labs(fill="Cluster", x=paste(categories, collapse=" x ")) +
    scale_fill_brewer(palette="Paired")
  print(p)
  
  if(length(categories)>1) {
    (p2 <- cluster_counts %>% filter(sequence.year !="2018") %>% mutate(cluster=factor(cluster), across(where(~ !is.numeric(.)),factor)) %>% 
     ggplot(aes_string(x=categories[[1]], y = categories[[2]], size="n", color="cluster")) + geom_point(shape=1, stroke=1) +
      facet_wrap(vars(sequence.year), scales="free") + theme_minimal()+ scale_size_area(max_size=15)+
      theme(axis.text.x = element_text(angle=90)) + scale_color_manual(values=c("darkgreen","magenta")))
    print(p2)
  }
}

ggplot(volsmeta, aes(x=Desorb.Start.Time, y=cluster))  + geom_line() + geom_point(aes(color=factor(mday(sampledate))))+
  facet_wrap(vars(sequence.year), scales="free_x", ncol=1) + scale_color_manual(values=sample(rainbow(30)))+
  scale_x_datetime(date_breaks="1 day", guide=guide_axis(angle=90)) + scale_y_discrete()
```


```{r gcms_date_ord, eval=FALSE}
for(yr in unique(year(volsmeta$sequence.start))) {
  print(yr)
  yrfilter <- year(volsmeta$sequence.start) == yr
  vols.cap <- capscale(sqrt(vols.cut[yrfilter,]) ~ factor(date(sequence.start)), data=volsmeta[yrfilter,],
                       distance="bray")
  print(vols.cap)
  print(anova(vols.cap, by="term"))
  # vols.cap.points <-  fortify(vols.cap) %>% as_tibble()
  # vols.cap.samples <- vols.cap.points %>% filter(Score=="sites") %>% 
  #   left_join(volsmeta, by=c("Label"="index"))
  # vols.cap.chems <- vols.cap.points %>% filter(Score=="species")
  # p <- vols.cap.samples %>% 
  #   mutate(sequence.date = date(sequence.start)) %>% 
  #   filter(sequence.year == yr) %>% 
  #   ggplot(aes(x=CAP1, y=CAP2)) + theme_classic()+
  #   geom_point(aes(color=factor(sequence.date)), size=3) + scale_color_hue("Run date")
  # print(p)
}
```

## Ordinations

### NMDS

```{r nmds, fig.height=9, fig.width=6}
(vols.nmds <- metaMDS(sqrt(vols.cut), autotransform=FALSE, trace=F))
vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
  left_join(volsmeta, by=c("Label"="index"))
vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")

(scents.nmds <- metaMDS(sqrt(vols.good), autotransform=FALSE, trace=F)) #had noshare = FALSE
nmds.points <-  fortify(scents.nmds) %>% as_tibble()
nmds.samples <- nmds.points %>% filter(Score=="sites") %>% left_join(volsmeta, by=c("Label"="index"))
nmds.chems <- nmds.points %>% filter(Score=="species") %>% left_join(ipochemsf, by=c("Label"="name"))

# By year
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() +
  geom_point(aes(color=year)) + scale_color_manual(values=year_pal)+
  labs(title="Unfiltered NMDS", color="Year")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  #geom_text(data=nmds.chems, aes(label=Label)) +
  geom_point(aes(color=year))+ scale_color_manual(values=year_pal)+
  #geom_point(aes(color=water, shape=year)) + scale_color_manual("Precipitation", values=water_pal) +
  labs(title="Filtered NMDS", color="Year")
grid.arrange(vols.plot, vols.good.plot)

# By a-Pinene
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2, shape=year)) + theme_classic() +
  geom_point(aes(color=vols.good$`a-pinene`^(1/6))) + 
  labs(title="Unfiltered NMDS", color="a-Pinene", shape="Year") + scale_color_viridis_c()
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2, shape=year)) + theme_classic() + 
  geom_point(aes(color=vols.good$`a-pinene`^(1/6)))+ 
  labs(title="Filtered NMDS", color="a-Pinene", shape="Year")+ scale_color_viridis_c()
grid.arrange(vols.plot, vols.good.plot)
```

```{r nmds_chems, fig.height=12, fig.width=12}
ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=year))+ scale_color_manual(values=year_pal)+
    geom_text(data=nmds.chems, aes(label=Label, alpha=freq.floral^(1/3))) +
  labs(title="Filtered NMDS", color="Year")
```


```{r nmds_clusters, eval=FALSE}
#Cluster based on filtered data
volsmeta %>% count(cluster.good, year) %>% arrange(year) %>% 
  pivot_wider(names_from="year", values_from="n") %>% 
  kable(caption = "Contaminated samples cluster = 1 based on filtered data")
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.good, year)))+ 
  labs(title="Unfiltered NMDS by year", color="Filtered cluster")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.good, year)))+ 
  labs(title="Filtered NMDS by year", color="Filtered cluster")
grid.arrange(vols.plot, vols.good.plot)

#cluster based on unfiltered data
set.seed(1)
volsmeta$cluster.cut <-factor(kmeans((vols.cut)^(1/3), centers=3, nstart=3)$cluster)
volsmeta %>% count(cluster.cut, year) %>% arrange(year) %>% 
  pivot_wider(names_from="year", values_from="n") %>% 
  kable(caption = "Contaminated samples cluster = 1 based on unfiltered data")
vols.plot <- ggplot(vols.nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.cut, year)))+ 
  labs(title="Unfiltered NMDS by year", color="Unfiltered cluster")
vols.good.plot <- ggplot(nmds.samples, aes(x=NMDS1, y=NMDS2)) + theme_classic() + 
  geom_point(aes(color=paste(cluster.cut, year)))+ 
  labs(title="Filtered NMDS by year", color="Unfiltered cluster")
grid.arrange(vols.plot, vols.good.plot)

# Ordinations for each year
for(yr in unique(year(volsmeta$sequence.start))) {
  thisyr <- year(volsmeta$sequence.start)==yr
  (vols.nmds <- metaMDS(sqrt(vols.cut[thisyr,]), autotransform=FALSE, trace=F))
  vols.nmds.points <-  fortify(vols.nmds) %>% as_tibble()
  vols.nmds.samples <- vols.nmds.points %>% filter(Score=="sites") %>% 
    left_join(volsmeta, by=c("Label"="index")) %>% filter(sequence.year == yr)
  vols.nmds.chems <- vols.nmds.points %>% filter(Score=="species")
  
  (scents.nmds <- metaMDS(sqrt(vols.good[thisyr,]), autotransform=FALSE, trace=F))
  nmds.points <-  fortify(scents.nmds) %>% as_tibble()
  nmds.samples <- nmds.points %>% filter(Score=="sites") %>% 
    left_join(volsmeta, by=c("Label"="index")) %>% filter(sequence.year == yr)
  nmds.chems <- nmds.points %>% filter(Score=="species")
  
  vols.plot <- vols.nmds.samples %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Unfiltered NMDS run",yr))+
    geom_text(aes(color=factor(sequence.date), label=cluster), size=4) + scale_color_hue("Run date")
  vols.good.plot <- nmds.samples %>% 
    ggplot(aes(x=NMDS1, y=NMDS2)) + theme_classic()+ labs(title = paste("Filtered NMDS run",yr))+
        geom_text(aes(color=factor(sequence.date), label=cluster), size=4)  + scale_color_hue("Run date")
  grid.arrange(vols.plot, vols.good.plot)
}
```

## Correlations

```{r corr, fig.height=8, fig.width=10}
corr.annot <- ipochemsf %>% select(name, Class=class, Frequency=freq.floral) %>% #`Retention time`=RT,
  column_to_rownames("name") %>%  
  filter(Frequency>0.1)
corr.vols <- cor(vols.good[, rownames(corr.annot)], method="pearson")
pheatmap(corr.vols, 
         scale="none", clustering_method = "mcquitty",
         color=colorRampPalette(c("blue","white","red"))(200), breaks=seq(-1,1,by=0.01),
         annotation_col=corr.annot, annotation_row=corr.annot, 
         annotation_colors = list(Class=class_pal))

(corr.vols*lower.tri(corr.vols)) %>% as.data.frame() %>% rownames_to_column("name1") %>% 
  pivot_longer(cols=-name1, values_to="cor") %>% arrange(desc(cor)) %>% filter(cor!=1, cor!=0) %>% 
  mutate(cor.cut = cut_width(cor, width=0.25, boundary=0)) %>% count(cor.cut)

```


# Plasticity analyses

## Plasticity to treatments

### CAP

```{r cap}
(scents.cap <- capscale(sqrt(vols.good) ~ year * water * snow, data=volsmeta, distance="bray"))
anova(scents.cap, by="term") %>% kable()

scents.cap <- capscale(sqrt(vols.good) ~ year + water, data=volsmeta, distance="bray")
plot(scents.cap, type="n")
points(scents.cap, display = "sites", col=water_pal[volsmeta$water], pch=as.integer(volsmeta$year)+14)
text(scents.cap, display="cn", labels=c(levels(volsmeta$year), names(water_pal)), col=c(rep("black",3),water_pal), font=2, cex=1.5)

plot(scents.cap, type="n")
text(scents.cap, display="species", col="black", cex=0.5)

scents.cap <- capscale(sqrt(vols.good) ~ year + snow, data=volsmeta, distance="bray")
plot(scents.cap, type="n")
points(scents.cap, display = "sites", col=snow_pal[volsmeta$snow], pch=as.integer(volsmeta$year)+14)
text(scents.cap, display="cn", labels=c(levels(volsmeta$year), names(snow_pal)), col=c(rep("black",3),snow_pal), font=2, cex=1.5)

plot(scents.cap, type="n")
text(scents.cap, display="species", col="black", cex=0.5)
```

### Joint species distribution model

```{r jsdm_splitplot, eval=F}
library(Hmsc)

# model volatile abundance (emission rates) plasticity to year, water, and snow
# include random effect of plot
model.abu <- Hmsc(Y=vols.good,
                XData = volsmeta %>% select(year, water, snow) %>% mutate_all(as.factor), 
                XFormula= ~ year + water + snow, 
                studyDesign = volsmeta %>% select(plot) %>% mutate_all(as.factor),
                ranLevels = list(plot = HmscRandomLevel(units = levels(treatments$plot))),
                distr="lognormal poisson")

# Fit model with MCMC
thin <- 1
(model.abu <- sampleMcmc(model.abu, thin = thin, samples = 100, transient = thin*50, nChains = 3, nParallel = 3,
                        initPar="fixed effects")) #chains start from ML solution to fixed-effects model
```

```{r jsdm_splitplot_results, eval=F}
# Model convergence
mpost <- convertToCodaObject(model.abu)[c("Beta","Gamma","V","Sigma")]
ess <- map(mpost, effectiveSize)
par(mfrow=c(2,2))
walk(names(ess), ~hist(ess[[.x]], xlab = expression("Effective sample size" ~ beta ~ ""), main=.x))
psrf <- map(mpost, gelman.diag, multivariate=FALSE)
walk(names(psrf), ~hist(psrf[[.x]]$psrf, xlab = expression("Potential scale reduction factor"), main=.x))

# Parameters
postBeta = getPostEstimate(model.abu, parName="Beta")
plotBeta(model.abu, post=postBeta, param="Support", supportLevel = 0.95, 
         spNamesNumbers = c(T,F), covNamesNumbers = c(T,F))

Gradient = constructGradient(model.abu,focalVariable = "water")
predY = predict(model.abu,Gradient = Gradient, expected = TRUE)
prob = c(0.25,0.5,0.75)
plotGradient(model.abu, Gradient, pred=predY, measure="S", showData = TRUE, q = prob, ylabel="Total emissions") 

Gradient = constructGradient(model.abu,focalVariable = "snow")
predY = predict(model.abu,Gradient = Gradient, expected = TRUE)
plotGradient(model.abu, Gradient, pred=predY, measure="S", showData = TRUE, q = prob, ylabel="Total emissions")

# Model fit (explanatory power)
predY = computePredictedValues(model.abu)
MF = evaluateModelFit(hM=model.abu, predY=predY)
tibble(name = ipogood, SR2 = MF$SR2) %>% mutate(name=fct_reorder(name, SR2)) %>% 
  filter(SR2>0.05) %>% 
  ggplot(aes(y=name, x=SR2))+ geom_point() + xlim(c(0,NA))+
  labs(x="squared Spearman correlation between obs and pred values (SR2)", 
       title="Volatiles with highest R2 in plasticity model") + theme_minimal()

# cross validation (predictive power)
partition = createPartition(model.abu, nfolds = 2)
#predY = computePredictedValues(model.abu, partition = partition, nParallel = 3)
#MFCV = evaluateModelFit(hM=model.abu, predY=predY)
#plot(MF$SR2, MFCV$SR2)
#abline(0,1)
```

### Boxplots

```{r boxplots, fig.width=16, fig.height=10}
treatment_boxplots <- function(volsmeta.vol, treatment) {
  ggplot(volsmeta.vol %>% 
           mutate(water_snow = factor(paste(water, snow), levels=names(water_snow_pal))),
         aes_string(x = "year", fill = treatment, y = "value")) + 
    facet_wrap(vars(name), scales="free_y") + geom_boxplot(outlier.size=0.2) + scale_y_sqrt() +
    scale_fill_manual(values=list(water=water_pal, snow=snow_pal,
                                  water_snow=water_snow_pal)[[treatment]]) + theme_bw() + 
    labs(x="Year", y ="Emissions (ng/hr/flower)", 
         fill=c(water="Precipitation",snow="Snowmelt", 
                water_snow="Precipitation, snowmelt")[[treatment]]) 
}
treatment_boxplots(volsmeta.good.top, "water_snow")
```

### Emissions by class

```{r boxplots_class}
treatment_boxplots(volsmeta.class, "water_snow")

options(contrasts = c("contr.sum", "contr.poly"))
treatment_effects <- function(class) {
  lmer(sqrt(value) ~ year * water * snow + (1|plot) + (1|plotid), 
       data = filter(volsmeta.class, name==class))
}

mod.split <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, treatment_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.split %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to treatments")
```

## Plasticity to snowmelt date and precipitation

### CAP

```{r cap_abs}
(scents.cap <- capscale(sqrt(vols.good) ~ sun_date * precip_est_mm, data=volsmeta, distance="bray"))
anova(scents.cap, by="term") %>% kable()

library(ggvegan)
cap.df <- fortify(update(scents.cap, ~ sun_date + precip_est_mm)) #drop non-signif interaction for plotting

ggplot(mapping=aes(x=CAP1, y=CAP2, label=Label)) + 
  geom_point(data=cap.df %>% filter(Score=="sites") %>% bind_cols(volsmeta), aes(color=year))+
  geom_segment(data=cap.df %>% filter(Score=="biplot"), aes(x=0,y=0,xend=CAP1*3.7, yend=CAP2*3.7), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(data=cap.df %>% filter(Score=="biplot") %>% 
              mutate(Label=c("Snowmelt", "Precipitation")),
            aes(x=CAP1*4, y=CAP2*4, angle=atan(CAP2/CAP1)*180/pi), hjust=0, fontface="bold") +
  geom_text(data=cap.df %>% left_join(ipochemsf, by=c("Label"="name")) %>% filter(Score=="species", mean.floral>1.5), aes(x=CAP1*3, y=CAP2*3), size=3)+
  scale_color_manual("Year", values=year_pal) + coord_fixed(xlim=c(-3,NA)) + 
  theme_bw() + theme(panel.grid = element_blank())
```

### Emissions by class

```{r mod_abs}
options(contrasts = c("contr.sum", "contr.poly"))
abs_effects <- function(class) {
  lmer(sqrt(value) ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), 
       data = filter(volsmeta.class, name==class))
}

mod.abs <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, abs_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.abs %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to snowmelt date and estimated precipitation")
```

### Scatterplots

#### Snowmelt date

```{r vols_abs}
precip.breaks <- seq(25,175, by=25)
grid.points <- list(sun_date=range(treatments$sun_date), precip_est_mm=precip.breaks)
options(contrasts = c("contr.sum", "contr.poly"))
#TODO this code partially repeats chunk above
mod.class.abs <- volsmeta.class %>% group_by(name) %>% nest() %>% 
  mutate(model = map(data, ~lmer(value ~ sun_date * precip_est_mm + (1|plot) + (1|plotid), data=.)),
         test = map(model, ~tidy(anova(.x))),
         emm.grid = map(model, ~ summary(emmeans(.x, ~ precip_est_mm * sun_date, at=grid.points))))

ggplot(volsmeta.class, aes(x = sun_date, y = value, color=precip_est_mm)) + facet_wrap(vars(name), scales="free_y") + 
  geom_point(aes(shape=year), size=2) + 
    geom_line(data = select(mod.class.abs, name, emm.grid) %>% unnest(emm.grid),
            aes(y=emmean, group=precip_est_mm))+
  theme_minimal()+ scale_y_sqrt()  + 
  labs(x="Snowmelt date", y ="Emissions (ng/hr/flower)", shape="Year", color="Precipitation") +
  scale_color_gradientn("Summer precipitation (mm)", colors=rev(water_pal), 
                        values=rev(c(1, (treatments %>% group_by(water) %>% summarize(across(precip_est_mm, mean)) %>%
                                           pull(precip_est_mm) %>% scales::rescale(from=range(precip.breaks)))[2],0)),
                        breaks=precip.breaks)
```

## Plasticity to plant-level soil moisture

### CAP

```{r cap_sm}
hasVWC <- !is.na(volsmeta$VWC)
(scents.cap <- capscale(sqrt(vols.good[hasVWC,]) ~ year * VWC, data=volsmeta[hasVWC,], distance="bray"))
anova(scents.cap, by="term") %>% kable()

cap.df <- fortify(scents.cap)
ggplot(mapping=aes(x=CAP1, y=CAP2, label=Label)) + 
  geom_point(data=cap.df %>% filter(Score=="sites") %>% bind_cols(volsmeta[hasVWC,]), aes(color=year))+
  geom_segment(data=cap.df %>% filter(Score=="biplot"), aes(x=0,y=0,xend=CAP1*3.7, yend=CAP2*3.7), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(data=cap.df %>% filter(Score=="biplot") ,
            aes(x=CAP1*4, y=CAP2*4, angle=atan(CAP2/CAP1)*180/pi), hjust=0, fontface="bold") +
  geom_text(data=cap.df %>% left_join(ipochemsf, by=c("Label"="name")) %>% filter(Score=="species", mean.floral>1.5), aes(x=CAP1*3, y=CAP2*3), size=3)+
  scale_color_manual("Year", values=year_pal) + coord_fixed(xlim=c(-3,NA)) + 
  theme_bw() + theme(panel.grid = element_blank())
```

### Emissions by class

```{r mod_sm}
options(contrasts = c("contr.sum", "contr.poly"))
sm_effects <- function(class) {
  lmer(sqrt(value) ~ year * VWC + (1|plotid), 
       data = filter(volsmeta.class, name==class))
}

mod.sm <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, sm_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.sm %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to soil moisture")
```

### Scatterplots

```{r sm_vols_top, fig.width=14, fig.height=10}
ggplot(volsmeta.good.top, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") +
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt()  + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year") +
  scale_color_manual(values=year_pal)
```

```{r sm_vols}

ggplot(volsmeta.class, aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year")+
  scale_color_manual(values=year_pal)

ggplot(volsmeta, aes(x = VWC, color = year, y = total)) + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ xlim(c(0,15))+ #TODO cuts off a few high VWC values
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Total filtered emissions (ng/hr/flower)", color="Year") +
  scale_color_manual(values=year_pal)
```

## Plasticity to subplot-level soil moisture

### CAP

```{r cap_sm_subplot}
(scents.cap <- capscale(sqrt(vols.good) ~ year * VWC, data=left_join(select(volsmeta,-VWC), sm.subplotyear),
                        distance="bray"))
anova(scents.cap, by="term") %>% kable()
cap.df <- fortify(scents.cap)
ggplot(mapping=aes(x=CAP1, y=CAP2, label=Label)) + 
  geom_point(data=cap.df %>% filter(Score=="sites") %>% bind_cols(volsmeta), aes(color=year))+
  geom_segment(data=cap.df %>% filter(Score=="biplot"), aes(x=0,y=0,xend=CAP1*3.7, yend=CAP2*3.7), 
               arrow=arrow(length=unit(0.5, "lines")))+
  geom_text(data=cap.df %>% filter(Score=="biplot") ,
            aes(x=CAP1*4, y=CAP2*4, angle=atan(CAP2/CAP1)*180/pi), hjust=0, fontface="bold") +
  geom_text(data=cap.df %>% left_join(ipochemsf, by=c("Label"="name")) %>% filter(Score=="species", mean.floral>1.5), aes(x=CAP1*3, y=CAP2*3), size=3)+
  scale_color_manual("Year", values=year_pal) + coord_fixed(xlim=c(-3,NA)) + 
  theme_bw() + theme(panel.grid = element_blank())
```

### Emissions by class

```{r mod_sm_subplot}
options(contrasts = c("contr.sum", "contr.poly"))
sm_effects <- function(class) {
  lmer(sqrt(value) ~ year * VWC + (1|plotid), 
       data = filter(left_join(select(volsmeta.class,-VWC), sm.subplotyear), name==class))
}

mod.sm <- tibble(class=levels(ipochemsf$class)) %>% 
  mutate(model = map(class, sm_effects),
         test = map(model, ~ tidy(anova(.x, ddf="Ken")))) %>% 
  select(-model) %>% unnest(test)

mod.sm %>% select(class, term, p.value) %>% 
  pivot_wider(names_from=term, values_from=p.value) %>%
  kable(caption="Plasticity of emissions to subplot-level soil moisture")
```

### Scatterplots

```{r sm_subplot_vols_top, fig.width=14, fig.height=10}
ggplot(left_join(select(volsmeta.good.top,-VWC), sm.subplotyear), aes(x = VWC, color = year, y = value)) +  facet_wrap(vars(name), scales="free_y") +
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ 
  scale_y_sqrt()  + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year") +
  scale_color_manual(values=year_pal)
```

```{r sm_subplot_vols}
ggplot(left_join(select(volsmeta.class,-VWC), sm.subplotyear), aes(x = VWC, color = year, y = value)) + 
  facet_wrap(vars(name), scales="free_y") + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ 
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Emissions (ng/hr/flower)", color="Year")+
  scale_color_manual(values=year_pal)

ggplot(left_join(select(volsmeta,-VWC), sm.subplotyear), aes(x = VWC, color = year, y = total)) + 
  geom_point() + geom_smooth(se=F, span=1)+ theme_minimal()+ 
  scale_y_sqrt() + labs(x="Soil moisture (%VWC)", y ="Total filtered emissions (ng/hr/flower)", color="Year") +
  scale_color_manual(values=year_pal)
```


# Selection analyses

## Fitness measures

```{r fitness_corr}
fitnesstraits <- c("seeds_est", "seeds_per_flower", "prop_uninfested", "eggs_per_flower")
fitnessnames <- c(traitnames, prop_uninfested = "Prop. uneaten fruits",
                  eggs_per_flower="Fly eggs per flower")[fitnesstraits]

library(GGally)
mnps.plantyr.traits %>% select(all_of(fitnesstraits)) %>% 
  ggpairs(mapping=aes(color=mnps.plantyr$year), lower=list(continuous="smooth_loess")) + 
  scale_color_manual(values=year_pal)+ scale_fill_manual(values=year_pal)
```

## Plots of selection in treatments and years

```{r selection, fig.width=16, fig.height=10}
walk(fitnesstraits, ~ 
       print(ggplot(traits.good.top %>% filter(ifelse(.x == "prop_uninfested", fruits_nonaborted > 4, TRUE)), 
                    aes_string(y=.x, x="value", color="year")) + 
               geom_point() + geom_smooth(method="lm", se=F) + 
               facet_wrap(vars(name), scales="free_x") + 
               labs(x="Emissions (ng/hr/flower)", y=fitnessnames[.x], color="Year")+
               theme_bw() + scale_x_sqrt() + scale_color_manual(values=year_pal)))


walk(fitnesstraits, ~ 
       print(ggplot(traits.good.top %>% filter(ifelse(.x == "prop_uninfested", fruits_nonaborted > 4, TRUE)), 
                    aes(x=value, color=paste(water,snow), linetype=year, shape=year)) + 
               geom_point(aes_string(y=.x)) + geom_smooth(aes_string(y=.x), method="lm",se=F) + 
               facet_wrap(vars(name), scales="free_x") + 
               labs(x="Emissions (ng/hr/flower)", y=fitnessnames[.x], 
                    color="Precipitation and snowmelt", linetype="Year", shape="Year")+
               theme_bw() + scale_x_sqrt() + scale_color_manual(values=water_snow_pal)))
```

### Emissions by class

```{r selection_class}
walk(fitnesstraits, ~ 
       print(ggplot(traits.class %>% filter(ifelse(.x == "prop_uninfested", fruits_nonaborted > 4, TRUE)), 
                    aes(x=value, color=paste(water,snow), linetype=year, shape=year)) + 
               geom_point(aes_string(y=.x)) + geom_smooth(aes_string(y=.x), method="lm",se=F) + 
               facet_wrap(vars(name), scales="free_x") + 
               labs(x="Emissions (ng/hr/flower)", y=fitnessnames[.x], 
                    color="Precipitation and snowmelt", linetype="Year", shape="Year")+
               theme_bw() + scale_x_sqrt() + scale_color_manual(values=water_snow_pal)))
```

```{r selection_class_gradient, fig.width=11, fig.height=8}
trait.class.wide <- bind_cols(volsmeta, vols.class) %>% left_join(mnps.plantyr.traits) %>% 
  select(plantid, year, water, snow, all_of(fitnesstraits), all_of(names(class_pal))) %>% 
  mutate(across(all_of(names(class_pal)), sqrt))

selection_effects <- function(fitnesstrait) {
  options(contrasts = c("contr.sum", "contr.poly"))
  lm(relfitness ~ year * water * snow + 
     Monoterpene * snow + Monoterpene * water + Monoterpene * year +
     Sesquiterpene * snow + Sesquiterpene * water + Sesquiterpene * year + 
     #Benzenoid * snow + Benzenoid * water + Benzenoid * year +
     Aliphatic * snow + Aliphatic * water + Aliphatic * year,
       data = trait.class.wide %>% rename(fitness=fitnesstrait) %>% 
       mutate(relfitness = fitness/mean(fitness, na.rm=T)))
}

fitnesstraits.sd <- map_dbl(fitnesstraits, ~sd(mnps.plantyr.traits[[.x]], na.rm=T)) %>% set_names(fitnesstraits)
class.sd <- map_dbl(names(class_pal), ~sd(trait.class.wide[[.x]], na.rm=T)) %>% set_names(names(class_pal)) 
#note this is after sqrt transformation

get_emt_class <- function(mod) {
  map_dfr(set_names(names(class_pal)), ~emtrends(mod, c("year","water","snow"), var=.x) %>% 
            summary %>% as_tibble %>% rename("trend" = ends_with("trend")), .id="class")
}

mod.selection <- tibble(fitnesstrait=fitnesstraits) %>% 
  mutate(model = map(fitnesstrait, selection_effects),
         emt = map(model, get_emt_class),
         test = map(model, ~ tidy(car::Anova(.x, type=3))))

stars <- function(p.values) {
  symnum(p.values, corr = FALSE, na = FALSE, 
         cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
         symbols = c("***", "**", "*", ".", " "))
}

mod.selection %>% unnest(test) %>% select(fitnesstrait, term, p.value) %>% 
  mutate(p.value= paste(format(round(p.value,4),nsmall=4), stars(p.value))) %>% 
  pivot_wider(names_from=fitnesstrait, values_from=p.value) %>%
  filter(term!="Residuals") %>% 
  kable(caption="Selection of emissions in each treatment")

mod.selection %>% select(fitnesstrait, emt) %>% unnest(emt) %>% 
  mutate(classsd = class.sd[class],
         trend.std = trend * classsd,
         uSE = (trend + SE) * classsd,
         lSE = (trend - SE) * classsd,
         water_snow = factor(paste(water, snow), levels=names(water_snow_pal))) %>% 
  ggplot(aes(x=class, color=water_snow, y=trend.std, ymin=lSE, ymax=uSE)) +
  facet_grid(cols=vars(fitnesstrait), rows=vars(year), labeller=as_labeller(c(set_names(2018:2020), fitnessnames))) +
  geom_hline(yintercept=0)+
  geom_pointrange(position=position_dodge(width=0.5), size=0.2) +
  scale_color_manual(values=water_snow_pal) +
  labs(x="Compound class", y = "Standardized selection gradient", color="Precipitation, snow") +
  theme_bw() + theme(axis.text.x = element_text(angle=90, hjust=1))
```


## Joint species distribution model

```{r jsdm_selection, eval=F}
library(Hmsc)

# model Y "species" (seeds, seeds per flower, prop fly fruits, fly eggs) 
# vs. XRRRData "environment" (volatiles) with reduced-rank regression

fitness.matrix <- volsmeta.traits %>% select(all_of(fitnesstraits))
fitness.complete <- !is.na(rowSums(fitness.matrix)) & !is.infinite(rowSums(fitness.matrix))
  
(model.sel <- Hmsc(Y = fitness.matrix %>% filter(fitness.complete) %>% as.matrix(),
                   XData = volsmeta %>% filter(fitness.complete) %>% select(year),
                   XFormula = ~ year,
                   XRRRData = vols.good %>% filter(fitness.complete), 
                   XRRRScale = TRUE, ncRRR = 4, 
                   studyDesign = volsmeta %>% filter(fitness.complete) %>% select(plot) %>%
                     mutate_all(as.factor),
                   ranLevels = list(plot = HmscRandomLevel(units = levels(treatments$plot))),
                   distr=c("poisson", "poisson", "probit","poisson")))

# Fit model with MCMC
thin <- 1
(model.sel <- sampleMcmc(model.sel, thin = thin, samples = 100, transient = thin*50, 
                         nChains = 3, nParallel = 3)) 
```

```{r jsdm_selection_results, eval=F}
# Model convergence
mpost <- convertToCodaObject(model.sel)[c("Beta","Gamma","V","Sigma")]
ess <- map(mpost, effectiveSize)
par(mfrow=c(2,2))
walk(names(ess), ~hist(ess[[.x]], xlab = expression("Effective sample size" ~ beta ~ ""), main=.x))

# Parameters
postBeta = getPostEstimate(model.sel, parName="Beta")
plotBeta(model.sel, post=postBeta, param="Support", supportLevel = 0.95, 
         spNamesNumbers = c(T,F), covNamesNumbers = c(T,F))

# Model fit (explanatory power)
predY = computePredictedValues(model.sel)
MF = evaluateModelFit(hM=model.sel, predY=predY)
tibble(name = fitnesstraits, SR2 = MF$SR2) %>% 
  ggplot(aes(y=name, x=SR2))+ geom_point() +
  labs(x="squared Spearman correlation between obs and pred values (SR2)", 
       title="Model fit (pseudo R2)") + theme_minimal()
```

## Selection gradients

Following method of [Chong et al. 2018](https://doi.org/10.1002/evl3.63) for reconstructing selection estimates from estimates of selection on PCs: Beta = E * A.

```{r gradients, fig.width=14, fig.height=10}
# previously split the data before running PCA - now run PCA on the whole dataset

# Run PCA on traits
vols.PCA <- vols.good.top %>% mutate(across(everything(), ~ (.x-mean(.x, na.rm=T))/sd(.x, na.rm=T))) %>% 
  prcomp() 
summary(vols.PCA) # PCs 1 - 5 explain 56% of the variation, 20 PCs total (20 top volatiles)
vols.PCA.E <- vols.PCA$rotation[,1:5] #min(4, ncol(vols.PCA$rotation))

calculate_beta <- function(E, slopes) {
  data.frame(B = E %*% slopes$A,
             SE = sqrt(E ^ 2 %*% slopes$SE ^ 2)) %>% 
    mutate(not_zero = abs(B) - SE > 0) %>% 
    rownames_to_column("trait")
}

selection_pca <- function(year, water=names(water_pal), snow=names(snow_pal), fitnesstrait = "seeds_est") {
  vols.std.PCA <- volsmeta %>% bind_cols(as.data.frame(vols.PCA$x)) %>% 
    filter(year == .env$year, water %in% .env$water, snow %in% .env$snow) %>% 
    left_join(mnps.plantyr.traits %>% select(year, plotid, plant, all_of(fitnesstrait)), by=c("year", "plotid", "plant")) %>% 
    select(all_of(fitnesstrait), any_of(names(treatments)), starts_with("PC")) %>% 
    rename("fitness" = fitnesstrait) %>% 
    group_by(year) %>% mutate(relfitness = fitness/mean(fitness, na.rm=T), .keep="unused") %>% ungroup %>% 
    drop_na(relfitness) %>% 
    select(where(~sum(is.na(.))==0))
  
  # Regress relative fitness on PCs
  vols.PCA.regression <- lm(relfitness ~ PC1 + PC2 + PC3 + PC4 + PC5, data = vols.std.PCA)
  vols.PCA.slopes <- tidy(vols.PCA.regression) %>% filter(term != "(Intercept)") %>% rename(A=estimate, SE=std.error)
  
  # Convert selection gradients on PCs to gradients on traits
  calculate_beta(vols.PCA.E, vols.PCA.slopes)
}

plot_beta <- function(betas, colorby="year") {
  bs <- betas %>% left_join(ipochemsf, by=c("trait"="name")) %>% 
    mutate(trait = fct_reorder(trait, mean.floral))#order graph by mean emissions
  if(colorby=="water_snow")  bs <- bs %>%  mutate(water_snow = paste(water, snow))
  ggplot(bs, aes(y=trait, x = B, xmin=B-SE, xmax=B+SE)) + 
    geom_vline(xintercept=0)+ 
    geom_pointrange(mapping=aes_string(color=colorby), position=position_dodge(width=0.5)) + 
    labs(x = "Selection gradient (beta) estimated from gradients on PCs", y = "Trait",
         color=list(year="Year", water="Precipitation", snow="Snowmelt", 
                    water_snow = "Precipitation and snowmelt")[[colorby]]) +
    scale_color_manual(values=list(year=year_pal, water=water_pal, snow=snow_pal,
                                   water_snow=water_snow_pal)[[colorby]]) +
    theme_bw()
}

vols.beta.yr <- expand_grid(year=unique(volsmeta$year), fitnesstrait=fitnesstraits)  %>% 
 mutate(betas = pmap(., selection_pca)) %>% unnest(betas)
plot_beta(vols.beta.yr)+ facet_wrap(vars(fitnesstrait), nrow=1, labeller=as_labeller(fitnessnames))

vols.beta.water <- expand_grid(year=unique(volsmeta$year), water=names(water_pal), fitnesstrait=fitnesstraits)  %>% 
 mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

vols.beta.snow <- expand_grid(year=unique(volsmeta$year), snow=names(snow_pal), fitnesstrait=fitnesstraits)  %>% 
 mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

vols.beta.water.snow <- expand_grid(year=unique(volsmeta$year), water=names(water_pal), snow=names(snow_pal),  fitnesstrait=fitnesstraits)  %>% 
  mutate(betas = pmap(., selection_pca)) %>% unnest(betas)

walk(fitnesstraits, ~print(plot_beta(filter(vols.beta.water.snow, fitnesstrait==.x), colorby="water_snow") + 
                             facet_wrap(vars(year), nrow=1) + ggtitle(fitnessnames[.x])))
```


## Adaptive plasticity

### Selection gradient vs. plasticity

```{r adaptive, fig.height=7, fig.width=9}
#TODO change 1|plot to 1|snow:plot throughout?
#TODO not knitting signif  size right, but output here OK (not anymore - nothing significant)
# TODO include interactions btn snow and water?
# TODO this is wrong, for labelling the graph we want models within each year, but those don't all run
mod.top.split <- volsmeta.good.top %>% group_by(name, year) %>% nest() %>% 
  mutate(model = map(data, ~lm(value ~ snow + water, data=.)), # + (1|plotid) + (1|plot)
         emm = map(model, ~ summary(emmeans(.x, ~ snow + water))%>% filter(water=="Control", snow=="Normal")), 
         test =  map(model, ~tidy(multcomp:::summary.glht(multcomp::glht(.x, 
                           linfct = multcomp::mcp(water = c("Addition - Control = 0",
                                                            "Reduction - Control = 0"),
                                                  snow = c("Early - Normal = 0"))), multcomp::adjusted(type="fdr"))))) %>% 
  select(name, emm, test) %>% unnest(c(emm, test)) %>% 
  mutate(estimate=estimate/emmean, std.error=std.error/emmean) %>% #divide estimated difference in means by emm in Normal Control subplots
  select(year, name, contrast, adj.p.value) %>%
  mutate(signif = adj.p.value < 0.05, .keep="unused") %>% 
  pivot_wider(names_from=contrast, values_from=signif)

#TODO fix this ggplot?
#ggplot(mod.top.split, aes(x=estimate, xmin=estimate-std.error, xmax=estimate+std.error, y=name, color=year, alpha=adj.p.value < 0.05)) +
#  geom_pointrange() +  facet_wrap(vars(contrast))
  
plasticity.water <- volsmeta.good.top %>% group_by(year, water, name) %>% summarize(value=mean(value)) %>% 
  pivot_wider(names_from=water, values_from=value) %>% left_join(ipochemsf) %>% 
  mutate(A_C = (Addition-Control)/Control, 
         R_C = (Reduction-Control)/Control, 
         A_R = (Addition-Reduction)/Reduction,
         R_A = (Reduction-Addition)/Addition, .keep="unused") %>% 
  left_join(mod.top.split)

plasticity.snow <- volsmeta.good.top %>% group_by(year, snow, name) %>% summarize(value=mean(value)) %>% 
  pivot_wider(names_from=snow, values_from=value) %>% left_join(ipochemsf) %>% 
  mutate(E_N = (Early-Normal)/Normal, .keep="unused")%>% 
  left_join(mod.top.split)

adaptive.plasticity.data <- list(  
  plasticity.water %>% left_join(vols.beta.water %>% rename(name=trait) %>% 
                                   filter(water=="Addition", fitnesstrait!="eggs_per_flower")),
  plasticity.water %>% left_join(vols.beta.water %>% rename(name=trait) %>%
                                   filter(water=="Reduction", fitnesstrait!="eggs_per_flower")),
  plasticity.snow %>% left_join(vols.beta.snow %>% rename(name=trait) %>% 
                                  filter(snow=="Early", fitnesstrait!="eggs_per_flower")))

adaptive.plasticity.data[[1]] %>% group_by(year, fitnesstrait) %>%
  summarize(adaptive.cor = cor(A_C, B, use="complete"),
            adaptive.P = cor.test(A_C, B, use="complete")$p.value,n=n())
adaptive.plasticity.data[[2]] %>% group_by(year, fitnesstrait) %>%
  summarize(adaptive.cor = cor(R_C, B, use="complete"),
            adaptive.P = cor.test(R_C, B, use="complete")$p.value,n=n())
adaptive.plasticity.data[[3]] %>% group_by(year, fitnesstrait) %>%
  summarize(adaptive.cor = cor(E_N, B, use="complete"),
            adaptive.P = cor.test(E_N, B, use="complete")$p.value,n=n())

adaptive.plasticity.plots <- list(
  ggplot(adaptive.plasticity.data[[1]], aes(x=A_C, size=`Addition - Control`)) + 
    labs(x="Emissions change with additional precipitation", 
         y="Selection gradient (beta) under additional precipitation"), 
  ggplot(adaptive.plasticity.data[[2]], aes(x=R_C, size=`Reduction - Control`)) + 
    labs(x="Emissions change with reduced precipitation", 
         y="Selection gradient (beta) under reduced precipitation"), 
  ggplot(adaptive.plasticity.data[[3]], aes(x=E_N, size=`Early - Normal`)) + 
    labs(x="Emissions change with early snowmelt", 
         y="Selection gradient (beta) under early snowmelt"))
#TODO Early-Normal is just TRUE/FALSE, should be a number?

library(ggpubr)
walk(adaptive.plasticity.plots, 
     ~ print(.x + facet_grid(rows=vars(year), cols=vars(fitnesstrait), scales="fixed", 
                             labeller=as_labeller(c(fitnessnames, set_names(levels(treatments$year)))))+
               geom_vline(xintercept=0) + geom_hline(yintercept=0) +
               annotate("rect", xmin = 0, xmax = Inf, ymin = 0, ymax = -Inf, fill= "black", alpha=0.1) + 
               annotate("rect", xmin = 0, xmax = -Inf, ymin = Inf, ymax = 0, fill= "black", alpha=0.1) +
               geom_smooth(aes(y=B, group=1), color="grey30", method="lm", se=F, size=1)+
               geom_point(aes(y=B, alpha=name=="a-pinene"), 
                          color=class_pal["Monoterpene"], shape="X", size=4)+
               scale_alpha_manual(values=c(0,1), guide="none")+
               geom_pointrange(aes(y=B, color=class, ymin=B-SE, ymax=B+SE), size=0.2) + # size=mean.floral
               #scale_size("Mean emissions\n(ng/hr/flower)", range=c(1,3)) +
               stat_cor(aes(y=B), size=3, label.x=0.1)+
               scale_size_manual(values=c(1,2), labels=c("NS", "Signif."))+
               scale_x_continuous(labels=scales::percent, limits=c(-1.5,NA), breaks=c(-2:5)) + 
               scale_color_manual("Compound class", values=class_pal, 
                                  guide=guide_legend(override.aes = list(size=1))) + 
               theme_bw() + theme(axis.text = element_text(size=7))))
```

```{r adaptive_quadrants}
plasticity.water %>% left_join(vols.beta.water %>% rename(name=trait) %>%
                                 filter(water=="Reduction", fitnesstrait!="eggs_per_flower")) %>% 
  mutate(adaptive = sign(B) == sign(R_C)) %>% count(adaptive) %>% kable()
plasticity.water %>% left_join(vols.beta.water %>% rename(name=trait) %>%
                                 filter(water=="Reduction", fitnesstrait!="eggs_per_flower")) %>% 
  mutate(adaptive = sign(B) == sign(R_C)) %>% count(adaptive) %>% kable()
plasticity.snow %>% left_join(vols.beta.snow %>% rename(name=trait) %>% filter(snow=="Early", fitnesstrait!="eggs_per_flower")) %>% 
  mutate(adaptive = sign(B) == sign(E_N)) %>% count(adaptive) %>% kable()
```


# Plasticity to snowmelt vs. precipitation

```{r snow_v_water}
ggplot(left_join(plasticity.snow, plasticity.water), aes(x=R_A, y=E_N)) + 
  facet_wrap(vars(year), ncol=1)+
  geom_vline(xintercept=0) + geom_hline(yintercept=0) +
  annotate("rect", xmin = 0, xmax = Inf, ymin = 0, ymax = -Inf, fill= "black", alpha=0.1) + 
  annotate("rect", xmin = 0, xmax = -Inf, ymin = Inf, ymax = 0, fill= "black", alpha=0.1) +
  geom_point(aes(alpha=name=="a-pinene"), color=class_pal["Monoterpene"], shape="X", size=4)+
  scale_alpha_manual(values=c(0,1), guide="none")+
  #eom_smooth(method="lm", color="black", se=F, size=0.5)+
  geom_point(aes(color=class)) + # size=mean.floral
  #scale_size("Mean emissions\n(ng/hr/flower)", range=c(1,5)) +
  scale_x_continuous(labels=scales::percent) + 
  scale_y_continuous(labels=scales::percent) + 
  scale_color_manual("Compound class", values=class_pal, 
                     guide=guide_legend(override.aes = list(size=5))) + theme_bw() +
  labs(x="Plasticity to precipitation reduction (vs. addition)",
       y="Plasticity to early snowmelt (vs. normal)")

left_join(plasticity.snow, plasticity.water) %>% filter(!is.infinite(R_A)) %>% 
  group_by(year) %>%   summarize(snow_water_cor = cor(R_A, E_N, use="complete"),
                                 snow_water_p = cor.test(R_A, E_N, use="complete")$p.value,
                                 n=n())
```


```{r adaptive_diff, fig.height=14, fig.width=10}
### Change in selection gradient vs. plasticity
diffbeta <- vols.beta.water %>% rename(name=trait) %>% 
  pivot_wider(names_from=water, values_from=c(B,SE)) %>% 
  mutate(B_A_R = B_Addition - B_Reduction)

plasticity.water %>% left_join(diffbeta) %>% 
  ggplot(aes(x=A_R, y=B_A_R, color=class)) + facet_wrap(vars(year), ncol=1)+
  geom_vline(xintercept=0)+geom_hline(yintercept=0) + geom_text(aes(label=name)) + 
  scale_x_continuous(labels=scales::percent, limits=c(NA,NA)) + theme_bw()  + theme(legend.position = "top") +
  labs(x="Emissions change in precipitation addition vs. reduction", y="Change in selection gradient (beta) in precipitation addition vs. reduction")
```

## Selection vs. absolute snowmelt, precip

```{r selection_abs_model}
selection_mod_abs <- function(trait, response) {
  vm.abs <- bind_cols(volsmeta.traits, vols.good) %>% 
    select(all_of(c(response, trait)), sun_date, precip_est_mm) %>% 
    rename("trait" = trait, "response" = response) %>% 
    filter(!is.infinite(response), !is.nan(response), !is.na(response)) %>% 
    mutate(response = response/mean(response, na.rm=T), trait = trait/sd(trait, na.rm=T))
  options(contrasts = c("contr.sum", "contr.poly"))
  lm(response ~ trait * sun_date + trait * precip_est_mm, data = vm.abs)
}

mod.selection.abs <- expand.grid(list(response = fitnesstraits, trait = ipogood)) %>%
  mutate(model = pmap(., selection_mod_abs),
         coefs = map(model, tidy))

mod.selection.tests <- mod.selection.abs %>% 
  left_join(ipochemsf, by=c("trait"="name")) %>% 
  filter(freq.floral>0.1) %>% 
    mutate(tests = map(model, ~ tidy(car::Anova(.x, type=3)))) %>% 
  select(-model) %>% unnest(coefs) %>% 
  filter(str_detect(term, "trait"))
```

```{r selection_abs, fig.width=10, fig.height=9}
mod.selection.abs %>% select(-model) %>% unnest(coefs) %>% 
  ggplot(aes(x=response, y=estimate)) + facet_wrap(vars(term), scales="free_y") +
  geom_hline(yintercept=0) + geom_boxplot()

library(ggrepel)
ggplot(mod.selection.tests, aes(x=response, y=estimate, color=class, label=trait)) + 
  facet_wrap(vars(term), labeller = as_labeller(c(trait="A. Trait (beta)",
                                                  `trait:precip_est_mm`=
                                                    "B. Trait x summer precipitation (beta / mm)",
                                                  `trait:sun_date`=
                                                    "C. Trait x snowmelt date (beta / day)")),
             scales="free_y", ncol=1) + theme_bw() +
  geom_point(aes(size=mean.floral)) + geom_hline(yintercept=0) + 
  geom_text_repel(data = filter(mod.selection.tests, p.value < 0.01), hjust=0, nudge_x = 0.15, show.legend = FALSE) + 
  scale_size_area(max_size = 10) + scale_x_discrete(labels=fitnessnames) +
  scale_color_manual(values=class_pal, guide=guide_legend(override.aes = list(size=5))) + 
  labs(x="Aspect of fitness", y = "Estimate", size="Mean emissions\n(ng/hr/flower)", color="Class")

mod.selection.abs %>% filter(trait=="a-pinene") %>% 
  mutate(tests = map(model, ~ tidy(car::Anova(.x, type=3)))) %>% 
  select(-c(model, coefs)) %>% unnest(tests) %>% filter(str_detect(term, "trait")) %>% kable()
```

# Volatiles vs. floral traits

```{r flrsize}
walk(floraltraits, ~
print(ggplot(traits.class, aes_string(x=.x, y="value", color="year")) + 
  geom_point() + geom_smooth(se=F, span=1) + 
  facet_wrap(vars(name), scales="free_y") + 
  labs(y="Emissions (ng/hr/flower)", x=traitnames.units[.x], color="Year")+
  theme_bw() + scale_y_sqrt() + scale_color_manual(values=year_pal)))
```

